%% leanOSLF: A Verified OSLF Formalization in Lean 4
%%
%% Render with:
%%   pdflatex leanOSLF.tex && bibtex leanOSLF && pdflatex leanOSLF.tex && pdflatex leanOSLF.tex

\documentclass[]{article}
\usepackage{url}
\usepackage[hyperindex,breaklinks]{hyperref}
\usepackage{breakurl}
\usepackage{listings}
\lstset{basicstyle=\ttfamily\footnotesize,breaklines=true,frame=single}
\usepackage{float}
\restylefloat{table}
\usepackage{longtable}
\usepackage{graphicx}
\usepackage[font=small,labelfont=bf]{caption}
\usepackage[skip=0pt]{subcaption}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{amsthm}

\newtheorem{theorem}{Theorem}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{definition}[theorem]{Definition}

\newcommand{\code}[1]{\texttt{#1}}
\newcommand{\codepath}[1]{\path{#1}}
\newcommand{\rhocalc}{\ensuremath{\varrho}}
\newcommand{\Dia}{\ensuremath{\diamondsuit}}
\newcommand{\Bx}{\ensuremath{\blacksquare}}
\newcommand{\OSLF}{\textup{OSLF}}
\newcommand{\MeTTaIL}{\textup{MeTTaIL}}
\newcommand{\GSLT}{\textup{GSLT}}

\title{Verified Operational Semantics in Logical Form:\\A Lean 4 Formalization of the \OSLF{} Algorithm\\
  \texttt{(DRAFT --- \today)}}
\author{Zar \and Oru\v{z}i (Claude Anthropic)}

\begin{document}
\maketitle

\begin{abstract}
We present a comprehensive Lean~4 formalization of Operational Semantics in Logical
Form (\OSLF{}), the algorithm that mechanically derives spatial-behavioral type systems
from rewrite rules.  Our formalization spans 22{,}300+ lines across 58 Lean files,
with 0~sorries across the entire core pipeline (29~sorries remain only in auxiliary
$\pi$-to-\rhocalc{} encoding correctness proofs).
The formalization covers:
(i) \MeTTaIL{}, a meta-language for defining process calculi with capture-avoiding
substitution and a totalized pattern matcher;
(ii) the \rhocalc{}-calculus with full reduction semantics, structural congruence,
and modal type system;
(iii) the abstract \OSLF{} framework instantiated for \emph{four} languages
(\rhocalc{}-calculus, lambda calculus, Petri nets, and TinyML), each with
\emph{fully proven} Galois connections $\Dia \dashv \Bx$;
(iv) executable rewrite engines (specialized and generic) proven sound with respect
to a declarative reduction specification;
(v) a \emph{constructor category} built from sort-crossing constructors with a
\code{SubobjectFibration} and \code{ChangeOfBase}, connecting to the GSLT
categorical infrastructure;
(vi) \emph{derived typing rules} where the modal operator ($\Dia$ or $\Bx$)
assigned to each constructor is determined automatically by its position in the
constructor category;
(vii) a \emph{Beck--Chevalley analysis} of substitution as change-of-base, with a
proven counterexample showing the strong condition fails; and
(viii) a MeTTa Core interpreter specification with confluence and progress.
All Galois connections, the type soundness theorem, the engine soundness theorem,
the constructor fibration, the derived typing rules, and the Beck--Chevalley
analysis carry zero sorries.
\end{abstract}

\section{Introduction}

The \OSLF{} algorithm~\cite{meredith-stay-oslf} takes a rewrite system as input and
produces a spatial-behavioral type system as output.  The core insight is that every
reduction relation induces a pair of adjoint modal operators:
\begin{align}
  \Dia \varphi &= \{ p \mid \exists q.\; p \rightsquigarrow q \;\wedge\; q \in \varphi \}
    \quad\text{(step-future / possibly)} \\
  \Bx \varphi &= \{ q \mid \forall p.\; p \rightsquigarrow q \;\Rightarrow\; p \in \varphi \}
    \quad\text{(step-past / rely)}
\end{align}
and that $\Dia \dashv \Bx$ forms a Galois connection.  Combined with the
spatial decomposition from parallel composition, this yields a type system where types
are ``behavioral neighborhoods'' and typing is substitutability under bisimulation.

Previous treatments of \OSLF{} were paper-only.  We give the first machine-checked
formalization, connecting:
\begin{itemize}
\item a \emph{generic} abstract framework (any rewrite system),
\item a \emph{concrete} \rhocalc{}-calculus instance with all rules proven,
\item a \emph{categorical} derivation via a constructor category with fibered
  change-of-base and derived typing rules,
\item an \emph{executable} reduction engine proven sound w.r.t.\ the spec, and
\item \emph{four language instances} validating the full pipeline.
\end{itemize}

\paragraph{Contributions.}
\begin{enumerate}
\item A Lean~4 formalization of the \OSLF{} algorithm as an abstract structure
  (\code{RewriteSystem} $\to$ \code{OSLFTypeSystem}) and its full \rhocalc{}-calculus
  instance (\code{rhoOSLF}) with a proven Galois connection.
\item A categorical proof that the Galois connection arises from the adjoint triple
  $\exists_f \dashv f^* \dashv \forall_f$ applied to the reduction span, with the
  result shown equal to the direct \rhocalc{}-calculus modalities.
\item A \emph{constructor category} built from sort-crossing constructors of any
  \code{LanguageDef}, with a \code{SubobjectFibration} and \code{ChangeOfBase}
  connecting to the GSLT infrastructure.
\item \emph{Derived typing rules}: the modal operator ($\Dia$/$\Bx$/id) assigned to
  each constructor is determined automatically by its classification
  (quoting/reflecting/neutral), and the assignment is proven correct for the
  \rhocalc{}-calculus.
\item A \emph{Beck--Chevalley analysis} of the COMM rule as change-of-base along the
  substitution map, with a proven counterexample showing the GSLT strong
  Beck--Chevalley condition does not hold for the constructor fibration.
\item An executable reduction engine (\code{reduceStep}) handling COMM, DROP, and PAR
  rules, with a machine-checked soundness theorem.
\item A declarative reduction relation (\code{DeclReduces}) independent of the engine,
  with proven soundness and completeness of the generic engine.
\item \emph{Four} \OSLF{} instantiations validating generality:
  \rhocalc{}-calculus, lambda calculus, Petri nets, and TinyML (a multi-sort CBV
  $\lambda$-calculus with booleans, pairs, and thunks).
\item \MeTTaIL{}: a meta-language for defining process calculi, with totalized pattern
  matching, 29 proven theorems about capture-avoiding substitution, and a complete
  \rhocalc{}-calculus language definition.
\item A verified bounded model checker for \OSLF{} formulas, with support for
  predecessor-based $\Bx$ checking and proven soundness.
\item A MeTTa Core interpreter specification with progress, confluence, and barbed
  bisimulation properties (97 proven theorems, 0 sorries).
\end{enumerate}

\section{Background: The \rhocalc{}-Calculus}

The reflective higher-order calculus~\cite{meredith-radestock-rhocalc} extends the
$\pi$-calculus with:
\begin{itemize}
\item \emph{Quoting}: any process $P$ can be turned into a name $@P$ (name = quoted process).
\item \emph{Dequoting}: $*x$ recovers the process quoted by name $x$.
\item \emph{No built-in names}: all names arise from quoting, giving the calculus a
  reflective character.
\end{itemize}

The reduction rules are:
\begin{align}
  \text{COMM:}\quad & \{n!(q) \mid \mathbf{for}(x \leftarrow n)\{p\} \mid \mathit{rest}\}
    \;\rightsquigarrow\; \{p[@q/x] \mid \mathit{rest}\} \\
  \text{DROP:}\quad & *(@P) \;\rightsquigarrow\; P
\end{align}
plus structural congruence (11 rules) and contextual reduction under parallel composition.

\section{Formalization Architecture}

\subsection{Module Structure}

The formalization is organized in seven directories plus standalone files:

\begin{center}
\begin{tabular}{lrrl}
\textbf{Directory} & \textbf{Lines} & \textbf{Sorries} & \textbf{Content} \\
\hline
\code{MeTTaIL/}        & 2{,}929 & 0  & Meta-language AST, substitution, matching, declarative reduction \\
\code{MeTTaCore/}      & 2{,}946 & 0  & Interpreter specification \\
\code{Framework/}      & 4{,}400 & 0  & Abstract \OSLF{} + categorical bridge + 4 instances \\
\code{RhoCalculus/}    & 3{,}893 & 0  & Concrete \rhocalc{}-calculus + engine \\
\code{PiCalculus/}     & 6{,}582 & 29 & $\pi$-calculus + \rhocalc{}-encoding \\
\code{NativeType/}     &   263   & 0  & Native type construction \\
\code{Formula.lean}    &   582   & 0  & Verified bounded model checker \\
\code{Main.lean}       &   387   & 0  & Re-exports \\
\hline
\textbf{Total}         & \textbf{22{,}320} & \textbf{29} & \textbf{Core: 0 sorries}
\end{tabular}
\end{center}

All 29 remaining sorries are in the $\pi$-to-\rhocalc{} encoding correctness proofs
(\code{PiCalculus/RhoEncodingCorrectness.lean}), with one major theorem (Proposition~2:
substitution invariance) already fully proven.
The entire core \OSLF{} pipeline---\MeTTaIL{}, Framework, RhoCalculus, Formula,
and MeTTa Core---carries \textbf{0 sorries and 0 axioms}.

The \code{Framework/} directory (4{,}400 lines, 15~files) is the largest component,
containing:

\begin{center}
\begin{tabular}{lrl}
\textbf{File} & \textbf{Lines} & \textbf{Content} \\
\hline
\code{ConstructorCategory.lean}    & 460 & Sort quiver + free category \\
\code{TinyMLInstance.lean}         & 528 & CBV $\lambda$-calculus with booleans/pairs/thunks \\
\code{BeckChevalleyOSLF.lean}      & 449 & Substitution as change-of-base \\
\code{DerivedTyping.lean}          & 346 & Generic typing rules from constructor category \\
\code{ModalEquivalence.lean}       & 311 & Constructor change-of-base $\leftrightarrow$ modalities \\
\code{GeneratedTyping.lean}        & 294 & \code{GenHasType} typing rules \\
\code{SynthesisBridge.lean}        & 282 & Three-layer bridge \\
\code{ConstructorFibration.lean}   & 251 & \code{SubobjectFibration} + \code{ChangeOfBase} \\
\code{DerivedModalities.lean}      & 250 & Adjoint triple derivation \\
\code{CategoryBridge.lean}         & 247 & GaloisConnection $\to$ Adjunction \\
\code{PetriNetInstance.lean}       & 233 & Petri net OSLF instance \\
\code{LambdaInstance.lean}         & 218 & Lambda calculus OSLF instance \\
\code{TypeSynthesis.lean}          & 201 & \code{langOSLF} pipeline \\
\code{RewriteSystem.lean}          & 196 & Abstract \OSLF{} input/output \\
\code{RhoInstance.lean}            & 134 & \rhocalc{}-calculus instance \\
\end{tabular}
\end{center}

\subsection{Abstract \OSLF{} Framework}

The abstract layer defines two key structures:

\begin{lstlisting}[caption={The \OSLF{} input and output (RewriteSystem.lean)}]
structure RewriteSystem where
  Sorts    : Type*
  procSort : Sorts
  Term     : Sorts -> Type*
  Reduces  : Term procSort -> Term procSort -> Prop

structure OSLFTypeSystem where
  Sorts    : Type*
  procSort : Sorts
  Term     : Sorts -> Type*
  Pred     : Sorts -> Type*
  [frame   : (S : Sorts) -> Frame (Pred S)]
  satisfies  : (S : Sorts) -> Term S -> Pred S -> Prop
  diamond    : Pred procSort -> Pred procSort
  box        : Pred procSort -> Pred procSort
  galois     : GaloisConnection diamond box
\end{lstlisting}

\subsection{The Galois Connection}

The central theorem: $\Dia \dashv \Bx$.

\begin{theorem}[Galois Connection, 0 sorries]
\label{thm:galois}
For all predicates $\varphi, \psi$ on \rhocalc{}-calculus processes:
$$\Dia\varphi \le \psi \iff \varphi \le \Bx\psi$$
where $\Dia\varphi(p) = \exists q.\; p \rightsquigarrow q \wedge \varphi(q)$
and $\Bx\psi(q) = \forall p.\; p \rightsquigarrow q \to \psi(p)$.
\end{theorem}

In Lean:
\begin{lstlisting}[caption={The Galois connection (Reduction.lean)}]
theorem galois_connection :
    GaloisConnection possiblyProp relyProp := by
  intro phi psi
  constructor
  . intro h q hrely p hred
    exact h (hrely p hred)
  . intro h p hposs
    exact h.2 p hposs.1 hposs.2
\end{lstlisting}

\subsection{Categorical Derivation via Adjoint Triples}

The Galois connection arises from the general theory of change-of-base along
a span.

\begin{definition}[Reduction Span]
A span $\mathcal{S} \xleftarrow{\mathrm{src}} E \xrightarrow{\mathrm{tgt}} \mathcal{S}$
where $E$ is the set of reduction edges, $\mathrm{src}$ extracts the source, and
$\mathrm{tgt}$ extracts the target.
\end{definition}

From any such span we derive three operations on predicates:
\begin{align}
  f^*(\psi)(e) &= \psi(\mathrm{tgt}(e))
    && \text{(pullback)} \\
  \exists_f(\varphi)(q) &= \exists e.\; \mathrm{tgt}(e) = q \wedge \varphi(e)
    && \text{(direct image)} \\
  \forall_f(\varphi)(q) &= \forall e.\; \mathrm{tgt}(e) = q \to \varphi(e)
    && \text{(universal image)}
\end{align}

\begin{theorem}[Derived Galois, 0 sorries]
\label{sec:derived}
For any \code{ReductionSpan}, the composition $\Dia = \exists_{\mathrm{src}} \circ
\mathrm{tgt}^*$ and $\Bx = \forall_{\mathrm{tgt}} \circ \mathrm{src}^*$ form a
Galois connection.  Furthermore, for the \rhocalc{}-calculus span, the derived operators
equal the concrete \code{possiblyProp} and \code{relyProp}.
\end{theorem}

\begin{lstlisting}[caption={Derived modalities equal concrete (DerivedModalities.lean)}]
theorem derived_diamond_eq_possiblyProp :
    derivedDiamond rhoSpan = possiblyProp := ...

theorem derived_box_eq_relyProp :
    derivedBox rhoSpan = relyProp := ...

theorem rho_galois_from_span :
    GaloisConnection (derivedDiamond rhoSpan)
                     (derivedBox rhoSpan) :=
  derived_galois rhoSpan
\end{lstlisting}

\section{Constructor Category and Fibration}
\label{sec:constructor-cat}

A key contribution of this formalization is the \emph{constructor category}: a
non-discrete category built from the sort-crossing constructors of any
\code{LanguageDef}, replacing the discrete \code{Discrete R.Sorts} from the
earlier categorical lift.

\subsection{Sort Quiver and Free Category}

Given a \code{LanguageDef}, we extract the \emph{unary sort-crossing constructors}:
grammar rules with exactly one \code{.simple} parameter whose base sort differs
from the constructor's output sort.  These become the arrows of a quiver on the
language's sorts.

\begin{lstlisting}[caption={Constructor category (ConstructorCategory.lean)}]
-- Sort type: valid sort names
def LangSort (lang : LanguageDef) :=
  { s : String // s IN lang.types }

-- Sort-crossing arrows
structure SortArrow (lang : LanguageDef)
    (dom cod : LangSort lang) where
  label : String
  valid : (label, dom.val, cod.val) IN unaryCrossings lang

-- Free category: paths of sort-crossing arrows
inductive SortPath (lang : LanguageDef)
    : LangSort lang -> LangSort lang -> Type where
  | nil  : SortPath lang s s
  | cons : SortPath lang s t -> SortArrow lang t u
         -> SortPath lang s u
\end{lstlisting}

For the \rhocalc{}-calculus: 2 objects (Proc, Name), 2 arrows
(\code{NQuote}: Proc $\to$ Name, \code{PDrop}: Name $\to$ Proc),
and composites \code{PDrop} $\circ$ \code{NQuote} and
\code{NQuote} $\circ$ \code{PDrop}.

Each arrow has a \emph{semantic function} \code{arrowSem}: wrapping a pattern
in the constructor's \code{.apply} node (e.g., $p \mapsto \mathtt{NQuote}(p)$).
This extends to paths via \code{pathSem}, with a proven composition law
\code{pathSem\_comp}.

A \emph{universal property} (free category lifting) is proven:
any assignment of objects and arrows to a target category $\mathcal{C}$
lifts uniquely to a functor \code{liftFunctor}, with uniqueness proven in
\code{lift\_map\_unique}.

\subsection{SubobjectFibration and ChangeOfBase}

Over the constructor category we build a fibration and change-of-base
(\codepath{Framework/ConstructorFibration.lean}, 251 lines, 0~sorries):

\begin{lstlisting}[caption={Constructor fibration (ConstructorFibration.lean)}]
-- Each sort has fiber Pattern -> Prop (a Frame)
def constructorFibration (lang : LanguageDef) :
    SubobjectFibration (ConstructorObj lang) where
  Sub   := fun _ => Pattern -> Prop
  frame := fun _ => Pi.instFrame

-- Full change-of-base with proven adjunctions
def constructorChangeOfBase (lang : LanguageDef) :
    ChangeOfBase (constructorFibration lang) where
  pullback f       := pb (pathSem lang f)
  directImage f    := di (pathSem lang f)
  universalImage f := ui (pathSem lang f)
  direct_pullback_adj f := di_pb_adj (pathSem lang f)
  pullback_universal_adj f := pb_ui_adj (pathSem lang f)
  ...
\end{lstlisting}

The adjunctions $\exists_f \dashv f^* \dashv \forall_f$ are proven (not axiomatized),
following from the generic \code{di\_pb\_adj} / \code{pb\_ui\_adj} in
\code{DerivedModalities.lean}.

Key proven properties:
\begin{itemize}
\item \textbf{Pullback functoriality}: $(f \circ g)^* = g^* \circ f^*$
  (from \code{pathSem\_comp}), $\mathit{id}^*(\varphi) = \varphi$.
\item \textbf{Frame morphism}: $f^*(\varphi \wedge \psi) = f^*(\varphi) \wedge f^*(\psi)$
  and $f^*(\top) = \top$ (both by \code{rfl}).
\item \textbf{Monotonicity} of all three operations (from adjunctions).
\end{itemize}

\subsection{Modal Equivalence}

The file \codepath{Framework/ModalEquivalence.lean} (311 lines) connects
the constructor change-of-base to the \OSLF{} modalities:

\begin{theorem}[Modal = Change-of-Base, 0 sorries]
The \OSLF{} modalities are Set-level change-of-base along the reduction span:
\begin{align*}
  \Dia_{\mathit{lang}} &= \exists_{\mathrm{src}} \circ \mathrm{tgt}^*
    && \text{(definitional)} \\
  \Bx_{\mathit{lang}} &= \forall_{\mathrm{tgt}} \circ \mathrm{src}^*
    && \text{(definitional)}
\end{align*}
\end{theorem}

For the \rhocalc{}-calculus, this gives the \emph{typing actions}:
\begin{itemize}
\item \code{NQuote} (Proc $\to$ Name): $\varphi \mapsto \Dia\varphi$
  \quad (``can reduce to $\varphi$'')
\item \code{PDrop} (Name $\to$ Proc): $\alpha \mapsto \Bx\alpha$
  \quad (``all predecessors satisfy $\alpha$'')
\end{itemize}
The composite $\code{PDrop} \circ \code{NQuote}$ gives $\Bx \circ \Dia$
and $\code{NQuote} \circ \code{PDrop}$ gives $\Dia \circ \Bx$.
The typing action Galois connection $\Dia \dashv \Bx$ is proven as an
instance of the general language Galois connection.

\subsection{Derived Typing Rules}
\label{sec:derived-typing}

The file \codepath{Framework/DerivedTyping.lean} (346 lines, 0~sorries)
derives typing rules generically from the constructor category structure.

Each sort-crossing arrow is automatically classified:
\begin{lstlisting}[caption={Constructor classification (DerivedTyping.lean)}]
inductive ConstructorRole where
  | quoting    -- domain = procSort: introduces diamond
  | reflecting -- codomain = procSort: introduces box
  | neutral    -- neither: identity

def classifyArrow (lang : LanguageDef) (procSort : String)
    (arr : SortArrow lang dom cod) : ConstructorRole :=
  if dom.val = procSort then .quoting
  else if cod.val = procSort then .reflecting
  else .neutral
\end{lstlisting}

\begin{theorem}[Classification Correctness, 0 sorries]
For the \rhocalc{}-calculus:
\begin{itemize}
\item \code{NQuote} is classified as \emph{quoting}; its typing action equals $\Dia$.
\item \code{PDrop} is classified as \emph{reflecting}; its typing action equals $\Bx$.
\end{itemize}
\end{theorem}

The \code{DerivedHasType} judgment provides a generic typing rule for unary
sort-crossing constructors: apply the constructor's typing action
($\Dia$/$\Bx$/id) to the argument's predicate, then tag the result at the
output sort.

\subsection{Beck--Chevalley for Substitution}
\label{sec:beck-chevalley}

The file \codepath{Framework/BeckChevalleyOSLF.lean} (449 lines, 0~sorries)
analyzes the COMM rule's substitution $p[@q/x]$ as a change-of-base map.

\begin{definition}[COMM Substitution Map]
$\sigma_q : \mathit{Pattern} \to \mathit{Pattern}$ defined by
$\sigma_q(\mathit{pBody}) = \mathit{openBVar}\;0\;(\mathtt{NQuote}(q))\;\mathit{pBody}$.
\end{definition}

This induces the adjoint triple
$\exists_{\sigma_q} \dashv \sigma_q^* \dashv \forall_{\sigma_q}$
via the same \code{pb}/\code{di}/\code{ui} infrastructure, and the modal+substitution
Galois connections compose:

\begin{theorem}[Composed Galois, 0 sorries]
$\Dia \circ \exists_{\sigma_q} \dashv \sigma_q^* \circ \Bx$
\end{theorem}

The COMM rule's type preservation (\code{comm\_preserves\_type} from
\code{Soundness.lean}) is re-expressed categorically as a pullback inequality:

\begin{theorem}[Substitutability as Pullback, 0 sorries]
For any typing context $\Gamma$, type $\tau$, variable $x$, value $q$,
and type $\sigma$ with $\Gamma \vdash q : \sigma$:
$$\code{typedAt}(\Gamma[x \mapsto \sigma], \tau)
  \;\le\; \sigma_q^*(\code{typedAt}(\Gamma, \tau))$$
\end{theorem}

The COMM substitution map factors through the constructor semantics:
$\sigma_q(p) = \mathit{openBVar}\;0\;(\mathit{pathSem}\;\mathtt{nquoteMor}\;q)\;p$.

\begin{theorem}[Strong Beck--Chevalley Fails, 0 sorries]
\label{thm:bc-fails}
The GSLT's universal Beck--Chevalley condition
$f^* \circ \exists_g = \exists_{\pi_1} \circ \pi_2^*$
does \textbf{not} hold for the constructor fibration.

Concretely, for the commuting square
$\code{PDrop} \circ \code{NQuote} = \code{PDrop} \circ \code{NQuote}$:
$$\code{PDrop}^*(\exists_{\code{PDrop}}(\top))(\code{fvar}\;x) = \top$$
but
$$\exists_{\code{NQuote}}(\code{NQuote}^*(\top))(\code{fvar}\;x) = \bot$$
because $\code{NQuote}(q) \ne \code{fvar}\;x$ for all $q$.
\end{theorem}

The counterexample is proven by exhibiting a concrete witness at
$\code{fvar}\;``x"$: the LHS is inhabited by $\langle \code{fvar}\;x, \mathit{rfl},
\top \rangle$ while the RHS requires some $p$ with
$\code{NQuote}(p) = \code{fvar}\;x$, which is impossible since
$\code{NQuote}(p) = \code{.apply}\;\text{``NQuote''}\;[p]$.

\section{Executable Rewrite Engine}

A formalization that only \emph{specifies} reduction is incomplete for verification
of actual implementations.  We provide \code{reduceStep}, a computable function that
enumerates all one-step reducts, proven sound w.r.t.\ the propositional \code{Reduces}.

\subsection{Engine Design}

\begin{lstlisting}[caption={The executable engine (Engine.lean)}]
def reduceStep (p : Pattern) (fuel : Nat := 100)
    : List Pattern :=
  match fuel with
  | 0 => []
  | fuel + 1 =>
    match p with
    | .collection .hashBag elems none =>
      let commReducts := findAllComm elems
      let parReducts :=
        reduceElemsAux (reduceStep . fuel) elems
        |>.map fun (i, elem') =>
          .collection .hashBag (elems.set i elem') none
      commReducts ++ parReducts
    | .apply "PDrop" [.apply "NQuote" [inner]] =>
      [inner]
    | _ => []
\end{lstlisting}

The engine handles COMM (all output-input pairs on matching channels), DROP
($*(@P) \rightsquigarrow P$), and PAR (recursive reduction under parallel composition).
Non-deterministic races produce multiple reducts in the output list.

\subsection{Soundness}

\begin{theorem}[Engine Soundness, 0 sorries]
Every reduct computed by \code{reduceStep} corresponds to a valid \code{Reduces}:
$$q \in \code{reduceStep}(p, \mathit{fuel}) \implies \exists d : p \rightsquigarrow q$$
\end{theorem}

The proof proceeds by case analysis:
\begin{itemize}
\item \textbf{COMM}: Each output-input pair is located by \code{findAllComm}, whose
  specification is proven to identify valid COMM redex positions.  The soundness chain
  is: list permutation (\code{perm\_extract\_two}) $\to$ structural congruence
  (\code{par\_perm}) $\to$ \code{Reduces.comm} $\to$ \code{Reduces.equiv}.
\item \textbf{DROP}: Direct pattern match yields \code{Reduces.drop}.
\item \textbf{PAR}: The \code{reduceElemsAux\_spec} lemma extracts the sub-element
  index and recursive reduct.  List decomposition via
  \code{List.set\_eq\_take\_append\_cons\_drop} connects \code{List.set} to the
  $\mathit{before} \mathbin{+\!\!+} [p] \mathbin{+\!\!+} \mathit{after}$ form
  required by \code{Reduces.par\_any}.
\end{itemize}

\section{Generic MeTTaIL Rewrite Framework}

The specialized \rhocalc{}-calculus engine is lifted to a language-parametric
framework.  Given any \code{LanguageDef} (a list of sorts, constructors, equations,
and rewrite rules), the generic engine automatically:
\begin{enumerate}
\item matches concrete terms against rule LHS patterns (multiset matching with rest variables),
\item produces variable bindings via alpha-renaming for binders,
\item applies bindings to rule RHS patterns (including substitution evaluation), and
\item iterates under congruence (subterm rewriting inside parallel compositions).
\end{enumerate}

\begin{lstlisting}[caption={Generic rewrite step (Match.lean)}]
def applyRule (rule : RewriteRule) (term : Pattern)
    : List Pattern :=
  if rule.premises.isEmpty then
    (matchPattern rule.left term).map
      fun b => applyBindings b rule.right
  else []

def rewriteStep (lang : LanguageDef) (term : Pattern)
    : List Pattern :=
  lang.rewrites.flatMap fun rule => applyRule rule term
\end{lstlisting}

\paragraph{Multiset matching.}
The key algorithmic contribution is \code{matchBag}: for a collection pattern with
$n$ elements and an optional rest variable, it enumerates all ways to match the $n$
pattern elements against term elements (backtracking search over permutations),
binding unmatched elements to the rest variable.

\paragraph{Declarative reduction.}
The file \codepath{MeTTaIL/DeclReduces.lean} provides a declarative inductive reduction
relation independent of the executable engine.  The engine is proven both \emph{sound}
and \emph{complete} with respect to this specification (0~sorries).

\section{MeTTaIL and MeTTa Core}

\subsection{MeTTaIL: The Meta-Language}

\MeTTaIL{} (``MeTTa Intermediate Language'') defines the AST shared by all process
calculi in the formalization, using locally nameless representation.  The core type is
\code{Pattern} with 7 constructors:

\begin{lstlisting}[caption={Pattern AST --- locally nameless (Syntax.lean)}]
inductive Pattern where
  | bvar       : Nat -> Pattern         -- bound variable (de Bruijn)
  | fvar       : String -> Pattern      -- free variable / metavariable
  | apply      : String -> List Pattern -> Pattern
  | lambda     : Pattern -> Pattern     -- binder (no name)
  | multiLambda: Nat -> Pattern -> Pattern
  | subst      : Pattern -> Pattern -> Pattern
  | collection : CollType -> List Pattern
                  -> Option String -> Pattern
\end{lstlisting}

Key proven properties of substitution (29 theorems, 0 sorries):
\begin{itemize}
\item \code{subst\_empty}: empty substitution is the identity;
\item \code{subst\_fresh}: substitution on a fresh variable is the identity;
\item \code{commSubst}: the COMM-rule substitution $p[@q/x]$ respects freshness.
\end{itemize}

\subsection{MeTTa Core Interpreter}

The \code{MeTTaCore/} directory provides a complete interpreter specification for
Hyperon Experimental MeTTa (2{,}946 lines, 0~sorries), including:
\begin{itemize}
\item \code{Atom}: the universal term type with \code{DecidableEq};
\item \code{Bindings}: variable resolution with merge and transitive lookup;
\item \code{MeTTaState}: the 4-register $\langle i, k, w, o \rangle$ machine;
\item \code{PatternMatch}: bidirectional unification;
\item \code{MinimalOps}: grounded operations ($+, -, *, /, <$, etc.);
\item \code{RewriteRules}: equation-driven rewriting;
\item \code{Atomspace}: multiset-based knowledge base with query operations;
\item \code{Properties}: progress, confluence, and barbed bisimulation.
\end{itemize}

\section{Type Soundness}

\begin{theorem}[Substitutability, 0 sorries]
If $P$ and $Q$ are bisimilar processes, then they have the same native types:
$$P \sim Q \implies \forall \tau.\; P : \tau \iff Q : \tau$$
\end{theorem}

\begin{theorem}[Type Preservation, 0 sorries]
The COMM and DROP rules preserve types:
$$\Gamma \vdash P : \tau \;\wedge\; P \rightsquigarrow Q \implies \Gamma \vdash Q : \tau$$
\end{theorem}

Both theorems are proven in \codepath{RhoCalculus/Soundness.lean} with
a 10-constructor typing judgment \code{HasType} and explicit typing contexts.

\section{\rhocalc{}-Calculus Instance}

The \rhocalc{}-calculus is formalized with full reduction semantics (COMM, DROP),
structural congruence (11 rules), and multi-step reduction. The OSLF algorithm
derives its spatial-behavioral type system, with the Galois connection
$\Dia \dashv \Bx$ proven in \code{RhoCalculus/Soundness.lean} (0 sorries).

\section{Type Synthesis: \code{LanguageDef} $\to$ \code{OSLFTypeSystem}}

The culmination of the formalization is the \emph{type synthesis pipeline}: given any
\code{LanguageDef}, mechanically produce a full \code{OSLFTypeSystem} with a
\textbf{proven Galois connection}.

\subsection{The Pipeline}

The pipeline proceeds in five steps, all implemented in
\codepath{Framework/TypeSynthesis.lean}:
\begin{enumerate}
\item \code{langReduces lang p q} wraps the executable engine:
  $q \in \mathtt{rewriteWithContext}\;\mathit{lang}\;p$.
\item \code{langRewriteSystem lang} assembles a \code{RewriteSystem}.
\item \code{langSpan lang} builds the reduction span (edges = reductions).
\item \code{langDiamond}/\code{langBox} derive modal operators via
  \code{derivedDiamond}/\code{derivedBox} from the adjoint triple.
\item \code{langOSLF lang} packages everything into an \code{OSLFTypeSystem},
  with the Galois connection proven by \code{derived\_galois}.
\end{enumerate}

\begin{theorem}[Automatic Galois Connection, 0 sorries]
For any \code{LanguageDef lang}:
$$\Dia_{\mathit{lang}} \dashv \Bx_{\mathit{lang}}$$
where $\Dia_{\mathit{lang}} = \exists_{\mathrm{src}} \circ \mathrm{tgt}^*$
and $\Bx_{\mathit{lang}} = \forall_{\mathrm{tgt}} \circ \mathrm{src}^*$
are derived from the reduction span.  No manual proof is needed per language.
\end{theorem}

\section{Language Instances}
\label{sec:instances}

The pipeline is validated by four language instances of increasing complexity:

\subsection{Lambda Calculus (1 sort, 0 crossings)}

Untyped lambda calculus with $\beta$-reduction
(\codepath{Framework/LambdaInstance.lean}, 218 lines).
One sort (\code{Term}), two constructors (\code{App}, \code{Lam}), one reduction rule.
The constructor category is discrete (no sort-crossing constructors).
Six executable demos verify $\beta$-reduction, multi-step normalization,
and formula checking.

\subsection{Petri Net (1 sort, 0 crossings)}

A simple Petri net with four places and two transitions
(\codepath{Framework/PetriNetInstance.lean}, 233 lines).
Validates that multiset (bag) matching works correctly without any
abstraction or substitution machinery.  Key properties proven by
\code{native\_decide}:
\begin{itemize}
\item $\{D\}$ is a dead marking (0 reducts);
\item $\{A, B\}$ has exactly 1 reduct via transition $T_1$.
\end{itemize}

\subsection{\rhocalc{}-Calculus (2 sorts, 2 crossings)}

The primary instance with sorts Proc and Name, constructors
\code{NQuote}: Proc $\to$ Name and \code{PDrop}: Name $\to$ Proc.
This is the most extensively developed instance, with the full
type soundness proof, specialized engine, and Beck--Chevalley analysis
(Sections~\ref{sec:constructor-cat}--\ref{sec:beck-chevalley}).

\subsection{TinyML (2 sorts, 2 crossings, 6 rules)}

\emph{New:} A call-by-value $\lambda$-calculus with booleans, pairs, and thunks
(\codepath{Framework/TinyMLInstance.lean}, 528 lines, 0~sorries).

\begin{lstlisting}[caption={TinyML language definition (TinyMLInstance.lean)}]
-- Sorts: Expr (process sort), Val (data sort)
-- Constructors:
--   Expr: App, If, Fst, Snd, Inject(v:Val)
--   Val:  BoolT, BoolF, Lam(^body), PairV, Thunk(e:Expr)
-- Reductions:
--   Beta:     App(Inject(Lam(^body)), Inject(v)) ~> body[v/x]
--   Force:    Inject(Thunk(e)) ~> e
--   IfTrue:   If(Inject(BoolT), t, e) ~> t
--   IfFalse:  If(Inject(BoolF), t, e) ~> e
--   FstPair:  Fst(Inject(PairV(a, b))) ~> Inject(a)
--   SndPair:  Snd(Inject(PairV(a, b))) ~> Inject(b)
\end{lstlisting}

TinyML mirrors the \rhocalc{}-calculus sort structure:
\begin{center}
\begin{tabular}{lll}
\textbf{TinyML} & \textbf{\rhocalc{}-Calculus} & \textbf{Role} \\
\hline
Expr & Proc & Process sort (carries reduction) \\
Val & Name & Data sort \\
\code{Thunk}: Expr $\to$ Val & \code{NQuote}: Proc $\to$ Name & Quoting ($\Dia$) \\
\code{Inject}: Val $\to$ Expr & \code{PDrop}: Name $\to$ Proc & Reflecting ($\Bx$) \\
\code{Beta} & \code{COMM} & Main computation rule \\
\code{Force} & \code{DROP} & Quoting/reflecting cancel \\
\end{tabular}
\end{center}

The CBV strategy is encoded syntactically: $\beta$-reduction requires both the
function and argument to be wrapped in \code{Inject(-)}, ensuring arguments
are evaluated to values before substitution.

Key proven results:
\begin{itemize}
\item \code{tinyML\_crossings}: exactly 2 sort-crossing constructors
  (\code{Inject}, \code{Thunk}) -- \code{native\_decide}.
\item \code{thunk\_is\_quoting}: \code{Thunk} classified as quoting; typing action = $\Dia$.
\item \code{inject\_is\_reflecting}: \code{Inject} classified as reflecting; typing action = $\Bx$.
\item \code{tinyML\_typing\_action\_galois}: $\Dia \dashv \Bx$ for TinyML typing actions.
\item 11 executable demos including a 3-step reduction chain
  (\code{force} $\to$ \code{ifTrue} $\to$ \code{fstPair}).
\end{itemize}

\section{Verified Formula Checker}

The file \codepath{Formula.lean} provides a verified bounded model checker for
\OSLF{} formulas (582 lines, 0~sorries).  The formula type \code{OSLFFormula}
supports $\Dia$, $\Bx$, $\wedge$, $\vee$, $\implies$, $\top$, $\bot$,
and atomic predicates.

\begin{theorem}[Checker Soundness, 0 sorries]
If the checker returns \code{.sat} for formula $\varphi$ at term $p$,
then $p \models \varphi$ in the denotational semantics.
\end{theorem}

The checker supports:
\begin{itemize}
\item \code{checkWithPred}: checking with external predecessor functions,
  enabling bounded $\Bx$ verification;
\item \code{aggregateBox}: universal checking of $\Bx\varphi$ over a
  predecessor list, with proven soundness (\code{aggregateBox\_sat}).
\end{itemize}

\section{Categorical Lift}

The file \codepath{Framework/CategoryBridge.lean} lifts the Set-level \OSLF{}
construction to Mathlib's categorical infrastructure (247~lines, 0~sorries).

\subsection{Modal Adjunction}

The predicate type \code{Pattern $\to$ Prop} carries conflicting category
instances (\code{Preorder.smallCategory} vs.\ \code{CategoryTheory.Pi}).
We introduce a type wrapper \code{PredLattice} to disambiguate, then lift:

\begin{lstlisting}[caption={Modal adjunction (CategoryBridge.lean)}]
noncomputable def langModalAdjunction (lang : LanguageDef) :
    (langGaloisL lang).monotone_l.functor |-
    (langGaloisL lang).monotone_u.functor :=
  (langGaloisL lang).adjunction
\end{lstlisting}

This provides a categorical \code{Adjunction} between the $\Dia$ and $\Bx$
endofunctors on the predicate preorder category, for any \code{LanguageDef}.

\section{Status and Roadmap}

\subsection{Milestones}

\begin{enumerate}
\item[\checkmark] \textbf{Milestone 1}: Executable \rhocalc{}-calculus engine.
  \code{reduceStep} computes all one-step reducts; \code{reduceStep\_sound} proven
  with 0~sorries; 6 executable tests pass.

\item[\checkmark] \textbf{Milestone 2}: Generic \MeTTaIL{} framework.
  Language-parametric pattern matching (\code{Match.lean}) and rewriting
  (\code{MeTTaIL/Engine.lean}) for any \code{LanguageDef}.
  8-test agreement suite confirms equivalence with specialized engine.

\item[\checkmark] \textbf{Milestone 3}: \OSLF{} type synthesis.
  \code{langOSLF} mechanically generates an \code{OSLFTypeSystem} from any
  \code{LanguageDef} with auto-proven Galois connection.  \code{GenHasType}
  provides generated typing rules.  Three-layer bridge established.

\item[\checkmark] \textbf{Milestone 4}: Correctness infrastructure.
  Totalized \code{Match.lean} (no \code{partial def} in core matching).
  Declarative \code{DeclReduces} with engine soundness and completeness.
  Enhanced formula checker with predecessor-based $\Bx$ checking.

\item[\checkmark] \textbf{Milestone 5}: Language instances.
  Four languages---lambda calculus, Petri nets, \rhocalc{}-calculus, TinyML---each
  with auto-generated \code{OSLFTypeSystem} and Galois connection.

\item[\checkmark] \textbf{Milestone 6}: Categorical lift.
  \code{langModalAdjunction}: Galois connection $\to$ Mathlib \code{Adjunction}.
  \code{predFibration}: \code{SubobjectFibration} over \code{Discrete R.Sorts}.
  Unconditional specialized engine bridges (\code{specialized\_possibly}, etc.).

\item[\checkmark] \textbf{Milestone 7}: Constructor category and GSLT bridge.
  \code{ConstructorCategory}: free category on sort-crossing constructors.
  \code{ConstructorFibration}: \code{SubobjectFibration} + \code{ChangeOfBase}
  with proven adjunctions.
  \code{ModalEquivalence}: constructor change-of-base $\leftrightarrow$ modalities.
  \code{DerivedTyping}: typing rules derived from constructor classification.
  \code{BeckChevalleyOSLF}: substitution as change-of-base, with counterexample.
  All 0~sorries.
\end{enumerate}

\section{Conclusion}

We have presented the first machine-checked formalization of the \OSLF{} algorithm,
spanning 22{,}300+ lines of Lean~4 across 58 files with 0~sorries in the core pipeline.
The formalization demonstrates that the entire pipeline from rewrite rules to
spatial-behavioral types can be made rigorous in a modern proof assistant.

The Galois connection at the heart of \OSLF{} is proven at three levels:
(1)~directly for the \rhocalc{}-calculus, (2)~categorically for any reduction span via
adjoint composition, and (3)~lifted to a Mathlib \code{Adjunction} between endofunctors
on the predicate preorder category.  All three levels are shown to agree.

The constructor category (Section~\ref{sec:constructor-cat}) provides a genuine
categorical backbone: sort-crossing constructors become morphisms, the fibered
change-of-base gives the adjoint triple $\exists_f \dashv f^* \dashv \forall_f$
for each constructor, and the modal operators $\Dia$/$\Bx$ are shown to be
the typing actions of quoting/reflecting constructors.  The derived typing rules
(Section~\ref{sec:derived-typing}) make explicit that the assignment
$\code{NQuote} \mapsto \Dia$, $\code{PDrop} \mapsto \Bx$ is not ad hoc but
follows from the constructor's position in the category.  The Beck--Chevalley
analysis (Section~\ref{sec:beck-chevalley}) shows that while substitution
commutes with change-of-base in the COMM-specific case (proven via
\code{comm\_preserves\_type}), the universal condition fails---a mathematically
interesting negative result proven by explicit counterexample.

The full type synthesis pipeline is validated by four language instances
(\rhocalc{}-calculus, lambda calculus, Petri nets, and TinyML), each with
auto-proven Galois connections.  TinyML demonstrates the framework's ability
to handle multi-sort CBV languages with binders, conditionals, and
sort-crossing constructors, mirroring the \rhocalc{}-calculus's
NQuote/PDrop structure with Thunk/Inject.

A declarative reduction relation provides an engine-independent specification,
and the executable engines are proven both sound and complete with respect to it.
All seven milestones are complete with 0~sorries in the core pipeline.

\bibliographystyle{plain}
\bibliography{references}

\end{document}
