%% Native Grammatical Formalism: GF x OSLF in Lean 4
%%
%% Render with:
%%   pdflatex native-grammatical-formalism && bibtex native-grammatical-formalism && pdflatex native-grammatical-formalism && pdflatex native-grammatical-formalism

\documentclass[11pt]{article}
\usepackage[margin=1in]{geometry}
\usepackage{url}
\usepackage[hyperindex,breaklinks]{hyperref}
\usepackage{listings}
\lstset{
  basicstyle=\ttfamily\small,
  breaklines=true,
  frame=single,
  columns=fullflexible,
  keepspaces=true,
  inputencoding=utf8,
  extendedchars=true,
  literate={→}{$\to$}1 {↔}{$\leftrightarrow$}1 {∧}{$\land$}1 {∨}{$\lor$}1
           {∀}{$\forall$}1 {∃}{$\exists$}1 {⊨}{$\models$}1
           {≠}{$\neq$}1 {¬}{$\lnot$}1 {⊣}{$\dashv$}1
           {π}{$\pi$}1 {σ}{$\sigma$}1 {φ}{$\phi$}1 {ψ}{$\psi$}1
           {∈}{$\in$}1 {⊓}{$\sqcap$}1 {⨅}{$\bigsqcap$}1 {⨆}{$\bigsqcup$}1
           {⟨}{$\langle$}1 {⟩}{$\rangle$}1
}
\usepackage{float}
\restylefloat{table}
\usepackage{longtable}
\usepackage{booktabs}
\usepackage{graphicx}
\usepackage[font=small,labelfont=bf]{caption}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{stmaryrd}
\usepackage{tikz}
\usetikzlibrary{arrows.meta,positioning}

\newtheorem{theorem}{Theorem}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{definition}[theorem]{Definition}

\newcommand{\code}[1]{\texttt{#1}}
\newcommand{\Dia}{\ensuremath{\diamondsuit}}
\newcommand{\Bx}{\ensuremath{\blacksquare}}
\newcommand{\OSLF}{\textup{OSLF}}
\newcommand{\GF}{\textup{GF}}
\newcommand{\RGL}{\textup{RGL}}
\newcommand{\NTT}{\textup{NTT}}
\newcommand{\MeTTaIL}{\textup{MeTTaIL}}

\title{Native Grammatical Formalism:\\
  Verified Multilingual Semantics via GF and \OSLF{} in Lean~4\\[4pt]
  \large\texttt{DRAFT --- \today}}
\author{Zar \and Oru\v{z}i\footnote{``Oru\v{z}i'' denotes AI-assisted collaboration using ChatGPT, Codex, and Claude Code.}}

\begin{document}
\maketitle

\begin{abstract}
We present a comprehensive Lean~4 formalization connecting the Grammatical
Framework (\GF{}) to Operational Semantics in Logical Form (\OSLF{}), yielding
verified multilingual semantics for natural language grammars.  Our
formalization covers: (i)~the \GF{} Resource Grammar Library core abstract
syntax (985~function signatures, 112~categories); (ii)~substantial English and Czech concrete
grammar subsets---English with morphology, clause construction, and linearization
for ${\sim}284$ abstract functions; Czech with 14~declension paradigms,
adjective/verb morphology, and agreement, but not yet sentence-level linearization; (iii)~an automatic bridge
from \GF{} grammars to \OSLF{} type systems with proven Galois connections
$\Dia \dashv \Bx$; (iv)~a three-layer semantic pipeline: internal syntax
rewrites, temporal policy, and visible semantic rules for scope,
binding, and anaphora; and (v)~a highlighted semantic spine of
120+ paper-facing theorems covering scope completeness via linear
extensions, pronoun-binding locality, store-to-environment invariance,
evidence-valued world-model semantics, lexical compositional monotonicity,
and garden-path disambiguation within a
broader \GF{} Lean subtree currently containing 664 theorem/lemma declarations.
The broader \GF{} subtree spans 43,162 lines across 91~Lean files, with
\textbf{zero \code{sorry} declarations} and zero axiom/opaque declarations.
\end{abstract}

\section*{What Is Formally Proved}
\begin{center}
\small
\begin{tabular}{@{}p{0.26\linewidth}p{0.22\linewidth}p{0.14\linewidth}p{0.28\linewidth}@{}}
\toprule
\textbf{Lean declaration} & \textbf{Paper claim} & \textbf{Status} & \textbf{Why it matters} \\
\midrule
\code{translation\_preserves\_type} &
Abstract-tree typing is translation-invariant &
\textbf{Proved (D)} &
Meaningful grammatical types live on shared abstract syntax, not on surface form. \\
\code{montague\_thesis} &
Language-parametric architectural invariance &
\textbf{Proved (D)} &
The grammar-to-semantics interface is correct by construction. \\
\code{adjModification\_preserves\_lexical} &
Compositional lexical-containment monotonicity &
\textbf{Proved (Sem)} &
Adjectival modification preserves previously present lexical material. \\
\code{garden\_path\_lexical\_disambiguation} &
Ambiguous parses can differ in lexical content &
\textbf{Proved (Sem)} &
The system distinguishes genuine lexical ambiguity, not just constructor shape. \\
\code{translation\_preserves\_evidence\_allW} &
Evidence is translation-invariant in all world models &
\textbf{Proved (Sem)} &
Abstract-tree equality already propagates into evidence semantics. \\
\code{translation\_preserves\_strength\_allW} &
World-model strength is translation-invariant &
\textbf{Proved (Sem)} &
Behavioral and evidential semantics stay aligned across concrete languages. \\
\code{checker\_sat\_implies\_wm\_semantics\_general} &
Checker soundness lifts to WM semantics &
\textbf{Proved (Sem)} &
Executable checking is connected to denotational evidence-grounded semantics. \\
\code{checker\_sat\_atom\_implies\_wm\_query\_judgment} &
Checker success yields WM query judgment on atoms &
\textbf{Proved (Sem)} &
The bounded checker reaches the world-model judgment layer. \\
\code{scopeOrderingNTBridge} &
Visible scope ordering reaches the \NTT{} layer &
\textbf{Proved (Sem)} &
Quantified/discourse structure already feeds into native-type semantics. \\
\code{closedFormula\_envInvariant\_NT} &
Closed quantified formulas are environment-invariant &
\textbf{Proved (Sem)} &
The quantified layer has genuine semantic stability theorems. \\
\bottomrule
\end{tabular}
\end{center}
\noindent
The main paper highlights a coherent theorem spine of 120+ paper-facing results.
The broader \code{Mettapedia/Languages/GF} subtree currently contains 664
theorem/lemma declarations across 91~Lean files.

\tableofcontents

%% ===================================================================
\section{Introduction}
\label{sec:intro}

\subsection{Roadmap for Non-Experts}

This paper connects four layers that are often studied separately:

\begin{enumerate}
\item \textbf{\GF{} (grammar layer).}  \GF{} provides typed abstract syntax trees
  (language-independent structure) plus concrete linearizations (surface forms
  in English, Czech, etc.).
\item \textbf{\OSLF{} (behavioral logic layer).}  Given rewrite rules on those
  trees, \OSLF{} derives modal operators and a type/predicate logic of what
  terms can reduce to.
\item \textbf{\NTT{}/GSLT (categorical semantics layer).}  The same construction
  is interpreted in the presheaf/topos viewpoint: predicates become fibers,
  substitution is functorial, and modal operators arise from adjoint structure.
\item \textbf{World-model/evidence layer.}  A stateful world model assigns
  evidence to queries; this grounds formula evaluation in evidence-valued
  semantics (\S\ref{sec:worldmodel}).
\end{enumerate}

In one line:
\[
\text{\GF{} tree} \to \text{pattern} \to \text{\OSLF{} formula}
\to \text{world model} \to \text{evidence} \to \text{truth}.
\]

The key distinction is between \emph{behavioral} semantics and
\emph{evidence-grounded} semantics.  \OSLF{} by itself gives a logic of
rewriting behavior; the world-model layer chooses how atomic predicates are
interpreted and supplies evidence values used by \code{semE}.

\paragraph{Roadmap.}  The paper proceeds from the \GF{} formalization
(\S\ref{sec:formalization}) to the \GF{}$\to$\OSLF{} bridge
(\S\ref{sec:bridge}), then to abstract-tree invariance
(\S\ref{sec:invariance}), evidence-grounded semantics
(\S\ref{sec:worldmodel}), the visible layer (\S\ref{sec:visible}), and
the interface/\NTT{} extensions (\S\ref{sec:extensions}).

Two independent research programs have developed powerful abstractions for
natural language: the Grammatical Framework (\GF{})~\cite{Ranta2004,Ranta2011}
provides language-independent abstract syntax with language-specific
linearization, while Native Type Theory (\NTT{})~\cite{WilliamsStay2021} and
\OSLF{}~\cite{MeredithStayOSLF2020} mechanically derive spatial-behavioral type systems from
rewrite rules.  This paper connects them: \textbf{every \GF{} grammar
automatically receives a verified \OSLF{} type system}.

The key insight is that \GF{}'s abstract syntax trees define a rewrite system
(via constructor-elimination rules like $\code{UseN}(x) \rightsquigarrow x$),
and \OSLF{} transforms any rewrite system into a type system equipped with
modal operators $\Dia$ (step-future) and $\Bx$ (step-past), connected by a
Galois adjunction $\Dia \dashv \Bx$.  The resulting type system lets us
\emph{prove} properties of natural language that linguists care about:

\begin{itemize}
\item \textbf{Scope completeness}: Reachable scope orders for a
  quantifier fragment are exactly the linear extensions of the
  dependency partial order.
\item \textbf{Pronoun-binding locality}: Binding a pronoun preserves
  the evidence of any formula that does not mention the bound pronoun free.
\item \textbf{Store-environment invariance}: Stores that agree on
  bind/ref membership produce identical variable environments.
\item \textbf{Evidence-grounded semantics}: Evidence-valued world-model
  semantics with checker soundness, cross-linguistic evidence invariance,
  and strength transport.
\item \textbf{Lexical compositional monotonicity}: Adding an adjective
  to a noun phrase preserves previously present lexical material
  (universally quantified, not just for concrete examples).
\item \textbf{Garden-path disambiguation}: The two parses of ``the old man
  the boats'' have provably different lexical profiles---not just different
  constructors, but different \emph{words}.
\end{itemize}

\paragraph{Scale.}  The broader \code{Mettapedia/Languages/GF} subtree currently
comprises 43,162 lines of Lean~4 across 91~files with
\textbf{zero \code{sorry} declarations} and zero axiom/opaque declarations.
The theorem spine highlighted in this paper is a 120+ result semantic slice
inside that larger development.

\paragraph{Contributions.}
\begin{enumerate}
\item To our knowledge, the first formalization of the \GF{} \RGL{} core abstract
  syntax in a proof assistant (985~function signatures, 112~categories).
\item Substantial English and Czech concrete grammar subsets (morphology,
  clause construction, linearization) with 80+ proven properties.
\item An automatic bridge from any \GF{} grammar to an \OSLF{} type system
  with proven Galois connection.
\item A three-layer semantic pipeline: internal syntax rewrites + temporal
  policy + visible rules for quantifier scope, discourse referents,
  and pronoun binding.  Key results include scope completeness via
  linear extensions, pronoun-binding locality, and store-to-environment
  invariance.
\item Evidence-valued world-model semantics with checker soundness and
  cross-linguistic evidence/strength invariance.
\item A highlighted semantic spine of 120+ paper-facing theorems,
  situated inside a broader \GF{} subtree with 664 theorem/lemma
  declarations and classified by claim strength in
  Table~\ref{tab:claims}.
\end{enumerate}

\begin{table}[t]
\centering
\caption{Claim strength classification.  \textbf{D}~=~definitional
  (holds by construction, proof is \code{rfl}/\code{Iff.rfl});
  \textbf{S}~=~structural (follows from inductive structure, proved by
  \code{simp}/\code{decide});
  \textbf{B}~=~behavioral (depends on specific rewrite rules);
  \textbf{Sem}~=~semantic (connects linguistic content to formal
  properties).  ``$\forall$'' marks universally quantified theorems.}
\label{tab:claims}
\small
\begin{tabular}{@{}llcl@{}}
\toprule
\textbf{Theorem family} & \textbf{Strength} & $\forall$? & \textbf{Proof method} \\
\midrule
Galois connection $\Dia \dashv \Bx$        & D  &         & construction \\
Montague thesis (cross-ling.\ invariance)  & D  & $\checkmark$ & \code{Iff.rfl} \\
Sort assignment (\code{nounSort}, etc.)    & D  &         & \code{rfl} \\
Frege compositionality                     & S  & $\checkmark$ & \code{simp} \\
Constructor exclusivity / partition        & S  & $\checkmark$ & \code{intro}/\code{decide} \\
Parse disambiguation (type-level)          & S  &         & \code{simp} \\
Lexical entailment (concrete)              & Sem &        & \code{simp} \\
Lexical compositional monotonicity          & Sem & $\checkmark$ & \code{unfold}/\code{simp} \\
General compositionality                   & Sem & $\checkmark$ & induction \\
Garden-path lexical disambiguation         & Sem &        & \code{simp} \\
Subcategorization as entailment            & Sem &        & \code{simp} \\
Czech syncretism invariants (25+)          & S  & $\checkmark$ & \code{decide}/\code{simp} \\
\midrule
\multicolumn{4}{@{}l}{\textit{World-model and visible-layer theorems (\S\ref{sec:worldmodel}, \S\ref{sec:visible})}} \\
\midrule
World-model grounding (67 thms)            & Sem & $\checkmark$ & induction/\code{simp} \\
Temporal policy laws                       & S  & $\checkmark$ & \code{simp}/construction \\
Store monotonicity (scope/binding steps)   & S  & $\checkmark$ & \code{cases}/\code{Multiset.le\_add} \\
Scope choice nondeterminism                & S  & $\checkmark$ & construction \\
Independent store commutation              & S  & $\checkmark$ & \code{add\_comm} \\
\bottomrule
\end{tabular}
\end{table}

%% ===================================================================
\section{Background}
\label{sec:background}

\subsection{Grammatical Framework}

The Grammatical Framework~\cite{Ranta2004,Ranta2011} separates natural
language description into two layers:

\begin{itemize}
\item \textbf{Abstract syntax}: A typed signature of \emph{categories}
  (grammatical types like \code{N}, \code{CN}, \code{NP}, \code{VP},
  \code{Cl}) and \emph{functions} (grammar rules like
  $\code{DetCN} : \code{Det} \to \code{CN} \to \code{NP}$).  Abstract
  syntax is \emph{language-independent}.
\item \textbf{Concrete syntax}: Language-specific linearization rules mapping
  abstract trees to surface strings, parameterized by morphological features
  (case, number, gender, tense, etc.).
\end{itemize}

The \GF{} Resource Grammar Library (\RGL{}) provides a shared abstract syntax
with concrete grammars for 30+ languages.  A sentence like ``the cat sleeps''
is represented as:
\[
  \code{PredVP}(\code{DetCN}(\code{the\_Det},\, \code{UseN}(\code{cat})),\,
  \code{UseV}(\code{sleep}))
\]
The same tree linearizes to English ``the cat sleeps'' or Czech
``ko\v{c}ka sp\'{i}''.

\subsection{OSLF and Native Type Theory}

The \OSLF{} algorithm~\cite{MeredithStayOSLF2020} takes a \emph{language
definition}---a set of sorts, constructors, equations, and directed rewrite
rules---and produces a type system.  The key construction:

\begin{enumerate}
\item A \textbf{rewrite system} defines one-step reduction $p \rightsquigarrow q$.
\item The reduction graph forms a \textbf{span}:
  $X \xleftarrow{\mathrm{src}} E \xrightarrow{\mathrm{tgt}} X$
  where $E$ is the set of reduction edges.
\item \textbf{Change-of-base functors} on predicate lattices yield:
  \begin{align*}
    \Dia\varphi(p) &\iff \exists q.\; p \rightsquigarrow q \;\wedge\; \varphi(q)
    & \text{(step-future)} \\
    \Bx\varphi(p) &\iff \forall q.\; q \rightsquigarrow p \;\Rightarrow\; \varphi(q)
    & \text{(step-past)}
  \end{align*}
\item The adjoint composition gives a \textbf{Galois connection}
  $\Dia \dashv \Bx$ automatically.
\end{enumerate}

A \emph{native type}~\cite{WilliamsStay2021} is a pair $(\text{sort},
\text{predicate})$ where predicates at each sort form a \emph{Frame}
(complete Heyting algebra).

\subsection{Lean 4}

Lean~4~\cite{Moura2021} is a dependently-typed proof assistant and
programming language.  We use Mathlib~\cite{mathlib2020} for
\code{GaloisConnection}, \code{Order.Frame}, and lattice infrastructure.
Proofs use \code{decide} (kernel evaluation), \code{simp} (equational
rewriting), and \code{rfl} (definitional equality).

%% ===================================================================
\section{GF Formalization in Lean 4}
\label{sec:formalization}
\label{sec:gf}

\subsection{Core Abstractions}

The core layer (\code{Core.lean}, 267~lines) defines:

\begin{lstlisting}[language=ml,caption={Core GF types}]
inductive Category where
  | base : String → Category
  | arrow : Category → Category → Category

structure FunctionSig where
  name : String
  type : Category

inductive AbstractNode where
  | leaf : String → Category → AbstractNode
  | apply : FunctionSig → List AbstractNode → AbstractNode
\end{lstlisting}

Categories are the types of \GF{}: \code{base "N"} for nouns, \code{arrow
(base "Det") (arrow (base "CN") (base "NP"))} for determiners.
\code{AbstractNode} represents abstract syntax trees.

\subsection{Abstract Syntax: 985 Functions}

\code{Abstract.lean} (1,552~lines) defines the complete modeled \GF{} \RGL{} core abstract
syntax as Lean definitions.  Table~\ref{tab:rgl} shows the module breakdown.

\begin{table}[ht]
\centering
\caption{GF RGL module breakdown}
\label{tab:rgl}
\begin{tabular}{@{}lrl@{}}
\toprule
\textbf{Module} & \textbf{\#Functions} & \textbf{Key examples} \\
\midrule
Noun      & 42 & \code{DetCN}, \code{UseN}, \code{AdjCN}, \code{RelCN} \\
Verb      & 29 & \code{UseV}, \code{ComplSlash}, \code{SlashV2a}, \code{PassV2} \\
Adjective & 11 & \code{PositA}, \code{ComparA}, \code{AdjOrd} \\
Sentence  & 19 & \code{PredVP}, \code{UseCl}, \code{SSubjS} \\
Question  & 17 & \code{QuestCl}, \code{QuestVP}, \code{QuestSlash} \\
Relative  &  5 & \code{RelCl}, \code{RelVP}, \code{RelSlash} \\
Phrase    & 19 & \code{PhrUtt}, \code{UttS}, \code{UttNP} \\
Conjunction & 27 & \code{ConjNP}, \code{ConjAP}, \code{BaseS}/\code{ConsS} \\
Tense     &  9 & \code{TTAnt}, \code{TPres}, \code{TPast} \\
Structural & 50+ & \code{and\_Conj}, \code{every\_Det}, \code{he\_Pron} \\
Other     & 700+ & Numerals, Idiom, Extra, Lexicon \\
\midrule
\textbf{Total} & \textbf{985} & \\
\bottomrule
\end{tabular}
\end{table}

\subsection{English Concrete Grammar}

The English grammar (20~files, 5,890~lines) linearizes approximately 284 of
the 985~abstract functions---covering the core Noun, Verb, Adjective,
Sentence, Question, Relative, Phrase, Conjunction, Adverb, and Tense
modules but not yet the full Extend, Construction, or Symbol modules.
The runtime linearization layer is split into a core API
(\code{Linearization.lean}), typed domain helpers
(\code{Linearization/Types.lean}), composition helpers
(\code{Linearization/Compose.lean}), rendering
(\code{Linearization/Render.lean}), witness proofs
(\code{Linearization/Witnesses.lean}), and coverage accounting
(\code{Linearization/Coverage.lean}).
The grammar implements:

\paragraph{Noun morphology.}  Three paradigms: \emph{regular} (8 pluralization
rules handling sibilants, consonant+y, final~-o, etc.), \emph{irregular}
(man$\to$men, child$\to$children, etc., 8~defined), and \emph{compound}
(baby boom$\to$baby booms).

\paragraph{Verb conjugation.}  Five base forms (infinitive, 3sg present,
past, past participle, present participle) with 12~irregular verbs and
3~auxiliary paradigms (be/have/do with contracted negation forms).

\paragraph{Sentence construction.}  16 tense/aspect/polarity combinations
($4 \times 2 \times 2$: present/past/future/conditional $\times$
simple/perfect $\times$ positive/negative), with do-support for questions
and negation, and relative clause formation.

\paragraph{Adjective comparison.}  Three strategies: synthetic
(big$\to$bigger$\to$biggest), analytic (beautiful$\to$more
beautiful$\to$most beautiful), and irregular (good$\to$better$\to$best).

\subsection{Czech Concrete Grammar}

The Czech grammar (12~files, 2,865~lines) formalizes the morphological
layer---declension, conjugation, and agreement---but does not yet include
sentence-level linearization comparable to the English grammar.
It implements:

\paragraph{Declension system.}  14~complete noun paradigms across 4~genders
(masculine animate/inanimate, feminine, neuter) and 7~cases (nominative,
genitive, dative, accusative, vocative, locative, instrumental).
Table~\ref{tab:czech} summarizes the syncretism analysis.

\begin{table}[ht]
\centering
\caption{Czech noun paradigms: syncretism analysis (14 theoretical slots)}
\label{tab:czech}
\small
\begin{tabular}{@{}llccc@{}}
\toprule
\textbf{Paradigm} & \textbf{Gender} & \textbf{Distinct} & \textbf{Compression} & \textbf{Key syncretism} \\
\midrule
p\'{a}n     & Masc.Anim.  & 10 & 29\% & Gen.Sg = Acc.Sg \\
p\v{r}edseda & Masc.Anim. & 10 & 29\% & Nom.Sg $\neq$ Acc.Sg \\
soudce      & Masc.Anim.  &  6 & 57\% & Nom=Gen=Acc=Voc (Sg) \\
mu\v{z}     & Masc.Anim.  &  7 & 50\% & Soft declension \\
hrad        & Masc.Inan.  &  8 & 43\% & Nom.Sg = Acc.Sg \\
stroj       & Masc.Inan.  &  7 & 50\% & Soft declension \\
\v{z}ena    & Fem.        & 10 & 29\% & Gen.Pl = bare stem \\
r\r{u}\v{z}e & Fem.       &  6 & 57\% & Soft, high syncretism \\
p\'{i}se\v{n} & Fem.      &  7 & 50\% & Gen=Dat=Voc=Loc (Sg) \\
kost        & Fem.        &  6 & 57\% & Gen=Dat=Voc=Loc = -i \\
m\v{e}sto   & Neut.       &  8 & 43\% & Nom=Acc=Voc \\
ku\v{r}e    & Neut.       &  9 & 36\% & Pl. -ata paradigm \\
mo\v{r}e    & Neut.       &  6 & 57\% & Maximal syncretism \\
staven\'{i} & Neut.       &  4 & 71\% & Near-invariant \\
\bottomrule
\end{tabular}
\end{table}

\paragraph{Phonological rules.}  Palatalization (k$\to$c, h$\to$z,
ch$\to$\v{s}, r$\to$\v{r} before~-i), vowel shortening in vocatives
(p\'{a}n$\to$pane), and fleeting~`e' (pes$\to$psa).

\paragraph{Adjective paradigms.}  Five patterns: hard (-\'{y}), soft (-\'{i}),
feminine possessive (-in), masculine possessive (-\r{u}v), and invariable.

\paragraph{Agreement.}  Numeral-noun agreement with three size classes:
Num1 (singular), Num2\_4 (plural), Num5+ (genitive plural with neuter
singular verb agreement).

\paragraph{Proven properties.}  25+ universally quantified syncretism
invariants (valid for \emph{all} lemmas, not just test data), exact distinct
form counts for all 14~paradigms, and phonological regression tests.

%% ===================================================================
\section{The OSLF Bridge}
\label{sec:bridge}

\subsection{GF as a Language Definition}

The bridge (\code{OSLFBridge.lean}, 581~lines) converts \GF{} abstract syntax
into an \OSLF{} \code{LanguageDef}.  The key transformation:

\begin{lstlisting}[language=ml,caption={Abstract tree to pattern conversion}]
def gfAbstractToPattern : AbstractNode → Pattern
  | .leaf name _ => .fvar name
  | .apply f args => .apply f.name (args.map gfAbstractToPattern)
\end{lstlisting}

Each \code{FunctionSig} becomes a \code{GrammarRule} with fresh argument
names (\code{arg0}, \code{arg1}, \ldots).  The full \GF{} \RGL{} becomes
\code{gfRGLLanguageDef} with 169+ core grammar rules.

\subsection{Rewrite Rules: Non-Vacuous Modalities}

The bridge defines two families of directed rewrite rules.

\paragraph{Identity-wrapper eliminations} (\code{allIdentityRewrites}).
Six rules eliminate constructor wrappers that carry no independent semantic
content:
\begin{align*}
  \code{UseN}(x) &\rightsquigarrow x & \text{(noun identity)} \\
  \code{PositA}(x) &\rightsquigarrow x & \text{(positive adjective)} \\
  \code{UseV}(x) &\rightsquigarrow x & \text{(verb identity)} \\
  \code{UseComp}(x) &\rightsquigarrow x & \text{(complement identity)} \\
  \code{UseN2}(x) &\rightsquigarrow x & \text{(relational noun identity)} \\
  \code{UseA2}(x) &\rightsquigarrow x & \text{(relational adjective identity)}
\end{align*}
These give $\Dia$ and $\Bx$ non-vacuous behavioral content:
$\Dia(\text{is\_cat})$ holds for $\code{UseN}(\code{cat})$ because it reduces
to \code{cat}, but not for bare \code{cat} (which is irreducible).

\paragraph{Semantic entailment rewrites} (\code{allSemanticRewrites}).
Four rules capture genuine entailment or tense-grounding:
\begin{align*}
  \code{PredVP}(np_1,\code{ComplSlash}(\code{SlashV2a}(v),np_2))
    &\rightsquigarrow \code{PredVP}(np_2,\code{PassV2}(v))
    & \text{(active$\to$passive)} \\
  \code{UseCl}(\code{TTAnt}(\code{TPres},\code{ASimul}),\code{PPos},cl)
    &\rightsquigarrow \mathtt{{\otimes}temporal}(cl,0)
    & \text{(present)} \\
  \code{UseCl}(\code{TTAnt}(\code{TPast},\code{ASimul}),\code{PPos},cl)
    &\rightsquigarrow \mathtt{{\otimes}temporal}(cl,-1)
    & \text{(past)} \\
  \code{UseCl}(\code{TTAnt}(\code{TFut},\code{ASimul}),\code{PPos},cl)
    &\rightsquigarrow \mathtt{{\otimes}temporal}(cl,1)
    & \text{(future)}
\end{align*}
The active$\to$passive rule holds in one direction only.  Tense rewrites tag
clauses with a temporal offset ($0 =$ present, $-1 =$ past, $1 =$ future)
for the evidence-semantics layer.

\subsection{Automatic Galois Connection}

The \OSLF{} pipeline produces a type system with zero manual proof effort:
\begin{lstlisting}[language=ml,caption={Automatic type system construction}]
-- Zero-effort Galois connection
theorem langGalois (lang : LanguageDef) :
    GaloisConnection (langDiamond lang) (langBox lang)
\end{lstlisting}

This is a \emph{design theorem}: the proof follows automatically from the
composition of adjoint functors $\exists_{\text{src}} \dashv
\text{src}^* \dashv \forall_{\text{src}}$ applied to the reduction span.

\begin{figure}[ht]
\centering
\begin{tikzpicture}[>=Stealth, node distance=2cm]
  \node (an) {\code{AbstractNode}};
  \node (pat) [right=2.5cm of an] {\code{Pattern}};
  \node (ld) [right=2.5cm of pat] {\code{LanguageDef}};
  \node (rs) [below=1.5cm of ld] {\code{RewriteSystem}};
  \node (span) [left=2.5cm of rs] {Reduction Span};
  \node (oslf) [left=2.5cm of span] {$\Dia \dashv \Bx$ + Galois};
  \draw[->] (an) -- node[above]{\small\code{gfAbstractToPattern}} (pat);
  \draw[->] (pat) -- node[above]{\small + rules} (ld);
  \draw[->] (ld) -- node[right]{\small\code{langRewriteSystem}} (rs);
  \draw[->] (rs) -- node[above]{\small\code{langSpan}} (span);
  \draw[->] (span) -- node[above]{\small\code{langOSLF}} (oslf);
\end{tikzpicture}
\caption{The GF $\to$ OSLF pipeline}
\label{fig:pipeline}
\end{figure}

\subsection{Type-Checking Layer}

The typing layer (\code{Typing.lean}, 592~lines) builds native types from
\GF{} constructors:

\begin{lstlisting}[language=ml,caption={Constructor-based native types}]
def gfConstructorCheck (label : String) : Pattern → Bool
  | .apply f _ => f == label
  | _ => false

def gfConstructorType (sort : String) (label : String) :
    GFNativeType :=
  { sort := sort
    pred := fun p => gfConstructorCheck label p = true }
\end{lstlisting}

This yields 38~theorems, including:

\begin{theorem}[Constructor exclusivity]
For any pattern $p$ and distinct labels $l_1 \neq l_2$, if $p$ passes the
constructor check for~$l_1$, it fails for~$l_2$.
\end{theorem}

\begin{theorem}[Parse disambiguation]
The two abstract trees for ``the old man \ldots'' satisfy different native
type predicates: Parse~1's CN subtree satisfies \code{AdjCN}-type but not
\code{UseN}-type; Parse~2's CN subtree satisfies \code{UseN}-type but not
\code{AdjCN}-type.
\end{theorem}

\begin{theorem}[Frege compositionality]
If two applications use the same function signature and their argument lists
produce the same patterns, then their \OSLF{} patterns are identical.
Formally, equal argument patterns imply equal result patterns.
\end{theorem}

%% ===================================================================
\section{Lexical and Structural Invariance Theorems}
\label{sec:invariance}

This section presents warm-up theorems about lexical containment and
structural invariance (\code{LinguisticInvariance.lean},
362~lines, 15~theorems).  These results demonstrate that the
\GF{}$\to$\OSLF{} bridge carries genuine linguistic content; the deeper
semantic results (scope completeness, pronoun-binding locality,
store-environment invariance) follow in \S\ref{sec:visible} and
\S\ref{sec:extensions}.

\subsection{Lexical Containment}

We define a recursive predicate testing whether a free variable name
appears anywhere in a pattern tree:

\begin{lstlisting}[language=ml,caption={Recursive lexical containment}]
def containsLexical (name : String) : Pattern → Bool
  | .fvar n => n == name
  | .apply _ args => containsLexical.go name args
  | .bvar _ => false
  | .lambda body => containsLexical name body
  | _ => false  -- other cases
where
  go (name : String) : List Pattern → Bool
    | [] => false
    | p :: ps => containsLexical name p || go name ps
\end{lstlisting}

This predicate is more expressive than $\Dia$ for linguistic purposes:
$\Dia$ reaches only \emph{one-step} reducts ($\code{UseN}(\code{cat})
\rightsquigarrow \code{cat}$), while \code{containsLexical} reaches
\emph{arbitrarily deep} subterms.

\subsection{Lexical Entailment}

Concrete sentences entail the presence of specific lexical items:

\begin{lstlisting}[language=ml,caption={Lexical entailment (proved by \code{simp})}]
-- "the cat sleeps" entails "cat" and "sleep"
theorem catSleeps_entails :
    containsLexical "cat" (gfAbstractToPattern theCatSleeps) = true ∧
    containsLexical "sleep" (gfAbstractToPattern theCatSleeps) = true ∧
    containsLexical "dog" (gfAbstractToPattern theCatSleeps) = false
\end{lstlisting}

\subsection{Monotonicity of Modification}

The key compositional lexical-containment invariance theorem is:

\begin{lstlisting}[language=ml,caption={Monotonicity (universally quantified)}]
theorem adjModification_preserves_lexical
    (adjName : String) (cn : AbstractNode) (name : String)
    (h : containsLexical name (gfAbstractToPattern cn) = true) :
    containsLexical name
      (gfAbstractToPattern (addAdjModifier adjName cn)) = true
\end{lstlisting}

This says: for \emph{any} adjective and \emph{any} common noun, adding the
adjective preserves previously present lexical items under
\code{containsLexical}.  This is a \emph{lexical monotonicity} theorem
for adjectival modification---it establishes that symbol occurrence in
the abstract-tree-to-pattern encoding is compositionally monotone,
a Frege-style compositionality result at the lexical level~\cite{Frege1892}.
It does \emph{not} by itself imply full truth-conditional semantic
monotonicity in a world model; that stronger claim requires additional
semantic assumptions and is handled separately in the world-model semantics
layer.

A companion theorem proves the adjective is also \emph{added}:

\begin{lstlisting}[language=ml]
theorem adjModification_adds_adjective (adjName : String)
    (cn : AbstractNode) :
    containsLexical adjName
      (gfAbstractToPattern (addAdjModifier adjName cn)) = true
\end{lstlisting}

\subsection{General Compositionality}

The monotonicity principle generalizes to \emph{any} constructor:

\begin{lstlisting}[language=ml,caption={General compositionality}]
theorem constructor_preserves_lexical (fname : String)
    (ft : Category) (args : List AbstractNode)
    (name : String) (a : AbstractNode)
    (hmem : a ∈ args)
    (h : containsLexical name (gfAbstractToPattern a) = true) :
    containsLexical name
      (gfAbstractToPattern (.apply ⟨fname, ft⟩ args)) = true
\end{lstlisting}

Any \GF{} constructor that keeps its arguments as subterms preserves
their lexical content.  This is Frege's principle made formal: compound
expressions contain their parts.

\subsection{Garden-Path Disambiguation}

The sentence ``the old man the boats'' is structurally ambiguous:
\begin{itemize}
\item \textbf{Parse 1} (adjective reading): ``the [old man] walks''\\
  \code{PredVP(DetCN(the, AdjCN(PositA(old), UseN(man))), UseV(walk))}
\item \textbf{Parse 2} (verb reading): ``the old [man the boats]''\\
  \code{PredVP(DetCN(the, UseN(old)), ComplSlash(SlashV2a(man), DetCN(the, UseN(boat))))}
\end{itemize}

We prove that these parses have provably different \emph{lexical content}:

\begin{lstlisting}[language=ml,caption={Garden-path lexical disambiguation}]
theorem garden_path_lexical_disambiguation :
    -- Parse 1 has "walk" but not "boat"
    containsLexical "walk" (gfAbstractToPattern gp_parse1) = true ∧
    containsLexical "boat" (gfAbstractToPattern gp_parse1) = false ∧
    -- Parse 2 has "boat" but not "walk"
    containsLexical "boat" (gfAbstractToPattern gp_parse2) = true ∧
    containsLexical "walk" (gfAbstractToPattern gp_parse2) = false ∧
    -- Both share the common material
    containsLexical "old" (gfAbstractToPattern gp_parse1) = true ∧
    containsLexical "old" (gfAbstractToPattern gp_parse2) = true ∧
    containsLexical "man" (gfAbstractToPattern gp_parse1) = true ∧
    containsLexical "man" (gfAbstractToPattern gp_parse2) = true
\end{lstlisting}

This goes beyond constructor-type disambiguation: even though both parses
are clauses (\code{Cl}), their \emph{lexical profiles} provably differ.
The words ``walk'' and ``boat'' serve as semantic witnesses that distinguish
the readings.

\subsection{Cross-Linguistic Invariance}

\GF{}'s architecture guarantees that abstract syntax is language-independent.
We formalize this with cross-linguistic pairs:

\begin{lstlisting}[language=ml,caption={Cross-linguistic invariance}]
structure CrossLingPair where
  tree : AbstractNode
  englishSurface : String
  czechSurface : String

-- "the cat" / "kocka" share the same abstract tree
theorem cross_ling_surfaces_differ :
    theCat_pair.englishSurface ≠ theCat_pair.czechSurface

-- Same tree → same lexical containment
theorem cross_ling_lexical_invariance :
    containsLexical "cat"
      (gfAbstractToPattern theCat_pair.tree) = true
\end{lstlisting}

The cleanest architectural invariance endpoint is Montague's
thesis~\cite{Montague1973} formalized:

\begin{lstlisting}[language=ml]
theorem montague_thesis (tree : AbstractNode)
    (phi : Pattern → Prop) :
    phi (gfAbstractToPattern tree) ↔
    phi (gfAbstractToPattern tree) := Iff.rfl
\end{lstlisting}

The proof is \code{Iff.rfl}---definitional equality.

\begin{quote}
\textbf{Design Theorem Flag.}  The \code{Iff.rfl} proofs of
\code{translation\_preserves\_type} and \code{montague\_thesis} are
\emph{not} discovered semantic facts.  They are \textbf{design
certifications}: the proof is trivial \emph{because}
\code{gfAbstractToPattern} takes no language parameter.  The trivial
proof \emph{is} the content---it certifies that the architecture
guarantees type-preservation by construction.  A failed design (e.g.,
one where patterns depended on linearization) would make these
theorems non-trivial or false.
\end{quote}

\subsection{Subcategorization as Entailment}

The V/V2 distinction (intransitive/transitive verbs) has provable semantic
consequences:

\begin{lstlisting}[language=ml,caption={Subcategorization determines entailment}]
theorem subcat_entailment :
    -- V2: VP contains the object
    containsLexical "her" (gfAbstractToPattern vp_transitive) = true ∧
    -- V: VP does NOT contain any object
    containsLexical "her" (gfAbstractToPattern vp_intransitive) = false ∧
    -- Both contain their respective verbs
    containsLexical "love" (gfAbstractToPattern vp_transitive) = true ∧
    containsLexical "walk" (gfAbstractToPattern vp_intransitive) = true ∧
    -- But not each other's verbs
    containsLexical "walk" (gfAbstractToPattern vp_transitive) = false ∧
    containsLexical "love" (gfAbstractToPattern vp_intransitive) = false
\end{lstlisting}

This is the formal content of subcategorization: the argument structure of a
verb \emph{determines} what entities are semantically accessible in the VP.

%% ===================================================================
\section{World-Model Semantics}
\label{sec:worldmodel}

The behavioral semantics of \S\ref{sec:invariance} captures structural
properties (lexical containment, reduction accessibility) but not
\emph{grounded} meaning.  \code{WorldModelSemantics.lean} (1,352~lines,
70~theorems) now already implements both tree-level denotation
\code{gfEvidenceDenote} and formula-level evidence semantics
\code{gfWMFormulaSemE}; the threshold-Prop semantics \code{gfWMFormulaSem}
is retained as a projection for checker soundness rather than as the primary
semantic layer.

\subsection{Primary Evidence-Valued Semantics}

The core definition assigns each \OSLF{} formula an evidence value (from a
complete lattice) at each pattern:

\begin{lstlisting}[language=ml,caption={Evidence-valued formula semantics}]
noncomputable def semE (R : Pattern → Pattern → Prop)
    (I : AtomSem) (φ : OSLFFormula) (p : Pattern) : E :=
  match φ with
  | .atom a   => I a p              -- atom interpretation
  | .and φ ψ  => semE R I φ p ⊓ semE R I ψ p
  | .dia φ    => ⨆ q ∈ {q | R p q}, semE R I φ q
  | .box φ    => ⨅ q ∈ {q | R q p}, semE R I φ q
  | ...
\end{lstlisting}

At the tree level, denotation is
\[
  \texttt{gfEvidenceDenote}\; W\; t
  = \texttt{WorldModel.evidence}\; W\; (\texttt{gfAbstractToPattern}\; t).
\]
At the formula level, the primary endpoint is
\[
  \texttt{gfWMFormulaSemE}\; W\; \varphi\; p
  := \texttt{semE}(\texttt{langReduces}\;\texttt{gfRGLLanguageDef},
      \texttt{gfEvidenceAtomSemFromWM}\; W,\; \varphi,\; p).
\]
The paper-facing bridge theorems are already proved in Lean:
\code{translation\_preserves\_evidence\_allW},
\code{translation\_preserves\_strength\_allW},
\code{checker\_sat\_implies\_wm\_semantics\_general}, and
\code{checker\_sat\_atom\_implies\_wm\_query\_judgment}.

\subsection{Rewrite Rules: 10 Directed Rules}

The bridge defines 10~directed rewrite rules in two groups
(\code{allIdentityRewrites} + \code{allSemanticRewrites}):

\begin{enumerate}
\setlength{\itemsep}{0pt}
\item \code{UseN}$(x) \rightsquigarrow x$ (noun identity)
\item \code{PositA}$(x) \rightsquigarrow x$ (positive adjective)
\item \code{UseV}$(x) \rightsquigarrow x$ (verb identity)
\item \code{UseComp}$(x) \rightsquigarrow x$ (complement identity)
\item \code{UseN2}$(x) \rightsquigarrow x$ (relational noun identity)
\item \code{UseA2}$(x) \rightsquigarrow x$ (relational adjective identity)
\item Active$\to$passive: $\code{PredVP}(np_1, \code{ComplSlash}(\code{SlashV2a}(v), np_2)) \rightsquigarrow \code{PredVP}(np_2, \code{PassV2}(v))$
\item Present tense: $\code{UseCl}(\code{TTAnt}(\code{TPres},\code{ASimul}),\code{PPos},cl) \rightsquigarrow \mathtt{{\otimes}temporal}(cl,0)$
\item Past tense: \ldots $\rightsquigarrow \mathtt{{\otimes}temporal}(cl,-1)$
\item Future tense: \ldots $\rightsquigarrow \mathtt{{\otimes}temporal}(cl,1)$
\end{enumerate}

\subsection{Temporal Policy}

Temporal expressions like tense and aspect are handled by a
\code{TemporalPolicy} that governs temporal-node rewriting:

\begin{lstlisting}[language=ml,caption={Temporal policy (extensible)}]
inductive TemporalPolicy where
  | syntaxOnly                                    -- no temporal steps (default)
  | withStep (step : Pattern -> Pattern -> Prop)  -- user-supplied relation
\end{lstlisting}

The \code{TemporalStepLaws} record packages well-scopedness: temporal
steps only transform temporal nodes.  The trivial policy
\code{syntaxOnly} satisfies these laws automatically.

\subsection{Modal Properties}

The file proves antitone and monotone relationships between the modal
operators and the reduction relation:

\begin{lstlisting}[language=ml,caption={Box is antitone in R for modal-free formulas}]
theorem sem_antitone_box {R1 R2} (hR : ∀ p q, R2 p q → R1 p q)
    (I : AtomSem) {φ} (hmf : modalFree φ) {p}
    (h : sem R1 I (.box φ) p) : sem R2 I (.box φ) p
\end{lstlisting}

\noindent The \code{modalFree} predicate characterizes the fragment of
formulas without $\Dia$/$\Bx$ nesting, where semantics is independent of
the accessibility relation.

%% ===================================================================
\section{Scope and Binding Step Layer}
\label{sec:visible}

The central semantic extension implements an explicit \emph{scope-and-binding}
transition layer.
While internal syntax rewrites are silent (they simplify tree structure
without semantic choice), \emph{visible rules} make meaning-bearing
decisions: scope ordering, referent introduction, pronoun binding.
Our design is compatible with earlier state-based semantic proposals while
keeping neutral terminology.

\subsection{Grammar State}

The key insight is that state extends from a bare term to a
\emph{term plus semantic store}:

\begin{lstlisting}[language=ml,caption={Grammar state = term + store}]
inductive StoreAtom where
  | quant (q : String) (det : Pattern) (restr : Pattern)
  | scope (q1 q2 : String)    -- q1 scopes over q2
  | ref (r : String) (pos : Pattern)
  | bind (pr r : String)       -- pronoun pr → antecedent r
  deriving Repr, DecidableEq

structure GrammarState where
  term : Pattern
  store : Multiset StoreAtom
\end{lstlisting}

The store is a \code{Multiset} (resource-sensitive, order-independent),
which is a natural fit for linear-style resource tracking.

\subsection{Visible Step Relation}

Four meaning-bearing rules, parameterized by a grammar-independent
\code{VisibleCfg} that provides NP replacement:

\begin{itemize}
\item \textbf{Quantifier introduction}: At an NP position, introduce
  quantifier handle $q$, replace the NP subtree with $\code{NPVar}(q)$,
  and record $\code{Quant}(q, \text{det}, \text{restr})$ in the store.
\item \textbf{Scope choice}: Given two distinct quantifiers $q_1 \neq
  q_2$ with no relative ordering, commit to $q_1$ scoping over $q_2$.
  The alternative $\code{Scope}(q_2, q_1)$ is a separate nondeterministic
  step---this is where scope ambiguity lives.
\item \textbf{Referent introduction}: Introduce a discourse referent
  $r$ at a position, making it available as a potential antecedent.
\item \textbf{Pronoun binding}: Resolve pronoun $\code{pr}$ to an
  accessible antecedent referent $r$.  Precondition: $r$ must have been
  introduced via referent introduction.
\end{itemize}

\subsection{Combined Relation}

The full reduction composes all three layers:

\begin{lstlisting}[language=ml,caption={Three-layer combined relation}]
def gfReducesFull (cfg : VisibleCfg) (π : TemporalPolicy) :
    GrammarState → GrammarState → Prop :=
  fun s1 s2 =>
    -- Layer 1: internal syntax rewrites (store unchanged)
    (langReduces gfRGLLanguageDef s1.term s2.term ∧ s1.store = s2.store)
    -- Layer 2: temporal policy steps (store unchanged)
    ∨ (temporalStep π s1.term s2.term ∧ s1.store = s2.store)
    -- Layer 3: scope/binding semantic steps
    ∨ VisibleStep cfg s1 s2
\end{lstlisting}

\subsection{Structural Theorems}

Five theorems are proved about the visible layer (zero sorry):

\begin{theorem}[Store monotonicity]
Every visible step only adds atoms:
$\code{VisibleStep}\; s_1\; s_2 \implies s_1.\code{store} \le
s_2.\code{store}$.
\end{theorem}

\begin{theorem}[Scope nondeterminism]
Given two distinct unordered quantifiers $q_1, q_2$, both orderings
$\code{Scope}(q_1, q_2)$ and $\code{Scope}(q_2, q_1)$ are reachable
from the same state.
\end{theorem}

\begin{theorem}[Bind requires referent]
Pronoun binding requires a prior referent introduction:
the precondition $\exists p,\; \code{Ref}(r, p) \in s.\code{store}$
is part of the constructor.
\end{theorem}

\subsection{Independence and Commutativity}

Two store atoms are \emph{independent} when their rely footprints
(which resources they read/write) are disjoint.  Independent atoms commute:

\begin{lstlisting}[language=ml,caption={Independence via rely footprints}]
def independent (a1 a2 : StoreAtom) : Prop :=
  Disjoint (relyFootprint a1) (relyFootprint a2)

theorem independent_store_commute (σ : Multiset StoreAtom)
    (a1 a2 : StoreAtom) :
    σ + {a1} + {a2} = σ + {a2} + {a1}
\end{lstlisting}

\noindent Positive example: $\code{Quant}(q_1, \ldots)$ and
$\code{Ref}(r_1, \ldots)$ are independent (proved by \code{decide}).
Negative example: $\code{Scope}(q_1, q_2)$ and
$\code{Quant}(q_1, \ldots)$ are \emph{not} independent.

\subsection{GF-Specific Instance}

The abstract \code{NPReplacer} interface is instantiated for \GF{}
(\code{VisibleLayerGFInstance.lean}, 97~lines):

\begin{lstlisting}[language=ml,caption={GF NP recognition}]
def isNPConstructor : Pattern → Bool
  | .apply "DetCN" [_, _]  => true
  | .apply "MassNP" [_]    => true
  | .apply "UsePN" [_]     => true
  | .apply "UsePron" [_]   => true
  | _                      => false
\end{lstlisting}

\noindent Verification examples confirm correct behavior on concrete
\GF{} trees (proved by \code{simp}/\code{decide}).

%% ===================================================================
\section{Discussion}
\label{sec:discussion}

\subsection{Design Theorems and Trivial Proofs}

Several of our theorems have trivial proofs (\code{rfl}, \code{Iff.rfl}).
We argue these are \emph{more} informative than complex proofs: they show
that the property holds \emph{by construction} rather than by accident.
Montague's thesis (\code{montague\_thesis}) is \code{Iff.rfl} because the
\GF{} architecture was \emph{designed} so that types live on abstract trees.
The trivial proof certifies the design.

\subsection{Proof Methods}

Proofs use sound kernel reduction throughout:

\begin{itemize}
\item \textbf{\code{simp}}: Equational rewriting with definitional unfolding.
  Used for all concrete lexical containment proofs and universally quantified
  monotonicity theorems.
\item \textbf{\code{decide}}: Kernel evaluation of decidable propositions.
  Used for string inequality, sort comparisons, independence examples, and
  simple Bool evaluations.
\item \textbf{\code{rfl}}: Definitional equality.  Used for design theorems
  and sort assignment.
\item \textbf{\code{native\_decide}}: No remaining uses in the checked
  \GF{} theorem modules.
\end{itemize}

\subsection{Scope and Limits}

It is important to be precise about what our formalization \emph{does} and
\emph{does not} provide.

\paragraph{What ``semantics'' means here.}  The predicates in
\S\ref{sec:invariance} (\code{containsLexical}, \code{gfConstructorCheck})
capture \emph{lexical containment} and \emph{reduction accessibility}---they
determine which words appear in a parse tree and which constructor
wrappers can be eliminated by rewriting.  The \OSLF{} types are
\emph{behavioral} (what reductions are available).  However, the
formalization already contains the substrate for genuine
\emph{evidence-valued} denotational semantics; \S\ref{sec:worldmodel}
describes the implemented main layer and \S\ref{app:evidence} gives
supplementary details.

\paragraph{Cross-linguistic invariance.}  Theorem~\code{montague\_thesis}
is an \emph{architectural} guarantee, not a deep semantic result.  It
holds because \code{gfAbstractToPattern} has no language parameter---the
invariance is designed in, not discovered.  This is a strength (the design
is \emph{certified} correct) but users should not read it as proving
semantic equivalence of natural-language translations in the deep sense.

\paragraph{Diamond proofs.}  The \OSLF{} rewrite engine
(\code{rewriteWithContextWithPremisesUsing}) is not kernel-reducible in
Lean~4, so \code{decide} cannot evaluate modal properties like
$\Dia(\text{is\_cat})(\code{UseN}(\code{cat}))$.  The runtime checker
confirms these properties (returning \code{.sat}), and checker soundness is
proved, but kernel-level proofs require restructuring the engine for
definitional reduction.

\paragraph{Visible layer coverage.}  The current visible rules formalize
quantifier scope and anaphora-binding steps.  Voice/event structure and
tense/aspect commitment are partially covered by the
existing \code{activePassiveRewrite} and \code{TemporalPolicy} but not
yet formalized as visible steps with store atoms.  Presupposition
projection remains open (Table~\ref{tab:negative}).

\paragraph{Rewrite coverage.}  The bridge defines 10~directed rewrite rules
(identity wrappers, active--passive, tense reduction) plus a temporal policy
layer and visible scope/binding rules.
\paragraph{Context-closure commutation bridge.}
In \code{SemanticKernelConfluence.lean},
the theorem \code{gf\_context\_commuteAt\_useN\_pastTense}
records the context-closure commutation result used by the
GF$\to$\OSLF{} bridge, and its companion promotes it into the
observable-kernel interface.  A negative result shows that
voice-change and noun-use do \emph{not} commute at the context level,
confirming that commutation is genuinely label-dependent.
Table~\ref{tab:negative} tracks the status of linguistic properties that
were originally out of scope.

\begin{table}[ht]
\centering
\caption{Linguistic properties: progress from 6-rule baseline.
  \textbf{Addressed} = formalized infrastructure exists (definitions +
  theorems); \textbf{Open} = not yet formalized.}
\label{tab:negative}
\small
\begin{tabular}{@{}llll@{}}
\toprule
\textbf{Property} & \textbf{Status} & \textbf{Module} & \textbf{Mechanism} \\
\midrule
Scope ambiguity            & Addressed & \code{VisibleLayer}     & quantifier introduction + scope choice \\
  (every/some interaction) &           &                         & nondeterministic ordering \\
Anaphora resolution        & Addressed & \code{VisibleLayer}     & referent introduction + pronoun binding \\
  (he $\to$ John)          &           &                         & store-based binding \\
Active--passive equivalence & Addressed & \code{WorldModelSem.}  & \code{activePassiveRewrite} \\
  (``X loves Y'' $\approx$ ``Y is loved by X'') & &             & rule 9 of 10 \\
Tense entailment           & Addressed & \code{WorldModelSem.}   & \code{TemporalPolicy} \\
  (past $\Rightarrow$ $\lnot$future) & &                        & + \code{TemporalStepLaws} \\
Presupposition projection  & Open      &                         & Multi-sentence / DRT rules \\
$\Dia(\text{is\_cat})(\code{UseN}(\code{cat}))$ & Partial & \code{Formula}/\code{OSLFBridge} & Checker-sound; kernel-reduction pending \\
  (modal kernel proof)     &           &                         & (\code{checkLangUsing\_sat\_sound} proved) \\
\bottomrule
\end{tabular}
\end{table}

\paragraph{Grammar coverage.}  Our English and Czech grammars cover core
morphology and syntax but not the full \RGL{} concrete syntax (which
includes thousands of lexical entries and morphological exceptions).

\subsection{Relation to Other Approaches}

Our work connects three traditions:

\begin{itemize}
\item \textbf{Montague grammar}~\cite{Montague1973}: We formalize the
  thesis that meaning is preserved under translation, but at the level of
  abstract syntax trees rather than model-theoretic denotations.
\item \textbf{Type-Theoretical Grammar}~\cite{Ranta1994}: Ranta's original
  type-theoretic interpretation of \GF{} uses Martin-L\"{o}f type theory.
  Our approach uses \OSLF{}'s frame-valued predicates instead, giving modal
  operators automatically.
\item \textbf{Abstract Categorial Grammars}~\cite{deGroote2001}: ACGs also
  use $\lambda$-calculus as abstract syntax.  Our approach adds rewrite-rule
  semantics and automatic Galois connections.
\item \textbf{GF meets Coq}~\cite{BernardyChatzikyriakidis2017}:
  Bernardy and Chatzikyriakidis connect \GF{} to Coq for the FraCaS test
  suite, using type-theoretical semantics for textual inference.  Our
  contribution differs in scope: we formalize the full \RGL{} abstract
  signature (985 functions) and target \OSLF{} rewrite-rule semantics
  rather than model-theoretic denotations.
\end{itemize}

%% ===================================================================
\section{Extended Theorem Families}
\label{sec:extensions}

This section documents additional theorem families beyond the core
formalization, organized by module.
The interface refinement, UG-core quotient, and categorical packaging results
are developed in a companion paper~\cite{UGCores2026}.

\subsection{Modal Theorem Family (\code{SemanticHighlights.lean})}

Four theorems characterize how the accessibility relation interacts with the
formula fragment:

\begin{itemize}
\item \textbf{Positive-fragment monotonicity.}  If $R_1 \subseteq R_2$ and
  $\varphi$ is positive (no $\Bx$), then $\mathrm{sem}_{R_1} \varphi \; p$
  implies $\mathrm{sem}_{R_2} \varphi \; p$
  (\code{modal\_positive\_monotone}).
\item \textbf{Modal-free relation irrelevance.}  If $\varphi$ contains no
  $\Dia$/$\Bx$, its truth value is independent of $R$:
  $\mathrm{sem}_{R_1} \varphi \; p \Leftrightarrow \mathrm{sem}_{R_2} \varphi \; p$
  (\code{modal\_free\_relation\_irrelevant}).
\item \textbf{Box antitonicity.}  $\Bx$ is anti-monotone: if $R_2 \subseteq R_1$,
  then $\mathrm{sem}_{R_1} (\Bx\varphi) \; p$ implies $\mathrm{sem}_{R_2} (\Bx\varphi) \; p$
  for modal-free $\varphi$ (\code{modal\_box\_antitone}).
\item \textbf{Syntax lifts to temporal.}  Any positive theorem over syntax reductions
  transfers to the combined temporal relation
  (\code{modal\_syntax\_lifts\_to\_temporal}).
\end{itemize}

\subsection{Temporal Asymmetry (\code{SemanticHighlights.lean})}

A sharp positive/negative pair:

\begin{itemize}
\item \textbf{Negative.}  Under \code{syntaxOnly}, present temporal nodes
  have no $\Dia$-progress:
\begin{lstlisting}[language=ml]
theorem temporal_asymmetry_syntaxOnly :
    ¬ sem (gfReducesTemporal .syntaxOnly) I (.dia φ)
        (.apply "⊛temporal" [cl, .apply "0" []])
\end{lstlisting}
\item \textbf{Positive.}  If the temporal policy supplies an edge from
  present to past, $\Dia(\top)$ is obtainable
  (\code{temporal\_progress\_with\_policy\_edge}).
\end{itemize}

\subsection{Scope Ordering and Completeness}

\paragraph{Scope ordering as an inequality (\code{SemanticHighlights.lean}).}
For the ``every man loves some woman'' (EMLA) witness:
\begin{lstlisting}[language=ml]
theorem emla_scope_inverse_le_surface :
    qsemE2 R I Dom env (∃q2, ∀q1, ...) p ≤
    qsemE2 R I Dom env (∀q1, ∃q2, ...) p
\end{lstlisting}
$\exists\forall$ evidence is dominated by $\forall\exists$ evidence---the
surface scope is the ``stronger'' reading in the evidence order.

\paragraph{Scope completeness (\code{ScopeLinearExtension.lean}).}
For the two-quantifier EMLA fragment, reachable scope orders are
\emph{exactly} the linear extensions of the (dependency-free) partial order:
\begin{lstlisting}[language=ml]
theorem emla_scope_completeness_linearExtensions (ord : List String) :
    emlaReachableScopeOrder ord ↔ emlaLinearExtension ord
-- i.e. iff ord = ["q1","q2"] ∨ ord = ["q2","q1"]
\end{lstlisting}
This is also stated against the generic finite-DAG linear-extension notion
(\code{emla\_scope\_completeness\_dag}).

\subsection{Pronoun-Binding Locality (\code{SemanticHighlights.lean})}

Binding step V4 preserves the evidence of any formula that does not
mention the bound pronoun free---a dynamic-semantics locality statement:
\begin{lstlisting}[language=ml]
theorem v4_preserves_unmentioned_formula
    (pr r : String) ... (hnotfree : pr ∉ freeVarsQF2 φ) :
    let s' := ⟨s.term, s.store + {StoreAtom.bind pr r}⟩
    gfReducesFull cfg π s s' ∧
    gsemE2Full cfg π I Dom φ s = gsemE2Full cfg π I Dom φ s'
\end{lstlisting}
Contrast: V4 \emph{can} change evidence for formulas that mention the pronoun
(\code{V4\_post\_step\_semantic\_change}).

\subsection{Store-to-Environment Invariance (\code{SemanticHighlights.lean})}

If two stores agree on all \code{bind} and \code{ref} memberships, they
produce identical variable environments---a store-equivalence semantic
invariance result stronger than multiset commutativity:
\begin{lstlisting}[language=ml]
theorem store_env_invariant_equiv :
    (∀ pr r, bind pr r ∈ σ₁ ↔ bind pr r ∈ σ₂) →
    (∀ r pos, ref r pos ∈ σ₁ ↔ ref r pos ∈ σ₂) →
    storeToEnv σ₁ = storeToEnv σ₂
\end{lstlisting}

\subsection{NTT and Quantified Bridge (\code{GFWMConnections.lean})}

The quantified/discourse layer is no longer only background machinery.  Two
paper-facing endpoints show that visible and world-model semantics already reach
the \NTT{}/GSLT route:
\begin{lstlisting}[language=ml]
noncomputable def scopeOrderingNTBridge : ...
theorem closedFormula_envInvariant_NT : ...
\end{lstlisting}
\code{scopeOrderingNTBridge} transports visible scope-ordering information into
the native-type semantics, while
\code{closedFormula\_envInvariant\_NT} shows that closed quantified formulas are
invariant under environment change.  These endpoints connect the visible grammar semantics to the
native-type layer; the UG-core constructions that build on them
are developed in a companion paper~\cite{UGCores2026}.

%% ===================================================================
\section{Related Work}
\label{sec:related}

\paragraph{GF formalizations.}  To our knowledge, this is the first
formalization of the \GF{} \RGL{} abstract syntax in a proof assistant.
Previous work has formalized fragments of natural language grammar in
Coq and Agda, but not the complete \RGL{} signature.

\paragraph{OSLF/NTT.}  The \OSLF{} algorithm was introduced
by Meredith and Stay~\cite{MeredithStayOSLF2020}, with the categorical
foundations developed as Native Type Theory~\cite{WilliamsStay2021}.
Our earlier work formalized \OSLF{} for process calculi
($\varrho$-calculus, $\lambda$-calculus, Petri nets, TinyML) in
22,300+~lines of Lean~4.  The present paper extends this to natural
language grammars.

\paragraph{Verified NLP.}  There is growing interest in verified natural
language processing, including formalized parsing algorithms and grammar
formalisms.  Our contribution is orthogonal: we verify not the parsing
algorithm but the \emph{semantic properties} of parsed structures.

%% ===================================================================
\section{Conclusion}
\label{sec:conclusion}

We have presented a comprehensive Lean~4 formalization connecting the
Grammatical Framework to \OSLF{}, demonstrating that natural language
grammars automatically receive verified type systems with three semantic
layers: internal syntax rewrites, temporal policy, and visible rules for
scope and binding.  The formalization covers the \GF{} \RGL{}
core abstract syntax (985~function signatures, 112~categories), substantial English and Czech
concrete grammar subsets, evidence-valued world-model semantics (70~theorems),
and visible scope/binding rules.
The broader \GF{} subtree now contains 664 theorem/lemma declarations across
91~Lean files (43,162~lines), while the main paper highlights a 120+ theorem
semantic spine.  The interface refinement and UG-core quotient constructions
that build on this foundation are developed in a companion paper~\cite{UGCores2026}.

The key results are: scope completeness via linear extensions
(reachable scope orders are exactly the linear extensions of the
dependency partial order); pronoun-binding locality (binding preserves
evidence for formulas not mentioning the bound pronoun);
store-to-environment invariance (stores with the same bind/ref
membership produce identical environments); evidence-valued
world-model semantics with checker soundness; and compositional
lexical-containment monotonicity (modification preserves previously
present lexical material).

All code is available in the Mettapedia repository with zero
\code{sorry} declarations, zero axiom/opaque declarations, and sound proofs
throughout the checked \GF{} theorem modules.

\paragraph{Future work.}  We plan to: (1)~make the \OSLF{} rewrite engine
kernel-reducible to enable modal proofs via \code{decide}; (2)~add more
languages (German, Finnish) to exercise the cross-linguistic invariance
theorems with concrete linearizations; (3)~connect the lexical containment
predicate to the $\Dia$ modality via multi-step reduction;
(4)~implement explicit visible rules for voice/event structure and
tense/aspect commitment in the same store-based style;
(5)~formalize the presupposition projection layer (the remaining
open item from Table~\ref{tab:negative}); and
(6)~bridge the quantified formula semantics (\code{QFormula2}) to
PLN's \code{forAllEvalExt}, unifying the argument-aware quantifier
layer with evidence-valued inference.

\subsection*{Reproducibility}

\begin{quote}
\small
\begin{verbatim}
$ cd lean-projects/mettapedia
$ lean --version
Lean (version 4.28.0, ...)
$ export LAKE_JOBS=3 && nice -n 19 lake build \
    Mettapedia.Languages.GF
Build completed successfully.
$ rg '^\s*sorry\b' Mettapedia/Languages/GF/ --type lean | wc -l
0
$ rg '^\s*native_decide' Mettapedia/Languages/GF/ --type lean | wc -l
0
$ lake build 2>&1 | grep -c "^error:"
0
\end{verbatim}

\textbf{Lean 4.28.0} with \textbf{Mathlib v4.28.0}.
0~errors, 0~warnings, 0~\code{sorry},
0~\code{native\_decide}, 0~axiom/opaque declarations.
\end{quote}

\bibliographystyle{plain}
\bibliography{references}

\appendix
\section{Artifact Checksums}
\label{app:artifact}

\begin{quote}
\small
\begin{verbatim}
Repository commit: e0b3f54 (mettapedia, main)
Lean toolchain:   leanprover/lean4:v4.28.0
Mathlib pin:      v4.28.0
LeanHammer pin:   v4.28.0

Build command:
  export LAKE_JOBS=3 && nice -n 19 lake build

Build output (final lines):
  Build completed successfully.

Integrity checks:
  rg '^\s*sorry\b' Mettapedia/Languages/GF/ --type lean | wc -l  => 0
  rg '^\s*native_decide' Mettapedia/Languages/GF/ --type lean | wc -l  => 0
  rg '^(axiom|opaque)\s' Mettapedia/Languages/GF/ --type lean | wc -l  => 0
  lake env printPaths | grep -c error  => 0

GF subtree stats:
  Files: 91    Lines: 43,162
  Abstract functions: 985    Categories: 112
  English files: 20 (5,890 lines)
  Czech files: 12 (2,865 lines)
  WorldModelSemantics: 1,352 lines, 70 theorems
  VisibleLayer: 319 lines, 11 theorems
  VisibleLayerGFInstance: 97 lines
  SemanticKernelConfluence: 39 theorems
  WorldModelVisibleBridge: 34 theorems
  Total theorem/lemma declarations: 664
  Paper-facing theorem spine: 120+
\end{verbatim}
\end{quote}

\noindent
Executable English demonstration scripts are kept outside the
Mettapedia Lean library under \code{Demos/GF/English/}, so normal
library builds do not replay English demo \code{\#eval!} output.
English library modules retain example outputs only as comments or
theorem regressions.  (Some non-English utility modules such as
\code{CommentCNLEng.lean} still contain \code{\#eval} examples.)

%% ===================================================================
\section{Evidence-Grounded Semantics Details}
\label{app:evidence}
\label{sec:evidence}

The main text (\S\ref{sec:worldmodel}) already uses the implemented
evidence-grounded layer.  This appendix records the exact denotational
interfaces and theorem endpoints connecting \GF{}/\OSLF{} semantics to the
world-model/evidence calculus~\cite{goertzel2009pln}.

\paragraph{Status.}  The modules referenced below are implemented and
sorry-free.  The \GF{}/\OSLF{} evidence bridge is already formalized in Lean:
\code{gfEvidenceDenote} gives tree-level denotation,
\code{gfWMFormulaSemE} is the primary evidence-valued formula semantics,
\code{gfWMFormulaSem} is its thresholded Prop projection, and the checker-to-WM
bridge theorems are proved.  Table~\ref{tab:status} distinguishes fully proved
results from conditional generalizations and remaining expansion items.

\subsection{The Denotation Equation}

The implemented tree-level denotation assigns each \GF{} abstract tree $t$ an
evidence value relative to a world-model state $W$:
\begin{equation}
\label{eq:denote}
  \llbracket t \rrbracket_W
    \;:=\; \texttt{gfEvidenceDenote}\; W\; t
    \;=\; \texttt{WorldModel.evidence}\; W\;
           (\texttt{gfAbstractToPattern}\; t)
\end{equation}
where:
\begin{itemize}
\item \code{gfAbstractToPattern} converts a \GF{} tree to an \OSLF{}
  pattern (\code{OSLFBridge.lean}:96);
\item \code{WorldModel.evidence} extracts binary evidence $(n^+, n^-)
  \in \mathbb{R}_{\ge 0}^{\infty} \times \mathbb{R}_{\ge 0}^{\infty}$
  from the posterior state (\code{PLNWorldModel.lean}:48);
\item formula-level atom semantics uses the concrete query encoder
  \code{queryOfAtom} rather than a hypothetical generic \code{queryOf}.
\end{itemize}
World-model states $W$ replace possible worlds; \code{Evidence} values
$(n^+, n^-)$ replace truth values.  This is model-theoretic semantics
over \emph{evidence} rather than truth.

\subsection{Atom Interpretation Contract}

The \OSLF{} formula semantics (\code{Formula.lean}:104) interprets atoms
via \code{AtomSem := String $\to$ Pattern $\to$ Prop}.  We define an
evidence-grounded atom interpretation parameterized by a state $W$ and
per-atom thresholds $\tau$:
\[
  \texttt{atomSem}_W(a, p)
    \;:=\;
    \tau(a) \;\le\; \texttt{toStrength}(\texttt{evidence}\; W\;
                      (\texttt{queryOfAtom}\; a\; p))
\]
This is a \emph{chosen interface layer}: \OSLF{} does not prescribe how
atoms get their meaning; we plug in the world-model evidence channel.
The resulting formula semantics is then:
\[
  \texttt{sem}(\texttt{langReduces}\; \texttt{gfRGLLanguageDef},\;
               \texttt{atomSem}_W,\; \varphi,\; p)
\]
where the reduction relation $R$ provides the modal structure ($\Dia$, $\Bx$)
and the atom interpretation provides the evidential content.

\subsection{The 4-Layer Pipeline}

\begin{figure}[ht]
\centering
\begin{tikzpicture}[>=Stealth, node distance=1.8cm, font=\small]
  \node (gf) {\code{AbstractNode}};
  \node (pat) [right=2.2cm of gf] {\code{Pattern}};
  \node (oslf) [right=2.2cm of pat] {\code{OSLFFormula}};
  \node (q) [below=1.5cm of oslf] {\code{PLNQuery}};
  \node (ev) [left=2.2cm of q] {\code{Evidence}};
  \node (prop) [left=2.2cm of ev] {\code{Prop}};
  \draw[->] (gf) -- node[above]{\tiny\code{gfAbstractToPattern}} (pat);
  \draw[->] (pat) -- node[above]{\tiny rewrite rules} (oslf);
  \draw[->] (oslf) -- node[right]{\tiny atom interp.} (q);
  \draw[->] (q) -- node[above]{\tiny\code{evidence $W$}} (ev);
  \draw[->] (ev) -- node[above]{\tiny$\tau \le$ \code{toStrength}} (prop);
\end{tikzpicture}
\caption{Evidence-grounded semantics pipeline: \GF{} tree $\to$
  pattern $\to$ formula $\to$ query $\to$ evidence $\to$ truth.}
\label{fig:evidence-pipeline}
\end{figure}

\subsection{Checker-to-WM Soundness Chain}

The existing \code{checkLangUsing\_sat\_sound} theorem
(\code{Formula.lean}) bridges the bounded model checker to denotational
semantics.  With evidence-grounded atoms, the chain becomes:
\begin{align}
  \texttt{checkLangUsing}\;\ldots\; \texttt{atomCheck}_W\; \textit{fuel}\;
    p\; \varphi &= \texttt{.sat}
    \label{eq:checker} \\
  \Longrightarrow\quad
  \texttt{sem}(\texttt{langReduces}\;\texttt{gfRGLLanguageDef},\;
    \texttt{atomSem}_W,\; \varphi,\; p)
    \label{eq:sem} \\
  \Longrightarrow\quad
  \text{WM query judgment for threshold atoms}
    \label{eq:wmj}
\end{align}
Step~\eqref{eq:checker}$\to$\eqref{eq:sem} is the existing soundness
theorem.  Step~\eqref{eq:sem}$\to$\eqref{eq:wmj} follows by
instantiating the atom semantics definition and reading off the evidence
extraction.

\subsection{Design Choice: Prop-Threshold vs.\ Evidence-Valued}

The current code keeps \textbf{both} layers, with a clear priority order.
\textbf{Evidence-valued semantics} is primary:
\[
  \texttt{gfWMFormulaSemE} : \texttt{State} \to \texttt{OSLFFormula} \to
    \texttt{Pattern} \to \texttt{Evidence}
\]
\textbf{Threshold-Prop semantics} is a projection used by the checker soundness
pipeline:
\[
  \texttt{gfWMFormulaSem}\; W\; \tau\; \varphi\; p
  := \texttt{sem}(\ldots,\texttt{gfAtomSemFromWM}\; W\; \tau,\varphi,p)
\]
This split keeps the existing \OSLF{} checker interface intact while exposing a
strictly richer evidence-valued semantics in the same repository.  The relevant
agreement lemmas are \code{gfAtomSemFromWM\_eq\_threshold},
\code{gfRGLSemantics\_formulaSemE}, and the paper-facing checker-to-WM theorems
listed in Table~\ref{tab:status}.

\subsection{Assumptions}

The currently proved bridge fixes a concrete atom-query encoding
\code{queryOfAtom}.  Pushing beyond that concrete interface relies on three
further assumptions:

\begin{enumerate}
\item \textbf{Query extraction totality.}  Every \OSLF{} pattern maps to a
  well-defined \code{PLNQuery} (the encoding function
  \code{queryOf} must be total).
\item \textbf{Threshold monotonicity.}  If $\texttt{evidence}\; W\; q$
  increases (in the evidence lattice order), all threshold atoms remain
  satisfied.  This follows from monotonicity of \code{toStrength} with
  respect to the evidence order.
\item \textbf{Query encoding.}  Lexical atoms (``cat'', ``sleep'') map
  to \code{PLNQuery.prop}; constructor atoms may map to
  \code{PLNQuery.link} or remain purely structural.
\end{enumerate}

\subsection{Theorem Status}

\begin{table}[ht]
\centering
\caption{Status of evidence-grounded semantics results.
  \textbf{Proved} = formally verified in Lean~4 with 0~\code{sorry};
  \textbf{Conditional} = proved under a named assumption.}
\label{tab:status}
\small
\begin{tabular}{@{}llp{6.5cm}@{}}
\toprule
\textbf{Theorem} & \textbf{Status} & \textbf{Statement sketch} \\
\midrule
\code{translation\_preserves\_evidence\_allW}
  & \textbf{Proved}
  & Same abstract tree $\Rightarrow$ same evidence in all $W$
    (\code{WorldModelSemantics.lean}) \\
\code{checker\_sat\_atom\_implies\_wm\_query\_judgment}
  & \textbf{Proved}
  & Checker \code{.sat} on atom $\Rightarrow$
    WM formula semantics $\land$ WM query judgment
    (\code{WorldModelSemantics.lean}) \\
\code{adjCN\_preserves\_evidence\_if}
  & \textbf{Conditional}
  & \code{AdjCN} preserves evidence under
    \code{LexicalEvidenceMonotone}
    (\code{WorldModelSemantics.lean}) \\
\code{checker\_sat\_implies\_wm\_semantics\_general}
  & \textbf{Proved}
  & General checker soundness for arbitrary formulas
    (\code{WorldModelSemantics.lean}) \\
\code{visible\_layer\_extends\_base}
  & \textbf{Proved}
  & Base relation (syntax $+$ temporal) $\subseteq$ full relation
    (\code{WorldModelVisibleBridge.lean}) \\
\code{visible\_not\_base\_witness}
  & \textbf{Proved}
  & Scope choice is full-reachable but not base-reachable
    (\code{WorldModelVisibleBridge.lean}) \\
\code{anaphora\_not\_base\_witness}
  & \textbf{Proved}
  & referent-intro/pronoun-bind chain is full-reachable but not base-reachable
    (\code{WorldModelVisibleBridge.lean}) \\
\midrule
\code{checkLangUsing\_sat\_sound}
  & \textbf{Proved}
  & Checker soundness for arbitrary \code{AtomSem}
    (\code{Formula.lean}) \\
\code{montague\_thesis}
  & \textbf{Proved}
  & Cross-linguistic type invariance
    (\code{LinguisticInvariance.lean}) \\
\code{adjModification\_preserves\_lexical}
  & \textbf{Proved}
  & Compositional lexical-containment monotonicity
    (\code{LinguisticInvariance.lean}) \\
\code{evidence\_add'}
  & \textbf{Proved}
  & Evidence extraction commutes with revision
    (\code{PLNWorldModel.lean}) \\
\bottomrule
\end{tabular}
\end{table}

\paragraph{Base--visible separation.}
The base relation (syntax rewrites $+$ temporal policy) alone cannot
resolve scope ambiguity or anaphora---it preserves the store unchanged.
The visible layer supplies the additional reachability needed for these
phenomena.  This separation is formally witnessed: scope choice is
reachable via the full relation but not the base relation, and likewise
for the referent-introduction/pronoun-binding chain.  Three worked
examples demonstrate the complete pipeline: single quantifier
(\code{EveryManWalks}), scope ambiguity (\code{ScopeAmbiguity}),
and cross-sentential anaphora (\code{AnaphoraBinding}).

\end{document}
