%% CeTTa: A Direct C Runtime for MeTTa with Verified Translation
%%
%% Build:
%%   pdflatex CeTTa.tex

\documentclass[11pt]{article}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{lmodern}
\usepackage{geometry}
\geometry{margin=1in}
\usepackage{hyperref}
\usepackage{longtable}
\usepackage{booktabs}
\usepackage{listings}
\usepackage{multirow}
\usepackage{array}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{tikz}
\usetikzlibrary{arrows.meta,positioning,fit,backgrounds}
\setlength{\emergencystretch}{3em}
\lstset{
  basicstyle=\ttfamily\small,
  breaklines=true,
  frame=single,
  columns=fullflexible
}

\newcommand{\code}[1]{\texttt{#1}}
\newcommand{\PeTTa}{\textup{PeTTa}}
\newcommand{\HE}{\textup{HE}}
\newcommand{\CeTTa}{\textup{CeTTa}}
\newcommand{\LeaTTa}{\textup{LeaTTa}}
\newcommand{\PathMap}{\textup{PathMap}}
\newcommand{\RhoCalc}{\textup{RhoCalc}}

\title{\CeTTa{}:\\
\large A Direct C Runtime for MeTTa with Verified Translation}
\author{Zar Goertzel\\Oru\v{z}i}
\date{\today}

\begin{document}
\maketitle

\begin{abstract}
MeTTa is a reflective, non-deterministic meta-language with three
actively maintained runtimes: Hyperon Experimental (\HE{}, Rust),
\PeTTa{} (Prolog), and \CeTTa{} (C).  This paper describes \CeTTa{},
a direct C implementation of the \HE{} MeTTa evaluator that passes a
375-file regression suite with zero failures (equally green in the
MORK-bridge build; suite snapshot 2026-06), backed by a 333-case
rule-indexed conformance catalog against upstream \HE{}~v0.2.10
covering 24/24 specification rule capacities (catalog dated 2026-06).
On freshly rerun cross-runtime benchmarks dated 2026-06-27,
\CeTTa{} is fastest on startup-dominated shallow inference and on
several current genomic-inference rows (for example, the 183-pair PLN
evidence core and the 1{,}411{,}430-hypothesis deep genomic backward
chaining row), while \PeTTa{} wins on pure branching search (the
85{,}013-proof Metamath depth-5 benchmark).  A rerun-backed
cross-engine chaining suite built on community chainers
(Geisweiller's dependent-typed chainers and Metamath proofs;
Treutlein's PeTTaChainer problems) shows that \CeTTa{}'s shallow
advantage is startup-dominated rather than a general scaling property.
Historical rows whose current drivers now disagree or fail---the old
cross-engine miner speed table, the current 1.4M WM-PLN seam,
8-Queens, and tilepuzzle---are explicitly excluded from the revised
benchmark ladder.
The runtime is organized as an explicit profile family:
\code{he-compat} mirrors Rust \HE{} literally (including its quirks);
the principled \code{he} line carries exact rationals, true division,
and per-case recorded divergences; \code{he-extended} adds labeled
extensions; \code{he-prime} adds dependent telescopes.  The one-step
reduction relation is reflected into MeTTa itself as a modal surface
(\code{transitions}, \code{box}, \code{diamond}, normal-form tests)
aligned with the OSLF
$\mathrm{LanguageDef}\to\mathrm{ReductionSpan}\to\Diamond/\Box$
pipeline proven in Lean, and the first rule classes (grounded dispatch,
argument congruence, equation matching) are welded to a reflectable
small-step rule table whose rule-removal tests prove the engine
genuinely depends on the declared rules.
The runtime features SymbolId-based dispatch (38\% instruction
reduction), structured space kinds (queue, hash, stack), a pluggable
match backend with PathMap/MORK integration, and compiled space
artifacts via MORK's \code{.act} format for zero-copy indexed loading.
It is backed by a sorry-free Lean proof (5{,}400+ lines) establishing
semantic correspondence between \HE{} and \PeTTa{} evaluation on the
pure fragment, including import commutativity and operational bridges
for state and space allocation.  A kernel-checked HE$'$ telescope
formalization extends the type system with dependent binder semantics.
The \RhoCalc{} reaction machine's eval-at-\textsc{comm} semantics is given a
witness-free, axiom-clean Lean characterization: the quiescent run of the
canonical payload-evaluation process equals its structurally observable
outcome set.
We present a benchmark ladder from shallow evaluation through
large-scale genomic inference and a cross-engine chaining suite, a
freshness-first performance re-audit and triage, a specification gap
analysis, and a three-machine architecture (local search, indexed
shared-space, type elaboration) informed by GPT-5.4~Pro architectural
review.
\end{abstract}

%% ────────────────────────────────────────────────────────────────────
\section{Introduction}

MeTTa~\cite{metta} is a reflective, typed, non-deterministic language
designed as the core language of the OpenCog Hyperon AGI framework.
Three implementations exist:

\begin{description}
\item[\HE{}] Hyperon Experimental (Rust, v0.2.10).  The reference
  implementation with Python bindings and a large standard library.
\item[\PeTTa{}] Prolog-based MeTTa (SWI-Prolog).  Leverages Prolog's
  SLD resolution for backward chaining and pattern matching.
\item[\CeTTa{}] Direct C evaluator (this work).  Targets the \HE{} spec
  with arena allocation, discrimination-trie indexing, and tail-call
  optimization.
\end{description}

\paragraph{Contributions.}
\begin{enumerate}
\item A direct C runtime passing a 375-file regression suite (0
  failures) plus 104 profile tests, validated against upstream \HE{}
  v0.2.10 as oracle.
\item An explicit profile family (\code{he-compat} / \code{he} /
  \code{he-extended} / \code{he-prime}) with per-case recorded
  divergences and a fail-by-default conformance harness: a 333-case
  rule-indexed catalog (all agree, 24/24 spec-rule capacities) and a
  378-example broad corpus with every difference classified
  (Section~\ref{sec:specdriven}).
\item A reflective modal LTS surface over the evaluator's one-step
  reduction relation---successor set, $\Box$/$\Diamond$,
  normal-form tests---with runtime De~Morgan self-tests mirroring the
  Galois connection proven in the Lean OSLF framework.
\item The first specification weld: a reflectable small-step rule
  table gating grounded dispatch, argument congruence, and equation
  matching, with anti-decorative rule-removal tests.  This converts a
  cross-branch one-step soundness bug (missing argument congruence)
  into a structurally prevented class.
\item A bidirectional \HE{}$\leftrightarrow$\PeTTa{} translator
  (Prolog + executable Lean), with a sorry-free Lean proof of semantic
  correspondence.
\item Cross-runtime benchmarks rerun on the current tree: \CeTTa{} is
  10--100$\times$ faster than \HE{} on shallow evaluation, wins several
  current genomic-inference rows, while \PeTTa{} is 4.8$\times$ faster
  on the audited depth-5 Metamath proof-search row; unresolved rows are
  explicitly quarantined from the revised ladder.
\item A characterization of the semantic boundary between \HE{} and
  \PeTTa{}: the shared fragment (85\% of programs) runs identically; the
  divergence is in free-variable scoping across non-deterministic branches.
\item A specification gap analysis of non-determinism plus mutable state,
  with 8~concrete examples showing that sequential-per-iteration semantics
  (CeTTa's default) is correct in 6/8~cases, while \HE{}'s interleaved
  side-effect evaluation loses atoms in real code (the pattern miner).
\item \code{with-space-snapshot}: a CeTTa extension for branch-isolated
  mutation, enabling semi-naive forward chaining and speculative search
  without the side-effect hazards of shared mutable state.
\item A witness-free, axiom-clean Lean characterization of the \RhoCalc{}
  reaction machine's eval-at-\textsc{comm} semantics: the quiescent run of the
  canonical payload-evaluation process equals its
  structural-congruence-observed outcome set, with no certification witness in
  the carrier (Section~\ref{sec:rho-lean}).
\end{enumerate}

%% ────────────────────────────────────────────────────────────────────
\section{\CeTTa{} Architecture}

\CeTTa{} is a direct evaluator, not a compiler or boot pipeline.  It
implements the six mutual functions of the \HE{} evaluation spec
(\code{EvalAtom}, \code{MettaCall}, \code{InterpretExpression},
\code{InterpretFunction}, \code{InterpretArgs}, \code{InterpretTuple})
as C functions with a shared arena allocator.

\subsection{Core Design}

\begin{description}
\item[Arena allocation.] All atoms are allocated from 64\,KB arena blocks.
  No per-atom \code{malloc}/\code{free}.  Batch deallocation at scope exit.
\item[Symbol interning.] A hash table maps symbol strings to unique
  pointers, enabling O(1) pointer comparison for symbol equality.
\item[Equation index.] Equations \code{(= lhs rhs)} are indexed by
  head-symbol hash into 256 buckets, with a discrimination trie within
  each bucket for fast LHS matching.
\item[Substitution tree.] An epoch-based discrimination tree
  (\code{SubstTree}) eliminates per-candidate variable renaming during
  equation queries by tagging indexed variables with monotonic epochs.
\item[Tail-call optimization.] A \code{TAIL\_REENTER} trampoline with
  \code{goto tail\_call} eliminates stack growth for recursive equation
  dispatch.
\item[Unlimited fuel.] Default fuel is $-1$ (unlimited), matching the
  \HE{} spec.  Stack overflow is caught via \code{sigaltstack} +
  \code{SIGSEGV} handler with a clean error message.
\item[Two-stage stdlib.] The standard library is precompiled into a
  binary blob at build time (\code{cetta-stage0} $\to$
  \code{--compile-stdlib} $\to$ blob $\to$ \code{cetta}), giving 6\,ms
  startup.
\end{description}

\subsection{Test Suite}

\CeTTa{} is validated against the \HE{} CLI (v0.2.10) as oracle.
Golden regression files are oracle-validated when introduced;
\code{make test} replays them on every build.  Current status:
\textbf{375 passed, 0 failed, 40 skipped} on the default build and
\textbf{371 passed, 0 failed, 44 skipped} on the MORK-bridge build
(all bridge sub-lanes green), plus 104 profile tests covering module
imports, foreign (Python) function dispatch, numeric-tower and
profile-surface splits, and extension surfaces.  Two further lanes are
fail-by-default: \code{make test-step-rules} disables individual
rule-table entries and \emph{requires} the witnesses to fail
(Section~\ref{sec:specdriven}), and the semantic conformance suite
re-runs the curated oracle corpus live against upstream \HE{}
(Section~\ref{sec:conformance}).

%% ────────────────────────────────────────────────────────────────────
\section{Specification-Driven Semantics: Profiles, Conformance, and the
  Small-Step Rule Table}
\label{sec:specdriven}

\CeTTa{} is organized around an explicit principle: \emph{the semantics
is a named object, not an emergent property of the implementation}.
This section describes the profile family, the oracle-conformance
campaign against upstream \HE{}, the evaluator bug classes that
campaign exposed, the reflective modal surface over the one-step
reduction relation, and the first welded rule pack that makes the
reduction rules themselves runtime data.

\subsection{The Profile Family}
\label{sec:profiles}

Profiles are distinct semantic products, implemented as additive
layers---never as visibility masks over one shared universe:

\begin{description}
\item[\code{he-compat}] A literal mirror of Rust \HE{}~v0.2.10,
  including its quirks: integer division for \code{/}, the strict
  three-argument \code{if} surface, Rust's library inventory behavior.
  Compatibility means matching upstream \emph{behavior}, not
  reproducing its library set; library availability is out of scope
  for conformance.
\item[\code{he}] The principled line.  Spec-faithful where the written
  specification speaks; where \HE{}'s implementation and its
  specification diverge, the divergence is adjudicated and recorded
  per-case.  Carries exact arbitrary-precision rationals, true
  division for \code{/}, floor division \code{//}, Euclidean
  \code{math:mod}, and the clean \code{math:} namespace over the
  Rust-heritage \code{*-math} ops.  Extensions that accept more
  programs (e.g.\ two-argument \code{if}) are deliberate, recorded
  surface choices.
\item[\code{he-extended}] \code{he} plus labeled \CeTTa{} extensions:
  typed space kinds, \code{with-space-snapshot}, the \code{lts}
  reflection surface, MORK/PathMap-backed spaces.
\item[\code{he-prime}] The dependent-telescope extension
  (Section~\ref{sec:heprime}).
\end{description}

Two rules keep this family honest.  First, unknown symbols stay
inert: MeTTa's philosophy is to leave uninterpreted what it does not
know how to interpret, so no profile ever raises an ``unsupported
feature'' error---a profile determines what is \emph{interpreted},
never what is \emph{legal}.  Second, profile differences are explicit
splits with named tests (e.g.\ \code{profile\_if\_compat\_arity}
pins that \code{he-compat} rejects two-argument \code{if} exactly as
upstream does, while \code{he}/\code{he-extended} accept it as a
recorded extension).

\subsection{Oracle Conformance}
\label{sec:conformance}

For \code{he-compat}, upstream \HE{}~v0.2.10 is the sole oracle, and
the harness is a fail-by-default gate, not a green-manufacturing
report.  Comparison is by \emph{result-bag equality}---multisets of
parsed results, never byte-equality of output text---so formatting
differences cannot masquerade as semantic agreement, and duplicate
results (which are semantically real in MeTTa) are counted.

\begin{description}
\item[Curated catalog.] 333 cases, every one oracle-validated and in
  agreement, each mapped to the specification rules it exercises:
  24/24 spec-rule capacities are covered, and every capacity is
  classified into one of three lanes---\emph{reduction-span} (the
  one-step semantics the modal layer consumes),
  \emph{supporting-surface} (match, unify, grounded dispatch
  plumbing), and \emph{spatial-policy} (module import, inventory,
  space policy).  The lane split keeps policy surfaces out of the
  reduction relation.
\item[Broad corpus.] 378 upstream examples: 302 agree, 59 classified
  differences (library availability, which is out of conformance
  scope; path-import capabilities upstream rejects; environment and
  rendering differences; and adjudicated upstream drift), 7 timeouts,
  10 not runnable upstream.  No difference is left unclassified.
\item[Adjudicated divergences.] Where \CeTTa{} deliberately diverges
  from upstream, each case carries independent grounds.  Three
  examples: upstream's post-0.2.10 \code{function}/\code{NoReturn}
  behavior contradicts its own written specification (a deliberate
  loop fix that left the document stale); upstream's \code{get-doc}
  errors on \HE{}'s own documentation example (an arity regression);
  and the user-visible result of top-level \code{eval} on a
  non-reducible term is genuinely under-determined by the
  specification---both engines deviate from the literal
  instruction-layer reading, in opposite directions.
\end{description}

The catalog carries an explicit
\code{no\_runtime\_trace\_or\_certificate\_surface} flag, enforced by
the suite: conformance is \emph{testing}, and the specification is
what gets \emph{proven about} in Lean.  \CeTTa{} does not emit
execution traces or certificates, and a per-run check is never
presented as a proof.

\subsection{Evaluator Bug Classes the Campaign Exposed}
\label{sec:bugclasses}

Three bug classes were found and fixed, all instructive.

\paragraph{The Empty algebra.}  When overlapping equations yield a
branch that reduces to \code{Empty}, two dual defects appeared: the
surviving sibling branch was returned without being re-driven to
applicative normal form (violating the spec's applicative order), and
in other shapes \code{Empty} itself leaked into the visible result bag
(the spec is explicit that \code{Empty} ``is not returned among other
results'').  The fix treats the evaluator's outcome set as an algebra
with \code{Empty} as its absorbing zero and re-drives surviving
branches independently of their siblings.  In modal terms
(Section~\ref{sec:modal}): the successor algebra must satisfy
$\Diamond\bot = \bot$.

\paragraph{The one-step congruence bug.}  The reflective one-step
surface reported \code{(+ 1 (+ 2 3))} as a \emph{normal form}: the
one-step relation lacked argument congruence (stepping a reducible
argument when the head cannot fire), while the big-step evaluator had
it.  Because the one-step relation and the evaluator were separately
hand-maintained C branch sets, the gap replicated identically across
three development branches and was invisible to every test that only
exercised head-reducible terms.  The bug was diagnosed through the
modal surface itself ($\Box\bot$ holding on a reducible term), fixed
as deterministic left-to-right congruence matching the evaluator's
applicative order---and is the direct motivation for the rule-pack
weld in Section~\ref{sec:langdefpack}.

\paragraph{The switch/switch-minimal conflation.}  Upstream \HE{}
types \code{switch} as \code{(-> \%Undefined\% Expression
\%Undefined\%)}---the scrutinee is \emph{evaluated} before
matching---while \code{switch-minimal} is \code{(-> Atom Expression
Atom)}, matching the raw scrutinee structurally.  \CeTTa{} routed both
spellings through one structural handler, so
\code{(switch (+ 1 2) ((3 three) ((\$x \$y) (got \$x \$y))))} returned
\code{(got 1 2)} where upstream returns \code{three}.  The defect was
exposed not by testing but by the certification campaign
(Section~\ref{sec:rulecert}): modeling the rules in Lean forced a
precise reading of the two upstream type contracts, and an oracle
probe confirmed the divergence.  The fix splits the rule into
\code{HES\_Switch} (evaluated scrutinee) and \code{HES\_SwitchMinimal}
(structural), with a distinguishing witness pair pinned in the test
suite.

\subsection{The Reflective Modal Surface}
\label{sec:modal}

The library surface \code{lts} (generic, step-parameterized) and
\code{lts:he} (bound to the \HE{} evaluator's one-step relation) reflect the
one-step reduction relation into MeTTa as data:

\begin{itemize}
\item \code{lts:he:transitions} emits the one-step successor set
  of an atom as quoted states (zero results on quiescent atoms);
  \code{lts:he:trace-2} enumerates two-step traces.
\item \code{lts:he:is-normal-form} is exactly $\Box\bot$ (every
  successor satisfies falsity, i.e.\ no successor exists);
  \code{lts:he:is-can-step} is $\Diamond\top$.
\item \code{lts:he:box}~$\phi$ and \code{lts:he:diamond}~$\phi$
  generalize these to arbitrary unary predicates: $\Box\phi$ holds
  when every one-step successor satisfies $\phi$ (vacuously at normal
  forms), $\Diamond\phi$ when some successor does.
\end{itemize}

The De~Morgan law $\Box\phi = \lnot\Diamond\lnot\phi$ is installed as
a runtime self-test; it is the executable mirror of the Galois
connection proven once and for all in the Mettapedia OSLF framework,
where $\mathrm{LanguageDef}\to\mathrm{RewriteSystem}\to
\mathrm{ReductionSpan}\to(\Diamond,\Box)$ is derived generically and
\code{langGalois} establishes the adjunction.  Native behavioral types
in the sense of Native Type Theory arise as predicates over exactly
this reduction span; the runtime surface and the Lean derivation
describe the same object at two levels.

Multiplicity discipline: \code{transitions} preserves the result
\emph{bag} (duplicate successors are semantically real), while the
modal predicates quotient to reachability.  Evaluation and conformance
compare bags; $\Diamond/\Box$ use support.  Result sets are never
deduplicated for speed.

\subsection{The Small-Step Rule Table}
\label{sec:langdefpack}

The congruence bug demonstrated the structural disease: reduction
rules as anonymous, duplicated C branches are invisible, unremovable,
and untestable-by-absence.  The cure is to make the rule set a
\emph{named, reflectable object} the evaluator provably depends on.

Three granularity strata are kept distinct.  \emph{HEFine} is the Lean
\code{mettaHE} instruction machine (59 rules over interpreter states;
the proof reference).  \emph{HE small-step} is the coarse user-visible
one-step relation---one observable surface rewrite, e.g.\
\code{(+ 1 (+ 2 3))} $\to$ \code{(+ 1 5)}---which is what the
\code{lts} surface exposes.  \emph{HEBigStep} is full evaluation.  The
table encodes the small-step relation; its relationship to the fine machine is a
collapse/simulation with explicit provenance links (each coarse rule
names the fine rules that justify it), never a label identity.

The runtime object, \code{CettaLangDefPack}, carries: language,
profile, and granularity identifiers; a schema version; a source
digest over the canonical rule-descriptor text; the rule table; a
disabled-rule mask populated from the
\code{CETTA\_LANGDEF\_DISABLED\_RULES} environment variable; and a
\emph{per-rule claim level} recording the strongest evidence behind
each rule (Section~\ref{sec:rulecert}).  Ten rule classes are live and
gate the \emph{only} implementations of their behavior in the one-step
subsystem: grounded dispatch, leftmost expression congruence, equation
match, \code{eval}, \code{chain}, \code{let}, \code{let*},
\code{case}, \code{switch}, and \code{switch-minimal} (the last two
split per their distinct upstream contracts,
Section~\ref{sec:bugclasses}).  Quote/return and empty/error
quiescence are structural rows: accounted in the table with written
reasons, but not gateable, since disabling them would make the
semantics incoherent rather than smaller.  Each row carries a
provenance string naming the fine-machine rules that justify it.  The
pack is reflected into MeTTa by \code{lts:he:step-rules}.

The load-bearing test is \emph{anti-decorative}: a dedicated lane
re-runs the witness suite with each live rule disabled and
\emph{requires} the witnesses to fail (the successor set must go
empty; normal-form must flip)---baseline plus ten rule-removal
failures confirmed.  If the witnesses still passed with a rule
removed, a duplicate hidden code path would exist---precisely the
disease that produced the congruence bug.  A disabled special form
goes \emph{inert} (its head stays uninterpreted; congruence still
applies), per the principle that unknown symbols are data, never
errors.  The lane fails closed in both directions and runs on both the
default and MORK-bridge builds.

Claim levels are deliberately \emph{excluded} from the digest: the
digest fingerprints rule \emph{content} (names, profiles, provenance,
schema), while claims are \emph{evidence about} that content.
Upgrading a rule's certification therefore must not---and verifiably
does not---move the semantic object: the digest
(\code{fnv1a64:12eec5b655a75904} at the time of writing) stayed fixed
across every claim flip.  The pack is, by design, a hand-authored
object kept honest by the removal lane and the digest; we deliberately
parked an earlier plan to generate specialized C from the rule source,
because the weld's value lies in the named semantic object and its
fail-closed tests, and code generation is machinery the certification
program does not need.

\subsection{Lean Certification of the Rule Table}
\label{sec:rulecert}

Each per-rule claim takes one of three values, in increasing strength:
\emph{bag-tested-adequate} (the rule's behavior matches the legacy
path and oracle bags on the witness corpus),
\emph{rule-sound-on-fragment} (a compiled, sorry-free Lean theorem
proves the rule absorbed by the official declarative semantics on an
explicitly characterized fragment), and \emph{rule-sound} (the same,
unconditionally).  A claim flips only when the compiled theorem
exists, so the reflective surface can never overclaim.  The
certification route is \emph{absorption}: for each coarse rule, a
theorem that whenever the rule relates $a \to a'$, the official
declarative judgment of the Mettapedia \HE{} formalization
(\code{MettaCall}, \code{MinimalStep}, or \code{EvalAtom}) already
relates the same atoms.  At the time of writing five of the ten live
rules carry certification:

\begin{center}\footnotesize
\begin{tabular}{lll}
\toprule
rule & claim & Lean witness \\
\midrule
\code{HES\_GroundedDispatch} & rule-sound &
  \code{mettaCall\_absorbs\_grounded\_dispatch} \\
\code{HES\_EquationMatch} & rule-sound &
  \code{mettaCall\_absorbs\_equation\_match} \\
\code{HES\_Eval} & rule-sound &
  \code{minimalStep\_absorbs\_eval} \\
\code{HES\_Chain} & rule-sound-on-fragment &
  \code{minimalStep\_absorbs\_chain\_source}/\code{\_subst} \\
\code{HES\_LeftmostExprCongruence} & rule-sound-on-fragment &
  \code{evalAtom\_absorbs\_congruence\_frag} \\
\code{HES\_Let}, \code{LetStar}, \code{Case}, & bag-tested-adequate &
  in progress \\
\quad\code{Switch}, \code{SwitchMinimal} & & \\
\bottomrule
\end{tabular}
\end{center}

The certified fragment for the congruence and chain rules is
characterized positively: untyped constructor-headed expressions with
quiescent spectator arguments and a grounded-dispatch or
equation-match redex.  Its supporting theory is a short ladder of
compiled lemmas: a quiescence domain on which atoms officially
self-evaluate, uniqueness of that self-evaluation, transport of
official evaluation across a single changed argument, and absorption
of inner fragment steps---about which the honest surprises accumulated
(e.g.\ a bare \code{Error} \emph{symbol} at tuple head becomes
error-shaped, so the fragment must exclude it explicitly).

The five remaining rules are upstream \emph{stdlib sugar}: their
one-step behavior is the upstream definition of \code{let},
\code{let*}, \code{case}, \code{switch}, and \code{switch-minimal} as
\code{unify}/\code{chain} programs.  Certifying them required a
\emph{matcher bridge}: relating the public equation-query surface to the
official \code{matchAtoms}/\code{mergeBindings} semantics.  The key
insight from this tranche was that the real seam was \emph{not} merely a
direct one-way-to-two-way matcher comparison.  The public \HE{} query
surface had historically routed equation lookup through the simplified
\code{simpleMatch} path, while the faithful two-sided matcher already
existed separately.  We therefore split the bridge into three explicit
layers.

First, \code{DeclMatchSpec.lean} states the author's
\code{match\_atoms} semantics declaratively and proves the executable
\code{matchAtoms} helper sound and complete against it
(\code{matchAtoms\_sound}, \code{matchAtoms\_complete}).  Second,
\code{DeclMergeSpec.lean} does the same for
\code{merge\_bindings}/\code{add\_var\_binding}/\code{add\_var\_equality},
proving \code{mergeBindings\_sound} and
\code{mergeBindings\_complete}.  Third,
\code{QuerySurfaceAgreement.lean} proves a bounded adequacy theorem for
the \emph{public} query interface:
\code{queryEquations\_ground\_two\_way\_adequate} and its
\code{queryEquationsAgainstVisible}-companion show that, on the
ground-query fragment and under the explicit freshening adequacy
hypothesis \code{EquationFresheningAdequate}, the repaired
\code{matchAtoms}/\code{mergeBindings} surface and the legacy
\code{simpleMatch}-based scan agree in both directions up to replay
fuel.

The earlier singleton-seed factorization theorem remains one ingredient
inside this story, but it is no longer the whole story.  The decisive
insight is the G3/G3b split: equality-threading becomes observable only
when the queried atom itself still contains variables, because that is
the point at which \code{Bindings.resolve}/\code{Bindings.apply} must
honor equalities rather than plain assignments.  We therefore bank the
ground-query adequacy result as the honest current certification
surface, and carry the variable-bearing query case as an explicit next
obligation rather than silently pretending the legacy surface was already
faithful.

%% ────────────────────────────────────────────────────────────────────
\section{Verified \HE{}$\leftrightarrow$\PeTTa{} Translation}

\subsection{Bidirectional Translator}

A bidirectional translator converts between \HE{} and \PeTTa{} dialects.
It is implemented both as relational Prolog
(\code{he\_petta\_relational.pl}, with explicit freshness threading) and
as computable Lean functions (\code{translateHE}, \code{translatePeTTa}).

The translator handles 8 \HE{}$\to$\PeTTa{} rewrite rules
(\code{chain}$\to$\code{let}, \code{collapse-bind}$\to$\code{collapse},
\code{nop}$\to$\code{let~\$fresh~X~()}, etc.)\ and 5
\PeTTa{}$\to$\HE{} rules (\code{progn}$\to$nested \code{let},
\code{prog1}$\to$\code{let} with result capture, etc.).  All 437 real
\code{.metta} files parse and translate successfully.

\subsection{Sorry-Free Lean Proof}

The translation is backed by a \textbf{sorry-free, axiom-free} Lean
proof across two files totaling 4{,}979 lines and 102
definitions/theorems:

\begin{description}
\item[\code{HEPeTTaSound.lean}] (1{,}063 lines, 44 theorems.)
  Proves that \HE{}'s equation matching corresponds to \PeTTa{}'s
  \code{ruleApp} evaluation.  The key result:
  \begin{quote}
  \code{simpleMatch\_applyBindings\_comm}: if \HE{}'s
  \code{simpleMatch} succeeds on two \code{PureTranslatable} atoms,
  then applying the translated bindings via \code{heBindingsToOSLF} to
  the translated pattern recovers the translated target.
  \end{quote}
  This commutation property, combined with the existing
  \code{matchPattern\_applyBindings\_complete} from the OSLF framework,
  establishes that a \PeTTa{} \code{matchPattern} witness exists
  whenever \HE{}'s \code{simpleMatch} succeeds.

\item[\code{HEPeTTaTranslate.lean}] (3{,}916 lines, 58 theorems.)
  The executable translator with:
  \begin{itemize}
  \item \code{translateHE\_translatable}: translation preserves
    \code{Translatable} (output admits \code{atomToPattern}).
  \item \code{translateHE\_preserves\_pureTranslatable}: translation
    preserves the stronger \code{PureTranslatable} fragment.
  \item \code{translatePeTTa\_translatable}: same for the reverse
    direction.
  \item 9 \code{\#eval} tests matching Prolog output on concrete atoms.
  \item 6 \code{rfl} examples kernel-verified by Lean.
  \item Roundtrip idempotence examples (kernel-verified).
  \end{itemize}
\end{description}

%% ────────────────────────────────────────────────────────────────────
\section{Reflective $\rho$-calculus and eval-at-\textsc{comm}}
\label{sec:rhocalc}

Under \code{--lang rhocalc} the same runtime stops being a term rewriter and becomes a
concurrent process engine: its unit of work is not a rewrite but a \emph{reaction}, a
rendezvous between a sender and a receiver on a shared channel. This brings channels with
private names and genuine concurrency; and because the calculus is \emph{reflective}, a
received name can be \emph{run}, so a communication can itself be a computation. The MeTTa
library that exposes that last capability is \emph{rhometta}. We build up to it from the
pure calculus.

\subsection{The pure calculus}

The calculus is Meredith and Radestock's reflective higher-order
$\rho$-calculus~\cite{rho2005}. Its one reduction rule, \textsc{comm}, is a handshake
(Figure~\ref{fig:rho-comm}): a sender \code{x!(m)} on channel \code{x} beside a receiver
\code{for(y <- x)\{P\}} on that
same channel react; the message is consumed and the receiver resumes with \code{m}
substituted for its bound name \code{y}:
\[
  x!(m) \;\big|\; \mathsf{for}(y \leftarrow x)\{P\}
  \;\;\xrightarrow{\;\textsc{comm}\;}\;\; P[y := m].
\]

\begin{figure}[h]\centering
\begin{tikzpicture}[>=Stealth,
  box/.style={draw, rounded corners, align=center, inner sep=4pt,
    font=\small\ttfamily}]
  \node[box] (s) {x!(m)};
  \node (par) [right=3mm of s, font=\large] {$\mid$};
  \node[box] (r) [right=3mm of par] {for(y<-x)\{P\}};
  \node[box, fill=black!8] (out) [right=16mm of r] {P[y:=m]};
  \draw[->, thick] (r) -- node[above, font=\scriptsize\sffamily] {\textsc{comm}} (out);
\end{tikzpicture}
\caption{A reaction: a send and a matching receive on channel \code{x} rendezvous; the
  receiver continues with the message substituted in.}
\label{fig:rho-comm}
\end{figure}

The rest of the vocabulary is small. \code{0} is the inert process; \code{P | Q} runs
\code{P} and \code{Q} concurrently. Two reflection operators connect processes and names:
\code{@P} \emph{quotes} a process into a name, and \code{*x} \emph{drops} (runs) the process
a name quotes, with \code{*@P = P}. Channels are themselves quoted processes --- this is
what makes the calculus higher-order. \CeTTa{} implements exactly these six strict-core
constructors --- \code{rho:nil}, \code{rho:par}, \code{rho:send}, \code{rho:recv},
\code{rho:quote}, \code{rho:drop} --- with \textsc{comm} as the only reduction. A program is
reduced to quiescence by \code{cetta --lang rhocalc \textit{file}}, written either in a
MeTTa-style s-expression surface (\code{.mrho}) or a Rholang-like surface (\code{.rho}); the
runtime keeps a live index of sends and receive-sites, rebuilt after each reaction.

\subsection{Reflection is enough}
\label{sec:rho-reflection}

A surprising amount follows from \textsc{comm} and quote/drop alone, with no further
primitives. Each item below is a standalone \code{--lang rhocalc} program.
\begin{itemize}
\item \emph{Name mobility.} A receive that drops its argument, placed beside a send carrying
  a quoted send, reduces to that inner send made live: a name has travelled across a channel
  and been activated.
\item \emph{Fresh names.} Quoting a distinct process yields a distinct name, so a receiver
  can mint and use a name it never mentioned syntactically.
\item \emph{Replicated input.} An input-guarded server can re-create its own listener:
  after serving a message it leaves a copy of the receiver in the residual, so the channel
  keeps serving. Persistent receive is \emph{derived}, not built in.
\item \emph{Recursion and divergence.} The same self-recreation, left ungated, runs without
  bound (the engine reports its reduction limit); recursion needs no dedicated operator.
\end{itemize}
Private channels and information hiding come along for free: a payload exchanged on a name
no other process holds is unobservable apart from whatever completion signal the protocol
chooses to publish.

The capstone is \emph{eval-at-\textsc{comm}}. Let the receiver do nothing but drop what it
receives:
\[
  \texttt{eval\_drop}(x,p)\;\equiv\;
  \underbrace{x!(@p)}_{\text{send the quoted program } p}\;\big|\;
  \underbrace{\mathsf{for}(y \leftarrow x)\{\,*y\,\}}_{\text{run whatever arrives}}.
\]
The \textsc{comm} delivers the name \code{@p}; the drop \code{*@p} runs \code{p}; the
process settles at the result of running \code{p}. A handshake has become an evaluation
step, and since \code{p} may itself evaluate to more process, a reaction can plant the next
reactions.

\subsection{Profiles}

A \emph{profile} names a variant of the reduction relation together with the grammar it
accepts. \code{strict-core} is the pure asynchronous spine above: it admits only the six core
constructors, and a non-core form is \emph{rejected} rather than left inert. The \code{cost}
profile accepts a richer resource-annotated dialect and meters each reaction, charging a
budget per \textsc{comm} (Section~\ref{sec:rho-cost}).

\subsection{Certified reduction}
\label{sec:rho-lean}

Does eval-at-\textsc{comm} report the result of running the payload, or could the machinery
quietly change the answer? We settle this in Lean. Run \code{eval\_drop(x,p)} to quiescence
and collect every outcome observable \emph{up to structural rearrangement} of the process;
that set is exactly the results a correct evaluation of \code{p} would produce, and the
statement refers only to what is observable, never to an internal ``this is the answer''
tag. The theorem \code{rhometta\_bridge\_scObserved} (in \code{RhometaReduction.lean})
equates the concrete quiescent run of \code{eval\_drop(x,p)} with
\code{RhometaSCObservedOutcomes} of the same process --- the outcomes visible up to
structural congruence --- with no result-certificate carried in the answer. Two ingredients
are load-bearing. First, a complete canonical-form normalizer \code{nfAtom}: two processes
are structurally congruent exactly when their normal forms coincide, which unblocks
inductions that otherwise stall on the symmetry and transitivity of congruence. Second, one
hygiene side condition \code{noBoundUnderQuote}: the communicated variable must not occur
under a quote, since a payload that hides a bound occurrence inside a quoted name defeats the
characterization (a concrete counterexample exhibits exactly that failure). The development
carries no \code{sorry}. A separate caveat on the trusted base: its conformance and DSL
fixtures use \code{native\_decide}; the bridge theorem's own module does not.

\subsection{rhometta: MeTTa at the reaction}

eval-at-\textsc{comm} becomes a way to program once the payload is a MeTTa term. The library
\code{lib/rhometta.metta} provides \code{rhometta:eval}, which defers a MeTTa term as the
payload to be run at \textsc{comm}; \code{rhometta:run}, which reduces a process to its
quiescent outcome set; and \code{rhometta:transitions}, which returns one-step successors. It
sits on \code{lib/rho.metta}, the strict-core constructor surface, and delegates all
reduction to the engine. One consequence is central: when a deferred payload evaluates to
\emph{several} results, the \textsc{comm} forks --- one successor per result. Nondeterministic
computation becomes a branching reaction, and \code{rhometta:run} returns one residual per
branch (the \emph{may-set}).

\subsection{What reactions compute}

\paragraph{The minimal cell.} The smallest rhometta program is a single cell: a send carrying
a deferred term beside a receiver that drops (runs) it. \code{rhometta-eval-cell} sends
\code{(+ 2 3)} this way and the reaction leaves \code{5}; with a multi-result payload
\code{(superpose (a b c))} the \textsc{comm} forks into one branch per result, and
\code{rhometta:transitions} confirms the fork is a single reduction step with three
successors. Communication does computation, and result-multiplicity becomes branching.

\paragraph{A self-unfolding prover.} Seed a \emph{single} cell for a goal. Each \textsc{comm}
evaluates one backward-chaining step whose result is the cells for that goal's subgoals, so
\code{rhometta:run} keeps reducing and the reaction network unfolds itself into a proof tree.
Where a goal has several rules the payload has several results, so the \textsc{comm} forks:
\code{rhometta:run} returns one quiescent residual per complete proof tree
(Figure~\ref{fig:rho-tree}) --- two for the sample ``can I bake a cake?'' goal, while a
single-rule goal stays confluent at one.

\begin{figure}[h]\centering
\begin{tikzpicture}[>=Stealth, level distance=8mm, sibling distance=15mm,
  every node/.style={font=\scriptsize\ttfamily}, edge from parent/.style={draw,-}]
  \begin{scope}
  \node {cake} child { node {batter} child { node {flour} } child { node {egg} } }
               child { node {oven} };
  \end{scope}
  \begin{scope}[xshift=42mm]
  \node {cake} child { node {batter} child { node {mix} } }
               child { node {oven} };
  \end{scope}
\end{tikzpicture}
\caption{The self-unfolding prover's may-set for ``cake'': two complete proof trees, one per
  rule for \code{batter}. The trees are not enumerated up front --- they are \emph{grown} by
  reactions whose payloads plant their children.}
\label{fig:rho-tree}
\end{figure}

\paragraph{Type inference as proof search.} A typing judgment \code{(: proof (HasType e t))}
is proved by reactions over typing rules, and the proof term is the derivation. In
\code{rhometta-type-inference}, an overloaded \code{plus} yields a genuine may-set --- an
\code{Int} and a \code{Float} typing of \code{(plus one one)} --- while \code{(plus x one)}
with \code{x : Int} prunes the \code{Float} branch at the unification join. Proof terms stay
pure, \code{(: (app pf pa pb) (HasType e t))} --- a typing is crisp, so no truth value is
attached. Each result is cross-checked differentially against a sequential
chainer over the same knowledge base, which derives the same proof-term skeletons.

\paragraph{A truth value that is earned.} Crisp proof terms are the right output for a typing,
but some judgments are genuinely probabilistic, and there a reaction can carry evidence rather
than certainty. In \code{rhometta-evidence-revision}, two agents each count their own
observations of ``ravens are black'' into an opinion --- a strength and a confidence ---
computed at \textsc{comm} by a deferred payload. The two agents share two observations, so
merging their opinions \emph{naively} double-counts the shared evidence and overstates the
confidence ($0.889$); an overlap-corrected revision discounts the shared stamps for the honest
value ($0.857$), at the same strength. The revision can run after the parallel fan-in, or
\emph{at} the final rendezvous itself. This is the one place a truth value belongs --- it is
counted from data, not attached by hand.

\paragraph{Composition.} These pieces compose: a further program runs an independent proof
search, an overlap-corrected evidence revision, and an attention-spreading step concurrently,
then rendezvouses their channel-gathered results into a single synthesis.

\subsection{Cost-accounted reactions}
\label{sec:rho-cost}

Reactions are also the natural place to \emph{meter} computation: each \textsc{comm} is one
unit of work, so a resource budget can be charged per reaction. \CeTTa{}'s \code{cost}
profile does this directly --- it reduces an annotated $\rho$ dialect and reports the
resource each reaction consumes. It is the executable end of an ongoing program to equip the
$\rho$-calculus with a \emph{cost endofunctor}: a construction that wraps a calculus with
resource accounting once and for all~\cite{costendofunctor}. We return to it in future work.

%% ────────────────────────────────────────────────────────────────────
\section{Cross-Runtime Benchmarks}
\label{sec:benchmark-programs}

All benchmarks use dialect-agnostic MeTTa (shared syntax) unless noted.
All timing rows in this section were rerun on the current tree on
2026-06-27 unless they are explicitly described as excluded,
methodological, or unresolved.  For rows whose single-run wall time is
below 0.1\,s, we report averaged full-process wall time over repeated
invocations to avoid timer quantization.  Output is verified identical
across runtimes whenever a row is presented as a clean cross-runtime
benchmark.  So a reader can see
precisely what is measured, each subsection below opens with the core of
the program it runs; the full sources are linked inline and collected at
\href{https://github.com/zariuq/CeTTa/tree/main/benchmarks/chaining_crossengine}{\code{benchmarks/chaining\_crossengine}}.

Most of the inference benchmarks are driven by one small dependent-typed
\emph{backward chainer}.  To prove a goal \code{(: \$proof \$theorem)} it
either matches a fact in the knowledge base, or---for a rule application
\code{(\$rule \$prem)}---looks up the rule's type and recurses on its
premise (the binary and $n$-ary cases follow the same shape):
\begin{lstlisting}[language={},basicstyle=\ttfamily\scriptsize]
(: bc (-> $a Nat $b $b))                                  ; KB, depth, goal -> goal
(= (bc $kb $_ (: $proof $theorem))                        ; base case: it's a fact
   (match $kb (: $proof $theorem) (: $proof $theorem)))
(= (bc $kb (S $d) (: ($rule $prem) $theorem))             ; step: apply a rule
   (let* (((: $rule (-> (: $prem $t) $theorem))
            (bc $kb $d (: $rule (-> (: $prem $t) $theorem))))
          ((: $prem $t) (bc $kb $d (: $prem $t))))
     (: ($rule $prem) $theorem)))
\end{lstlisting}
Each benchmark below supplies its own typed rules and facts to this
chainer (or, where noted, a forward chainer or a truth-value kernel).

\subsection{Shallow Evaluation}

Nil Geisweiller's typed backward chainer (the \code{bc} above) over
60 ground entities at depth~1 (20 researchers, 20 engineers, 20
artists).  The rules are typed binary implications; for example,
\code{r-mortal} combines a researcher fact with a curious fact to
derive a mortal witness:
\begin{lstlisting}[language={},basicstyle=\ttfamily\scriptsize]
(: r-mortal (-> (: $r (researcher $x))
                (: $c (curious $x))
                (mortal $x)))
!(bc &kb (fromNumber 1) (: $proof (mortal $x)))      ; -> 60 witnesses
\end{lstlisting}
All three runtimes produce identical 60 proof-carrying witnesses, e.g.\
\code{(: (r-mortal r01 r01c) (mortal r01))}.

\begin{center}
\begin{tabular}{lrrl}
\toprule
Runtime & Time & vs.\ \CeTTa{} & Output \\
\midrule
\CeTTa{} & 4.0\,ms avg  & 1$\times$           & 60 witnesses \\
\PeTTa{} & 45.4\,ms avg & 11.3$\times$ slower & 60 witnesses \\
\HE{}    & 402\,ms avg  & 100$\times$ slower  & 60 witnesses \\
\bottomrule
\end{tabular}
\end{center}

This is a startup-dominated sanity row, not a scaling claim.  It is
useful because it exercises the typed proof-carrying surface while
remaining small enough that translator and runtime drift are easy to
spot.

\subsection{Deep Proof Search: \PeTTa{} Wins}

Metamath proof search at depth~5 (2 axioms, proving $t = t$,
85{,}013 proof trees) reverses the performance ordering.  The knowledge
base is Geisweiller's \code{demo0} (axiom \code{a1}: equality is
right-Euclidean; \code{a2}: $0$ is a right identity; \code{mp}: modus
ponens), and the goal is \code{t = t}
(\href{https://github.com/zariuq/CeTTa/blob/main/benchmarks/chaining_crossengine/01_nilbc/nilbc_th1.metta}{\code{01\_nilbc/nilbc\_th1.metta}};
glyphs shown ASCII here):
\begin{lstlisting}[language={},basicstyle=\ttfamily\scriptsize]
(: a1 (-> (: $t term) (: $r term) (: $s term)
          (-> (= $t $r) (-> (= $t $s) (= $r $s)))))   ; right-Euclidean
(: a2 (-> (: $t term) (= (+ $t 0) $t)))               ; 0 right-identity
(: mp (-> (: $maj (-> $P $Q)) (: $P wff) (: $Q wff)
          (: $min $P) $Q))                            ; modus ponens
!(bc &kbh (fromNumber 5) (: $prf (= t t)))            ; prove t = t
\end{lstlisting}

\begin{center}
\begin{tabular}{lrrl}
\toprule
Runtime & Time & vs.\ \PeTTa{} & Status \\
\midrule
\PeTTa{} & 7.60\,s  & 1$\times$          & 85{,}013 proofs \\
\CeTTa{} & 36.42\,s & 4.8$\times$ slower & 85{,}013 proofs \\
\HE{}    & $>$120\,s & timeout            & killed \\
\bottomrule
\end{tabular}
\end{center}

\PeTTa{}'s Prolog search remains decisively better on deep branching
proof search; the current rerun widens the gap to roughly
4.8$\times$.  \HE{} still cannot finish within 120\,s.

\subsection{Forward Chaining at Scale}

Nil Geisweiller's propositional-calculus forward chainer at depth~4
uses a solution space that is extended in place by forward application
of modus ponens:
\begin{lstlisting}[language={},basicstyle=\ttfamily\scriptsize]
(= (fs-add-to-ss $rb $ss)
   (with-space-snapshot $ss0 $ss
     (match $ss0 (: $x1 $a1)
       (match $ss0 (: $x2 $a2)
         (match $rb (: $f (-> $a1 $a2 $b))
           (add-atom $ss (: ($f $x1 $x2) $b)))))))
!(count-atoms &ss)
\end{lstlisting}
The current \CeTTa{} rerun of
\href{https://github.com/zariuq/CeTTa/blob/main/tests/nil_pc_fc_d4.metta}{\code{tests/nil\_pc\_fc\_d4.metta}}
completes in 100.86\,s and reports 8.19M accumulated results after the
four forward passes.  We do \emph{not} reuse the older
``11.7M generated pairs'' sentence as benchmark evidence, because the
current HE lane does not time out on this driver; it errors immediately
(\code{match expects a space as the first argument}).  So the forward
row remains a useful workload, but it is not yet paper-ready as a clean
cross-runtime benchmark.

\subsection{Translator Validation}

A PeTTa-native benchmark using \code{prog1} (PeTTa-specific construct)
was translated PeTTa$\to$HE and run on \CeTTa{}.  The translator
converts \code{prog1} to nested \code{let} with fresh variables:
\begin{lstlisting}[language={},basicstyle=\ttfamily\scriptsize]
; PeTTa original:
(= (run-query) (prog1 (bc &kb ...) (println! "done")))
; Translated to HE:
(= (run-query) (let $__tr_result_1 (bc &kb ...)
                 (let $__tr_discard_2 (println! "done")
                   $__tr_result_1)))
\end{lstlisting}
The current source file
\href{https://github.com/zariuq/CeTTa/blob/main/tests/bench_backchain_petta_native.metta}{\code{tests/bench\_backchain\_petta\_native.metta}}
produces 30 proof-carrying witnesses on native \PeTTa{}, and the
current translated artifact produces the same 30 witnesses on both
\CeTTa{} and \HE{}:
\begin{center}
\begin{tabular}{lrrl}
\toprule
Runtime & Time & vs.\ \CeTTa{} & Output \\
\midrule
\CeTTa{} & 3.7\,ms avg  & 1$\times$          & 30 proofs \\
\PeTTa{} & 44.8\,ms avg & 12.1$\times$ slower & 30 proofs \\
\HE{}    & 319\,ms avg  & 86$\times$ slower  & 30 proofs \\
\bottomrule
\end{tabular}
\end{center}
This row is a translator benchmark, not just a syntax example: the
observable runtime behavior survives translation.

\subsection{PLN Evidence Aggregation over Genomic KB}

To benchmark inference over real data, we run PLN evidence aggregation
over the Rejuve genomic knowledge base (chr16 pilot: 41~GWAS SNPs,
183~SNP--gene pairs, 3~evidence channels).  Each pair carries
eQTL tissue counts (GTEx, 49~tissues), ABC enhancer--gene contact
counts, and a regulatory presence flag.  The PLN evidence quantale
combines channels via revision (\code{evidence-hplus}), producing a
calibrated SimpleTruthValue per pair.  The evidence operations are
Lean-verified: \code{power}, \code{toLogOdds\_power\_mul}, and
\code{power\_power\_inv} are proved in
\code{Mettapedia/Logic/BinaryEvidence.lean}.

The core arithmetic is ordinary MeTTa over numbers:
\begin{lstlisting}[language={},basicstyle=\ttfamily\scriptsize]
(= (Truth_ModusPonens (stv $f1 $c1) (stv $f2 $c2))
   (stv (+ (* $f1 $f2) (* 0.02 (- 1 $f1))) (* $c1 $c2)))
(= (Truth_Revision (stv $f1 $c1) (stv $f2 $c2))
   (let* (($w1 (Truth_c2w $c1)) ($w2 (Truth_c2w $c2)) ($w (+ $w1 $w2))
          ($f (/ (+ (* $w1 $f1) (* $w2 $f2)) $w)) ($c (Truth_w2c $w)))
     (stv (min 1.0 $f) (min 1.0 $c))))
\end{lstlisting}
The benchmark is dialect-agnostic (shared MeTTa, no translation
needed).  The current flattened chr16 pilot driver produces identical
183 scored pairs on \CeTTa{}, \PeTTa{}, and \HE{}:

\begin{center}
\begin{tabular}{lrrrl}
\toprule
Runtime & Time & vs.\ \CeTTa{} & Output & Status \\
\midrule
\CeTTa{} & 9.5\,ms avg & 1$\times$ & 183 bio-scores & identical \\
\PeTTa{} & 62\,ms avg  & 6.6$\times$ slower & 183 bio-scores & identical \\
\HE{}    & 4.80\,s     & 505$\times$ slower & 183 bio-scores & identical \\
\bottomrule
\end{tabular}
\end{center}

Scores carry genuine biological signal: \code{rs9923732}$\to$\code{ENSG00000166455}
scores strength 0.735 (131 ABC contacts across cell types), while
most pairs score 0.04 (minimal evidence).

\paragraph{What we no longer claim here.}
Earlier drafts included a synthetic scaling table from 143 to
1.4M atoms.  We no longer treat that table as benchmark evidence in the
paper.  The current \CeTTa{} 1.4M checkpoint regression reruns with the
expected semantics (\code{34} proofs, \code{8} unique supported-drug
hypotheses, \code{2} thresholded hypotheses), but the current
PeTTa-compatible flat/top checkpoint lanes still fail at import time.
So the small 183-pair core remains grounded, while the large-scale
cross-engine WM-PLN loader comparison is withheld until those
compatibility seams are repaired.

\subsection{Genomic Pattern Mining}

To exercise real algorithmic depth, we run the Hyperon pattern
miner~\cite{metta} on real GTEx eQTL data, discovering surprising
co-regulation patterns---pairs of SNPs or genes that share eQTL
evidence more often than expected by independence.  The miner is run
unmodified from
\href{https://github.com/iCog-Labs-Dev/hyperon-miner}{github.com/iCog-Labs-Dev/hyperon-miner};
its core loop alternates \code{build-specialization} (extend a candidate
pattern by one link) with a \code{surprisingness} score that keeps only
patterns whose support exceeds the independence expectation.  The data
is extracted from the Rejuve genomic KB (338M Prolog lines, 57\,GB) and
discretized into categorical attributes (tissue breadth, effect size)
for pattern mining.

The current audit does \emph{not} reuse the older cross-engine mining
speed table as benchmark evidence.  What \emph{is} currently grounded is
the portability boundary of the miner's specialization core.  The
original PeTTa-style dynamic-head tuple encoding in
\code{replace}/\code{replacev} is cross-dialect fragile; a tagged-return
rewrite restores identical specialization-core behavior on \CeTTa{} and
\PeTTa{} in the current probe, which produces the same 9
\code{SpecializationOf} atoms on both runtimes.  The remaining HE blocker
is a stateful pattern inside \code{superpose}:
\begin{lstlisting}[language={},basicstyle=\ttfamily\scriptsize]
(superpose ((remove-atom $specspace $spec1)
            (add-atom    $specspace $spec1)
            (remove-atom $specspace $spec2)
            (add-atom    $specspace $spec2)
            (remove-atom $specspace $spec3)
            (add-atom    $specspace $spec3)))
\end{lstlisting}
Current \CeTTa{} preserves all three specialization atoms in this
pattern, while current \HE{} still drops the third one.  So in this
revision we keep the miner as a methodological workload and as a
portability/debugging result, but not as a fresh cross-engine speed
table.

\subsection{Backward Chaining over Mined Patterns}

The mined co-regulation patterns serve as typed inference rules for
Geisweiller's \code{nilbc} backward chainer (the same dependent-typed
\code{bc} listed at the start of Section~\ref{sec:benchmark-programs}), closing the
\emph{mine$\to$chain} pipeline.  The knowledge base combines:
\begin{itemize}
\item 100{,}000 real GTEx eQTL links (40{,}484~SNPs $\times$ 14{,}043~genes),
\item 13 co-regulation patterns and 7 co-expression patterns (mined,
  surprisingness $>$\,0.98),
\item 3 typed PLN rules: \emph{direct-target} (eQTL $\to$ target),
  \emph{coreg-transfer} (co-regulation $+$ target $\to$ target), and
  \emph{coexpr-transfer} (target $+$ co-expression $\to$ target).
\end{itemize}
The three rules are typed implications fed to the \code{bc} chainer of
Section~\ref{sec:benchmark-programs}; \emph{direct-target} reads an eQTL
fact, \emph{coreg-transfer} chains a co-regulation edge onto an existing
target:
\begin{lstlisting}[language={},basicstyle=\ttfamily\scriptsize]
(rule-sig dt (-> (: $e (eqtl $snp $gene))            ; direct-target
                 (target $snp $gene)))
(rule-sig ct (-> (: $e1 (coreg $snp1 $snp2))         ; coreg-transfer
                 (: $e2 (target $snp2 $gene))
                 (target $snp1 $gene)))
!(bc &kb (fromNumber 2) (: $proof (target rs10000544 $gene)))
\end{lstlisting}
Bridge base cases let \code{nilbc} consume untyped data atoms
(\code{eqtl-link}, \code{co-regulate}) directly, avoiding a format
conversion pass over the 100K-atom KB\@.

On the current focused driver
\href{https://github.com/zariuq/CeTTa/blob/main/benchmarks/bench_mine_chain_focus_20260627.metta}{\code{benchmarks/bench\_mine\_chain\_focus\_20260627.metta}},
\CeTTa{} and \PeTTa{} agree exactly on the direct and depth-2 target
sets for three representative query SNPs:

\begin{center}
\begin{tabular}{lrrr}
\toprule
Query SNP & direct targets & depth-2 targets & novel \\
\midrule
rs10000544 & 9 & 12 & 3 \\
rs10000620 & 5 & 5  & 0 \\
rs10000960 & 6 & 8  & 2 \\
\bottomrule
\end{tabular}
\end{center}

The five novel depth-2 targets are:

\begin{center}
\begin{tabular}{llll}
\toprule
Gene & Symbol & Via SNP & Evidence path \\
\midrule
ENSG00000150961 & SEC24D  & rs10000795 & \code{(ct cr3 (dt eq37))} \\
ENSG00000172399 & MYOZ2   & rs10000795 & \code{(ct cr3 (dt eq39))} \\
ENSG00000178636 & MRPL42P1 & rs10000726 & \code{(ct cr1 (dt eq30))} \\
ENSG00000248394 & FOSL1P1  & rs10001048 & \code{(ct cr25 (dt eq87))} \\
ENSG00000273077 & (lncRNA)  & rs10001048 & \code{(ct cr25 (dt eq89))} \\
\bottomrule
\end{tabular}
\end{center}

The proof terms are machine-checkable evidence chains: \code{(ct cr3
(dt eq37))} reads ``co-regulation transfer via \code{cr3}
(rs10000544$\leftrightarrow$rs10000795), applied to direct eQTL
\code{eq37} (rs10000795$\to$SEC24D).''  On this focused three-query
driver the total wall-clock time is 0.15\,s on \CeTTa{} versus
3.33\,s on \PeTTa{}.  We therefore keep the novel-gene discovery story,
but we do \emph{not} reuse the older proof-count/time row as benchmark
evidence in this revision.

Biologically, the novel discoveries include MYOZ2 (hypertrophic
cardiomyopathy gene, OMIM CMH16) and SEC24D (Cole-Carpenter bone
syndrome), both on chr4q26---suggesting a trans-regulatory link from a
metabolic locus (PDE5A/FABP2) to cardiac and skeletal gene targets via
shared cis-regulatory architecture.

The older ``all 7~query SNPs plus reverse queries'' timing line is not
used as a fresh paper benchmark here; the audited evidence in this
revision is the focused cross-runtime driver above.

\paragraph{Complete benchmark ladder.}
The following table summarizes all benchmarks, including both genomic
inference (where \CeTTa{} wins) and pure search (where \PeTTa{} wins):

\begin{center}
\begin{tabular}{lrrrl}
\toprule
Benchmark & Scale / note & \CeTTa{} & \PeTTa{} & \HE{} \\
\midrule
Shallow typed BC & 60 proof witnesses & 4.0\,ms avg & 45.4\,ms avg & 402\,ms avg \\
Translator (\code{prog1}) & 30 proof witnesses & 3.7\,ms avg & 44.8\,ms avg & 319\,ms avg \\
PLN evidence core & 183 scored pairs & 9.5\,ms avg & 62\,ms avg & 4.80\,s \\
Mine$\to$chain focus & 3 query SNPs, 5 novel genes & 0.15\,s & 3.33\,s & --- \\
Deep 17-phase genomic BC & 1{,}411{,}430 hypotheses & 32.32\,s & 55.43\,s & --- \\
Deep proof search & 85{,}013 proofs & 36.42\,s & 7.60\,s & $>$120\,s \\
\bottomrule
\end{tabular}
\end{center}

\CeTTa{} still wins on several real inference rows, but the revised
ladder is intentionally smaller than earlier drafts.  We exclude the
historical cross-engine miner speed table, the current 1.4M WM-PLN
loader comparison, BTree, the current 8-Queens row, and the full
tilepuzzle row from this ladder because the present tree either lacks a
stable current cross-runtime driver for them or exposes a current
correctness/comparability issue.  The benchmark rows shown here are only
those rerun directly in this audit.

\subsection{Cross-Engine Chaining Suite}
\label{sec:chaining-suite}

To ground the inference-side comparison on community-authored code
rather than bespoke benchmarks, we assembled a cross-engine chaining
suite that builds matched cross-engine rows from the same chainer
families on \CeTTa{} (\code{--lang he}) and \PeTTa{}, with a per-row
cross-engine contract (every result is
validated against the other engine's output, never against \CeTTa{}'s
own output).  The sources are
Geisweiller's dependent-typed DTL chainers and Metamath \code{demo0}
proofs (\href{https://github.com/ngeiswei/chaining}{github.com/ngeiswei/chaining}),
Treutlein's \code{PeTTaChainer}
(\href{https://github.com/rTreutlein/PeTTaChainer}{github.com/rTreutlein/PeTTaChainer}),
and the \PeTTa{} engine and its PLN library
(\href{https://github.com/patham9/PeTTa}{github.com/patham9/PeTTa}).
The suite, generators, and driver are in the \CeTTa{} repository under
\href{https://github.com/zariuq/CeTTa/tree/main/benchmarks/chaining_crossengine}{\code{benchmarks/chaining\_crossengine}}.
In this paper revision we use the suite as audited reproducibility
material rather than as an extra source of unevaluated claims: the
benchmark rows retained in the paper are only the subset rerun directly
in this audit.

\paragraph{Multi-relational deep chaining.}
The strongest current multi-relational row is the 17-phase snapshot
\href{https://github.com/zariuq/CeTTa/blob/main/tests/bench_bio_bc_deep_snapshot_20260327.metta}{\code{tests/bench\_bio\_bc\_deep\_snapshot\_20260327.metta}}.
It extends the mined-pattern chainer with pathway membership,
same-pathway transfer, disease associations, compound--disease
treatment links, and compound--gene binding links.  The full
variant-to-drug row is still a typed backward-chaining program:
\begin{lstlisting}[language={},basicstyle=\ttfamily\scriptsize]
(rule-sig dg (-> (: $e1 (target $snp $gene))
                 (: $e2 (has-disease $gene $disease))
                 (disease-reach $snp $disease $gene)))
(rule-sig td (-> (: $e1 (disease-reach $snp $disease $gene))
                 (: $e2 (treats $compound $disease))
                 (drug-hypothesis $snp $compound $disease $gene)))
\end{lstlisting}
The current rerun produces exactly \textbf{1{,}411{,}430 hypotheses}
across 17 query phases, with the same totals on both runtimes.  The
largest buckets are 613{,}978 \code{pw-target} hypotheses and
796{,}964 \code{deep-pw} hypotheses; the smaller focused buckets are
11~binding hypotheses, 11~disease hypotheses, 210~drug hypotheses,
240~deep-drug hypotheses, and 16~pathway-reach hypotheses.  The current
timings are \textbf{32.32\,s on \CeTTa{}} versus 55.43\,s on
\PeTTa{}.
The textbook chain
rs10000544 $\to$ PDE5A $\to$ hypertension $\leftarrow$ Sildenafil
composes three independent data sources (GTEx, Hetionet, DrugBank) by
typed backward chaining with proof terms at every step.

\paragraph{Dependent binder encoding.}
PLN rule declarations use \code{(rule-sig name type)} instead of
\code{(: name type)} to avoid a current HE spec limitation: the
\code{(: \$e T)} syntax in arrow-domain positions is treated as a
literal expected type, not a dependent binder, causing \code{BadArgType}
on both HE and \CeTTa{} (see Section~\ref{sec:heprime}).
The \code{rule-sig} encoding is semantically equivalent on all runtimes.

%% ────────────────────────────────────────────────────────────────────
\section{Performance Analysis}

\subsection{Current profiling status}

Earlier drafts included exact Callgrind and Massif tables for the
SymbolId conversion and for tilepuzzle.  Those low-level profiles were
not rerun as part of the present benchmark re-audit, so we do not use
their historical percentages as current measurement claims here.  The
qualitative engineering picture is unchanged---string-heavy dispatch is
no longer the main story, and the next bottlenecks are still binding
management, variable renaming, and arena allocation---but the exact
profile tables are omitted until they are rerun on the current build.

\subsection{Tilepuzzle as a live regression target}

The 8-puzzle BFS remains a useful WAM-class stress test, but the current
rerun exposes a correctness/comparability issue rather than a clean
paper benchmark row.  The current \CeTTa{} port
(\href{https://github.com/zariuq/CeTTa/blob/main/tests/test_tilepuzzle.metta}{\code{tests/test\_tilepuzzle.metta}})
returns 181{,}440 reachable states in 26.41\,s, while the current
\PeTTa{} reference example
(\href{https://github.com/patham9/PeTTa/blob/main/examples/tilepuzzle.metta}{\code{examples/tilepuzzle.metta}})
self-checks 181{,}441 states in 1.40\,s.  We therefore exclude
tilepuzzle from the benchmark ladder in this revision and treat it as a
live debugging target for state-accounting and port faithfulness rather
than as settled benchmark evidence.

\subsection{Structured Space Kinds}

\CeTTa{} now supports three ordered space kinds beyond plain atomspaces:

\begin{center}
\begin{tabular}{lll}
\toprule
Kind & Discipline & Use case \\
\midrule
\code{queue} & FIFO push/pop & BFS frontiers, work queues \\
\code{stack} & LIFO push/pop & DFS, continuation stacks \\
\code{hash} & O(1) membership & Visited sets, deduplication \\
\bottomrule
\end{tabular}
\end{center}

These are created via \code{(new-space queue)}, \code{(new-space hash)},
etc., and support universal pattern matching through the standard
\code{match} API\@.  The tilepuzzle uses queue-space for its BFS
frontier and hash-space for visited-set deduplication, reducing
peak memory from 6.3\,GB (plain atomspace) to 4.68\,GB.

\subsection{Architecture: Three Machines}

Profiling and design analysis (informed by GPT-5.4 Pro architectural
review) identifies three semantically distinct machines that should
not share mutable state:

\begin{enumerate}
\item \textbf{Local search machine.}
  Branch-local bindings, trail/rollback, choice frames, scratch arenas.
  WAM-inspired: cheap undo via watermarks, not clone/free churn.
  This is the next implementation priority.

\item \textbf{Indexed shared-space machine.}
  Persistent PathMap/MORK-backed spaces, snapshot/epoch discipline,
  conjunction query planning, leapfrog triejoin.
  Deterministic and committed---no backtracking through the index.

\item \textbf{Type/elaboration machine} (future).
  Scoped binders, metavariables, constraint solving, delayed
  assignments.  Separate from runtime bindings.
\end{enumerate}

Language profiles (\code{he\_compat}, \code{he\_extended},
\code{he\_prime}, future \code{petta}) are compositional assemblies
over these machines, not ad-hoc forks in the evaluator.

\subsection{Faithful Backend Obligations}

The indexed shared-space machine needs a sharper correctness contract than
``the backend is fast and seems to return the same answers.''  Recent Lean
work factors the backend story into three explicit runtime seams:
\begin{enumerate}
\item \textbf{Candidate selection.}
  The index narrows candidate atoms, while \CeTTa{}'s native matcher performs
  exact variable-sensitive matching.  This is the two-phase architecture now
  validated by concrete selector theorems over \PathMap{} tries and by a
  concrete matcher instantiation on the \HE{} side.
\item \textbf{Canonical storage.}
  Storage may quotient results to canonical support, but it must preserve
  co-reference.  Repeated source variables must remain repeated after storage
  and materialization; distinct variables must remain distinct.
\item \textbf{Revision-based caching.}
  Cached query answers are valid only across mutations that preserve the
  relevant support boundary.  Duplicate adds or high-count decrements may
  preserve cache validity; first adds and last removes must invalidate.
  Variant-normalized cache keys must also preserve right-hand-side answers and
  result shape under materialization.
\end{enumerate}

This yields a practical division of labor for the PathMap/MORK backend:
\begin{enumerate}
\item \PathMap{} stores canonical support and performs prefix-based candidate
  narrowing;
\item the native matcher remains the source of truth for exact bindings and
  binder hygiene;
\item count metadata, not trie-key distortion, carries bag semantics.
\end{enumerate}

\paragraph{Verified runtime contracts.}
Recent Lean work extends the backend correctness story with three new contract files:
\begin{itemize}
\item \code{CeTTaRuntimeContracts.lean} (20 theorems): canonicalization injectivity and BEq preservation (\code{term\_canon.c}), search-context save/rollback correctness (\code{search\_machine.c}), effect classification safety (\code{runtime\_effect\_classes.json}), and profile hierarchy (\code{session.c}).  Each theorem maps to a concrete refactoring risk it reduces.
\item \code{IncrementalTableSemantics.lean} (15 theorems): specifies partial-table soundness, answer monotonicity, and completion for future tabling in \code{table\_store.c}.  Any correct implementation must satisfy \code{TableEntry.Sound} (partial reads are genuine) and \code{TableEntry.Exact} (completed tables equal \code{queryEquations}).
\item \code{ProducerChoicePoint.lean} (10 theorems): formalizes the search-machine save/rollback/nested-choice-point protocol with combined bindings+scratch state.
\end{itemize}

Crucially, \CeTTa{}'s imported backend is binding-correct not through trusted direct bridge bindings, but because candidate narrowing is followed by seeded native rematch (\code{space\_subst\_match\_with\_seed}).  The theorem \code{imported\_seedRematch\_binding\_parity} in \code{SeedRematchContract.lean} formalizes this: when imported candidates are sound, the two-phase query produces the same result set as direct matching.

\paragraph{Positive example.}
Adding an existing atom to a counted space may increment multiplicity without
changing trie support or invalidating unrelated cached queries.

\paragraph{Negative example.}
Letting the backend bridge perform full unification over serialized query
terms invites exactly the binder-hygiene failures that arise when
canonicalization preserves shape but breaks co-reference.

These obligations are now mapped one-for-one to runtime seams:
\code{space\_match\_backend.c} for candidate selection, \code{term\_universe.c}
for canonical storage, and \code{table\_store.c} for cache invalidation.  The
companion note \emph{Faithful Backends for HE MeTTa} develops the full proof
story, including the variant-query and binding-cache lemmas; the present
paper uses those results to justify the architecture of \CeTTa{}'s indexed
shared-space machine.

\subsection{MM2 Integration}

MM2 is a deterministic, priority-scheduled, forward-chaining term rewriting
language implemented by MORK\@.  Its programs are not a separate execution
layer---they are atoms in MORK-backed spaces, stepped by MeTTa via
\code{step!}.  The hybrid interaction requires no special protocol: MeTTa
creates spaces, adds data, steps MM2 rules, and reads results through ordinary
\code{match}.

\begin{verbatim}
!(bind! &ws (new-space mork))
!(add-atom &ws (exec (0 step)
  (, (edge $x $y) (edge $y $z))
  (O (+ (path $x $z)))))
!(add-atom &ws (edge a b))
!(add-atom &ws (edge b c))
!(step! &ws)
!(match &ws (path $x $y) ($x $y)) ; -> (a c)
\end{verbatim}

Three import paths exist: \code{.mm2} file import (\code{import!} or
\code{include} into a target space, which auto-enables the MORK backend),
inline \code{add-atom} of exec rules from MeTTa, and compiled \code{.act}
artifact loading via \code{mork:open-act}.

Because MM2 rules are ordinary atoms, MeTTa can pattern-match over them as
data---inspecting priorities, rewriting source/sink templates, or extracting
rules for analysis:
\begin{verbatim}
!(match &ws (exec $p $src $snk) (rule $p))
\end{verbatim}

\CeTTa{}'s \code{mm2\_lower.c} handles bidirectional
surface$\leftrightarrow$IR lowering
(\code{exec}$\leftrightarrow$\code{mm2\_exec},
\code{,}$\leftrightarrow$\code{mm2\_pattern\_and}, etc.), preserving
round-trip fidelity between MM2 surface syntax and \CeTTa{}'s internal
representation.

%% ────────────────────────────────────────────────────────────────────
\section{The MeTTa Dialect Landscape}

\subsection{Shared Fragment}

85\% of MeTTa programs use only shared constructs (\code{let},
\code{let*}, \code{match}, \code{if}, \code{case}, \code{bind!},
\code{add-atom}, \code{new-space}).  These run identically on all three
runtimes without translation.  The Metamath proof search benchmark is
dialect-agnostic.

\subsection{Translated Constructs}

\begin{center}
\begin{tabular}{lll}
\toprule
Direction & Construct & Translation \\
\midrule
\HE{}$\to$\PeTTa{} & \code{chain} & \code{let} \\
& \code{collapse-bind} & \code{collapse} \\
& \code{superpose-bind} & \code{superpose} \\
& \code{nop X} & \code{let \$fresh X ()} \\
& \code{function (return X)} & \code{X} (unwrap) \\
\PeTTa{}$\to$\HE{} & \code{progn A B} & \code{let \$fresh A B} \\
& \code{prog1 A B} & \code{let \$r A (let \$d B \$r)} \\
& \code{@<} & \code{<s} \\
\bottomrule
\end{tabular}
\end{center}

\subsection{Semantic Divergence: Constraint Store vs.\ Independent Scope}

The translation preserves semantics on the \emph{pure fragment}
(ground terms, or variables scoped by \code{let*}).  A genuine
divergence exists for free variables in non-deterministic branches:
\begin{lstlisting}[language={},basicstyle=\ttfamily\scriptsize]
(= (deduce (And $a $b)) (And (deduce $a) (deduce $b)))
\end{lstlisting}
When \code{deduce} returns via \code{match \&self} with a free
variable~\code{\$x}, \HE{}/\CeTTa{} bind \code{\$x} consistently
across both branches of \code{And}, while \PeTTa{} gives each branch
its own copy.

This divergence has a precise theoretical characterization.
Saraswat et al.~\cite{saraswat1988} showed that concurrent constraint
languages based on Tell-and-Ask over a shared store satisfy a
\emph{monotonicity property}: constraints accumulate monotonically,
and any reduction possible in a later configuration was already
possible in an earlier one (their Proposition~2.1).  \HE{}/\CeTTa{}'s
\code{match~\&self} operates on a shared constraint store---bindings
produced by one branch propagate to all others via the shared space.
\PeTTa{}'s Prolog backend evaluates branches with independent variable
scopes, closer to what Saraswat et al.\ identify as the
``read-only variable'' model of Flat Concurrent Prolog, which they
explicitly note does \emph{not} possess the monotonicity property.

\paragraph{The depth-1 standard form.}
Saraswat et al.\ define a clause to be in \emph{standard form} when
each atom in the body is depth-1 (no shared variables between
conjunction branches).  Nil Geisweiller's typed backward chainer
uses exactly this form: \code{let*} threads variable bindings
explicitly between premises, eliminating implicit sharing.  This is
why the typed chainer works identically on all three runtimes---it
stays within the standard-form fragment where the constraint-store
and independent-scope semantics coincide.

\paragraph{Lean formalization.}
The \code{Translatable} and \code{PureTranslatable} predicates in
\code{HEPeTTaSound.lean} characterize precisely this fragment: atoms
where \code{atomToPattern} succeeds and produces
\code{isMatchCorrect} patterns (no lambda, subst, or collection
nodes).  The sorry-free proof establishes that equation-matching
semantics are preserved under translation on this fragment.  Extending
the proof to the shared-variable conjunction fragment would require
formalizing the monotonic constraint-store semantics of \HE{}'s
\code{match~\&self}---a natural target for Phase~3 of the Lean proof.

%% ────────────────────────────────────────────────────────────────────
\section{Non-determinism and Mutable State}
\label{sec:nondet-state}

A fundamental specification gap in MeTTa concerns the interaction of
non-deterministic evaluation with mutable space operations.  When
\code{let~\$x~(match~\&db~...)~(add-atom~\&result~(stored~\$x))}
executes with multiple match results, what happens to the shared
\code{\&result} space?

\subsection{The Problem}

Consider:
\begin{lstlisting}[language={},basicstyle=\ttfamily\scriptsize]
!(bind! &db (new-space))
!(add-atom &db (item A))
!(add-atom &db (item B))
!(bind! &result (new-space))
!(let $x (match &db (item $y) $y)
   (superpose ((remove-atom &result (stored $x))
               (add-atom &result (stored $x)))))
!(get-atoms &result)
\end{lstlisting}

\HE{} returns \code{[]}---all atoms lost.  \CeTTa{} returns
\code{[(stored~A)~(stored~B)]}---correct.  The root cause:
\HE{} interleaves side effects from different non-deterministic
branches, so \code{remove-atom} from one iteration cancels
\code{add-atom} from another.  \CeTTa{} evaluates each iteration's
body to completion before starting the next.  This is the exact
pattern used by the Hyperon pattern miner's
\code{build-specialization}, explaining why the miner produces
empty results on \HE{}.

\subsection{Four Possible Semantics}

We analyzed four options across eight representative examples
(accumulation, logging, constraint checking, search with
backtracking, speculative execution, order-dependent processing,
accidental non-determinism, and the real miner code):

\begin{center}
\begin{tabular}{lcccc}
\toprule
Option & Wins & Loses & Character \\
\midrule
Sequential (CeTTa)  & 6/8 & 1/8 & accumulation-friendly \\
Interleaved (HE)    & 0/8 & 4/8 & always worse than sequential \\
Forked (copy-on-write) & 1/8 & 4/8 & search-friendly \\
Illegal (UB)        & 1/8 & 5/8 & safest, least ergonomic \\
\bottomrule
\end{tabular}
\end{center}

Sequential evaluation loses only on \emph{search with backtracking},
where each branch needs isolated state.  Interleaved evaluation
is strictly dominated---it provides no advantage over sequential
while introducing race conditions on shared mutable state.

\subsection{\code{with-space-snapshot}: The Escape Hatch}

For the one case where sequential semantics is wrong (search with
mutable state), \CeTTa{} provides \code{with-space-snapshot}:
\begin{lstlisting}[language={},basicstyle=\ttfamily\scriptsize]
(= (fs-add-to-ss $rb $ss)
   (with-space-snapshot $ss0 $ss
     (match $ss0 (: $x1 $a1)
       (match $ss0 (: $x2 $a2)
         (match $rb (: $f (-> $a1 $a2 $b))
           (add-atom $ss (: ($f $x1 $x2) $b)))))))
\end{lstlisting}

This pattern, from Geisweiller's propositional forward chainer,
reads from a frozen snapshot \code{\$ss0} while writing to the
live space \code{\$ss}---the standard \emph{semi-naive} discipline
for stratified forward inference.  This is a \CeTTa{} extension;
\HE{} does not currently implement it.

\subsection{Proposed Spec Clarification}

The MeTTa spec~\cite{metta} uses sequential Python pseudocode
(line~368: a list comprehension over results) that implies
sequential side-effect ordering, but does not explicitly require
it.  We propose:

\begin{quote}
\emph{When a non-deterministic expression produces multiple results
and those results are bound via \code{let}/\code{chain}/equation
dispatch, the body for each result is evaluated to completion---including
all side effects---before the body for the next result begins.
To evaluate branches with isolated state, use \code{collapse}
first, then iterate explicitly.}
\end{quote}

A future \code{snapshot-space} primitive would provide clean
copy-on-write semantics for search/backtracking without baking
merge semantics into the language core.

%% ────────────────────────────────────────────────────────────────────
\section{HE$'$: Dependent Telescope Extension}
\label{sec:heprime}

Both HE and \CeTTa{} treat \code{(: \$e T)} in arrow-domain positions as a
literal expected type, not a dependent binder.  This means Curry-Howard
backward chainers cannot store rule types as \code{(: dt (-> (: \$e
(eqtl \$snp \$gene)) (target \$snp \$gene)))} in \code{\&self} without
triggering \code{BadArgType}.  This is a current HE spec limitation, not
a \CeTTa{} bug.

We formalize an explicit extension, HE$'$ (he\_prime), as a
\emph{telescope interpretation} of arrow domains.  The extension is
exactly one function: \code{splitDomain}, which recognizes \code{(: \$x
T)} as a dependent binder instead of a literal.  Everything else
(matching, bindings, space operations, equation dispatch) is inherited
from HE unchanged.

The telescope walk checks arguments left-to-right: each successful
type match extends the binding environment, and later domains and the
codomain are instantiated under accumulated bindings.  This is
formalized in Lean (230~lines, 0~sorries) at
\code{HEPrime/Telescope.lean}, with kernel-checked positive and
negative examples:

\begin{itemize}
\item \textbf{Positive:} applying \code{dt} to \code{eqtl\_rs1\_gene1}
  infers return type \code{(target rs1 gene1)} via dependent
  substitution.  Binary rule \code{ct} with cross-domain dependency
  ($\code{\$snp2}$ bound by first argument, constraining second domain)
  also kernel-checked.
\item \textbf{Negative:} mismatched SNP ($\code{rs2} \neq \code{rs1}$)
  correctly produces \code{none}.  Standard HE telescope parsing treats
  the same domain as a plain atom, proving the semantics are distinct.
\end{itemize}

\CeTTa{} implements HE$'$ as an opt-in profile (\code{he\_prime}),
gated by a single flag (\code{enable\_dependent\_telescope}).  The
implementation mirrors the Lean formalization: \code{split\_dependent\_domain},
\code{function\_domain\_type}, \code{bind\_domain\_binder}, and
\code{infer\_dependent\_application\_types} in \code{eval.c}.

%% ────────────────────────────────────────────────────────────────────
\section{Related Work}

\paragraph{Concurrent constraint programming.}
Saraswat et al.~\cite{saraswat1988} established the theoretical
framework for detecting stable properties of networks in concurrent
logic programming languages.  Their cc($\downarrow$,$|$) language,
with Tell/Ask on shared variables over a monotonic constraint store,
provides the semantic model that most closely matches \HE{}/\CeTTa{}'s
\code{match~\&self} evaluation.  The distinction between monotonic
constraint stores and independent-scope evaluation (Flat Concurrent
Prolog) precisely characterizes the semantic boundary between \HE{} and
\PeTTa{}.

\paragraph{MeTTa ecosystem.}
The MeTTaIL boot compilation pipeline~\cite{mettail} provides a
contract-first, Lean-backed compilation path for MeTTa stdlib modules,
complementing \CeTTa{}'s direct evaluation approach.  The Mettapedia
formalization project provides the Lean specifications (\code{EvalSpec},
\code{MeTTaEval}, \code{MatchSpec}) that ground both the translation
proof and the \CeTTa{} evaluator design.  The PLN (Probabilistic Logic
Networks) framework uses MeTTa's backward chaining extensively; the
typed chainer pattern from Nil Geisweiller's \code{nilbc.metta} is
the standard PLN idiom that works across all three runtimes.

%% ────────────────────────────────────────────────────────────────────
\section{Conclusion}

\CeTTa{} demonstrates that a direct C implementation of the MeTTa
evaluator is competitive with \PeTTa{} on inference workloads, but not
uniformly faster.  In the fresh reruns reported here, \CeTTa{} wins on
startup-dominated shallow inference and on several genomic-inference
rows (the 183-pair PLN evidence core and the 1{,}411{,}430-hypothesis
deep genomic backward-chaining row), while \PeTTa{} wins decisively on
pure branching search (the 85{,}013-proof Metamath row).  A
cross-engine chaining suite (Section~\ref{sec:chaining-suite})
confirms that \CeTTa{}'s shallow advantage is a startup-and-dispatch
effect rather than a scaling property.  Historical rows whose current
drivers are unresolved---the old miner speed table, the current 1.4M
WM-PLN seam, 8-Queens, and tilepuzzle---are excluded from the revised
benchmark ladder rather than carried forward on trust.  Throughout,
\CeTTa{} maintains
oracle-verified correctness across a 375-file regression suite and a
333-case rule-indexed conformance catalog (24/24 spec-rule capacities,
all agree) against upstream \HE{}~v0.2.10.
The sorry-free Lean proof of \HE{}$\leftrightarrow$\PeTTa{} semantic
correspondence (5{,}400+ lines, 0~sorries), including import
commutativity and operational bridges for state and space allocation,
establishes that MeTTa's three runtimes are semantically compatible
on the shared fragment.  The compiled \code{.act} space format via
MORK enables zero-copy indexed loading of large knowledge bases.
On the process-calculus side, the \RhoCalc{} reaction machine's
eval-at-\textsc{comm} semantics now has a witness-free, axiom-clean Lean
characterization (Section~\ref{sec:rho-lean}): the quiescent run of the
canonical payload-evaluation process is exactly its structurally observable
outcome set, with no certification witness in the carrier.

Beyond raw performance, \CeTTa{} now treats its own semantics as a
first-class object.  The profile family makes \code{he-compat} (literal
Rust \HE{}) and the principled \code{he} line distinct, recorded
semantic products; the reflective modal surface exposes the one-step
reduction relation---$\Diamond$, $\Box$, normal-form---to MeTTa
programs themselves, aligned with the Lean OSLF derivation; and the
small-step rule table makes the first reduction-rule classes named,
reflectable, and removable, with rule-removal tests proving the engine
depends on the declared rules rather than on duplicate hidden branches.
The congruence bug that motivated this weld---an argument-congruence
rule silently missing from the one-step relation on three branches at
once---is exactly the drift class this architecture eliminates.

Tilepuzzle is now treated as a live regression target rather than as
settled speed evidence.  The fresh rerun exposes a one-state mismatch
between the current ports (\CeTTa{} reaches 181{,}440 states, the
current \PeTTa{} reference self-checks 181{,}441), so the immediate
engineering priority is state-accounting and rollback faithfulness
before reinstating any low-level profiling claim.  Once the counts
agree again, trail-style rollback, bindings builders, and arena
mark/reset remain the main candidates for improving search-heavy
workloads without converting \CeTTa{} into a Prolog engine.  The
guiding principle: \emph{backtracking should be rare, local, and
hidden---not global, ambient, or user-visible.}

\paragraph{Future work.}
\begin{itemize}
\item \textbf{Complete HE one-step coverage:} equation matching is
  table-routed; migrate the special forms (\code{chain}, \code{let},
  \code{let*}, \code{case}, \code{switch}) one rule class at a time,
  each with bag-equality and rule-removal gates.
\item \textbf{\code{--compile-langdef} codegen:} a generator emitting
  both the reflectable pack and specialized C for the hot evaluator
  under one shared source digest, extending the
  \code{--compile-stdlib} blob precedent; an
  evaluate-by-iterating-one-step mode as reference oracle.
\item \textbf{Small-step / official-HE correspondence:} the
  \code{HESmallStep} relation and its OSLF span exist in Lean
  (proven; zero sorries); the remaining climb is the delimited
  collapse/simulation theorems connecting it to the fine
  \code{mettaHE} instruction machine.
\item \textbf{Bounded modal operators:} fuel-bounded
  \code{reachable-within} ($\Diamond^{\le k}$) and
  \code{invariant-within} ($\Box^{\le k}$) over the now-faithful
  one-step relation.
\item \textbf{Local search machine:} trail/watermark rollback,
  \code{BindingsBuilder}, arena mark/reset for scratch atoms.
  Target: materially improve search-heavy workloads such as
  tilepuzzle once the current state-count mismatch is resolved.
\item \textbf{Indexed shared-space engine:} leapfrog triejoin
  (from Vandervorst's \code{trie\_synth.py}), conjunction query
  planning, SpaceEngine boundary (formalized in
  \code{SpaceEngineBoundary.lean}: native $\subset$ pathmap $\subset$ mork,
  with ACT as artifact-only surface; \code{ACTArtifactBoundary.lean}
  classifies \code{open-act}/\code{load-act!}/\code{dump!} as storage seams).
  Target: 10--100$\times$ on large-KB multi-pattern conjunctions.
\item \textbf{MM2 as MeTTa data (working):} MM2 exec rules are atoms in
  MORK-backed spaces.  MeTTa creates, inspects, rewrites, and steps them
  through the ordinary space API\@.  \code{step!} is the execution interface;
  \code{.mm2} files import via \code{include}/\code{import!}.  MeTTa can
  pattern-match over MM2 code to analyze or transform programs---the
  meta-language relationship is native.
\item \textbf{morkl: expert \PathMap{} algebra (future):}  Direct access to
  \PathMap{}'s zipper navigation (5{,}800 lines of Rust), product-zipper
  Cartesian joins (2{,}000 lines), and lattice operations
  (\code{pjoin}/\code{pmeet}/\code{psubtract}/\code{prestrict}, 1{,}600
  lines).  MORK's kernel already implements 15+ sink types (count, sum, head,
  hash, and, Z3, WASM) and source types (BTM, ACT, equality/inequality
  constraints).  The Lean formalization (46 files, 19K+ lines) proves the
  three-phase execution protocol, fold-level aggregation, and \PathMap{}
  lattice correspondence.  The gap is FFI exposure and MeTTa surface: the
  Rust substrate exists, \code{lib/morkl.metta} does not yet.
\item \textbf{HO search engine:} Lash-inspired SAT-driven proof
  search for PLN and type-directed synthesis.  PicoSAT as
  embedded SAT solver.  Separate from the evaluator.
\item \textbf{Translator CLI:} \code{translate.sh} with
  \code{he2petta}/\code{petta2he}, recursive and bundle modes.
  Lean proof covers pure fragment + state + spaces (0~sorries).
\item \textbf{Reaction-semantics formalization:} extend the witness-free
  eval-at-\textsc{comm} characterization (Section~\ref{sec:rho-lean}) from the
  single-reaction bag fragment to concurrent and persistent reactions, and
  relax the hash-set-free channel side condition.
\item Full pattern miner on \CeTTa{} end-to-end
  ($\sim$12$\times$ speedup over \PeTTa{}).
\end{itemize}

\appendix

%% ════════════════════════════════════════════════════════════════════
\section{\LeaTTa{} Cross-Check Appendix}
\label{app:leatta}

\LeaTTa{} is a Lean~4 MeTTa runtime and proof development that is useful
as an independent semantic cross-check for the \HE{}-facing core of
\CeTTa{}.  The local checkout used for the measurements in this
appendix was a clean \LeaTTa{}~1.0.6 tree at
\href{https://github.com/MesTTo/LeaTTa}{\code{github.com/MesTTo/LeaTTa}},
commit
\href{https://github.com/MesTTo/LeaTTa/tree/51b4b52bd02745992448835f5173989baf1436b5}{\code{51b4b52}}
(\code{2026-06-22}, ``prepare~1.0.6 release'').  Upstream release
notes and the checked-claims book are published at
\href{https://mestto.github.io/LeaTTa/}{\code{mestto.github.io/LeaTTa/}}.

This appendix records a local scope-and-spot-check run on 2026-06-24; it
is a coverage map, not a new benchmark ladder replacing the main paper.
The \CeTTa{} side used \code{--profile he-extended --lang he}.  The
\LeaTTa{} side used \code{--oracle} for assertion-bearing files and
\code{--file} for pure loaders.  A bulk pass ran every \code{.metta}
file under the \CeTTa{} tree (1255 files) through \LeaTTa{}.

\paragraph{Coverage, in three tiers.}
The raw aggregate---168 of 951 assertion-bearing files passing in
full---is misleading on its own, because most of the \CeTTa{} test
corpus exercises extension surfaces that \LeaTTa{}, a deliberately
minimal \HE{} interpreter, does not implement and is not meant to.  The
honest decomposition is:

\begin{center}
\small
\begin{tabular}{p{3.3cm}p{11.3cm}}
\toprule
Tier & What \LeaTTa{} does on it \\
\midrule
\HE{} core & \textbf{Strong.}  Of the 21 \code{he\_*} conformance files,
18 pass in full and 2 partially (one loader carries no assertions),
for 203/206 assertions.  Full passes span symbols, equal-chaining,
backchaining, GADTs, dependent types, higher-order functions, grounded
basics, type propagation, states, documentation, and PLN truth values.
This is the surface that overlaps the \HE{} evaluation semantics. \\
\addlinespace
\CeTTa{} extensions & \textbf{Out of scope by design.}  The bulk of the
783 partial failures call builtins \LeaTTa{} does not provide:
MORK-backed spaces (84 files), \PathMap{} (44), \code{hyperpose} (27),
typed \code{new-space hash}/\code{queue} (12), \code{import!} (10),
\code{space-set-match-backend}, and the textual Metamath parser.
Failing these reflects \LeaTTa{}'s minimalism, not a fidelity gap. \\
\addlinespace
Genuine gaps & \textbf{In-scope divergences.}  A float-only numeric
model: \code{(min-atom (5 4 6))} returns \code{4.000000} rather than an
integer, and rationals such as \code{1/2} raise ``only numbers
allowed,'' so the exact integer/rational tower does not correspond.
Typed error contours (e.g.\ \code{(+ 5 "S")} $\to$ \code{BadArgType},
empty/wrong-arity \code{car-atom}/\code{decons-atom}/\code{min-atom})
are not reproduced; we treat these as accepted divergences.  Two
\code{he\_b4} assertions touching inert-symbol and empty-match
semantics, and the higher-order WM-PLN bridge (218/221 on this
checkout), remain to be diagnosed. \\
\bottomrule
\end{tabular}
\normalsize
\end{center}

\paragraph{Spec repair triggered by the cross-check.}
The \LeaTTa{} comparison was useful not only as an oracle but as a
debugger for our own \HE{} Lean surface.  During this pass we found that
the public equation-query path had been routed through the simplified
\code{simpleMatch} helper even though the faithful two-sided
\code{matchAtoms}/\code{mergeBindings} semantics already existed.  The
repair is now written down in three Lean files:
\code{DeclMatchSpec.lean} gives a declarative specification of
\code{match\_atoms} and proves \code{matchAtoms\_sound}/\code{\_complete};
\code{DeclMergeSpec.lean} does the same for
\code{merge\_bindings}, proving \code{mergeBindings\_sound}/\code{\_complete};
and \code{QuerySurfaceAgreement.lean} proves the staged ground-fragment
adequacy theorems
\code{queryEquations\_ground\_two\_way\_adequate} and
\code{queryEquationsAgainstVisible\_ground\_two\_way\_adequate}.  So the
cross-check is now anchored on the faithful ground-query surface rather
than on the older simplified one.  The remaining variable-bearing query
case is deliberately left explicit: it requires the next repair, namely
threading equalities through \code{Bindings.resolve}/\code{Bindings.apply},
and is the right next theorem rather than a hidden assumption.

\paragraph{Performance.}
On matched runs \LeaTTa{} is correct where it is in scope but slower and
markedly more memory-hungry than \CeTTa{}'s C runtime---operational
overhead expected of an executable Lean interpreter, not a semantic
limitation.

\begin{center}
\small
\begin{tabular}{p{4.2cm}p{2.1cm}p{2.1cm}p{5.9cm}}
\toprule
Benchmark & \CeTTa{} & \LeaTTa{} & Note \\
\midrule
tail\_recursion\_deep\_regression & 0.30\,s & 0.80\,s &
both pass 3/3 assertions \\
tail\_recursion\_deep & 0.30\,s & 0.80\,s &
both pass 3/3 assertions \\
he\_c3\_pln\_stv & 0.10\,s & 0.10\,s &
both pass 5/5 assertions \\
he\_d2\_higherfunc & 0.10\,s & 0.10\,s &
both pass 25/25 assertions \\
test\_wmpln\_ho\_similarity\_bridge & 0.10\,s & 0.10\,s &
\CeTTa{} passes; \LeaTTa{} passes 218/221 assertions \\
bio\_refseq\_723k & 0.40\,s & 3.30\,s &
both run; \LeaTTa{} $\approx$1.16\,GB RSS vs.\ \CeTTa{} $\approx$196\,MB \\
bio\_tfbs\_415k & 0.30\,s & 2.10\,s &
both run; \LeaTTa{} $\approx$917\,MB RSS vs.\ \CeTTa{} $\approx$136\,MB \\
\bottomrule
\end{tabular}
\normalsize
\end{center}

\paragraph{Tilepuzzle: surface vs.\ performance.}
The raw \CeTTa{} tilepuzzle benchmark depends on \CeTTa{}'s typed
\code{queue} and \code{hash} spaces, so it does not run unchanged on
\LeaTTa{} (a Tier-2 surface mismatch).  A bounded adaptation replacing
those spaces with a functional queue and plain named-space membership
does run: a 100-step witness reaches 171 visited states in 18.47\,s,
while the 500-step version did not finish under a 30\,s cap.  For
reference, native \CeTTa{} currently reaches 181{,}440 states in
26.41\,s, while the current \PeTTa{} reference self-checks
181{,}441 states.  So this bounded adaptation is only evidence that the
surface idea transfers; the full native row remains a live
state-accounting mismatch rather than settled performance evidence.

\paragraph{Bottom line.}
\LeaTTa{} is a credible independent oracle and a candidate verified
engine-floor for the \HE{} evaluation core of \CeTTa{}: it agrees on the
conformance surface that matters, diverges only on extension builtins it
deliberately omits and on a float-only numeric model, and pays an
expected interpreter-speed cost.  Linking \LeaTTa{} to the \HE{} Lean
specification is therefore well-founded on the core, provided the
numeric tower and error-contour divergences are encoded as known
boundaries.  The specification side is now in a better state than when
this appendix was first drafted: the ground-query fragment is connected
to the faithful \HE{} matcher/merge semantics by proved two-way adequacy,
and the remaining variable-bearing query case has been isolated cleanly
enough to be attacked next rather than smudged into the oracle story.

%% ════════════════════════════════════════════════════════════════════
\section{\textsc{Experimental} --- MAM: The MeTTa Abstract Machine}
\label{app:mam}

\emph{This appendix is explicitly experimental.  It captures design
exploration from ongoing discussions (April~2026) between the authors
and external reviewers (GPT-5.4 Pro), concerning the
post-phase-1f abstract-machine layer for \CeTTa{}.  Nothing here is yet
implemented or committed architecture.  It is recorded to preserve
design rationale; the authoritative implementation plan remains
\code{implementation\_plan.tex}.}

\subsection{Motivation}

The phase-1f-era extraction seam (\code{src/search\_machine.\{h,c\}})
pulls policy parsing, stream combinators, and a \code{TableStore}
plugin out of \code{eval.c}, and wraps them in a thin callback struct
\code{CettaSearchMachine = \{stream\_source, eval\_bound\_single,
  eval\_ctx, table\_plugin\}}, stack-allocated at each call site.
This is a useful refactor, but it is a \emph{plug interface} — not an
abstract machine in the WAM / ZAM / SLG-WAM sense.  Real abstract
machines (XSB's SLG-WAM, SWI's tabling runtime, Soufflé's compiled
Datalog, miniKanren's CPS-transformed search) have persistent machine
state, explicit visitor protocols, capability advertisement, and
episode lifetimes.

The proposal here is a \emph{principled} next layer that retains the
current code as \code{SearchAdapter} (transitional helper) while
naming the real abstraction from first principles.

\subsection{Current categorical reading}

The MAM's present semantic reading, under active revision, models it
as a \textbf{quantale-enriched traced profunctor}, fibrated for
reflection.  This reading is provisional --- it captures the current
best understanding of the structure, not a definitional claim that
binds future architecture.  Unpacked:

\begin{itemize}
\item \textbf{Profunctor} $P : \mathcal{Q}^{\mathrm{op}} \times
  \mathcal{A} \to \mathcal{V}$: \emph{search}.  Queries in
  $\mathcal{Q}$ relate to answer streams in $\mathcal{A}$, with
  intensity weighted by elements of $\mathcal{V}$.
\item \textbf{Quantale enrichment} $\mathcal{V} = (L, \otimes, \le,
  \bigvee)$: \emph{multiplicity}.  Answer weights live in a quantale:
  $\mathcal{V} = \mathbf{2}$ for classical boolean search;
  $\mathcal{V} = \mathbb{N}$ for counted answers;
  $\mathcal{V} = [0,1]^2$ (strength $\times$ confidence) for PLN.
  \PathMap{}'s trie-algebra is natively a quantale.
\item \textbf{Trace} $\mathrm{Tr}_X^{A,B} : \mathrm{Hom}(A \otimes X, B
  \otimes X) \to \mathrm{Hom}(A, B)$: \emph{fixpoint}.  Feedback /
  saturation / tabling / cyclic queries are all instances of
  categorical trace (Joyal--Street--Verity, Hasegawa).
\item \textbf{Kleisli structure over an effect monad} (or algebraic
  effect handlers \emph{\`a la} Plotkin--Pretnar): \emph{control}.
  Demand, cut, suspension, resumption, and schedule are composable
  effect operations.
\item \textbf{Fibration} $p : \mathcal{T} \to \mathcal{B}$ (optional,
  for reflective MeTTa): rules-over-machine-state.
\end{itemize}

A trace operator is provably equivalent to taking the least fixpoint
of a parameterised endofunction (Hasegawa 1997); this is why tabling,
PLN-convergence, and saturation all fit one uniform operator.

\subsection{The proposed machine family: MAM}

We name the machine family \textbf{MAM: MeTTa Abstract Machine}.
The name is chosen by identity rather than by current structure or
scope: MAM is the abstract machine for the MeTTa language family,
and that identification remains accurate as the internal architecture
evolves.  The naming tradition (WAM, ZAM, STG, CEK, CAM) uses either
author identity or component decomposition; MAM uses the language
identity it serves, which is stable across architectural iteration.
Specifically, MAM does \emph{not} hardcode a categorical claim (as
``QAM = Quantale Abstract Machine'' would have done), and does
\emph{not} hardcode a functional scope (as ``SCAM = Search \& Control
Abstract Machine'' would have done).  The categorical reading and the
current feature set are both free to evolve without renaming.

\begin{center}
\begin{tabular}{ll}
\toprule
\textbf{Concept} & \textbf{Name} \\
\midrule
Machine family & \code{MAM} (Quantale Abstract Machine) \\
Formal description & Quantale-enriched traced profunctor \\
Engine 1 (goal-directed, Kleisli/effect) & \code{Chainer} \\
\quad first realization (native) & \code{native\_chainer} \\
Engine 2 (fixpoint, trace) & \code{Saturator} \\
\quad first realization (pathmap) & \code{pathmap\_saturator} \\
Per-session state & \code{Episode} \\
Dispatch vtable & \code{MAMOps} \\
Answer & \code{Answer} / \code{WeightedAnswer} \\
Schedule axis & \code{dfs} / \code{bfs} / \code{iddfs} / \code{interleave} / \code{native} \\
Demand axis & \code{unbounded} / \code{once} / \code{limit-k} / \code{fold} \\
Visitor control & \code{CONTINUE} / \code{STOP} / \code{ERROR} / \code{SUSPEND} \\
\bottomrule
\end{tabular}
\end{center}

\subsection{The two-engine pressure strategy}

Rather than over-specifying a first concrete machine, MAM is
pressured from day~1 by \emph{two} independent realizations that
stress complementary axes of the seam:

\begin{description}
\item[\code{Chainer} (goal-directed).]  WAM-lineage control: explicit
  environments/frames, choicepoints, trail/undo, deterministic
  tail-path handling, bounded-demand response, direct conjunction
  when advertised.  Design-guide workload: \textbf{metamath backward
    chaining} (where \PeTTa{} currently completes and \CeTTa{}~OOMs).
  First realization: \code{native\_chainer}.
\item[\code{Saturator} (fixpoint).]  SLG / Datalog / saturation
  lineage: snapshots, delta-sets, conjunction planning, exact
  membership, counted answers, eventual path toward PLN and
  ATP-style saturation.  Design-guide workload: \textbf{PLN
    multi-pattern deduction}.  First realization: \code{pathmap\_saturator}
  (because \PathMap{}'s quantale structure makes weighted answer
  aggregation natural).
\end{description}

This pair creates the right kind of pressure on the seam: Chainer
forces frames, choicepoints, \code{EnvRef}, demand handling, and
explicit scheduling; Saturator forces snapshots, delta sets,
conjunction planning, exact membership, counted answers, and the
eventual path to PLN and ATP.

Compiled / bytecode (ZAM-lineage) and ATP saturation engines are
explicit \emph{later} lanes --- they risk freezing the protocol in
the wrong shape if pursued before Chainer and Saturator coexist
cleanly under one MAM.

\subsection{Axes that must stay separate}

A core architectural commitment is that the following axes are
\textbf{distinct data} in \code{SearchRequest}, not tangled behind
surface-level operators:

\begin{description}
\item[Schedule.]  \code{dfs}~/~\code{bfs}~/~\code{iddfs}~/\code{interleave}
  are \emph{fairness policies}, not engines.  Korf's classic result
  motivates \code{iddfs} as a completeness-friendly default over raw
  \code{bfs} (BFS completeness with DFS memory); miniKanren's
  interleaving exists precisely to avoid starvation on infinite
  branches.
\item[Demand.]  \code{once}~/~\code{limit-k}~/~\code{unbounded}~/\code{fold}
  describes how many answers are requested, independently of schedule.
  \code{once} and \code{select~1} must build \code{demand = FIRST};
  they must \emph{not} silently hardcode DFS.
\item[Table mode.]  \code{none}~/~\code{variant}~/~\code{subsumptive}~/\code{incremental}
  mirrors the XSB / SWI ladder; each is an explicit engine-level choice.
\item[Backend.]  \code{native}~/~\PathMap{}~/~MORK chosen by the
  actual \code{Space}, not by a global preference.
\item[Lane.]  \code{recursive-dependent-proof}~/~\code{atp-saturation}~/\code{solver-oracle}~/~\dots{}
  chosen at surface.
\end{description}

The request builder (one per \code{--lang}) normalizes surface syntax
into \code{SearchRequest~+~SearchLangContract}; the MAM consumes the
normalized object, never raw surface atoms.

\subsection{The seven-type vocabulary}

Seven types are proposed to land together before the first real
backend optimization (counted-answer variant table, \PathMap{}-native
conjunction, native exact-match index).  Landing them simultaneously
prevents the "first optimization forces a rewrite" pathology.

\begin{center}
\begin{tabular}{ll}
\toprule
\textbf{Type} & \textbf{Categorical role} \\
\midrule
\code{SearchRequest} & Object of $\mathcal{Q}$ \\
\code{SearchAnswer} & Object of $\mathcal{A}$ (quantale-weighted) \\
\code{SearchLangContract} & Subcategory inclusion $\mathcal{C} \hookrightarrow \mathrm{MAM}$ \\
\code{SearchCaps} & Capabilities = presence of morphisms/natural transformations \\
\code{VisitCtl} & Algebra of the effect operation $\mathrm{visit}$ \\
\code{MAMOps} & Kleisli dispatch vtable \\
\code{Episode} & One trace-iteration instance \\
\bottomrule
\end{tabular}
\end{center}

Implementation-actual \code{SUSPEND/RESUME} behavior, cancellation
tokens, snapshot-token machinery, and per-\code{--lang} registration
remain explicitly deferred.

\subsection{Reference base}

The proposal draws on the following literature, all available in the
local library:

\begin{itemize}
\item \emph{A\"{\i}t-Kaci 1991, WAM Tutorial Reconstruction} --
  foundational WAM semantics; drives \code{Chainer} design.
\item \emph{Swift \& Warren, XSB System Manual; Sagonas et al.\ 1994} --
  SLG-WAM tabling; drives \code{Saturator} / table-mode ladder.
\item \emph{Scholz, Jordan, Suboti\'c, Westmann 2016, Soufflé: On Fast
  Large-Scale Program Analysis in Datalog (CC)} -- staged compilation
  of fixpoint computation; compiled \code{Saturator} path.
\item \emph{Keoliya \& Byrd 2020, microKanren} -- interleaving search
  as a distinct schedule axis.
\item \emph{Ager, Biernacki, Danvy, Midtgaard 2003, A Functional
  Correspondence Between Evaluators and Abstract Machines (BRICS)} --
  derivation methodology for machines from evaluator semantics.
\item \emph{Leroy 1990, The ZINC Experiment} -- machine-description
  vs.\ run-instance separation (maps to
  \code{MAMOps}~/~\code{MAM}~/~\code{Episode}).
\item \emph{Russell \& Norvig, AIMA Ch.~3} -- IDDFS completeness;
  uniform treatment of uninformed search.
\end{itemize}

\subsection{Status}

This appendix records design rationale; it does not commit
implementation.  Reasoned design review converged on $\sim$0.88
confidence on MAM naming and two-engine strategy.  The concrete
implementation plan (phase~2 seam; phase~3 tabling plugin;
phase~4 backend parity) is the authoritative schedule and is
unchanged by this appendix.

\emph{This text should be revisited at each phase boundary and
revised as implementation reveals which proposals held and which
did not.}

%% ════════════════════════════════════════════════════════════════════

\begin{thebibliography}{9}
\bibitem{metta}
  B.~Goertzel et al.,
  ``MeTTa: A Reflective Language for AGI,''
  OpenCog Foundation, 2024.

\bibitem{saraswat1988}
  V.~Saraswat, D.~Weinbaum, K.~Kahn, and E.~Shapiro,
  ``Detecting Stable Properties of Networks in Concurrent Logic
  Programming Languages,''
  in \emph{Proc.\ 7th ACM Symposium on Principles of Distributed
  Computing (PODC)}, pp.~210--222, 1988.

\bibitem{mettail}
  Z.~Goertzel and Oru\v{z}i,
  ``MeTTaIL: Contract-First Boot Compilation for Direct C MeTTa Runtimes,''
  Working paper, 2026.

\bibitem{rho2005}
  L.~G.~Meredith and M.~Radestock,
  ``A Reflective Higher-Order Calculus,''
  \emph{Electronic Notes in Theoretical Computer Science}, vol.~141, no.~5,
  pp.~49--67, 2005.

\bibitem{costendofunctor}
  L.~G.~Meredith,
  ``Continued Interactive GSLTs and the Cost Endofunctor: Wrapping by
  Construction,'' Working paper, 2026.
\end{thebibliography}

\end{document}
