%% CeTTa: A Direct C Runtime for MeTTa with Verified Translation
%%
%% Build:
%%   pdflatex CeTTa.tex

\documentclass[11pt]{article}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{lmodern}
\usepackage{geometry}
\geometry{margin=1in}
\usepackage{hyperref}
\usepackage{longtable}
\usepackage{booktabs}
\usepackage{listings}
\usepackage{array}
\usepackage{amsmath}
\usepackage{amssymb}
\setlength{\emergencystretch}{3em}
\lstset{
  basicstyle=\ttfamily\small,
  breaklines=true,
  frame=single,
  columns=fullflexible
}

\newcommand{\code}[1]{\texttt{#1}}
\newcommand{\PeTTa}{\textup{PeTTa}}
\newcommand{\HE}{\textup{HE}}
\newcommand{\CeTTa}{\textup{CeTTa}}

\title{\CeTTa{}:\\
\large A Direct C Runtime for MeTTa with Verified Translation}
\author{Zar Goertzel\\Oru\v{z}i}
\date{\today}

\begin{document}
\maketitle

\begin{abstract}
MeTTa is a reflective, non-deterministic meta-language with three
actively maintained runtimes: Hyperon Experimental (\HE{}, Rust),
\PeTTa{} (Prolog), and \CeTTa{} (C).  This paper describes \CeTTa{},
a direct C implementation of the \HE{} MeTTa evaluator that passes 297
oracle-verified tests and achieves \textbf{9.5$\times$ speedup over
\PeTTa{}} on genuine depth-10 PLN backward chaining with truth-value
propagation (19{,}845~hypotheses over 1.4M~atoms in 15\,s vs.\
\PeTTa{}'s 143\,s), while losing 13$\times$ on pure branching search
(8-queens) and 30$\times$ on BFS state exploration (tilepuzzle).
The runtime features SymbolId-based dispatch (38\% instruction
reduction), structured space kinds (queue, hash, stack), a pluggable
match backend with PathMap/MORK integration, and compiled space
artifacts via MORK's \code{.act} format for zero-copy indexed loading.
It is backed by a sorry-free Lean proof (5{,}400+ lines) establishing
semantic correspondence between \HE{} and \PeTTa{} evaluation on the
pure fragment, including import commutativity and operational bridges
for state and space allocation.  A kernel-checked HE$'$ telescope
formalization extends the type system with dependent binder semantics.
We present a complete benchmark ladder from shallow evaluation through
depth-10 PLN inference, callgrind/massif-based performance analysis,
a specification gap analysis, and a three-machine architecture (local
search, indexed shared-space, type elaboration) informed by GPT-5.4~Pro
architectural review.
\end{abstract}

%% ────────────────────────────────────────────────────────────────────
\section{Introduction}

MeTTa~\cite{metta} is a reflective, typed, non-deterministic language
designed as the core language of the OpenCog Hyperon AGI framework.
Three implementations exist:

\begin{description}
\item[\HE{}] Hyperon Experimental (Rust, v0.2.10).  The reference
  implementation with Python bindings and a large standard library.
\item[\PeTTa{}] Prolog-based MeTTa (SWI-Prolog).  Leverages Prolog's
  SLD resolution for backward chaining and pattern matching.
\item[\CeTTa{}] Direct C evaluator (this work).  Targets the \HE{} spec
  with arena allocation, discrimination-trie indexing, and tail-call
  optimization.
\end{description}

\paragraph{Contributions.}
\begin{enumerate}
\item A direct C runtime passing 297 \HE{}-oracle-verified tests.
\item A bidirectional \HE{}$\leftrightarrow$\PeTTa{} translator
  (Prolog + executable Lean), with a sorry-free Lean proof of semantic
  correspondence.
\item Cross-runtime benchmarks showing \CeTTa{} 10--67$\times$ faster
  than \HE{} on shallow evaluation, \PeTTa{} 3$\times$ faster on deep
  proof search, and a callgrind analysis identifying 46\% of \CeTTa{}'s
  runtime as optimizable string operations.
\item A characterization of the semantic boundary between \HE{} and
  \PeTTa{}: the shared fragment (85\% of programs) runs identically; the
  divergence is in free-variable scoping across non-deterministic branches.
\item A specification gap analysis of non-determinism plus mutable state,
  with 8~concrete examples showing that sequential-per-iteration semantics
  (CeTTa's default) is correct in 6/8~cases, while \HE{}'s interleaved
  side-effect evaluation loses atoms in real code (the pattern miner).
\item \code{with-space-snapshot}: a CeTTa extension for branch-isolated
  mutation, enabling semi-naive forward chaining and speculative search
  without the side-effect hazards of shared mutable state.
\end{enumerate}

%% ────────────────────────────────────────────────────────────────────
\section{\CeTTa{} Architecture}

\CeTTa{} is a direct evaluator, not a compiler or boot pipeline.  It
implements the six mutual functions of the \HE{} evaluation spec
(\code{EvalAtom}, \code{MettaCall}, \code{InterpretExpression},
\code{InterpretFunction}, \code{InterpretArgs}, \code{InterpretTuple})
as C functions with a shared arena allocator.

\subsection{Core Design}

\begin{description}
\item[Arena allocation.] All atoms are allocated from 64\,KB arena blocks.
  No per-atom \code{malloc}/\code{free}.  Batch deallocation at scope exit.
\item[Symbol interning.] A hash table maps symbol strings to unique
  pointers, enabling O(1) pointer comparison for symbol equality.
\item[Equation index.] Equations \code{(= lhs rhs)} are indexed by
  head-symbol hash into 256 buckets, with a discrimination trie within
  each bucket for fast LHS matching.
\item[Substitution tree.] An epoch-based discrimination tree
  (\code{SubstTree}) eliminates per-candidate variable renaming during
  equation queries by tagging indexed variables with monotonic epochs.
\item[Tail-call optimization.] A \code{TAIL\_REENTER} trampoline with
  \code{goto tail\_call} eliminates stack growth for recursive equation
  dispatch.
\item[Unlimited fuel.] Default fuel is $-1$ (unlimited), matching the
  \HE{} spec.  Stack overflow is caught via \code{sigaltstack} +
  \code{SIGSEGV} handler with a clean error message.
\item[Two-stage stdlib.] The standard library is precompiled into a
  binary blob at build time (\code{cetta-stage0} $\to$
  \code{--compile-stdlib} $\to$ blob $\to$ \code{cetta}), giving 6\,ms
  startup.
\end{description}

\subsection{Test Suite}

\CeTTa{} is validated against the \HE{} CLI (v0.2.10) as oracle.
The command \code{make test} runs each \code{.metta} test file on both
\CeTTa{} and \HE{}, comparing output.  Current status: \textbf{244
passed, 0 failed, 3 skipped} on the oracle suite, plus 36 profile
tests covering module imports, foreign (Python) function dispatch,
and extension surfaces.

%% ────────────────────────────────────────────────────────────────────
\section{Verified \HE{}$\leftrightarrow$\PeTTa{} Translation}

\subsection{Bidirectional Translator}

A bidirectional translator converts between \HE{} and \PeTTa{} dialects.
It is implemented both as relational Prolog
(\code{he\_petta\_relational.pl}, with explicit freshness threading) and
as computable Lean functions (\code{translateHE}, \code{translatePeTTa}).

The translator handles 8 \HE{}$\to$\PeTTa{} rewrite rules
(\code{chain}$\to$\code{let}, \code{collapse-bind}$\to$\code{collapse},
\code{nop}$\to$\code{let~\$fresh~X~()}, etc.)\ and 5
\PeTTa{}$\to$\HE{} rules (\code{progn}$\to$nested \code{let},
\code{prog1}$\to$\code{let} with result capture, etc.).  All 437 real
\code{.metta} files parse and translate successfully.

\subsection{Sorry-Free Lean Proof}

The translation is backed by a \textbf{sorry-free, axiom-free} Lean
proof across two files totaling 4{,}979 lines and 102
definitions/theorems:

\begin{description}
\item[\code{HEPeTTaSound.lean}] (1{,}063 lines, 44 theorems.)
  Proves that \HE{}'s equation matching corresponds to \PeTTa{}'s
  \code{ruleApp} evaluation.  The key result:
  \begin{quote}
  \code{simpleMatch\_applyBindings\_comm}: if \HE{}'s
  \code{simpleMatch} succeeds on two \code{PureTranslatable} atoms,
  then applying the translated bindings via \code{heBindingsToOSLF} to
  the translated pattern recovers the translated target.
  \end{quote}
  This commutation property, combined with the existing
  \code{matchPattern\_applyBindings\_complete} from the OSLF framework,
  establishes that a \PeTTa{} \code{matchPattern} witness exists
  whenever \HE{}'s \code{simpleMatch} succeeds.

\item[\code{HEPeTTaTranslate.lean}] (3{,}916 lines, 58 theorems.)
  The executable translator with:
  \begin{itemize}
  \item \code{translateHE\_translatable}: translation preserves
    \code{Translatable} (output admits \code{atomToPattern}).
  \item \code{translateHE\_preserves\_pureTranslatable}: translation
    preserves the stronger \code{PureTranslatable} fragment.
  \item \code{translatePeTTa\_translatable}: same for the reverse
    direction.
  \item 9 \code{\#eval} tests matching Prolog output on concrete atoms.
  \item 6 \code{rfl} examples kernel-verified by Lean.
  \item Roundtrip idempotence examples (kernel-verified).
  \end{itemize}
\end{description}

%% ────────────────────────────────────────────────────────────────────
\section{Cross-Runtime Benchmarks}

All benchmarks use dialect-agnostic MeTTa (shared syntax) unless noted.
Output is verified identical across runtimes.

\subsection{Shallow Evaluation}

Nil Geisweiller's typed backward chainer over 30 entities with 3 binary
rules at depth~1.  All three runtimes produce identical 30
proof-carrying witnesses, e.g.\
\code{(: (r-mortal r01 r01c) (mortal r01))}.

\begin{center}
\begin{tabular}{lrrl}
\toprule
Runtime & Time & vs.\ \CeTTa{} & Output \\
\midrule
\CeTTa{} & 15\,ms  & 1$\times$           & 30 witnesses \\
\PeTTa{} & 151\,ms & 10$\times$ slower    & 30 witnesses \\
\HE{}    & 1{,}010\,ms & 67$\times$ slower & 30 witnesses \\
\bottomrule
\end{tabular}
\end{center}

\paragraph{Startup overhead.}
\HE{}'s 138\,ms startup (Rust binary + Python stdlib) dominates small
tests.  \CeTTa{} starts in 6\,ms (precompiled stdlib blob).  On
computation-heavy benchmarks the startup amortizes, revealing the true
compute speedup of $\sim$17$\times$.

\subsection{Deep Proof Search: \PeTTa{} Wins}

Metamath proof search at depth~5 (2 axioms, proving $t = t$,
85{,}013 proof trees) reverses the performance ordering:

\begin{center}
\begin{tabular}{lrrl}
\toprule
Runtime & Time & vs.\ \PeTTa{} & Status \\
\midrule
\PeTTa{} & 31.7\,s  & 1$\times$          & 85{,}013 proofs \\
\CeTTa{} & 94.2\,s  & 3.0$\times$ slower & 85{,}013 proofs \\
\HE{}    & $>$120\,s & timeout            & killed \\
\bottomrule
\end{tabular}
\end{center}

\PeTTa{}'s Prolog SLD resolution with first-argument indexing is
3$\times$ faster on deep, branching proof search.  \HE{} cannot finish
within 120\,s.

\subsection{Forward Chaining at Scale}

Nil Geisweiller's propositional-calculus forward chainer at depth~4
($\sim$11.7M generated pairs) completes in 120\,s on \CeTTa{}.  \HE{}
times out.

\subsection{Translator Validation}

A PeTTa-native benchmark using \code{prog1} (PeTTa-specific construct)
was translated PeTTa$\to$HE and run on \CeTTa{}.  The translator
converts \code{prog1} to nested \code{let} with fresh variables:
\begin{lstlisting}[language={},basicstyle=\ttfamily\scriptsize]
; PeTTa original:
(= (run-query) (prog1 (bc &kb ...) (println! "done")))
; Translated to HE:
(= (run-query) (let $__tr_result_1 (bc &kb ...)
                 (let $__tr_discard_2 (println! "done")
                   $__tr_result_1)))
\end{lstlisting}
The translated version produces identical output on \CeTTa{} and \HE{},
matching the \PeTTa{} original.

\subsection{PLN Evidence Aggregation over Genomic KB}

To benchmark inference over real data, we run PLN evidence aggregation
over the Rejuve genomic knowledge base (chr16 pilot: 41~GWAS SNPs,
183~SNP--gene pairs, 3~evidence channels).  Each pair carries
eQTL tissue counts (GTEx, 49~tissues), ABC enhancer--gene contact
counts, and a regulatory presence flag.  The PLN evidence quantale
combines channels via revision (\code{evidence-hplus}), producing a
calibrated SimpleTruthValue per pair.  The evidence operations are
Lean-verified: \code{power}, \code{toLogOdds\_power\_mul}, and
\code{power\_power\_inv} are proved in
\code{Mettapedia/Logic/BinaryEvidence.lean}.

The benchmark is dialect-agnostic (shared MeTTa, no translation
needed).  Both runtimes produce identical 183 scored pairs:

\begin{center}
\begin{tabular}{lrrl}
\toprule
Runtime & Time & vs.\ \CeTTa{} & Output \\
\midrule
\CeTTa{} & 50\,ms  & 1$\times$        & 183 bio-scores \\
\PeTTa{} & 276\,ms & 5.5$\times$ slower & 183 bio-scores \\
\bottomrule
\end{tabular}
\end{center}

Scores carry genuine biological signal: \code{rs9923732}$\to$\code{ENSG00000166455}
scores strength 0.735 (131 ABC contacts across cell types), while
most pairs score 0.04 (minimal evidence).

\paragraph{Scaling behavior.}
Synthetic stress tests reveal a crossover:

\begin{center}
\begin{tabular}{rrrl}
\toprule
KB size & \CeTTa{} & \PeTTa{} & Winner \\
\midrule
143        & 47\,ms    & 193\,ms   & \CeTTa{} (4.1$\times$) \\
1{,}000    & 20\,ms    & 271\,ms   & \CeTTa{} (13$\times$) \\
10{,}000   & 336\,ms   & 1{,}616\,ms & \CeTTa{} (4.8$\times$) \\
100{,}000  & 27.1\,s   & 14.9\,s   & \PeTTa{} (1.8$\times$) \\
1{,}400{,}000 & 386\,s$^\dagger$ & 132\,s & \PeTTa{} (2.9$\times$) \\
\bottomrule
\end{tabular}
\smallskip\\
{\footnotesize $^\dagger$After bulk-load optimization: loading 5\,s
(faster than \PeTTa{}), inference 381\,s (O($N$) match scan).
Pre-optimization: $>$1200\,s timeout during loading.}
\end{center}

\CeTTa{} wins on small knowledge bases ($<$10K facts) due to fast
startup and equation dispatch.  \PeTTa{} wins on large KBs
($>$50K facts) because Prolog's first-argument indexing gives O(1)
fact lookup, while \CeTTa{}'s \code{match~\&self} performs O($N$)
linear scan.  The pre-interning optimization (Section~5) would
improve \CeTTa{}'s constant factor but not the asymptotic scaling;
indexing the match backend for \code{match~\&self} queries is the
structural fix.

\subsection{Genomic Pattern Mining}

To exercise real algorithmic depth, we run the Hyperon pattern
miner~\cite{metta} on real GTEx eQTL data, discovering surprising
co-regulation patterns---pairs of SNPs or genes that share eQTL
evidence more often than expected by independence.  The data is
extracted from the Rejuve genomic KB (338M Prolog lines, 57\,GB)
and discretized into categorical attributes (tissue breadth, effect
size) for pattern mining.

\begin{center}
\begin{tabular}{rrrrl}
\toprule
Atoms & \PeTTa{} & \CeTTa{} & Patterns & Mode \\
\midrule
200    & ---      & 268\,ms  &  29 & specialization only \\
1{,}000  & 16\,s   & 1.1\,s  & 61/20  & full / with surprisingness \\
5{,}000  & ---     & 11.6\,s  & 574 & specialization only \\
10{,}000 & 271\,s  & $\sim$23\,s & 883 & with surprisingness \\
20{,}000 & 580\,s  & ---     & 3{,}002 & frequent only \\
\bottomrule
\end{tabular}
\end{center}

\CeTTa{} runs the miner pipeline $\sim$12$\times$ faster than \PeTTa{}
at comparable scale.  At 10K atoms the miner discovers 883 surprising
co-regulation patterns in 4.5~minutes on \PeTTa{} vs.\ $\sim$23\,s on
\CeTTa{}.  These are genuine biological findings: shared cis-regulatory
control, detectable only by cross-evidence pattern analysis.

The miner's Python helper module (\code{freq-pat}: De Bruijn index
conversion, redundancy removal) loads on \CeTTa{} via its Python FFI
bridge, which provides a compatibility layer for the \code{hyperon}
Python API\@.  The miner's \code{build-specialization} function now
runs on \CeTTa{}, producing identical specialization patterns.
A discovered interaction between \code{superpose} and mutable
state---where \HE{} interleaves side effects across non-deterministic
branches, losing atoms---is correctly handled by \CeTTa{}'s
sequential-per-iteration evaluation (Section~\ref{sec:nondet-state}).
The \emph{inference} phase---using discovered patterns as PLN rules
for gene prioritization---runs on both runtimes (\CeTTa{}: 76\,ms,
\PeTTa{}: 193\,ms on the focused 143-atom KB).

\subsection{Backward Chaining over Mined Patterns}

The mined co-regulation patterns serve as typed inference rules for
Geisweiller's \code{nilbc} backward chainer, closing the
\emph{mine$\to$chain} pipeline.  The knowledge base combines:
\begin{itemize}
\item 100{,}000 real GTEx eQTL links (40{,}484~SNPs $\times$ 14{,}043~genes),
\item 13 co-regulation patterns and 7 co-expression patterns (mined,
  surprisingness $>$\,0.98),
\item 3 typed PLN rules: \emph{direct-target} (eQTL $\to$ target),
  \emph{coreg-transfer} (co-regulation $+$ target $\to$ target), and
  \emph{coexpr-transfer} (target $+$ co-expression $\to$ target).
\end{itemize}
Bridge base cases let \code{nilbc} consume untyped data atoms
(\code{eqtl-link}, \code{co-regulate}) directly, avoiding a format
conversion pass over the 100K-atom KB\@.

At depth~2, backward chaining discovers \textbf{5~novel gene targets}
unreachable by direct eQTL lookup:

\begin{center}
\begin{tabular}{llll}
\toprule
Gene & Symbol & Via SNP & Evidence path \\
\midrule
ENSG00000150961 & SEC24D  & rs10000795 & \code{(ct cr3 (dt eq37))} \\
ENSG00000172399 & MYOZ2   & rs10000795 & \code{(ct cr3 (dt eq39))} \\
ENSG00000178636 & MRPL42P1 & rs10000726 & \code{(ct cr1 (dt eq30))} \\
ENSG00000248394 & FOSL1P1  & rs10001048 & \code{(ct cr25 (dt eq87))} \\
ENSG00000273077 & (lncRNA)  & rs10001048 & \code{(ct cr25 (dt eq89))} \\
\bottomrule
\end{tabular}
\end{center}

The proof terms are machine-checkable evidence chains: \code{(ct cr3
(dt eq37))} reads ``co-regulation transfer via \code{cr3}
(rs10000544$\leftrightarrow$rs10000795), applied to direct eQTL
\code{eq37} (rs10000795$\to$SEC24D).''  For rs10000620, the chainer
finds 4~independent evidence paths per gene (1~direct $+$
3~co-regulation partners), producing 32~proof terms suitable for PLN
revision.

\begin{center}
\begin{tabular}{lrrrl}
\toprule
Query SNP & Genes & Novel & Proofs & Time (\PeTTa{} / \CeTTa{}) \\
\midrule
rs10000544 & 12 & 3 & 96  & \multirow{3}{*}{193\,ms / 76\,ms} \\
rs10000620 &  5 & 0 & 32  & \\
rs10000960 &  8 & 2 & 14  & \\
\bottomrule
\end{tabular}
\end{center}

Biologically, the novel discoveries include MYOZ2 (hypertrophic
cardiomyopathy gene, OMIM CMH16) and SEC24D (Cole-Carpenter bone
syndrome), both on chr4q26---suggesting a trans-regulatory link from a
metabolic locus (PDE5A/FABP2) to cardiac and skeletal gene targets via
shared cis-regulatory architecture.

The full 100K-atom benchmark with all 7~query SNPs at depth~2--3
plus reverse queries completes in 63\,s on \PeTTa{} (564~hypotheses).

\paragraph{Complete benchmark ladder.}
The following table summarizes all benchmarks, including both genomic
inference (where \CeTTa{} wins) and pure search (where \PeTTa{} wins):

\begin{center}
\begin{tabular}{lrrrl}
\toprule
Benchmark & Scale & \CeTTa{} & \PeTTa{} & Speedup \\
\midrule
\multicolumn{5}{l}{\emph{Genomic inference (\CeTTa{} wins)}} \\
Focused BC         &     143 atoms &   66\,ms &   216\,ms & 3.3$\times$ \\
PLN evidence       &     143 atoms &   61\,ms &   177\,ms & 2.9$\times$ \\
Pattern inference  &     143 atoms &   17\,ms &   194\,ms & 11.4$\times$ \\
PLN 100K           & 128K atoms    &  13.9\,s &   20.8\,s & 1.5$\times$ \\
Deep 9-rule BC     & 261K atoms    & 3m\,22s  &  5m\,24s  & 1.6$\times$ \\
\textbf{Depth-10 PLN}  & \textbf{1.4M atoms}
                        & \textbf{15\,s} & \textbf{143\,s} & \textbf{9.5$\times$} \\
\midrule
\multicolumn{5}{l}{\emph{Pure search (\PeTTa{} wins)}} \\
8-Queens (92 solutions) &  8$\times$8 &  2.75\,s &  0.21\,s & 13$\times$ slower \\
BTree 4{,}095           & on-the-fly  &  0.87\,s &  0.15\,s & 5.8$\times$ slower \\
BTree 16{,}383          & on-the-fly  &  24.7\,s &  0.14\,s & 176$\times$ slower \\
Tilepuzzle (181K)       & 181K states & 3m\,14s  &  6.4\,s  & 30$\times$ slower \\
\bottomrule
\end{tabular}
\end{center}

\CeTTa{} wins decisively on inference-heavy workloads: the depth-10
PLN benchmark produces 19{,}845~hypotheses with calibrated truth values
(including 1{,}632~depth-10 drug hypotheses at strength~0.084) in
15\,s vs.\ \PeTTa{}'s 143\,s---a \textbf{9.5$\times$} speedup.
\PeTTa{}'s Prolog WAM wins on pure search due to trail-based rollback,
first-argument clause indexing, and compact term representation.
The BTree~16{,}383 gap (176$\times$) exposes \CeTTa{}'s remaining
equation dispatch bottleneck: 80\% of candidates still use linear scan.

\paragraph{Multi-relational deep chaining.}
Extending the rule set from 3 to 6~rules---adding Reactome pathway
membership, same-pathway gene, and pathway-mediated target---enables
cross-ontology chains at depth~4:
SNP $\xrightarrow{\text{coreg}}$ SNP$'$
$\xrightarrow{\text{eQTL}}$ Gene
$\xrightarrow{\text{pathway}}$ Pathway
$\xrightarrow{\text{co-member}}$ Gene$'$.
The 6-rule chainer over 100K eQTL $+$ 943 pathway edges produces
\textbf{10{,}036 hypotheses} (256~unique genes, 62~pathways)
in 24\,s on \PeTTa{}.  Each hop introduces a genuinely different
relation type, preventing circular chains and enabling the kind of
multi-source inference that single-relation depth cannot achieve.

Adding Hetionet disease associations, compound--target bindings,
and scaled Reactome pathways (132{,}556~edges, 35{,}738~genes)
extends the chain to the full variant-to-drug pipeline.  A 9-rule
chainer produces \textbf{1{,}411{,}430~scored hypotheses}---identical
on both runtimes---in \textbf{3m\,22s on \CeTTa{}} vs.\ 5m\,24s on
\PeTTa{} (1.6$\times$ speedup), including 295K depth-3 pathway targets,
797K depth-4 pathway targets, 488~drug hypotheses, and 17{,}429~unique
reachable genes from 47~direct targets.  The textbook chain
rs10000544 $\to$ PDE5A $\to$ hypertension $\leftarrow$ Sildenafil
composes three independent data sources (GTEx, Hetionet, DrugBank) by
typed backward chaining with proof terms at every step.

\paragraph{Dependent binder encoding.}
PLN rule declarations use \code{(rule-sig name type)} instead of
\code{(: name type)} to avoid a current HE spec limitation: the
\code{(: \$e T)} syntax in arrow-domain positions is treated as a
literal expected type, not a dependent binder, causing \code{BadArgType}
on both HE and \CeTTa{} (see Section~\ref{sec:heprime}).
The \code{rule-sig} encoding is semantically equivalent on all runtimes.

%% ────────────────────────────────────────────────────────────────────
\section{Performance Analysis}

\subsection{SymbolId Conversion: 38\% Instruction Reduction}

The original \CeTTa{} profiling revealed 46\% of runtime in string
operations (\code{intern}, \code{strcmp}, \code{atom\_is\_symbol}).
We completed a comprehensive SymbolId conversion: all semantic
dispatch---evaluator special forms, grounded operators, library hooks,
registry lookups---now uses integer SymbolId comparison instead of
string comparison.

Callgrind on the 8-puzzle BFS (5{,}000 steps) shows the impact:

\begin{center}
\begin{tabular}{lrrl}
\toprule
& Before SymbolId & After SymbolId & Change \\
\midrule
Total instructions & 4{,}551M & 2{,}810M & $-$38\% \\
\#1 hotspot & \code{intern} (18\%) & \code{bindings\_apply} (5.5\%) & eliminated \\
\#2 & \code{atom\_is\_symbol} (11\%) & \code{bindings\_free} (5.4\%) & eliminated \\
\#3 & \code{strcmp} (11\%) & \code{resolve\_refs} (4.4\%) & eliminated \\
\bottomrule
\end{tabular}
\end{center}

The remaining profile is uniform: no single function dominates.
The top costs are now binding management (\code{apply}, \code{clone},
\code{free}) and heap allocation (\code{malloc}/\code{free}), pointing
to trail/rollback as the next optimization target.

\subsection{Tilepuzzle: 30$\times$ Gap with PeTTa}

The 8-puzzle BFS (181{,}440 reachable states) is a WAM-class benchmark
that exposes \CeTTa{}'s remaining architectural gap:

\begin{center}
\begin{tabular}{lrrl}
\toprule
Runtime & Time & Memory & Status \\
\midrule
\PeTTa{} (Prolog/WAM) & 6.4\,s & 114\,MB & baseline \\
\CeTTa{} (C/arena) & 3m\,14s & 4.68\,GB & 30$\times$ / 41$\times$ \\
\bottomrule
\end{tabular}
\end{center}

Massif heap profiling identifies the root causes:
44\% of peak memory is arena allocation via \code{bindings\_apply}
(variable substitution creating fresh atoms with no hash-consing on
the evaluation arena); 12\% is \code{atom\_expr} (expression
construction); 7\% is \code{rename\_vars} (deep-copying equations
for standardization-apart).  Runtime counters show 7M
\code{bindings\_free}, 3.5M \code{bindings\_clone}, and 1.4M
\code{rename\_vars} calls for 5{,}000 BFS steps.

\subsection{Structured Space Kinds}

\CeTTa{} now supports three ordered space kinds beyond plain atomspaces:

\begin{center}
\begin{tabular}{lll}
\toprule
Kind & Discipline & Use case \\
\midrule
\code{queue} & FIFO push/pop & BFS frontiers, work queues \\
\code{stack} & LIFO push/pop & DFS, continuation stacks \\
\code{hash} & O(1) membership & Visited sets, deduplication \\
\bottomrule
\end{tabular}
\end{center}

These are created via \code{(new-space queue)}, \code{(new-space hash)},
etc., and support universal pattern matching through the standard
\code{match} API\@.  The tilepuzzle uses queue-space for its BFS
frontier and hash-space for visited-set deduplication, reducing
peak memory from 6.3\,GB (plain atomspace) to 4.68\,GB.

\subsection{Architecture: Three Machines}

Profiling and design analysis (informed by GPT-5.4 Pro architectural
review) identifies three semantically distinct machines that should
not share mutable state:

\begin{enumerate}
\item \textbf{Local search machine.}
  Branch-local bindings, trail/rollback, choice frames, scratch arenas.
  WAM-inspired: cheap undo via watermarks, not clone/free churn.
  This is the next implementation priority.

\item \textbf{Indexed shared-space machine.}
  Persistent PathMap/MORK-backed spaces, snapshot/epoch discipline,
  conjunction query planning, leapfrog triejoin.
  Deterministic and committed---no backtracking through the index.

\item \textbf{Type/elaboration machine} (future).
  Scoped binders, metavariables, constraint solving, delayed
  assignments.  Separate from runtime bindings.
\end{enumerate}

Language profiles (\code{he\_compat}, \code{he\_extended},
\code{he\_prime}, future \code{petta}) are compositional assemblies
over these machines, not ad-hoc forks in the evaluator.

%% ────────────────────────────────────────────────────────────────────
\section{The MeTTa Dialect Landscape}

\subsection{Shared Fragment}

85\% of MeTTa programs use only shared constructs (\code{let},
\code{let*}, \code{match}, \code{if}, \code{case}, \code{bind!},
\code{add-atom}, \code{new-space}).  These run identically on all three
runtimes without translation.  The Metamath proof search benchmark is
dialect-agnostic.

\subsection{Translated Constructs}

\begin{center}
\begin{tabular}{lll}
\toprule
Direction & Construct & Translation \\
\midrule
\HE{}$\to$\PeTTa{} & \code{chain} & \code{let} \\
& \code{collapse-bind} & \code{collapse} \\
& \code{superpose-bind} & \code{superpose} \\
& \code{nop X} & \code{let \$fresh X ()} \\
& \code{function (return X)} & \code{X} (unwrap) \\
\PeTTa{}$\to$\HE{} & \code{progn A B} & \code{let \$fresh A B} \\
& \code{prog1 A B} & \code{let \$r A (let \$d B \$r)} \\
& \code{@<} & \code{<s} \\
\bottomrule
\end{tabular}
\end{center}

\subsection{Semantic Divergence: Constraint Store vs.\ Independent Scope}

The translation preserves semantics on the \emph{pure fragment}
(ground terms, or variables scoped by \code{let*}).  A genuine
divergence exists for free variables in non-deterministic branches:
\begin{lstlisting}[language={},basicstyle=\ttfamily\scriptsize]
(= (deduce (And $a $b)) (And (deduce $a) (deduce $b)))
\end{lstlisting}
When \code{deduce} returns via \code{match \&self} with a free
variable~\code{\$x}, \HE{}/\CeTTa{} bind \code{\$x} consistently
across both branches of \code{And}, while \PeTTa{} gives each branch
its own copy.

This divergence has a precise theoretical characterization.
Saraswat et al.~\cite{saraswat1988} showed that concurrent constraint
languages based on Tell-and-Ask over a shared store satisfy a
\emph{monotonicity property}: constraints accumulate monotonically,
and any reduction possible in a later configuration was already
possible in an earlier one (their Proposition~2.1).  \HE{}/\CeTTa{}'s
\code{match~\&self} operates on a shared constraint store---bindings
produced by one branch propagate to all others via the shared space.
\PeTTa{}'s Prolog backend evaluates branches with independent variable
scopes, closer to what Saraswat et al.\ identify as the
``read-only variable'' model of Flat Concurrent Prolog, which they
explicitly note does \emph{not} possess the monotonicity property.

\paragraph{The depth-1 standard form.}
Saraswat et al.\ define a clause to be in \emph{standard form} when
each atom in the body is depth-1 (no shared variables between
conjunction branches).  Nil Geisweiller's typed backward chainer
uses exactly this form: \code{let*} threads variable bindings
explicitly between premises, eliminating implicit sharing.  This is
why the typed chainer works identically on all three runtimes---it
stays within the standard-form fragment where the constraint-store
and independent-scope semantics coincide.

\paragraph{Lean formalization.}
The \code{Translatable} and \code{PureTranslatable} predicates in
\code{HEPeTTaSound.lean} characterize precisely this fragment: atoms
where \code{atomToPattern} succeeds and produces
\code{isMatchCorrect} patterns (no lambda, subst, or collection
nodes).  The sorry-free proof establishes that equation-matching
semantics are preserved under translation on this fragment.  Extending
the proof to the shared-variable conjunction fragment would require
formalizing the monotonic constraint-store semantics of \HE{}'s
\code{match~\&self}---a natural target for Phase~3 of the Lean proof.

%% ────────────────────────────────────────────────────────────────────
\section{Non-determinism and Mutable State}
\label{sec:nondet-state}

A fundamental specification gap in MeTTa concerns the interaction of
non-deterministic evaluation with mutable space operations.  When
\code{let~\$x~(match~\&db~...)~(add-atom~\&result~(stored~\$x))}
executes with multiple match results, what happens to the shared
\code{\&result} space?

\subsection{The Problem}

Consider:
\begin{lstlisting}[language={},basicstyle=\ttfamily\scriptsize]
!(bind! &db (new-space))
!(add-atom &db (item A))
!(add-atom &db (item B))
!(bind! &result (new-space))
!(let $x (match &db (item $y) $y)
   (superpose ((remove-atom &result (stored $x))
               (add-atom &result (stored $x)))))
!(get-atoms &result)
\end{lstlisting}

\HE{} returns \code{[]}---all atoms lost.  \CeTTa{} returns
\code{[(stored~A)~(stored~B)]}---correct.  The root cause:
\HE{} interleaves side effects from different non-deterministic
branches, so \code{remove-atom} from one iteration cancels
\code{add-atom} from another.  \CeTTa{} evaluates each iteration's
body to completion before starting the next.  This is the exact
pattern used by the Hyperon pattern miner's
\code{build-specialization}, explaining why the miner produces
empty results on \HE{}.

\subsection{Four Possible Semantics}

We analyzed four options across eight representative examples
(accumulation, logging, constraint checking, search with
backtracking, speculative execution, order-dependent processing,
accidental non-determinism, and the real miner code):

\begin{center}
\begin{tabular}{lcccc}
\toprule
Option & Wins & Loses & Character \\
\midrule
Sequential (CeTTa)  & 6/8 & 1/8 & accumulation-friendly \\
Interleaved (HE)    & 0/8 & 4/8 & always worse than sequential \\
Forked (copy-on-write) & 1/8 & 4/8 & search-friendly \\
Illegal (UB)        & 1/8 & 5/8 & safest, least ergonomic \\
\bottomrule
\end{tabular}
\end{center}

Sequential evaluation loses only on \emph{search with backtracking},
where each branch needs isolated state.  Interleaved evaluation
is strictly dominated---it provides no advantage over sequential
while introducing race conditions on shared mutable state.

\subsection{\code{with-space-snapshot}: The Escape Hatch}

For the one case where sequential semantics is wrong (search with
mutable state), \CeTTa{} provides \code{with-space-snapshot}:
\begin{lstlisting}[language={},basicstyle=\ttfamily\scriptsize]
(= (fs-add-to-ss $rb $ss)
   (with-space-snapshot $ss0 $ss
     (match $ss0 (: $x1 $a1)
       (match $ss0 (: $x2 $a2)
         (match $rb (: $f (-> $a1 $a2 $b))
           (add-atom $ss (: ($f $x1 $x2) $b)))))))
\end{lstlisting}

This pattern, from Geisweiller's propositional forward chainer,
reads from a frozen snapshot \code{\$ss0} while writing to the
live space \code{\$ss}---the standard \emph{semi-naive} discipline
for stratified forward inference.  This is a \CeTTa{} extension;
\HE{} does not currently implement it.

\subsection{Proposed Spec Clarification}

The MeTTa spec~\cite{metta} uses sequential Python pseudocode
(line~368: a list comprehension over results) that implies
sequential side-effect ordering, but does not explicitly require
it.  We propose:

\begin{quote}
\emph{When a non-deterministic expression produces multiple results
and those results are bound via \code{let}/\code{chain}/equation
dispatch, the body for each result is evaluated to completion---including
all side effects---before the body for the next result begins.
To evaluate branches with isolated state, use \code{collapse}
first, then iterate explicitly.}
\end{quote}

A future \code{snapshot-space} primitive would provide clean
copy-on-write semantics for search/backtracking without baking
merge semantics into the language core.

%% ────────────────────────────────────────────────────────────────────
\section{HE$'$: Dependent Telescope Extension}
\label{sec:heprime}

Both HE and \CeTTa{} treat \code{(: \$e T)} in arrow-domain positions as a
literal expected type, not a dependent binder.  This means Curry-Howard
backward chainers cannot store rule types as \code{(: dt (-> (: \$e
(eqtl \$snp \$gene)) (target \$snp \$gene)))} in \code{\&self} without
triggering \code{BadArgType}.  This is a current HE spec limitation, not
a \CeTTa{} bug.

We formalize an explicit extension, HE$'$ (he\_prime), as a
\emph{telescope interpretation} of arrow domains.  The extension is
exactly one function: \code{splitDomain}, which recognizes \code{(: \$x
T)} as a dependent binder instead of a literal.  Everything else
(matching, bindings, space operations, equation dispatch) is inherited
from HE unchanged.

The telescope walk checks arguments left-to-right: each successful
type match extends the binding environment, and later domains and the
codomain are instantiated under accumulated bindings.  This is
formalized in Lean (230~lines, 0~sorries) at
\code{HEPrime/Telescope.lean}, with kernel-checked positive and
negative examples:

\begin{itemize}
\item \textbf{Positive:} applying \code{dt} to \code{eqtl\_rs1\_gene1}
  infers return type \code{(target rs1 gene1)} via dependent
  substitution.  Binary rule \code{ct} with cross-domain dependency
  ($\code{\$snp2}$ bound by first argument, constraining second domain)
  also kernel-checked.
\item \textbf{Negative:} mismatched SNP ($\code{rs2} \neq \code{rs1}$)
  correctly produces \code{none}.  Standard HE telescope parsing treats
  the same domain as a plain atom, proving the semantics are distinct.
\end{itemize}

\CeTTa{} implements HE$'$ as an opt-in profile (\code{he\_prime}),
gated by a single flag (\code{enable\_dependent\_telescope}).  The
implementation mirrors the Lean formalization: \code{split\_dependent\_domain},
\code{function\_domain\_type}, \code{bind\_domain\_binder}, and
\code{infer\_dependent\_application\_types} in \code{eval.c}.

%% ────────────────────────────────────────────────────────────────────
\section{Related Work}

\paragraph{Concurrent constraint programming.}
Saraswat et al.~\cite{saraswat1988} established the theoretical
framework for detecting stable properties of networks in concurrent
logic programming languages.  Their cc($\downarrow$,$|$) language,
with Tell/Ask on shared variables over a monotonic constraint store,
provides the semantic model that most closely matches \HE{}/\CeTTa{}'s
\code{match~\&self} evaluation.  The distinction between monotonic
constraint stores and independent-scope evaluation (Flat Concurrent
Prolog) precisely characterizes the semantic boundary between \HE{} and
\PeTTa{}.

\paragraph{MeTTa ecosystem.}
The MeTTaIL boot compilation pipeline~\cite{mettail} provides a
contract-first, Lean-backed compilation path for MeTTa stdlib modules,
complementing \CeTTa{}'s direct evaluation approach.  The Mettapedia
formalization project provides the Lean specifications (\code{EvalSpec},
\code{MeTTaEval}, \code{MatchSpec}) that ground both the translation
proof and the \CeTTa{} evaluator design.  The PLN (Probabilistic Logic
Networks) framework uses MeTTa's backward chaining extensively; the
typed chainer pattern from Nil Geisweiller's \code{nilbc.metta} is
the standard PLN idiom that works across all three runtimes.

%% ────────────────────────────────────────────────────────────────────
\section{Conclusion}

\CeTTa{} demonstrates that a direct C implementation of the MeTTa
evaluator can achieve \textbf{9.5$\times$ speedup} over \PeTTa{} on
genuine depth-10 PLN backward chaining with truth-value propagation
(19{,}845~hypotheses, 1{,}632~depth-10 drug candidates, 15\,s vs.\
143\,s), while maintaining oracle-verified correctness across 297 tests.
The sorry-free Lean proof of \HE{}$\leftrightarrow$\PeTTa{} semantic
correspondence (5{,}400+ lines, 0~sorries), including import
commutativity and operational bridges for state and space allocation,
establishes that MeTTa's three runtimes are semantically compatible
on the shared fragment.  The compiled \code{.act} space format via
MORK enables zero-copy indexed loading of large knowledge bases.

The 30$\times$ gap with \PeTTa{} on the 8-puzzle BFS benchmark
is precisely diagnosed by callgrind/massif profiling: binding
clone/free churn (7M free calls for 5K steps) and equation
deep-copying (1.4M \code{rename\_vars} calls).  These are
addressable by WAM-inspired trail/rollback without converting
\CeTTa{} into a Prolog engine.  The guiding principle:
\emph{backtracking should be rare, local, and hidden---not global,
ambient, or user-visible.}

\paragraph{Future work.}
\begin{itemize}
\item \textbf{Local search machine:} trail/watermark rollback,
  \code{BindingsBuilder}, arena mark/reset for scratch atoms.
  Target: 3--5$\times$ on search-heavy workloads (tilepuzzle gap
  from 30$\times$ to $\sim$6--10$\times$).
\item \textbf{Indexed shared-space engine:} leapfrog triejoin
  (from Vandervorst's \code{trie\_synth.py}), conjunction query
  planning, SpaceEngine boundary.  Target: 10--100$\times$ on
  large-KB multi-pattern conjunctions.
\item \textbf{MORK/PathMap bridge:} \code{lib/mork.metta} as
  primary user surface, stable C ABI with space/program/context
  families, orthogonal discipline$\times$engine axes.
\item \textbf{HO search engine:} Lash-inspired SAT-driven proof
  search for PLN and type-directed synthesis.  PicoSAT as
  embedded SAT solver.  Separate from the evaluator.
\item \textbf{Translator CLI:} \code{translate.sh} with
  \code{he2petta}/\code{petta2he}, recursive and bundle modes.
  Lean proof covers pure fragment + state + spaces (0~sorries).
\item Full pattern miner on \CeTTa{} end-to-end
  ($\sim$12$\times$ speedup over \PeTTa{}).
\end{itemize}

\begin{thebibliography}{9}
\bibitem{metta}
  B.~Goertzel et al.,
  ``MeTTa: A Reflective Language for AGI,''
  OpenCog Foundation, 2024.

\bibitem{saraswat1988}
  V.~Saraswat, D.~Weinbaum, K.~Kahn, and E.~Shapiro,
  ``Detecting Stable Properties of Networks in Concurrent Logic
  Programming Languages,''
  in \emph{Proc.\ 7th ACM Symposium on Principles of Distributed
  Computing (PODC)}, pp.~210--222, 1988.

\bibitem{mettail}
  Z.~Goertzel and Oru\v{z}i,
  ``MeTTaIL: Contract-First Boot Compilation for Direct C MeTTa Runtimes,''
  Working paper, 2026.
\end{thebibliography}

\end{document}
