%% MettaKernel: A MeTTa-hosted Proof/Inference Kernel for both
%% Dependent Type Theory and Higher-Order Tarski--Grothendieck Set Theory
%%
%% Build:
%%   pdflatex MettaKernel.tex
%%
%% Status: design + curriculum memo (prototype-first), with a realized-architecture
%% update (June 2026): the MeTTa kernel is now built and runnable. Lean PureKernel,
%% Megalodon, and the Lean-verified LeaTTa are the oracles. The universal
%% subject-reduction/confluence metatheorem remains the open soundness obligation.

\documentclass[11pt]{article}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{lmodern}
\usepackage{geometry}
\geometry{margin=1in}
\usepackage{hyperref}
\usepackage{longtable}
\usepackage{booktabs}
\usepackage{listings}
\usepackage{array}
\usepackage{amsmath}
\usepackage{amssymb}
\setlength{\emergencystretch}{3em}
\lstset{
  basicstyle=\ttfamily\small,
  breaklines=true,
  columns=fullflexible,
  keepspaces=true,
  frame=single,
  framesep=4pt,
}
\newcommand{\MeTTa}{MeTTa}
\newcommand{\CeTTa}{CeTTa}
\newcommand{\HOTG}{HOTG}
\newcommand{\DTT}{DTT}
\newcommand{\code}[1]{\texttt{#1}}

\title{MettaKernel: A \MeTTa-hosted Proof/Inference Kernel for both\\
Dependent Type Theory and Higher-Order Tarski--Grothendieck Set Theory\\
\large Design, Curriculum, and a Prototype-First Build Plan}
\author{Oru\v{z}i \and Zar Goertzel}
\date{2026-06-18 (design draft); realized-architecture update 2026-06-23}

\begin{document}
\maketitle

\begin{abstract}
We set out to build one small \emph{trusted} proof/inference kernel, written
\emph{in \MeTTa{}} (as an ordinary \MeTTa{} program, executed on the \CeTTa{}
engine), that can express and check proofs in \emph{both} foundations we care
about: CIC-grade \emph{dependent type theory} (the foundation of Lean and Coq)
and \emph{higher-order Tarski--Grothendieck set theory} (\HOTG{}, the classical
higher-order logic of Megalodon). The kernel is a \MeTTa{} program --- not a new
\CeTTa{} language substrate --- in the same spirit as \code{pverify}, an existing
Metamath proof checker written in \MeTTa{}. Two principles govern the work.
First, \textbf{prototype a fully working kernel before any deeper integration into
the engine}, so that the design is chosen empirically rather than frozen in
advance. Second, \textbf{conversion is the danger line}: the kernel's
$\beta\eta\delta\iota$ conversion must be its own small, decidable, resource-bounded
\MeTTa{} functions and must never be delegated to \CeTTa{}'s general runtime
rewriting, or proof checking ceases to be a terminating judgment. We present the
two reference kernels we mirror (the Lean \code{PureKernel} for \DTT{} and
Megalodon's \code{tp/tm/pf}+\code{check\_doc} core for \HOTG{}), a graded
\emph{curriculum} of the easiest examples that pin down the smallest fully
expressive kernel (a \DTT{} ladder following Smolka \& Brown's \emph{Introduction
to Computational Logic}, and an \HOTG{} ladder following Megalodon), and a
prototype-first build plan whose first milestone is an honest \emph{capacity ledger}
(\textsc{Agree}/\textsc{Diverge}/\textsc{Open}) rather than a claim of finality. The
\MeTTa{} kernel's verdicts are treated as a \emph{subject} to be measured against
the oracles, not as an oracle itself.
\end{abstract}

\section{Why a single kernel for two foundations}

Lean/Coq rest on the Calculus of Inductive Constructions: dependent function and
pair types, an inductive-definition mechanism with computation rules, a universe
hierarchy, and a \emph{decidable} definitional equality. Megalodon rests on
\HOTG{}: simple types with an impredicative \code{Prop}, proof terms, classical
logic, set membership, extensionality, foundational ($\in$-)induction, and a choice
operator. These are different mathematics, but a \emph{proof kernel} for either is
small: it is a term/type/proof object language plus a conversion relation plus a
checking judgment, with everything else (parsing, automation, optimization, the
ambient runtime) living \emph{outside} the trusted core.

Our thesis --- developed at length in the companion memo
\emph{metta\_proof\_kernel\_grounded} --- is that \MeTTa{} is an apt host for such a
kernel \emph{as a program}: terms, types, and proofs are \MeTTa{} atoms; the
checking and conversion judgments are \MeTTa{} functions; theories and imports are
\MeTTa{} data. The trusted waist is then a small auditable \MeTTa{} module, and the
two foundations appear as two \emph{checkers} sharing that waist. We are not
proposing to encode both foundations into one ambient rewriting theory (the
Dedukti route); we keep the foundations' judgments first-class and only share the
machinery beneath them.

\paragraph{What this document is.} A design + curriculum memo to drive a
prototype-first build. It records: the reference kernels we mirror
(\S\ref{sec:oracles}); the \MeTTa-program architecture (\S\ref{sec:arch}); the
graded curriculum that defines ``smallest fully expressive'' (\S\ref{sec:curric});
the build plan and its first milestone, a capacity ledger (\S\ref{sec:plan});
the conversion discipline that is the central hazard (\S\ref{sec:conv}); the
open questions we deliberately leave open (\S\ref{sec:open}); and --- added in this
revision --- the \emph{realized architecture and results} (\S\ref{sec:realized})
recording the as-built kernel that now exists, including where realization refined
the plan.

\section{The two reference kernels (our oracles and templates)}
\label{sec:oracles}

\subsection{\DTT{}: the Lean \code{PureKernel}}
A trusted, \emph{sorry-free} dependently-typed kernel already exists in Lean
(\code{Mettapedia/.../MeTTa/PureKernel/}, 28 files). Its term language is
\begin{lstlisting}
PureTm n ::= var (Fin n) | const DeclName | U0 | U1
           | pi A B | sigma A B | id A t u
           | lam B | app f a | pair t u | fst p | snd p | refl t
\end{lstlisting}
with two universes ($\code{U0}:\code{U1}$), $\Pi$, $\Sigma$, and identity types. It
provides one-step $\beta/\delta/\iota$ reduction, Church--Rosser confluence,
\code{HasType}, a \emph{decidable} definitional equality
(\code{defEqByNormalization?}), subject reduction, and a bidirectional checker.
Ordinary families (\code{Bool}, \code{Nat}, \code{Unit}) are \emph{not} hard-coded:
they enter through a general declaration mechanism (\code{DeclSpec},
\code{InductiveDecl}, \code{RecursorDecl}) with full well-formedness proofs. This is
both our \DTT{} \emph{oracle} and our design template; crucially, the existing
runtime-to-kernel bridge marks \code{directExecEquivalence} as \code{false}, so the
runtime is \emph{not} yet known to agree with this kernel.

\subsection{\HOTG{}: the Megalodon \code{mathdata} core}
Megalodon's trusted core (\code{repos/megalodon-1.13/src/mathdata.mli}) is genuinely
compact --- three small inductive types and a checker:
\begin{lstlisting}
tp ::= TpVar i | Prop | Base i | Ar(tp,tp) | TpAll tp        (* simple types + type-vars + type-forall *)
tm ::= DB i | TmH h | Prim i | Ap(tm,tm) | Lam(tp,tm)
     | Imp(tm,tm) | All(tp,tm) | TTpAp(tm,tp) | TTpLam tm | TTpAll tm
pf ::= Gpa h | Hyp i | Known h | PTmAp(pf,tm) | PPfAp(pf,pf)
     | PLam(tm,pf) | TLam(tp,pf) | PTpAp(pf,tp) | PTpLam pf
\end{lstlisting}
plus \code{check\_doc} (the proof checker), $\beta\eta\delta$ normalization with
resource limits (\code{BetaLimit}, \code{NonNormalTerm}), and hash-addressed
\code{doc}/\code{theory}/\code{signa} imports. Logic and set-theoretic content are
\emph{not} in this core: \code{True}/\code{False}/\code{not}/\code{and}/\code{or}
are impredicative definitions, and membership/extensionality/choice/$\in$-induction
enter as \emph{axioms} in a preamble (Egal mode uses \code{Eps} choice and Tarski's
axiom; Mizar mode uses \code{the} and ZF-style axioms). The seed file
\code{Mettapedia/megalodon/experiments/01\_basics.mg} proves eight propositional
theorems entirely by proof-term lambdas (e.g.\
\code{andEL := fun A B H => H A (fun a b => a)}).

\subsection{Precedent that \MeTTa{} can host a checker}
\code{hyperon/metamath/pverify} (with \code{lib/pverify-utils.metta}) is a Metamath
proof checker realized as a \MeTTa{} program. It establishes feasibility of the
\emph{host}; it is Metamath, not \HOTG{}, so it is precedent, not evidence about the
present design.

\section{Architecture: the kernel as a \MeTTa{} program}
\label{sec:arch}

\paragraph{Shared waist (written once in \MeTTa{}).} Atom encodings for binders
(de Bruijn or locally nameless), $\alpha$-safe substitution, typed contexts,
declarations (parameter / definition / axiom / theorem / inductive / import), proof
objects, the \textbf{conversion engine} ($\beta/\eta/\delta/\iota$ as explicit,
terminating, resource-capped \MeTTa{} functions), a \code{check}/\code{infer}
driver, and hashing/provenance for imports. The two checkers below are
\emph{instances} of this waist --- that sharing \emph{is} the early unification.

\paragraph{\HOTG{} checker.} Mirror \code{tp}/\code{tm}/\code{pf} + \code{check\_doc}
+ $\beta\eta\delta$ as \MeTTa{} atoms and functions. Proof-term checking over
impredicative \code{Prop}; classical/set/TG axioms are a \emph{profile} layered
above the core, admitted only after the bare proof-term checker is green.

\paragraph{\DTT{} checker.} Mirror \code{PureTm} ($\code{U0}/\code{U1}$, $\Pi$,
$\Sigma$, \code{Id}) with bidirectional infer/check, $\beta/\delta/\iota$, and
decidable definitional equality by normalization; inductive families via
\code{DeclSpec}, never hard-coded. Where the \CeTTa{} engine's \code{--profile
he-prime} already provides dependent type inference, the \MeTTa{} kernel may use it;
new C features are added to \code{he-prime} \emph{only} when the prototype proves a
concrete need.

\section{Curriculum: the examples that pin the smallest fully expressive kernel}
\label{sec:curric}

The curriculum \emph{defines} ``fully expressive'': the smallest kernel is the one
that checks every example below. The \DTT{} ladder follows the chapter order of
Smolka \& Brown's \emph{Introduction to Computational Logic} (a Coq course by
Megalodon's own author), which introduces kernel features in increasing dependency
order. The \HOTG{} ladder follows Megalodon.

\paragraph{L0 --- shared waist (both ladders ride on it).} Names/binding,
$\alpha$-safe substitution, typed contexts, declarations, proof objects,
conversion ($\beta/\eta/\delta/\iota$, decidable), hash imports + provenance,
negative tests.

\begin{longtable}{@{}p{0.06\textwidth}p{0.42\textwidth}p{0.46\textwidth}@{}}
\caption{\DTT{} ladder (ICL spine + \CeTTa{} \code{he-prime} seeds + \code{PureKernel}).}\\
\toprule
\textbf{Step} & \textbf{Example (source)} & \textbf{Kernel feature it pins}\\
\midrule
\endfirsthead
\toprule \textbf{Step} & \textbf{Example} & \textbf{Kernel feature}\\ \midrule
\endhead
D1 & \code{bool}/\code{negb}/\code{andb}/\code{andb\_com} (ICL \S1.1) & inductive \code{Bool}, match/recursor, def-eq by computation\\
D2 & \code{plus\_O}/\code{plus\_S}/\code{plus\_assoc} (ICL \S1.3--1.4) & \code{Nat}, recursor, structural induction\\
D3 & $\Pi$ / proofs-as-programs (ICL \S1.2, 2.2, 2.6) & $\Pi$ formation/intro/elim, application\\
D4 & $\Sigma$ / pairs (ICL \S6.6; he-prime \code{TutPackHuman}) & $\Sigma$ formation/intro/projection\\
D5 & \code{Id}/refl/cong/transport (ICL \S3.3; he-prime \code{TutEq}/\code{TutRefl}/\code{TutSuccCong}) & identity types, $J$/transport\\
D6 & \code{app\_assoc}/\code{length\_app} (ICL \S8) & parametric inductive + recursor\\
D7 & \code{Vec}/\code{TutIndex} safety; \textbf{\code{VecN -1} (open)} & indexed inductive families; index-domain question (\code{Number} vs \code{Nat})\\
D8 & \code{Even}/$\le$ (ICL \S7) & inductive predicates, elim restriction\\
D9 & recursors + elim restriction + universes (ICL \S5) & $\iota$-reduction, proof irrelevance, universe consistency\\
\bottomrule
\end{longtable}

\begin{longtable}{@{}p{0.06\textwidth}p{0.42\textwidth}p{0.46\textwidth}@{}}
\caption{\HOTG{} ladder (\code{01\_basics.mg} + Megalodon + \code{mathdata} \code{tp/tm/pf}).}\\
\toprule
\textbf{Step} & \textbf{Example (source)} & \textbf{Kernel feature it pins}\\
\midrule
\endfirsthead
\toprule \textbf{Step} & \textbf{Example} & \textbf{Kernel feature}\\ \midrule
\endhead
H1 & propositional encodings (\code{01\_basics.mg}, 8 thms) & \code{tp(Prop)}, \code{tm(All,Imp,Lam,Ap)}, \code{pf(PLam,PPfAp,Hyp)}, $\beta$-normalization, \code{check\_doc}\\
H2 & $\forall$-instantiation; $\to$/proof application & \code{PTmAp}, \code{PPfAp}\\
H3 & proof $\lambda$; $\beta\eta\delta$; definition unfolding & $\delta$ on \code{ThyDef}/\code{SignaDef}, $\eta$\\
H4 & equality/congruence (Leibniz \code{eq}) & polymorphic \code{eq} via \code{TpAll}\\
H5 & \textbf{membership/set/TG axioms} (\code{In}/\code{Empty}/\code{EmptyE}, set-ext, \code{In\_ind}, \code{Eps}) & \emph{axiom profile}, admitted only after H1--H4 green\\
\bottomrule
\end{longtable}

\paragraph{Cross-profile smoke.} $P\to P$;\ $P\to Q\to P$;\ $P\wedge Q\to P$;\
$\bot\to P$;\ equality reflexivity/congruence;\ pair projections;\ and \emph{one}
tiny \CeTTa{} operational-semantics theorem treated as a \emph{certificate}, never
as definitional equality.

\paragraph{Negative corpus (a fail-by-default gate --- the soundness backstop).}
Bad de Bruijn index;\ unknown \code{Known}/\code{Hyp};\ application type mismatch;\
a proposition-shape used as a proof object;\ cyclic/non-normal $\delta$;\
\textbf{an arbitrary \MeTTa{} rewrite admitted as conversion} (the central danger);\
stale import/hash;\ \code{VecN \dots -1} wrongly accepted. The kernel must
\emph{reject} every item here; a checker that accepts any of them is unsound.

\section{Prototype-first build plan}
\label{sec:plan}

\begin{description}
\item[Phase 0.] Shared \MeTTa{} waist + the ledger harness (three deciders:
  \MeTTa{}-kernel vs.\ Lean \code{PureKernel} oracle vs.\ \code{he-prime} C).
\item[Phase 1 (\HOTG{} prototype).] \MeTTa{} \code{tp/tm/pf} + proof-term checker +
  $\beta$; \emph{check the eight \code{01\_basics.mg} theorems green} (H1--H2).
\item[Phase 2 (\DTT{} prototype).] \MeTTa{} \code{PureTm} + bidirectional checker +
  $\beta/\delta/\iota$ + decidable def-eq; \emph{D1--D5 green}; ledger vs.\ the Lean
  oracle and \code{he-prime} C.
\item[Phase 3 (unify + harden).] Both checkers share the waist; cross-profile smoke
  green; \emph{negative corpus rejected}.
\item[Phase 4 (climb the ladders).] \HOTG{} H3--H5 (definitions, equality, then
  set/TG axioms); \DTT{} D6--D9 (lists, \code{Vec}, inductive predicates,
  recursors/universes); resolve the \code{VecN} index-domain question.
\item[Later.] Only after a fully working prototype and its lessons: consider deeper
  substrate integration --- add the \code{he-prime} C features the prototype proved
  necessary; refine \CeTTa{}. Not before.
\end{description}

\paragraph{Milestone 1 (the ledger, not a constitution).} The deliverable is the
working H1 + D1--D3 \MeTTa{} kernel plus a populated \emph{capacity ledger} with
buckets \textsc{Agree-Accept}, \textsc{Agree-Reject}, \textsc{Diverge},
\textsc{Open}, plus the negative-corpus rejections. Divergences and open cases are
first-class outputs, not blemishes to hide. Only then does external review become
useful, and its question is whether the fragment boundary is honest and which
divergences are bugs versus legitimate language semantics --- not whether the final
kernel is settled. The \MeTTa{}-kernel verdicts are labelled \textbf{subject}; the
Lean \code{PureKernel} and Megalodon are the \textbf{oracles}.

\section{Realized architecture and results (June 2026)}
\label{sec:realized}

The prototype is built and runs. This section records the \emph{as-built} kernel and,
honestly, where realization \emph{refined} the plan above; the earlier sections stand
as the original intent.

\paragraph{One $\lambda\Pi$-modulo checker; object logics as signatures.} Rather than
two separate term languages (\code{PureTm} and \code{tp/tm/pf}), the realized kernel is
a \emph{single} $\lambda\Pi$-style logical-framework checker
(\code{Srt}/\code{Con}/\code{Var}/\code{Pi}/\code{Lam}/\code{App}/\code{Def}) with one
\code{infer}/\code{check}/\code{conv}; each object logic enters as a \emph{signature} of
typed declarations over that one checker. \DTT{}, \HOTG{}, and HOL all appear in
\code{kernel/curriculum\_lf\_demo.metta} as signatures over the same checker, each with
positive proofs and a rejected negative. This keeps every logic's rules first-class (as
signature declarations) while sharing the trusted waist --- LF-hosting, deliberately
distinct from desugaring every logic into one ambient rewrite theory.

\paragraph{One generic inductive engine (\code{DIndG}).} The hard-coded inductive
families of the \DTT{} ladder (D6--D9) are subsumed by a single generic
\emph{indexed-inductive} declaration, \code{DIndG}: a telescope IR (parameters $\Delta$,
indices $\Xi$, constructors) from which the family kind, constructor types, the
eliminator, branch types, and constructor-headed $\iota$-rules are \emph{generated and
checked}. \code{Nat}, \code{Bool}, \code{List}, \code{Tree}, \code{Pair}, \code{Vec},
\code{Fin}, and \code{Id} are all instances; crucially \code{Id} is itself a
\code{DIndG} and \code{J} is its \emph{generated} (diagonal) recursor, so the equality
primitives and six special-case inductive slices retired onto the one engine. (This
resolves the original \code{VecN}/index-domain open question: indices are an arbitrary
telescope, not a single \code{Nat}.)

\paragraph{Binding as a declared property (the ABT waist).} Binding lives in one
import-free waist (\code{kernel\_binding\_waist\_v1}): de Bruijn under the hood, with
capture-avoiding \code{shift}/\code{subst} written once and \emph{arity-parameterized};
clients delegate to it rather than re-implementing substitution. A \code{BindingDecl}
layer (\code{kernel\_binding\_decl\_v1}) lets each hosted LanguageDef \emph{declare} its
binders (\code{Bind0}/\code{Bind1}/\code{BindFields}/\code{BindPi}: dependent $\Pi$,
$\forall$/$\exists$ bodies, comprehension predicates, proof binders) without disturbing
the hot waist. Capture-avoidance is mutation-gated.

\paragraph{Admission, checked-once theorems, and the admit-once lesson.} A signature is
\emph{admitted} only if every declaration is well-formed: constants' types check,
definitions' bodies check against their declared types (\code{wf-decl}), and inductive
declarations are fully generatable. Two declaration forms serve derived libraries:
\code{DAxiom} (an honest postulate, recorded in \emph{axiom provenance}) and \code{DThm}
(a \emph{checked} theorem constant --- its proof body is checked \emph{once} at
admission, then used opaquely by its type, transparent only for provenance). A memory
hazard forced the central engineering lesson: naively, \code{kernel-check} re-ran full
signature admission on \emph{every} theorem check, so a file checking 21 theorems
against a signature carrying two large derived-proof bodies re-normalized those bodies
21 times and exhausted memory ($\sim$48\,GiB, OOM). The fix is \emph{admit-once batch
checking} --- admission computed once, lexically, for a batch of jobs, with \emph{no}
forgeable ``checked-signature'' token exposed (in \MeTTa{} a token is plain data anyone
could fabricate). This is exactly HOL-Light's discipline (a derived rule is a program
calling primitives, never one giant proof term re-normalized from scratch) and it
dropped the same file to $\sim$8\,GiB.

\paragraph{Faithful object-logic hosting: HOL-Light's equality kernel.} HOL is hosted
not as a convenient natural-deduction surrogate but on HOL-Light's \emph{actual}
primitive basis --- \code{REFL}, \code{TRANS}, \code{MK\_COMB}, \code{ABS}, \code{BETA},
\code{ASSUME}, \code{EQ\_MP}, \code{DEDUCT\_ANTISYM} as LF rule certificates checked by
the imported kernel --- with $\top$/$\wedge$/$\Rightarrow$ defined over equality and
$\lambda$ (matching \code{bool.ml}) and derived \code{MP}/\code{DISCH} shards admitted as
checked \code{DThm}s. Each primitive and each shard carries a mutation tooth.
\emph{Faithfulness caveat:} the definitional equations are currently \emph{postulated}
as constants matching \code{bool.ml} rather than introduced through a checked
conservative-definition principle, and the derived \code{MP}/\code{DISCH} are specific
checked replays, not yet a fully general derived-rule library.

\paragraph{The curriculum as a calibration matrix.} The curriculum is now an explicit
ledger (\code{Curriculum/oracle\_cases.tsv}): each row is a kernel capacity $\times$
profile (\DTT{}/HOL/\HOTG{}/Metamath/\dots) $\times$ the SOTA system that calibrates the
verdict (Lean/Coq/Agda/Megalodon/HOL-Light), with status \textsc{checked} or
\textsc{planned} and a per-row gate. The external systems are the \emph{calibration
oracle}; the destination is to \emph{host each core logic as a LanguageDef and re-prove
the curriculum's theorems in our own kernel}. Verdicts are differential --- a row
borrows its expectation from the source system, not from our own guess.

\paragraph{Validation discipline.} Every checkpoint runs on three engines --- \CeTTa{}
(C), PeTTa (Prolog), and the Lean-verified \emph{LeaTTa} as a deterministic oracle ---
in parity, with per-feature mutation gates (corrupt the rule, require red) and the
fail-by-default negative corpus, all behind one \code{run\_all} gate; heavy runs are
memory-capped.

\paragraph{Soundness status (honest).} Admission plus provenance close the per-input
declaration holes (a \code{break\_probes} probe documents what the bare term-checker
would accept \emph{without} admission; the real kernel rejects those). The \emph{one}
genuinely-open soundness obligation is metatheoretic, not a per-input check:
\textbf{universal subject-reduction, confluence, and termination of the generated
$\iota$-rules are not yet machine-proven} --- we have an executable SR-wedge self-test
and the LeaTTa engine-floor, not the theorem. That metatheorem is the live Lean target.
The kernel is today \emph{checked-correct per input} but not yet \emph{proven-correct as
a calculus}.

\section{Conversion discipline (the central hazard)}
\label{sec:conv}
The single biggest architectural risk is identifying kernel conversion with general
\MeTTa{} evaluation. We hold five lines:
(A) decidable conversion $\ne$ general \MeTTa{} eval --- the kernel's
$\beta/\eta/\delta/\iota$ are its own bounded \MeTTa{} functions (resource caps in
the spirit of Megalodon's \code{BetaLimit});
(B) \MeTTa{}'s many-valued/nondeterministic typing vs.\ a functional \DTT{}
judgment --- the kernel commits to a functional infer/check;
(C) grounded/FFI values are a trusted boundary --- grounded equality is not kernel
reduction;
(D) universe consistency is not optional forever, because proofs \emph{about}
\MeTTa{} are a goal;
(E) the kernel must be small and separable --- a pure checking slice, not entangled
in effectful evaluation.

\section{Open questions (kept deliberately open)}
\label{sec:open}
\emph{Resolved or refined by the realized architecture (\S\ref{sec:realized}):}
the \code{VecN}/index-domain question (indices are an arbitrary telescope under
\code{DIndG}); the first dependent-typing fragment ($\lambda\Pi$ + \code{DIndG}
indexed inductives, predicative); the trust boundary (it is signature
\emph{admission}); and the convergence target (the kernel is its own $\lambda\Pi$-LF,
with \code{PureKernel}/Megalodon as templates and the Lean-verified LeaTTa as the
runnable oracle floor, rather than \code{PureKernel} itself).

\emph{Live and deliberately open:} (1)~the \textbf{universal subject-reduction /
confluence / termination metatheorem} for the generated $\iota$-rules --- the single
open soundness obligation, now a Lean target (the toolchain is unblocked at v4.31);
(2)~\textbf{engine-level term sharing} (hash-consing / DAG / normalization-by-evaluation
in \CeTTa{}) for deeper scaling, of which admit-once batch checking is the cheaper
architectural half already in hand; (3)~a \textbf{checked conservative-definition
principle} so hosted definitional equations are derived rather than postulated;
(4)~a \textbf{fully general derived-rule library} (beyond specific checked replays); and
(5)~\textbf{breadth} --- hosting further core logics as LanguageDefs (fuller \HOTG{}
via Megalodon, then Lean-core/CIC) over the now-shared, now-affordable waist.

\section*{References (local anchors)}
\begin{itemize}
\item G.\ Smolka and C.\ E.\ Brown, \emph{Introduction to Computational Logic},
  Lecture Notes, Saarland University, 2014. (DTT curriculum spine.)
\item \emph{metta\_proof\_kernel\_grounded} (AIrxiv, 2026): framework core,
  object-theory profiles, conversion discipline, hash imports.
\item Megalodon: \code{repos/megalodon-1.13/src/mathdata.mli} (the \HOTG{}
  \code{tp/tm/pf} + \code{check\_doc} core); \code{examples/} (Egal/Mizar
  preambles); \code{experiments/01\_basics.mg} (seed).
\item Lean \code{PureKernel/README.md} and modules (the \DTT{} oracle/template).
\item \code{hyperon/metamath/pverify} (\MeTTa-hosted checker precedent).
\end{itemize}

\end{document}
