%% PLN credal degrees of freedom
%%
%% Render with:
%%   lualatex -shell-escape pln-credal-degrees-of-freedom.tex \
%%     && bibtex pln-credal-degrees-of-freedom \
%%     && lualatex -shell-escape pln-credal-degrees-of-freedom.tex \
%%     && lualatex -shell-escape pln-credal-degrees-of-freedom.tex

\documentclass[]{article}
\usepackage{url}
\usepackage[hyperindex,breaklinks]{hyperref}
\IfFileExists{breakurl.sty}{\usepackage{breakurl}}{}
\usepackage{listings}
\lstset{basicstyle=\ttfamily\footnotesize,breaklines=true,frame=single,
        xleftmargin=1em,xrightmargin=0.5em}
\usepackage{float}
\restylefloat{table}
\usepackage{longtable}
\usepackage{booktabs}
\usepackage{graphicx}
\usepackage[font=small,labelfont=bf]{caption}
\usepackage[skip=0pt]{subcaption}
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{unicode-math}
\IfFileExists{newunicodechar.sty}{
  \usepackage{newunicodechar}
  \newunicodechar{∘}{\ensuremath{\circ}}
}{}
\IfFontExistsTF{Stix Two Math}{\setmathfont{Stix Two Math}}{\setmathfont{STIX Math}}

% Compatibility fallback for older TeX bundles.
\makeatletter
\providecommand{\headerps@out}{}
\makeatother

\newtheorem{theorem}{Theorem}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{remark}[theorem]{Remark}
\newtheorem{observation}[theorem]{Observation}

\newcommand{\nuPLN}{\texorpdfstring{$\nu$}{nu}\textup{PLN}}
\newcommand{\xiPLN}{\texorpdfstring{$\xi$}{xi}\textup{PLN}}
\newcommand{\code}[1]{\texttt{#1}}
\newcommand{\codepath}[1]{\path{#1}}
\newcommand{\lid}[1]{{\ttfamily\nolinkurl{#1}}}
\newcommand{\lpath}[1]{\texttt{\nolinkurl{#1}}}
\newcommand{\ty}[1]{\ensuremath{\mathtt{#1}}}
\newcommand{\STV}[2]{\langle #1,\, #2 \rangle}
\newcommand{\strength}{\mathrm{strength}}
\newcommand{\confidence}{\mathrm{confidence}}
\newcommand{\Eq}{\;\Leftrightarrow\;}

\emergencystretch=3em

\title{Credal degrees of freedom for PLN truth values:\\
       a layered characterization of strength and confidence}
\author{Mettapedia Working Notes}
\date{\today}

\begin{document}
\maketitle

\begin{abstract}
PLN's two-number truth value $\STV{s}{c}$ is the compression of a richer
object: a credal set of admissible precise previsions over an outcome space.
We give a layered formal characterization of where $\STV{s}{c}$ is
\emph{determined} by the spec --- the strength is a single real, the
confidence is at its information ceiling --- and where the underlying
credal set has \emph{degrees of freedom} that the two-number summary
cannot see.

The characterization rests on a recently-landed Lean dichotomy in
\codepath{Mettapedia.ProbabilityTheory.ImpreciseProbability.ProjectiveCredal}:
the predicates \lid{credalSetDetermines} and \lid{credalSetHasStrictWidth}
on a credal set $C$ and a gamble $X$ are the formal locators of, respectively,
``$\STV{s}{c}$ is fully determined on $X$'' and ``$\STV{s}{c}$ has
genuine interval width on $X$.'' The collapse theorem
\lid{lower\_eq\_upperEnvelope\_of\_credalSetDetermines} closes the
``determined $\Rightarrow$ precise'' direction; the width theorem
\lid{lower\_upperEnvelope\_nontrivial\_of\_strictWidth} closes the
``disagreement $\Rightarrow$ envelope opens'' direction.

We lift this dichotomy to the projective credal level
(\lid{determinesGlobalGamble} / \lid{hasStrictGlobalWidth}) and onward to
the infinite Markov-logic setting through DLR completions
(\lid{dlrQueryDetermined} / \lid{dlrQueryHasStrictWidth}). Under the
Dobrushin uniform small-influence hypothesis,
\lid{PaperUniformSmallTotalInfluence} implies \lid{dlrQueryDetermined},
and therefore the MLN query envelope is precise. The low-temperature
two-dimensional symmetric-grid theorem now supplies the complementary
strict-width crown:
\lid{symmetricGridZeroField\_originPLNStrictIntervalCrown\_of\_axisAnchoredContourCode\_twentyFour}.
On the exchangeable side, the public de~Finetti theorem
\lid{impreciseDeFinetti\_analyticMixingFamily\_sharpCompatibilityCrown}
separates the unconditional external mixing-family readout from the raw
all-gambles crown, which holds exactly at the lower-prevision
compatibility boundary.

We organize the degrees of freedom into three slots:
(A)~specification-level freedom in the local credal sets;
(B)~completion-level freedom in extreme-point choice within a fixed credal
set (aleatoric, formally invisible to $\STV{s}{c}$);
(C)~phase-transition freedom from non-uniqueness of DLR completions
in infinite MLNs. The endpoint is now a closed Lean package rather than a
list of open infinite-side obligations; the remaining freedom is named by
the predicates and hypotheses, not hidden in the notation $\STV{s}{c}$.
\end{abstract}

\tableofcontents

\section{Introduction}

PLN's atomic truth-value object is a pair $\STV{s}{c}$ with $s, c \in [0,1]$
called \emph{strength} and \emph{confidence}. Standard PLN parametrises
$s$ as a probability estimate and $c$ as a function of evidence count,
typically $c = N/(N+\kappa)$ for a system-level constant $\kappa$ and an
evidence-count integer $N$. The pair is what travels through PLN's
rules (revision, deduction, induction, abduction); it is what is read off
by query atoms; it is the type signature of every PLN belief.

But $\STV{s}{c}$ is a two-number summary. The same pair can be produced by
very different underlying epistemic objects: a Beta-distribution
second-order Bayesian; a Walley-style coherent lower prevision with a
specific interval width; an exchangeable de Finetti mixture with a
particular concentration. PLN's compositional rules treat $\STV{s}{c}$
\emph{as} the belief object, and the algebra reduces to numerical update
formulas; but the question of \emph{what is and is not pinned down by
$\STV{s}{c}$} is rarely written explicitly in the PLN literature.

This paper makes that explicit. We organise the question as follows.
The strength and confidence of a query in PLN are determined by an
underlying credal set $C$ of admissible precise previsions. The
\emph{degrees of freedom} that $\STV{s}{c}$ summarises but does not
expose live in three slots:
\begin{itemize}
\item Specification freedom in the local credal sets that build $C$;
\item Completion freedom (extreme-point choice) inside a fixed $C$;
\item Phase-transition freedom from non-uniqueness of global completions
      when $C$ is the projective limit of a Markov-logic / DLR family.
\end{itemize}
Each slot has a formally locatable status. Recent landings in
the \codepath{Mettapedia.ProbabilityTheory.ImpreciseProbability} and
\codepath{Mettapedia.Logic} libraries (the \emph{credal dichotomy}
predicates and the DLR query lifts; see Sec.~\ref{sec:dichotomy} and
\ref{sec:dlr}) make this dichotomy an actual Lean theorem rather than
informal prose. Wherever possible we quote the Lean theorem statements
verbatim from their source files.

\paragraph{Scope.}
We treat strength and confidence as the read-off of credal-envelope width
and centre. We do \emph{not} address the propagation of $\STV{s}{c}$
through PLN's compositional rules (deduction, induction, abduction,
revision); that material is in the existing PLN corpus
(\xiPLN, \nuPLN, the WM-PLN book; see Sec.~\ref{sec:related}). The paper's
contribution is the determinacy / freedom characterization at the level
of a single truth value, with formal Lean hooks. Each named degree of
freedom additionally comes with a small runnable PeTTa/CeTTa canary paired
with its Lean oracle, in the companion package
\lpath{papers/pln-confidence-examples/} (one-command build and run via
\lpath{run_all.sh}; see Appendix~\ref{app:examples}).

\paragraph{Sibling papers.}
This paper is a sibling of \xiPLN\ and the \emph{PLN Review}
(\lpath{papers/xiPLN.tex}, \lpath{papers/pln-review.tex}). The
WM-PLN book \lpath{papers/wm-pln-book.tex} contains an informal section
on degrees of freedom versus forcing (\S 7.4 in the current draft); the
present paper offers the formal characterization that the book section
calls for.

\section{The credal stack: types and definitions}\label{sec:stack}

We work in the formal stack of
\codepath{Mettapedia.ProbabilityTheory.ImpreciseProbability.ProjectiveCredal}
(referenced throughout as \emph{ProjectiveCredal.lean}, file path
\lpath{Mettapedia/ProbabilityTheory/ImpreciseProbability/ProjectiveCredal.lean}).
Outcome spaces are arbitrary types $\Omega$; gambles are bounded real-valued
functions on $\Omega$.

\subsection{Gambles and previsions}

A \emph{gamble} on $\Omega$ is a real-valued function on $\Omega$:
\begin{lstlisting}
abbrev Gamble (Omega : Type*) := Omega -> Real
\end{lstlisting}

A \emph{precise prevision} on $\Omega$ is a Walley-coherent linear
functional from gambles to reals: the structure
\lid{PrecisePrevision} in ProjectiveCredal.lean is the standard linear
expectation with constant-preservation, positive homogeneity, and
additivity built in:
\begin{lstlisting}
structure PrecisePrevision (Omega : Type*) where
  toFun       : Gamble Omega -> Real
  lower_bound : ... -- c <= X (pointwise) implies c <= toFun X
  pos_homog   : ... -- toFun (r . X) = r * toFun X for r >= 0
  add         : ... -- toFun (X + Y) = toFun X + toFun Y
\end{lstlisting}

A \emph{lower prevision} relaxes additivity to superadditivity, and is
the Walley credal-summary object:
\begin{lstlisting}
structure LowerPrevision (Omega : Type*) where
  toFun       : Gamble Omega -> Real
  lower_bound : ... -- c <= X pointwise implies c <= toFun X
  pos_homog   : ... -- toFun (r . X) = r * toFun X for r >= 0
  superadd    : ... -- toFun (X + Y) >= toFun X + toFun Y
\end{lstlisting}
Coherence (positive homogeneity, superadditivity, and avoidance of sure
loss) is the Walley condition for the lower prevision to be the
lower envelope of some non-empty set of precise previsions.

\subsection{Credal sets and envelopes}

A \emph{credal set} is just a set of precise previsions:
\begin{lstlisting}
abbrev CredalPrevisionSet (Omega : Type*) := Set (PrecisePrevision Omega)
\end{lstlisting}

For a credal set $C$ and a gamble $X$, the \emph{lower} and \emph{upper}
envelopes are the standard inf / sup:
\begin{lstlisting}
def lowerEnvelope (C : CredalPrevisionSet Omega) (X : Gamble Omega) : Real :=
  sInf ((fun P : PrecisePrevision Omega => P X) '' C)

def upperEnvelope (C : CredalPrevisionSet Omega) (X : Gamble Omega) : Real :=
  sSup ((fun P : PrecisePrevision Omega => P X) '' C)
\end{lstlisting}
The construction
\lid{lowerEnvelopePrevision} packages \lid{lowerEnvelope} together with
the proofs of constant-preservation, positive homogeneity, and
superadditivity, exhibiting a \lid{LowerPrevision} object out of any
non-empty bounded-below credal set. The finite case theorem
\lid{lowerEnvelopePrevision\_isCoherent\_of\_finite}
confirms Walley coherence on finite
nonempty $\Omega$.

\subsection{PLN $\STV{s}{c}$ as the credal-envelope summary}\label{sec:stv-summary}

The canonical Lean STV type lives in
\codepath{Mettapedia.PLN.RuleFamilies.FirstOrder.PLNDeduction}:
\begin{lstlisting}
/-- Simple Truth Value: (strength, confidence) pair -/
structure STV where
  strength   : Real
  confidence : Real
  strength_nonneg  : 0 <= strength
  strength_le_one  : strength <= 1
  confidence_nonneg : 0 <= confidence
  confidence_le_one : confidence <= 1
\end{lstlisting}
Its distributional model is in \codepath{Mettapedia.PLN.TruthValues.PLNDistributional}:
given a Beta$(\alpha, \beta)$ distribution, the canonical PLN STV reads
\begin{equation*}
   s \;=\; \frac{\alpha}{\alpha + \beta},
   \qquad
   c \;=\; f(\alpha + \beta)
\end{equation*}
for an evidence-count map $f$ (typically $f(N) = N/(N+\kappa)$).

The two-number compression from a credal set $C$ and a query gamble
$X_q$ (e.g.\ the indicator $\mathbf{1}_A$ of a query atom $A$) is the
following implicit map:
\begin{equation}\label{eq:summary}
   C, X_q \;\longmapsto\;
   \left\{
   \begin{array}{l}
   s_q \;=\; \tfrac{1}{2}\bigl(\,\mathrm{lowerEnvelope}\, C\, X_q\,
                              +\,\mathrm{upperEnvelope}\, C\, X_q\,\bigr) \\[4pt]
   c_q \;=\; g\bigl(\,\mathrm{upperEnvelope}\, C\, X_q\,
                   -\,\mathrm{lowerEnvelope}\, C\, X_q\,\bigr)
   \end{array}
   \right.
\end{equation}
for a monotone-decreasing $g : [0,1] \to [0,1]$ with $g(0) = 1$ (no
width, full confidence) and $g(1) = 0$ (maximum width, no information).
The specific choice of midpoint and $g$ is an interface design decision
(PLN traditionally uses midpoint for $s$ and a sigmoidal $g$ derived from
the Beta variance); the freedom characterization of this paper is
independent of this choice. What matters is that \eqref{eq:summary}
is the read-off, and that the credal set $C$ may have degrees of
freedom that the pair $(s_q, c_q)$ cannot expose.

\begin{remark}
The summary map \eqref{eq:summary} loses two kinds of information.
First, \emph{which} gambles have positive credal-width and which do
not: $(s_q, c_q)$ summarises a single query but a credal set with
non-zero width on \emph{some} queries may have zero width on others.
Second, the \emph{shape} of $C$: a credal set on $\Omega$ whose extreme
points are clustered (high agreement) and one whose extreme points are
dispersed (low agreement) can produce the same midpoint and width on a
particular query, but compose differently under PLN's inference rules.
The width as a function on gambles --- which we will study below --- is
the right object; $(s_q, c_q)$ is its evaluation at one point.
\end{remark}

\section{Finite-case determinacy}\label{sec:finite}

We first treat the case where the outcome space $\Omega$ is finite. Here
the Walley theory is complete in the formalization; we summarise the
load-bearing results.

\subsection{Uniform positivity and avoidance of sure loss}

For finite $\Omega$, a pointwise-strictly-positive gamble has a uniform
positive lower bound :
\begin{lstlisting}
theorem finite_strictlyPositive_uniformLowerBound
    {Omega : Type*} [Fintype Omega] [Nonempty Omega]
    (X : Gamble Omega) (hX : forall omega, 0 < X omega) :
    exists epsilon : Real, 0 < epsilon and forall omega, epsilon <= X omega
\end{lstlisting}
This is the technical hinge that lets every nonempty finite credal set
\emph{avoid sure loss} in the strict Walley sense:
\begin{lstlisting}
theorem lowerEnvelopePrevision_avoidsSureLoss_of_finite
    {Omega : Type*} [Fintype Omega] [Nonempty Omega]
    (C : CredalPrevisionSet Omega) (hC : C.Nonempty)
    (hBdd : forall X : Gamble Omega,
      BddBelow ((fun P : PrecisePrevision Omega => P X) '' C)) :
    (lowerEnvelopePrevision C hC hBdd).avoidsSureLoss
\end{lstlisting}
and to be \emph{coherent} :
\begin{lstlisting}
theorem lowerEnvelopePrevision_isCoherent_of_finite
    {Omega : Type*} [Fintype Omega] [Nonempty Omega]
    (C : CredalPrevisionSet Omega) (hC : C.Nonempty)
    (hBdd : forall X : Gamble Omega,
      BddBelow ((fun P : PrecisePrevision Omega => P X) '' C)) :
    (lowerEnvelopePrevision C hC hBdd).isCoherent
\end{lstlisting}
The take-away for PLN: on a finite outcome space, the credal-envelope
construction never fails. The map from spec to $\STV{s}{c}$ is total.

\subsection{Singleton collapse}

The case where the credal set has exactly one element is the case where
PLN's $\STV{s}{c}$ has no freedom at all; everything is pinned. The
theorem is direct:
\begin{lstlisting}
theorem lowerEnvelope_singleton
    {Omega : Type*} (P : PrecisePrevision Omega) (X : Gamble Omega) :
    lowerEnvelope ({P} : CredalPrevisionSet Omega) X = P X
\end{lstlisting}
and similarly for \lid{upperEnvelope\_singleton}. On a singleton credal
set, lower = upper on every gamble. PLN-side: every query has $c$ at its
information ceiling, and $s$ is the single value $P X_q$.

\section{The credal dichotomy: where strength is determined and where it has width}\label{sec:dichotomy}

The contribution of the recent landing in ProjectiveCredal.lean is to
make the ``determined / has width'' distinction \emph{predicate-level},
not theorem-level. The collapse and non-triviality theorems following
those definitions pin down the PLN-side consequence.

\subsection{The two predicates}

\begin{definition}[determined credal set on a gamble]\label{def:determines}
The predicate \lid{credalSetDetermines} :
\begin{lstlisting}
def credalSetDetermines {Omega : Type*}
    (C : CredalPrevisionSet Omega) (X : Gamble Omega) : Prop :=
  forall P : PrecisePrevision Omega, P in C ->
    forall Q : PrecisePrevision Omega, Q in C -> P X = Q X
\end{lstlisting}
$C$ \emph{determines} $X$ when every two admissible precise previsions in
$C$ assign the same expectation to $X$.
\end{definition}

\begin{definition}[strict width]\label{def:strictwidth}
The predicate \lid{credalSetHasStrictWidth} :
\begin{lstlisting}
def credalSetHasStrictWidth {Omega : Type*}
    (C : CredalPrevisionSet Omega) (X : Gamble Omega) : Prop :=
  exists P : PrecisePrevision Omega, P in C and
    exists Q : PrecisePrevision Omega, Q in C and P X < Q X
\end{lstlisting}
$C$ has \emph{strict width} on $X$ when there exist two precise previsions
in $C$ that strictly disagree on $X$.
\end{definition}

The two predicates are mutually exclusive :
\begin{theorem}[mutual exclusion]\label{thm:mutex}
\lid{not\_credalSetDetermines\_of\_strictWidth}:
\begin{lstlisting}
theorem not_credalSetDetermines_of_strictWidth
    {Omega : Type*}
    {C : CredalPrevisionSet Omega} {X : Gamble Omega}
    (hWidth : credalSetHasStrictWidth C X) :
    not (credalSetDetermines C X)
\end{lstlisting}
\end{theorem}

The endpoint instances cover the trivial cases. Subsingleton credal sets
collapse to determinacy
(\lid{credalSetDetermines\_of\_subsingleton}), and so do
singletons (\lid{credalSetDetermines\_singleton}):
\begin{lstlisting}
theorem credalSetDetermines_singleton {Omega : Type*}
    (P : PrecisePrevision Omega) (X : Gamble Omega) :
    credalSetDetermines ({P} : CredalPrevisionSet Omega) X
\end{lstlisting}

\subsection{Collapse and width theorems}

The PLN-side consequence of \lid{credalSetDetermines} is precisely
that strength is pinned: the lower and upper envelopes coincide:
\begin{theorem}[envelope collapse from determinacy]\label{thm:collapse}
\lid{lower\_eq\_upperEnvelope\_of\_credalSetDetermines}:
\begin{lstlisting}
theorem lower_eq_upperEnvelope_of_credalSetDetermines
    {Omega : Type*}
    (C : CredalPrevisionSet Omega) (X : Gamble Omega)
    (hC : C.Nonempty)
    (hBddBelow : BddBelow ((fun P => P X) '' C))
    (hBddAbove : BddAbove ((fun P => P X) '' C))
    {P : PrecisePrevision Omega} (hP : P in C)
    (hDet : credalSetDetermines C X) :
    lowerEnvelope C X = upperEnvelope C X
\end{lstlisting}
This is the formal statement of ``$\STV{s}{c}$ on $X_q$ is precise''
when $C$ determines $X_q$.
\end{theorem}

The PLN-side consequence of \lid{credalSetHasStrictWidth} is the opposite:
the envelope is genuinely non-trivial :
\begin{theorem}[envelope width from strict disagreement]\label{thm:width}
\lid{lower\_upperEnvelope\_nontrivial\_of\_strictWidth}:
\begin{lstlisting}
theorem lower_upperEnvelope_nontrivial_of_strictWidth
    {Omega : Type*}
    (C : CredalPrevisionSet Omega) (X : Gamble Omega)
    (hBddBelow : BddBelow ((fun P => P X) '' C))
    (hBddAbove : BddAbove ((fun P => P X) '' C))
    (hWidth : credalSetHasStrictWidth C X) :
    lowerEnvelope C X < upperEnvelope C X
\end{lstlisting}
This is the formal statement of ``$\STV{s}{c}$ on $X_q$ is genuinely an
interval'' when two completions in $C$ strictly disagree on $X_q$.
\end{theorem}

\subsection{PLN read-off}\label{sec:plnreadoff}

Putting the dichotomy into the summary map \eqref{eq:summary}: for any
credal set $C$ on which both envelope-boundedness conditions hold, and any
query gamble $X_q$, exactly one of the following holds:
\begin{center}
\begin{tabular}{lll}
\toprule
predicate & envelope & PLN $\STV{s}{c}$ \\
\midrule
\lid{credalSetDetermines} $C$ $X_q$ & $L = U$ & precise $s = L\,X_q$; $c$ at ceiling \\
\lid{credalSetHasStrictWidth} $C$ $X_q$ & $L < U$ & $s$ in interval $[L, U]$; $c < $ ceiling \\
\bottomrule
\end{tabular}
\end{center}
The Lean dichotomy is the formal locator of which case the PLN truth-value
read-off lives in.

\section{Projective credal specs: three slots for freedom}\label{sec:projective}

Once $\Omega$ is large (or infinite), one builds the credal set
\emph{compositionally} from finite-window data. The standard packaging
is a projective family of local credal sets, indexed by ``windows''
(finite subsets, finite cylinders) of the global outcome space, glued
by restriction maps.

\subsection{The projective machinery}

The cylinder structure (ProjectiveCredal.lean, the
\lid{ProjectiveCylinderSystem} structure) packs the window-to-local
projection and the local restriction maps:
\begin{lstlisting}
structure ProjectiveCylinderSystem (Window Global : Type*) [LE Window] where
  Local             : Window -> Type*
  project           : forall i : Window, Global -> Local i
  restrict          : forall {i j : Window}, i <= j -> Local j -> Local i
  project_restrict  : ...   -- compatibility on the diagram
\end{lstlisting}
The full spec layers a local credal choice on top:
\begin{lstlisting}
structure ProjectiveLocalCredalSpec (Window Global : Type*) [LE Window] where
  cylinders   : ProjectiveCylinderSystem Window Global
  localCredal : forall i : Window, CredalPrevisionSet (cylinders.Local i)
\end{lstlisting}
The \emph{global credal set} associated to a spec is the
\lid{projectiveLimitCredalSet} (the set of $P$ in \lid{PrecisePrevision}
\lid{Global} that marginalise into \lid{localCredal}{} $i$ at every window
$i$).

\subsection{Three identifying cases and the dichotomy at the global level}

The codebase ships two endpoints and one general construction:
\begin{itemize}
\item \textbf{Singleton} :
\begin{lstlisting}
def singletonIdentityProjectiveSpec {Omega : Type*}
    (P : PrecisePrevision Omega) : ProjectiveLocalCredalSpec PUnit Omega
\end{lstlisting}
Every window's local credal is the supplied singleton $\{P\}$. The
projective limit is $\{P\}$
(\lid{singletonIdentityProjectiveSpec\_projectiveLimitCredalSet}). PLN $\STV{s}{c}$ on every gamble is precise; this is the
``no freedom'' case.

\item \textbf{Unrestricted} :
\begin{lstlisting}
def finiteUnrestrictedProjectiveSpec (Omega : Type*) :
    ProjectiveLocalCredalSpec PUnit Omega
\end{lstlisting}
Every window's local credal is \lid{Set.univ}. The projective limit
contains every precise prevision on $\Omega$. PLN $\STV{s}{c}$ on most
gambles has \lid{credalSetHasStrictWidth}; this is the ``maximum freedom''
case. The non-emptiness of the projective limit is witnessed by
\lid{finiteUnrestricted\_finiteEvaluationCompact\_hasCompatibleCompletion}
via a compact carrier in the finite-evaluation topology.

\item \textbf{Genuine intermediate}: any spec whose local credals are
non-trivial proper subsets of \lid{Set.univ}. Whether
\lid{credalSetDetermines} or \lid{credalSetHasStrictWidth} holds on a
particular gamble depends on the spec and the gamble.
\end{itemize}

The dichotomy lifts to the projective level. ProjectiveCredal.lean defines
\lid{determinesGlobalGamble} and \lid{hasStrictGlobalWidth}
as the global analogues of \lid{credalSetDetermines} and
\lid{credalSetHasStrictWidth}, with the mutual exclusion theorem
\lid{not\_determinesGlobalGamble\_of\_strictWidth}. The
natural-extension lifting theorems connect
the projective predicates to the lower / upper envelopes on the projective
limit.

\subsection{The three formal slots, structurally}\label{sec:slots-structural}

We can now name the three slots in which PLN credal freedom lives. Each
is formally locatable through one of the predicates above.

\begin{description}
\item[(A) Specification-level freedom.] The choice of \lid{localCredal}{} $i$
at each window. This is the modeller's freedom: evidence width, prior
choices, and what classes of measure are taken to be admissible at each
local scale.

\item[(B) Completion-level freedom.] Given a fixed credal set $C$, the
choice of $P \in C$ that actually governs the world. This freedom is
purely aleatoric: it is invisible to $\STV{s}{c}$ because $\STV{s}{c}$ is
a property of $C$, not of any specific $P \in C$.

\item[(C) Phase-transition freedom.] For globally-glued projective specs,
the projective limit may itself have multiple admissible elements that
correspond to different \emph{global} solutions of the same local
problem. This is the slot Markov-logic / DLR exemplifies (next section).
\end{description}

Slots (A) and (C) are the only ones $\STV{s}{c}$ can see. Slot (B) is
formally invisible at the truth-value level.

\section{Infinite Markov logic: Dobrushin collapse and DLR width}\label{sec:dlr}

For the infinite case --- countably-many ground atoms, Markov-logic
network with weights, density specified locally --- the projective
machinery specialises into the Dobrushin--Lanford--Ruelle framework. The
file
\codepath{Mettapedia.Logic.MarkovLogicInfiniteCredalBridge}
(\lpath{Mettapedia/Logic/MarkovLogicInfiniteCredalBridge.lean}) packages this.

\subsection{Classical MLN spec and DLR completions}

A \lid{ClassicalInfiniteGroundMLNSpec} packages a countable atom set, a
clause-id set, the clauses, their log-weights, and the local
finite-volume region structure. A \emph{DLR completion} of such a spec
is a probability measure on the infinite world satisfying the
DLR consistency condition at every finite volume:
\begin{lstlisting}
abbrev DLRCompletion (M : ClassicalInfiniteGroundMLNSpec Atom ClauseId) :=
  { mu : ProbabilityMeasure (InfiniteWorld Atom)
  // FixedRegionCylinderDLR M.toStrictlyPositiveInfiniteGroundMLNSpec
       (mu : Measure (InfiniteWorld Atom)) }
\end{lstlisting}
The set of DLR completions of $M$ is the ``credal set'' for the infinite
spec, and the projective-credal lift packages it as
\lid{DLRProjectiveCredalSpecialization} in
\codepath{Mettapedia/Logic/MarkovLogicInfiniteCredalBridge.lean}.

A \emph{query} is a finite-volume event; its \emph{probability} under a
DLR completion is its measure:
\begin{lstlisting}
noncomputable def dlrCompletionQueryProb
    (M : ClassicalInfiniteGroundMLNSpec Atom ClauseId)
    (q : ConstraintQuery Atom) (mu : DLRCompletion M) : Real :=
  ENNReal.toReal
    ((infiniteMLNMassSemantics M mu.1 mu.2).queryProb q)
\end{lstlisting}

\subsection{DLR-level dichotomy}\label{sec:dlrdichotomy}

The credal dichotomy lifts directly to the DLR setting:
\begin{lstlisting}
def dlrQueryDetermined
    (M : ClassicalInfiniteGroundMLNSpec Atom ClauseId)
    (q : ConstraintQuery Atom) : Prop :=
  forall mu nu : DLRCompletion M,
    dlrCompletionQueryProb M q mu = dlrCompletionQueryProb M q nu

def dlrQueryHasStrictWidth
    (M : ClassicalInfiniteGroundMLNSpec Atom ClauseId)
    (q : ConstraintQuery Atom) : Prop :=
  exists mu : DLRCompletion M, exists nu : DLRCompletion M,
    dlrCompletionQueryProb M q mu < dlrCompletionQueryProb M q nu
\end{lstlisting}
\lid{dlrQueryDetermined} says all DLR completions agree on the query;
\lid{dlrQueryHasStrictWidth} says two of them strictly disagree. The
mutual exclusion theorem
\lid{not\_dlrQueryDetermined\_of\_strictWidth}:
\begin{lstlisting}
theorem not_dlrQueryDetermined_of_strictWidth
    (M : ClassicalInfiniteGroundMLNSpec Atom ClauseId)
    (q : ConstraintQuery Atom)
    (hWidth : dlrQueryHasStrictWidth M q) :
    not (dlrQueryDetermined M q)
\end{lstlisting}

\subsection{Dobrushin uniqueness as the collapse-direction hypothesis}

The classical Dobrushin uniform-influence condition is packaged as the
predicate \lid{PaperUniformSmallTotalInfluence} on the MLN spec
in \codepath{Mettapedia.Logic.MarkovLogicInfiniteUniqueness}.
The collapse direction of the DLR-level dichotomy is the theorem
\lid{dlrQueryDetermined\_of\_uniform}:
\begin{lstlisting}
theorem dlrQueryDetermined_of_uniform
    (M : ClassicalInfiniteGroundMLNSpec Atom ClauseId)
    (hM : M.PaperUniformSmallTotalInfluence)
    (q : ConstraintQuery Atom) :
    dlrQueryDetermined M q
\end{lstlisting}
\emph{Under the Dobrushin uniform small-influence hypothesis, every finite
query is DLR-determined.}
Combining with the dichotomy gives the precision of the MLN query
envelope:
\begin{lstlisting}
theorem infiniteMLN_queryEnvelope_precise_of_uniform
    (M : ClassicalInfiniteGroundMLNSpec Atom ClauseId)
    [Nonempty (DLRCompletion M)]
    (hM : M.PaperUniformSmallTotalInfluence)
    (q : ConstraintQuery Atom) :
    infiniteMLNLowerQueryEnvelope M q
      = infiniteMLNUpperQueryEnvelope M q
\end{lstlisting}
\emph{Under the same hypothesis, and assuming the DLR completion family is
inhabited, the PLN $\STV{s}{c}$ for every finite query is precise.}

\subsection{Strict width as the phase-transition canary}

The opposite direction is the theorem
\lid{infiniteMLN\_queryEnvelope\_nontrivial\_of\_disagreement}
and the lift through
\lid{infiniteMLN\_queryEnvelope\_nontrivial\_of\_strictWidth}:
\begin{lstlisting}
theorem infiniteMLN_queryEnvelope_nontrivial_of_strictWidth
    (M : ClassicalInfiniteGroundMLNSpec Atom ClauseId)
    (q : ConstraintQuery Atom)
    (hBddBelow : BddBelow (Set.range (dlrCompletionQueryProb M q)))
    (hBddAbove : BddAbove (Set.range (dlrCompletionQueryProb M q)))
    (hWidth : dlrQueryHasStrictWidth M q) :
    infiniteMLNLowerQueryEnvelope M q
      < infiniteMLNUpperQueryEnvelope M q
\end{lstlisting}
\emph{If two DLR completions disagree on a finite query, the PLN
$\STV{s}{c}$ for that query is genuinely an interval.}

\subsection{Dobrushin collapse and strict-width witnesses}

Combining the two proved directions gives a one-sided collapse theorem
and a separate strict-width witness theorem. The Dobrushin condition
\lid{PaperUniformSmallTotalInfluence} is a formally sufficient hypothesis
for query determinacy and precise PLN/Walley envelopes. Separately, if
two DLR completions are exhibited that disagree on a query, then that
query has strict width. The Lean development does not prove that failure
of the Dobrushin hypothesis entails non-uniqueness or strict width.

\section{Exchangeable de~Finetti families: the F2 compatibility boundary}\label{sec:definetti-f2}

The second infinite lane is the exchangeable IID/de~Finetti lane. Here the
formal point is not merely that an external mixture representation exists:
it does, unconditionally, for any nonempty credal family of Bernoulli
mixtures. The stronger all-gambles process-law crown has a sharper
boundary.

The public theorem
\lid{impreciseDeFinetti\_analyticMixingFamily\_sharpCompatibilityCrown}
states:
\begin{lstlisting}
theorem impreciseDeFinetti_analyticMixingFamily_sharpCompatibilityCrown
    (C : Set BernoulliMixture) (hC : C.Nonempty) :
    ImpreciseDeFinettiCanonicalExternalMixingFamily C hC /\
      (ImpreciseDeFinettiAnalyticMixingFamilyProcessLawCrown C hC <->
        exists L : LowerPrevision (Nat -> Bool),
          (bernoulliMixturePrefixProcessLowerSpec C ... hC).
            respectsLocalLower L)
\end{lstlisting}
Thus the canonical sigma-additive external readout is always present, while
the raw all-gambles analytic crown is pinned exactly by existence of a
respecting lower prevision for the analytic prefix lower envelope. This is
the Walley natural-extension boundary in the exchangeable process setting.

A useful negative theorem prevents over-reading the result. The theorem
\lid{exists\_impreciseDeFinetti\_analyticMixingFamilyProcessLawCrown\_not\_pointwiseZeroInterior}
exhibits a nonempty mixing family whose raw all-gambles crown exists even
though not every member has zero interior mixing mass. Equivalently,
\lid{not\_forall\_impreciseDeFinetti\_analyticMixingFamilyProcessLawCrown\_imp\_pointwiseZeroInterior\_closed}
closes the refutation of the too-strong pointwise-zero necessity claim.
Pointwise zero-interior is a sufficient concrete route; it is not the
right family-level characterization.

\section{Worked examples already in the formalization}\label{sec:examples}

The codebase now ships both finite toy witnesses and a concrete infinite
DLR/PLN contrast crown, plus the low-temperature symmetric-grid strict
interval crown. The finite two-spin examples remain useful sanity witnesses
for strict credal width.

\subsection{Two-spin magnet}

In \lpath{ProjectiveCredal.lean}, the type
\lid{TwoSpin}, defined as $\mathrm{Bool} \times \mathrm{Bool}$, carries the two-spin
magnet credal set
\begin{lstlisting}
def twoSpinMagnetCredalSet : CredalPrevisionSet TwoSpin :=
  { PrecisePrevision.dirac twoSpinAllDown,
    PrecisePrevision.dirac twoSpinAllUp }
\end{lstlisting}
The credal set contains two Dirac measures: all spins down, and all
spins up. The first-spin-up gamble
\begin{lstlisting}
def twoSpinFirstUpGamble : Gamble TwoSpin :=
  fun s => if s.fst then 1 else 0
\end{lstlisting}
is rewarded by the all-up Dirac and unrewarded by the all-down Dirac;
the disagreement gives the strict width witness. The packaged theorem
\lid{twoSpinMagnetEnvelope\_nontrivial} closes:
\begin{lstlisting}
theorem twoSpinMagnetEnvelope_nontrivial :
    lowerEnvelope twoSpinMagnetCredalSet twoSpinFirstUpGamble
      < upperEnvelope twoSpinMagnetCredalSet twoSpinFirstUpGamble
\end{lstlisting}
PLN-side: on this two-Dirac credal, the magnetization query has lower
envelope $0$, upper envelope $1$, and confidence at its floor.

\subsection{Zero-temperature aligned magnet}

An incremental sibling
\lid{twoSpinZeroTemperatureMagnetEnvelope\_nontrivial} adds the
zero-temperature Gibbs constraint (spins must be aligned), and shows
the envelope is still strictly non-trivial: any measure supported on
the aligned-spin pair $\{(\downarrow,\downarrow), (\uparrow,\uparrow)\}$
still admits both extremes. The structural constraint by itself does
not close down credal width; the Dobrushin condition has to be checked
\emph{quantitatively} on the local-influence row-sums.

\subsection{Concrete infinite crown: reinforced line versus zero-weight grid}

The theorem
\lid{reinforcedLineGeometric\_strictWidth\_and\_zeroWeightGrid\_collapse}
packages the first concrete DLR/PLN contrast. The strict-width side uses
\lid{reinforcedLineGeometric\_originSpinUp\_plnStrictInterval}: a
geometrically reinforced half-line has positive PLN strict-width at the
origin-spin query. The collapse side uses
\lid{gridZeroField\_originSpinUp\_queryEnvelope\_precise\_of\_dobrushin}:
the zero-weight zero-field grid lies in the Dobrushin collapse regime for
the analogous origin-spin query. This remains a useful canary for the
collapse direction.

\subsection{Low-temperature symmetric-grid strict interval crown}

The full two-dimensional low-temperature strict-width side is supplied by
the symmetric-grid Peierls development in
\codepath{Mettapedia.Logic.MarkovLogicInfiniteSymmetricGridExample}. The
public theorem
\lid{symmetricGridZeroField\_originPLNStrictIntervalCrown\_of\_axisAnchoredContourCode\_twentyFour}
states:
\begin{lstlisting}
theorem symmetricGridZeroField_originPLNStrictIntervalCrown_of_axisAnchoredContourCode_twentyFour :
    SymmetricGridZeroFieldOriginPLNStrictIntervalCrown (24 : Real)
\end{lstlisting}
The crown packages positive scalar query-envelope width, width-complement
below one, and positive binary query-outcome credal width for the origin-spin
query. The proof path is the classical Peierls route specialized to the
finite box: axis-anchored contour codes, a finite-box $\mathbb{F}_2$
separator/potential argument, a geometric contour count, and the resulting
plus/minus boundary separation. The theorem is independent of the refuted
Markov de~Finetti branch and does not rely on the older conditional
\lid{CutWalkCode} adapters.

\section{Three kinds of degree of freedom: the synthesis}\label{sec:synthesis}

We now name the three formally-locatable freedom slots, each with the
predicate that witnesses it.

\paragraph{(A) Specification-level: \lid{localCredal} choice.}
A wider local credal at any cylinder window produces a wider projective
limit. The PLN-side question ``does my $\STV{s}{c}$ on query $X_q$ have
genuine width?'' reduces to ``does \lid{hasStrictGlobalWidth}
\lid{spec} on $X_q$ hold?''. The modeller controls this slot by varying
\lid{localCredal}{} $i$. The slot is finite or infinite depending on
whether $\Omega$ is.

\paragraph{(B) Completion-level: extreme-point choice.}
Given a credal set $C$, the actual world chooses one $P \in C$ (the
``true'' precise prevision). This freedom is aleatoric. It is formally
invisible to $\STV{s}{c}$ because $\STV{s}{c}$ is a property of $C$, not
of any specific $P$. PLN inference rules do not see it; only the
verification of a query against actual data sees it.

\paragraph{(C) Phase-transition: DLR non-uniqueness.}
In infinite Markov-logic specs, the projective limit is the set of DLR
completions, which may itself be non-singleton. The slot is opened
exactly when \lid{dlrQueryHasStrictWidth} on $(M, q)$ holds for the
query $q$ of interest, and closed when \lid{dlrQueryDetermined} on $(M, q)$
holds. The Dobrushin condition \lid{PaperUniformSmallTotalInfluence}
is a sufficient condition for the closed case; the low-temperature
symmetric-grid Peierls theorem is the complementary proved strict-width
case.

\begin{observation}
Slot (A) is the freedom the PLN modeller intends to give. Slot (B) is
the freedom the world has within the modeller's spec. Slot (C) is the
freedom that the infinite structure of the spec itself produces, without
any modeller intent and without any world-side aleatoric residual.
\end{observation}

The three slots are independent. A modeller can pick (A) tight, the
world can pick (B) freely, and the spec can still have (C) freedom from
a phase transition. The PLN $\STV{s}{c}$ sees (A) and (C); it does not
see (B).

\section{Closed endpoint and remaining polish}\label{sec:open}

The current Lean endpoint is the profile
\lid{confidenceCharacterizationEndpointProfile}, exported through
\codepath{Mettapedia.PLN.Core.PLNCanonicalAPI}. It packages the finite forcing
results, the typed-STV canaries showing residual degrees of freedom, the
Walley width-complement forcing theorem for the PLN odds coordinate, the
finite exact-predictive ITV collapse, the DLR/Dobrushin collapse lane, the
F2 exchangeable mixing-family compatibility boundary, and the F4
low-temperature symmetric-grid strict-interval crown.

The characterization should therefore be read as closed relative to its
explicit semantic gates:
\begin{itemize}
\item \textbf{Finite forcing.}
      Once evidence counts are read through a Walley width-complement
      compatible evidence-weight coordinate, the odds form of PLN
      confidence is forced by
      \lid{walley\_width\_complement\_forces\_pln\_odds}. The free
      parameter is the evidence-scale constant, not the shape of the
      coordinate.

\item \textbf{Finite exact collapse.}
      Singleton and exact posterior carriers have lower envelope equal to
      upper envelope. Their ITV readout has width zero, credibility one,
      and midpoint equal to the single predictive prevision.

\item \textbf{F2, exchangeable IID/de~Finetti.}
      \lid{impreciseDeFinetti\_analyticMixingFamily\_sharpCompatibilityCrown}
      proves that the external sigma-additive mixing-family readout is
      unconditional, while the raw all-gambles process-law crown holds
      exactly at the lower-prevision compatibility boundary. The theorem
      \lid{not\_forall\_impreciseDeFinetti\_analyticMixingFamilyProcessLawCrown\_imp\_pointwiseZeroInterior\_closed}
      prevents the common but false strengthening to pointwise zero-interior
      necessity.

\item \textbf{F3, Dobrushin collapse.}
      \lid{infiniteMLN\_paperUniformSmallTotalInfluence\_implies\_uniqueMeasure}
      gives the Markov-logic/DLR collapse direction: uniform small influence
      forces a unique infinite measure, hence precise finite-query envelopes.

\item \textbf{F4, low-temperature Ising width.}
      \lid{symmetricGridZeroField\_originPLNStrictIntervalCrown\_of\_axisAnchoredContourCode\_twentyFour}
      gives the 2D symmetric-grid strict-interval crown at low temperature,
      by the finite-box Peierls/$\mathbb{F}_2$ contour route.
\end{itemize}

What remains is integration polish rather than a missing crown theorem: the
book chapter should import the closed endpoint, and future work can broaden the
menu of examples and sharpen constants. A runnable companion now accompanies
this paper at \lpath{papers/pln-confidence-examples/}: it pairs each degree of
freedom with a small PeTTa/CeTTa canary and its Lean oracle (see
Appendix~\ref{app:examples}), with a one-command harness \lpath{run_all.sh}.
Those refinements do not change the status of the confidence-formula
characterization itself.

\section{Related work}\label{sec:related}

\paragraph{Walley.}
The credal-set lower-prevision framework is Walley (1991). Our
\lid{LowerPrevision} structure is the Walley structure with
constant-preservation, positive homogeneity, and superadditivity, and
our \lid{lowerEnvelopePrevision\_isCoherent\_of\_finite} is the
classical Walley coherence theorem for finite envelopes.

\paragraph{de~Finetti.}
The de~Finetti exchangeability theorem connects exchangeable distributions
on infinite sequences to mixtures over IID distributions. The codebase now
has the finite-prefix imprecise de~Finetti payoff theorems, the canonical
external sigma-additive mixing-family object, and the raw all-gambles
process-law compatibility boundary. The important caution is negative:
family-level crown existence does not force pointwise zero-interior mass for
every mixture in the family.

\paragraph{Dobrushin--Lanford--Ruelle.}
The DLR framework characterises infinite Gibbs measures by their
finite-volume conditional distributions. The Dobrushin uniqueness
condition (uniform smallness of total influence) implies a unique DLR
completion. Singla and Domingos have a Markov-logic specific
uniformity-of-influence framing. Our \lid{PaperUniformSmallTotalInfluence}
is the codified hypothesis.

\paragraph{PLN literature.}
The two-component truth value $\STV{s}{c}$ is canonical in the PLN
literature (Goertzel et al.\ \emph{Probabilistic Logic Networks}). The
sibling papers \xiPLN\ (\lpath{papers/xiPLN.tex}) and the PLN Review
(\lpath{papers/pln-review.tex}) treat the evidence-algebra and
formal-rule-coherence axes. The WM-PLN book
(\lpath{papers/wm-pln-book.tex}) has the informal predecessor of the
degrees-of-freedom characterization in its \S 7.4
(``Degrees of freedom versus forcing'').

\paragraph{Markov logic networks.}
Richardson and Domingos (Machine Learning, 2006) introduced MLNs as
finite log-linear models on first-order logic. The infinite
specialisation in
\codepath{Mettapedia.Logic.MarkovLogicInfinite*}
is cited inline above: Dobrushin uniqueness supplies the collapse side, and
the symmetric-grid Peierls theorem supplies the low-temperature strict-width
side.

\paragraph{Imprecise probability and second-order Bayesianity.}
The PLN tradition of treating $c$ as ``confidence-in-strength'' has a
natural Bayesian second-order reading (a Beta distribution over the
true probability), which the file
\codepath{Mettapedia.PLN.TruthValues.PLNDistributional} formalises. The credal
reading we develop here is the Walley-side complement; in the
finite case the two readings overlap on the singleton case and
diverge on the strict-width case.

\section{Conclusion}\label{sec:conclusion}

PLN truth values $\STV{s}{c}$ are summary statistics of a credal set,
and the question of where their information lives reduces, formally,
to a dichotomy on the underlying credal set. The codebase landings in
\codepath{Mettapedia.ProbabilityTheory.ImpreciseProbability.ProjectiveCredal}
(the \lid{credalSetDetermines} / \lid{credalSetHasStrictWidth}
predicates and their collapse / width theorems),
\codepath{Mettapedia.ProbabilityTheory.Exchangeability.DeFinettiProjectiveCredalBridge} (the F2
compatibility boundary), and
\codepath{Mettapedia.Logic.MarkovLogicInfiniteSymmetricGridExample} (the
F4 strict-width crown) make this dichotomy a checkable Lean endpoint rather
than informal commentary.

The three freedom slots --- specification, completion, phase-transition
--- partition the question along orthogonal axes. (A) is the modeller's
freedom. (B) is the world's freedom. (C) is the structural freedom
the spec produces by itself. PLN $\STV{s}{c}$ sees (A) and (C); it
does not see (B). Where it sees (C), Dobrushin uniqueness pins the
high-temperature collapse side and the Peierls theorem pins the
low-temperature strict-width side.

The infinite half is no longer a list of named crown gaps. The exchangeable
F2 lane is pinned by lower-prevision compatibility rather than pointwise
zero-interior mass; the DLR F3 lane is pinned on the collapse side by
Dobrushin uniqueness; and the Ising F4 lane is pinned on the width side by
the low-temperature symmetric-grid strict-interval crown. What remains is
the ordinary work of presentation, API consolidation, and additional
examples, not a missing theorem in the confidence-formula characterization.

\appendix

\section{Degrees-of-freedom zoo}\label{app:dof-zoo}

This appendix lists the degrees of freedom that should remain visible in
PLN truth-value practice. Defaults are useful; unnecessary fixation is not.
Each entry names the freedom, the formal gate that can collapse it, and why a
real system may want to keep the knob available.

\begin{description}
\item[Evidence scale / prior strength.]
      In the evidence-count view, confidence is commonly
      $c = N/(N+\kappa)$, where $N$ is total retained evidence and
      $\kappa > 0$ is the effective prior or context size. The Walley
      width-complement bridge forces the odds form once the scale is chosen,
      but it does not choose the scale. A curated clinical fact, a noisy
      high-throughput association, and an indirect pathway-mediated
      hypothesis should not usually accumulate confidence at the same rate.
      Positive example: a hand-curated contraindication can use small
      $\kappa$. Negative example: a single weak omics hit should not receive
      the same confidence merely because it has the same positive/negative
      ratio.

\item[Strength projection / prior context.]
      The same positive and negative counts can be read by maximum
      likelihood, by a uniform Beta posterior mean, by Jeffreys' prior, or by
      another explicit context. The formal tower records this as a projection
      choice rather than pretending that ``strength'' is context-free.
      Positive example: sparse medical counts may need shrinkage toward a
      population baseline. Negative example: a mature benchmark with thousands
      of observations can safely use a near-MLE projection.

\item[Evidence total versus evidence ratio.]
      Strength reads the positive/negative ratio; confidence reads the total
      amount of evidence. Thus $1/1$ and $100/100$ have the same strength but
      should not have the same confidence. Conversely, $1/1$ and $2/0$ can
      have the same total evidence and different strengths. This is the basic
      reason the two coordinates are separate.

\item[Confidence chart.]
      Count reconstruction only requires an invertible chart on total
      evidence; it does not by itself force the PLN odds chart. The
      width-complement bridge is the extra semantic law that narrows the
      chart to the PLN/NARS odds coordinate. Positive example: user-facing
      risk displays may want a compressed visual confidence scale. Negative
      example: revision must operate in evidence-weight space, not in an
      arbitrary display chart.

\item[Width-complement convention.]
      For unit gambles, the convention $c = 1 - (U-L)$ makes confidence the
      complement of interval width. For a general bounded gamble with range
      $[a,b]$, the normalized version is
      $c = 1 - (U-L)/(b-a)$. This convention is natural and minimal, but it is
      still a convention. Positive example: binary prediction has a canonical
      unit scale. Negative example: dollar-valued utilities must be
      range-normalized before width-complement confidence is meaningful.

\item[Interval point selector.]
      A nontrivial interval $[L,U]$ does not force a single displayed point:
      lower, midpoint, upper, or a risk-adjusted selector can all be
      reasonable. The paper uses midpoint when it wants a neutral PLN
      strength display. Positive example: safety-critical planning may use
      the lower envelope. Negative example: exploratory ranking may use the
      midpoint with the interval width shown separately.

\item[Credal shape beyond one interval.]
      A single query interval loses the shape of the credal set. Two credal
      sets can agree on $L$ and $U$ for one gamble but differ on other
      gambles, on correlations, or under composition. Positive example: two
      drug hypotheses can share the same marginal response interval but differ
      in adverse-event dependence. Negative example: reporting only an STV
      for each marginal can hide incompatible joint structure.

\item[Backend for uncertainty.]
      Bayesian credible intervals, Walley IDM predictive intervals, finite
      lower envelopes, projective credal envelopes, and DLR completion
      envelopes all produce interval-like readouts but carry different
      provenance. Positive example: a posterior credible interval is right
      for a parametric lab assay. Negative example: a robust safety envelope
      should not be collapsed into a single Bayesian posterior merely for
      convenience.

\item[Credible level / tolerance.]
      Bayesian credible intervals carry a chosen level, e.g. 90\%, 95\%, or
      99\%. This is not the same knob as $\kappa$: $\kappa$ governs evidence
      accumulation; credible level governs the interval requested from a
      posterior. Positive example: exploratory dashboards can show 90\%
      intervals. Negative example: clinical decision support may require 99\%
      or a one-sided lower bound.

\item[Categorical query choice.]
      In multinomial or categorical evidence, total concentration can be
      shared while each queried category has its own lower, upper, width, and
      strength. Positive example: a phenotype classifier may have high total
      evidence but low confidence for a rare subclass. Negative example:
      forcing one binary query can erase category-specific uncertainty.

\item[Feature/evidence discounting.]
      Evidence can be reweighted by source quality, feature informativeness,
      IDF-like rarity, route length, recency, or dependence corrections before
      it enters the count tower. Positive example: independent curated
      replications should count more than near-duplicate text-mined hits.
      Negative example: treating correlated pathway hits as independent
      overstates confidence.

\item[Local credal specification.]
      In projective credal models, the modeller chooses local credal sets at
      finite windows. This is the intended imprecision knob: how much local
      ambiguity the model admits before global completion. Positive example:
      sensor calibration uncertainty belongs in local credal sets. Negative
      example: hiding local ambiguity by picking a single precise prior can
      make later confidence appear unjustifiably high.

\item[Compatible completion / natural extension.]
      A family of finite-window lower previsions must be globally compatible
      before it gives a raw all-gambles process law. F2 shows that the
      exchangeable external readout is unconditional, while the raw crown is
      exactly the lower-prevision compatibility boundary. Positive example:
      finite-prefix forecasts can be useful before full natural extension is
      available. Negative example: calling those forecasts a full process law
      without compatibility is assumption laundering.

\item[Completion-level extreme point.]
      Given a fixed credal set, the world may instantiate one precise
      completion. This is real aleatoric freedom, but it is invisible to a
      truth value because $\STV{s}{c}$ is a property of the whole credal set.
      Positive example: scenario simulation samples an extreme completion.
      Negative example: inference rules should not pretend to know which
      extreme completion the world selected.

\item[Phase-transition / infinite boundary freedom.]
      Infinite local specifications can have multiple DLR completions. F3
      pins a collapse regime by Dobrushin uniqueness; F4 pins a strict-width
      regime by low-temperature Ising separation. Positive example: weakly
      coupled large sensor networks can have precise local-query envelopes.
      Negative example: strongly coupled spatial contagion or image domains
      can keep interval width because distant boundary conditions matter.

\item[Exchangeability / de~Finetti family freedom.]
      In the IID lane, a mixing family can have a canonical external
      sigma-additive readout while the raw all-gambles crown depends on
      lower-prevision compatibility. Pointwise zero-interior mass is
      sufficient in some kernel cases but not necessary at the family level.
      Positive example: a family with one compatible conservative member can
      support the raw crown. Negative example: requiring every member to have
      zero interior mass rejects legitimate imprecise families.

\item[Rule-level confidence transport.]
      PLN rule formulas must decide whether confidence is combined by
      addition in weight space, minimum in weight space, multiplication, or a
      rule-specific effective evidence term. Positive example: revision adds
      independent evidence weights. Negative example: feeding a raw confidence
      through \lid{w2c} as if it were already a weight double-damps the result.

\item[Provenance and route semantics.]
      Two derivations with the same final STV may have different provenance:
      direct observation, pathway transfer, abduction, or long inference
      chain. Positive example: biomedical drug support should distinguish
      direct disease evidence from pathway-mediated support. Negative example:
      merging all routes before preserving provenance makes audit and
      discounting impossible.

\item[Extra structure beyond classical STV.]
      Standard STV forgets structures such as phase, dependence witnesses, or
      higher-order distributional shape. The formal amplitude/phase boundary
      shows that such information can be invisible to the classical strength
      and confidence view. Positive example: interference-like composition
      requires extra structure. Negative example: carrying phase in an
      ordinary binary classifier may be needless complexity.
\end{description}

\section{Example coverage plan for PeTTa/CeTTa}\label{app:examples}

These examples accompany the paper as a runnable companion package,
\lpath{papers/pln-confidence-examples/} (build and run via \lpath{run_all.sh}
on CeTTa(HE) and PeTTa; \lpath{DOF-INDEX.md} is the full map). Each degree of
freedom falls in one of three honesty tiers: \emph{Runnable} (a real executable
canary with a positive and a negative case), \emph{Structural} (a finite shadow
of an infinite or joint fact --- the Lean theorem is the real claim), and
\emph{Lean-ref} (the witness is a theorem; no standalone \texttt{.metta}).
Thirteen runnable canaries (DoF 1--8, 10, 11, 13, 17, 18) are portable across
CeTTa(HE) and PeTTa; the three new Lean canaries (DoF 8, 11, 18) live in
\lpath{Mettapedia/lean/mettapedia/Mettapedia/Logic/PLNConfidenceDoFCanary.lean}
and are axiom-clean. DoF 14 deliberately points to the existing strict-width /
non-determination theorem rather than inventing a misleading ``extreme point''
executable. The table
below records each row's current artifact and what that artifact demonstrates;
for Lean-ref rows, it names the example story that would illustrate the theorem.

\begin{longtable}{p{0.22\linewidth}p{0.28\linewidth}p{0.36\linewidth}}
\toprule
Freedom & Current artifact & Demonstration \\
\midrule
\endhead
Evidence total versus ratio &
Raven example; \lpath{metta/dof03_ratio_vs_total.metta} &
Shows the raven asymmetry pattern and then isolates the ratio/total split:
$5/5$, $50/50$, and $500/500$ have the same strength but rising confidence. \\

Evidence scale $\kappa$ &
\lpath{metta/dof01_kappa_calibration.metta} &
Shows the same evidence count under $\kappa=1,10,50$, matching the intuition
that curated clinical, noisy omics, and pathway-mediated labels may need
different evidence scales. \\

Strength prior context &
\lpath{metta/dof02_strength_prior_contexts.metta};
Lean canaries in \lid{PLNTruthTower} &
Shows MLE, uniform-prior, and Jeffreys-prior strengths from the same sparse
counts, so the prior context is visible rather than silently fixed. \\

Confidence chart freedom &
\lpath{metta/dof04_confidence_chart.metta};
Lean chart-separation theorem &
Compares PLN odds, reserve-half, and smooth arctangent charts on the same
decoded evidence, demonstrating chart freedom without changing the evidence. \\

Revision as evidence addition &
\lpath{metta/dof17_revision_transport.metta} &
Shows \lid{Truth\_Revision} by adding decoded evidence weights and re-encoding
the result, while contrasting it with separate prior choices. \\

Rule-level confidence transport &
Lean transport analysis &
The Lean analysis records the weight-space transport discipline; a PeTTa
regression can show the old raw-min double-damping bug as an executable shadow. \\

Interval point selector &
\lpath{metta/dof06_interval_selector.metta};
Lean generic ITV selector canary &
Shows lower, midpoint, and upper selectors producing different displayed
strengths on the same interval. \\

Width-complement range normalization &
\lpath{metta/dof05_width_complement_range.metta};
paper convention &
Shows binary width as $1-(U-L)$ and bounded-utility width as
$1-(U-L)/(b-a)$, including the negative example where raw dollar width is
misleading. \\

Credal shape beyond one interval &
\lpath{metta/dof07_credal_shape.metta};
Lean credal projection tower canaries &
Shows two finite credal shapes with the same marginal interval and different
joint/correlation behavior. \\

Backend choice &
\lpath{metta/dof08_backend_beta_vs_idm.metta};
\lid{beta\_mean\_strictly\_inside\_walley\_idm\_interval} &
Compares a Bayesian beta posterior point with the Walley IDM interval from the
same counts, keeping backend provenance explicit. \\

Categorical query choice &
\lpath{metta/dof10_categorical_query.metta};
Lean categorical evidence tower &
Shows a three-category classifier where total evidence is high but a rare
category still has low confidence. \\

Feature/evidence discounting &
\lpath{metta/dof11_evidence_discounting.metta} &
Shows evidence-power discounting and amplification, matching the Lean theorem
that the log-odds scale linearly with the evidence power. \\

Biomedical route provenance &
\lpath{metta/dof18_route_provenance.metta} &
Shows direct, pathway-mediated, and weak-literature routes with the same
strength and route-specific confidence scales. \\

Exchangeable de~Finetti F2 &
\lpath{metta/dof13_f2_finite_shadow.metta};
Lean F2 crown &
Uses a finite-prefix shadow to show the always-available external readout,
while the raw all-gambles crown remains the Lean compatibility theorem. \\

Dobrushin F3 collapse &
Lean F3 crown &
The intended finite story is a weakly coupled sensor grid: all finite-query
completions agree under the Dobrushin budget, so the interval collapses. \\

Ising F4 strict width &
Lean F4 crown &
The intended finite story is spatial contagion or image segmentation:
low-temperature coupling and boundary conditions keep the origin query
interval nontrivial. \\

Completion-level extreme point &
\lid{boundedMeasurableCredalSetHasStrictWidth} and
\lid{boundedMeasurableCredalSet\_not\_determines\_of\_strictWidth} &
The theorem-level layer says two admissible precise completions can disagree
and therefore no single determined readout is justified. \\

Extra hidden structure &
Amplitude/phase Lean boundary &
The Lean boundary marks hidden phase/amplitude structure that shares classical
STV but can differ under a composition rule that sees the hidden coordinate. \\
\bottomrule
\end{longtable}

\bibliographystyle{plain}
\begin{thebibliography}{10}

\bibitem{walley91}
Peter Walley.
\newblock \emph{Statistical Reasoning with Imprecise Probabilities}.
\newblock Chapman \& Hall, 1991.

\bibitem{definetti37}
Bruno de Finetti.
\newblock La pr\'evision: ses lois logiques, ses sources subjectives.
\newblock \emph{Annales de l'Institut Henri Poincar\'e}, 7:1--68, 1937.

\bibitem{dobrushin68}
R.\ L.\ Dobrushin.
\newblock Description of a random field by means of conditional
        probabilities and conditions of its regularity.
\newblock \emph{Theory of Probability and its Applications},
        13(2):197--224, 1968.

\bibitem{lanford-ruelle69}
O.\ Lanford and D.\ Ruelle.
\newblock Observables at infinity and states with short-range correlations
        in statistical mechanics.
\newblock \emph{Communications in Mathematical Physics}, 13:194--215, 1969.

\bibitem{richardson-domingos06}
Matthew Richardson and Pedro Domingos.
\newblock Markov logic networks.
\newblock \emph{Machine Learning}, 62:107--136, 2006.

\bibitem{singla-domingos05}
Parag Singla and Pedro Domingos.
\newblock Discriminative training of Markov logic networks.
\newblock In \emph{AAAI 2005}.

\bibitem{goertzel-pln}
Ben Goertzel, Matthew Ikl\'e, Izabela Goertzel, and Ari Heljakka.
\newblock \emph{Probabilistic Logic Networks: A Comprehensive Framework
        for Uncertain Inference}.
\newblock Springer, 2008.

\bibitem{xipln}
Mettapedia Working Notes.
\newblock $\xi$PLN: a correct and complete foundation for probabilistic
        logic networks.
\newblock In the Mettapedia corpus: \lpath{papers/xiPLN.tex}.

\bibitem{wmpln-book}
Mettapedia Working Notes.
\newblock The WM-PLN book.
\newblock In the Mettapedia corpus: \lpath{papers/wm-pln-book.tex}.

\bibitem{pln-review}
Mettapedia Working Notes.
\newblock PLN review and corrections.
\newblock In the Mettapedia corpus: \lpath{papers/pln-review.tex}.

\end{thebibliography}

\end{document}
