\documentclass[11pt]{article}

\usepackage[margin=1in]{geometry}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{lmodern}
\usepackage{microtype}
\usepackage{amsmath,amssymb,amsthm}
\usepackage{mathtools}
\usepackage{hyperref}
\usepackage{booktabs}
\usepackage{enumitem}
\usepackage{listings}
\usepackage{xcolor}

\hypersetup{
  colorlinks=true,
  linkcolor=blue!50!black,
  citecolor=blue!50!black,
  urlcolor=blue!50!black
}

\lstdefinelanguage{Lean4}{
  morekeywords={
    import,namespace,end,open,section,noncomputable,abbrev,
    def,theorem,lemma,where,by,fun,let,in,have,show,exact,
    intro,apply,simpa,calc
  },
  sensitive=true,
  morecomment=[l]{--},
  morecomment=[s]{/-}{-/},
  morestring=[b]"
}

\lstset{
  language=Lean4,
  basicstyle=\ttfamily\small,
  keywordstyle=\bfseries\color{blue!50!black},
  commentstyle=\itshape\color{green!40!black},
  stringstyle=\color{red!50!black},
  showstringspaces=false,
  breaklines=true,
  frame=single,
  framerule=0.3pt,
  xleftmargin=0.5em,
  xrightmargin=0.5em,
  keepspaces=true,
  columns=fullflexible,
  literate=
    {¬}{{$\neg$}}1
    {∃}{{$\exists$}}1
    {→}{{$\to$}}1
    {≤}{{$\leq$}}1
}

\newtheorem{theorem}{Theorem}[section]
\newtheorem{lemma}[theorem]{Lemma}
\theoremstyle{remark}
\newtheorem{remark}[theorem]{Remark}

\title{P vs NP Grassroots:\\
Ground-Up Background Theory for a Structured Switching Route}
\author{}
\date{\today}

\begin{document}
\maketitle

\begin{abstract}
This note records the \emph{grassroots} side of the current Lean work around a
proposed route to $P \neq NP$.  It is intentionally not the blocker note.
The companion crux-first analysis lives separately in \texttt{pnp\_crux.tex}.

The goal here is narrower and more constructive: build reusable background theory
from the ground up, using mathlib and existing Mettapedia infrastructure wherever
possible, so that if the high-level proof strategy can be repaired, it lands in a
clean machine-checked framework rather than in ad hoc prose.

The current grassroots state now includes explicit encoded hypothesis classes,
finite empirical-risk minimization, finite deceptive-sample bounds, threshold
recovery lemmas, a rational uniform-rate layer, and explicit uniform success
bounds for exact ERM recovery under finite sampling.  It now also includes the
first weighted finite-sampling steps beyond uniform draws: an exact
one-predictor agreement law and a family-level weighted union bound over finite
encoded classes, together with weighted exact-recovery lower bounds for ERM.
It also now packages the exact \emph{same-route interface}: a repair that still
intends to use switching plus ERM must produce a bit-encoded family over an
available input surface and therefore inherits the current finite weighted
recovery theorem.
It also now isolates the exact post-switch visible surface and proves a clean
finite-summary kernel: any switched family that truly factors through a finite
summary surface automatically inherits an explicit truth-table bit budget equal
to the cardinality of that summary type.
It now also ties the four current exact-surface affine candidate classes
directly to the weighted recovery backbone: once the target lies in one of
those classes, the finite-sampling lower bound is available immediately with the
matching explicit bit budget.
It now also isolates a sharper shared-basis regime: if one fixed affine feature
extractor is reused across the switched family and only the downstream combiner
varies, then the code budget collapses to the combiner alone.
The Meredith-side limit is now considerably sharper.  The currently formalized
weakness bridge exports only global evidence scalars, so any predictor factoring
only through those summaries is constant on the whole state space.  The current
exact HML layer is infinitary by default: if the certified minimal-context type
is infinite, then even the depth-1 fragment is infinite, and the concrete rho
instance already realizes that case.  And one layer below exact HML, the
concrete rho present-moment surface already realizes arbitrary finite
membership patterns on a natural probe slice: a single-output probe state
against a flat bag of input guards records exactly which channels are present in
the environment.
On the semantic side, it now includes a depth-bounded GSLT/HML observational
kernel, a bridge from bisimulation quotients to quantale weakness, and a first
concrete minimal-context instance for the rho calculus.  This isolates the exact
positive shape that a successful switching theorem would need: not mere
locality, but locality together with a concrete compression theorem, a clean
finite learning interface, and an explicit observational/evidence semantics.
\end{abstract}

\tableofcontents

\section{Scope}

The October 2025 manuscript
\emph{``P != NP: A Non-Relativizing Proof via Quantale Weakness and Geometric
Complexity''} proposes a route from a structured Unique-SAT ensemble to a
distributional lower bound, and then to a contradiction with the upper bound that
would follow from $P=NP$.

This document does \emph{not} ask where that route fails.  That is the job of the
companion crux note.  Instead it asks a different question:

\begin{quote}
What background theory would we want in Lean even if the final proof still needs
repair?
\end{quote}

That question leads to a program with three design constraints:
\begin{enumerate}[leftmargin=1.5em]
  \item Prefer general infrastructure over one-off rebuttal lemmas.
  \item Reuse mathlib computability and existing Mettapedia semantics rather than
  inventing parallel foundations.
  \item State optimistic bridge theorems in a way that makes missing hypotheses
  explicit, so later fixes can slot into them cleanly.
\end{enumerate}

\section{Existing Lean Base}

Several ingredients already exist in the local formalization and are useful for a
ground-up PNP effort.

\begin{center}
\begin{tabular}{ll}
\toprule
\textbf{Module} & \textbf{Grassroots role} \\
\midrule
\texttt{Mathlib.Computability.Partrec} & partial recursive computation infrastructure \\
\texttt{KolmogorovComplexity/Basic.lean} & universal algorithm and plain complexity \\
\texttt{Hyperseed/Ultrainfinitism.lean} & observer-relative available-region semantics \\
\texttt{Hyperseed/Basic.lean} & closure/cascade interfaces on accessible observations \\
\texttt{GSLT/Logic/LogicalMetric.lean} & finite-depth observational distinction kernel \\
\texttt{GSLT/Meredith/WeaknessBridge.lean} & operational distinguishability $\to$ weakness/evidence \\
\texttt{GSLT/Meredith/RhoMinimalContext.lean} & first concrete minimal-context / HML anchor \\
\texttt{PNP/HypothesisClass.lean} & encoded classifier families and code-budget bounds \\
\texttt{PNP/FiniteERM.lean} & empirical-risk minimization kernel for finite encoded families \\
\texttt{PNP/FiniteConsistencyBound.lean} & finite uniform consistency bound for deceptive samples \\
\texttt{PNP/FiniteUniformRecovery.lean} & threshold corollary guaranteeing exact recovery on some sample \\
\texttt{PNP/FiniteUniformRate.lean} & rational uniform-rate bounds for deception and exact recovery \\
\texttt{PNP/FiniteUniformSuccess.lean} & miss-ratio decay and explicit uniform success bounds \\
\texttt{PNP/FiniteIIDAgreement.lean} & exact weighted one-predictor agreement mass under a finite PMF \\
\texttt{PNP/FiniteIIDFamilyBound.lean} & weighted deceptive-mass union bound for finite encoded families \\
\texttt{PNP/FiniteIIDRecovery.lean} & weighted exact ERM recovery lower bounds under finite PMFs \\
\texttt{PNP/SameRouteInterface.lean} & exact certificate for when a repair still counts as the same switching/ERM route \\
\texttt{PNP/ExactSwitchedFamily.lean} & exact post-switch surface plus finite-summary compression kernel \\
\texttt{PNP/ABVisibleSurface.lean} & reduced raw visible surface $(a,b)$ and its explicit cardinality \\
\texttt{PNP/AffineColumnFamily.lean} & concrete affine GF(2)-style column family with $k+1$ bit budget \\
\texttt{PNP/AffineFeatureFamily.lean} & bounded affine-feature family with budget $r(k+1)+2^r$ \\
\texttt{PNP/SparseThresholdAffineFamily.lean} & sparse-threshold affine-feature family with linear budget $r(k+3)$ \\
\texttt{PNP/AffineDecisionListFamily.lean} & affine decision-list family with linear budget $r(k+2)+1$ \\
\texttt{PNP/ExactAffineRecovery.lean} & exact-surface weighted recovery bounds for the current affine candidate classes \\
\texttt{PNP/SharedAffineFeatureFamilies.lean} & shared-basis exact-surface families with combiner-only budgets and recovery bounds \\
\texttt{PNP/ExactABDecisionListFamily.lean} & concrete raw exact-surface decision-list family on $(a,b)$ with budget $2k+1$ \\
\texttt{PNP/ABDecisionListRoute.lean} & pullback certificate from reduced $(a,b)$ decision lists to exact-surface compression/recovery \\
\texttt{PNP/ExactABAffineFamilies.lean} & affine-feature, sparse-threshold, and decision-list ladders on the raw $(a,b)$ surface \\
\texttt{PNP/SharedExactABFeatureFamilies.lean} & shared-basis affine ladders on the raw $(a,b)$ surface with combiner-only budgets \\
\texttt{PNP/SharedABFeatureRoutes.lean} & quotient-route transfer from reduced $(a,b)$ families to shared-basis exact-surface budgets \\
\texttt{PNP/SharedABTargetInterface.lean} & unified target data packages for the shared raw $(a,b)$ route \\
\texttt{PNP/CanonicalABTargetRoute.lean} & canonical-coordinate specialization reducing the target interface back to raw $(a,b)$ decision lists \\
\texttt{PNP/CanonicalABCandidateInterface.lean} & single candidate-data package for the canonical raw $(a,b)$ decision-list route, with recovery consequence \\
\texttt{PNP/CanonicalABCodeWitness.lean} & explicit code-witness reduction: the route follows automatically from one raw $(a,b)$ decision-list code assignment per index \\
\texttt{PNP/CanonicalABRecoveryInterface.lean} & final generic interface: explicit code assignment plus agreement bound imply compression and recovery \\
\texttt{PNP/BitFamilyERM.lean} & ERM transport layer: one fixed local class yields one selected code per sample/index \\
\texttt{PNP/ExactZABDecisionListFamily.lean} & exact-surface decision-list family on $(zfeat(z),a,b)$ with budget $r+2k+1$ \\
\texttt{PNP/ExactZABTargetInterface.lean} & target and recovery interfaces for the shared $(zfeat(z),a,b)$ decision-list route \\
\texttt{PNP/ExactZABERMRoute.lean} & ERM specialization: the selected code itself witnesses the shared $(zfeat(z),a,b)$ route \\
\texttt{PNP/SharedExactZABFeatureFamilies.lean} & shared-basis affine ladders on $(zfeat(z),a,b)$ with combiner-only budgets \\
\texttt{PNP/SharedExactZABTargetInterface.lean} & unified target and recovery interfaces for the shared-basis $(zfeat(z),a,b)$ route \\
\texttt{PNP/SharedExactZABERMRoute.lean} & ERM specialization of the shared-basis $(zfeat(z),a,b)$ route with combiner-only budget \\
\texttt{PNP/CanonicalZABTargetRoute.lean} & canonical-coordinate specialization reducing the shared-basis exact route back to raw $(zfeat(z),a,b)$ decision lists \\
\texttt{PNP/CanonicalZABCandidateInterface.lean} & concrete candidate package for the direct exact $(zfeat(z),a,b)$ route, with recovery consequence \\
\texttt{PNP/CanonicalZABCodeWitness.lean} & explicit code-witness reduction: one exact $(zfeat(z),a,b)$ decision-list code assignment per index suffices \\
\texttt{PNP/CanonicalZABRecoveryInterface.lean} & final direct exact-route interface: code assignment plus agreement bound imply compression and recovery \\
\texttt{PNP/CanonicalZABERMRoute.lean} & ERM-to-canonical bridge: the exact ERM wrapper itself supplies the canonical code witness and recovery interface \\
\texttt{PNP/CanonicalZABERMInterface.lean} & final manuscript-facing ERM interface: samples, equality to the ERM wrapper, and an agreement bound imply compression and recovery \\
\texttt{PNP/BitVecZABVisibleSurface.lean} & when $z$ is already bit-valued, the full exact visible surface $(z,a,b)$ becomes one raw bitvector of length $r+2k$ \\
\texttt{PNP/BitVecZABERMInterface.lean} & concrete identity-extractor specialization of the final exact ERM interface on the raw full visible bit surface \\
\texttt{PNP/ProjectedZABERMInterface.lean} & coordinate-projection specialization: the extractor burden is reduced to choosing finitely many coordinates of a bit-valued $z$ \\
\texttt{PNP/GlobalWeaknessObstruction.lean} & current weakness summaries alone cannot yield a nonconstant predictor surface \\
\texttt{PNP/InfinitaryHMLObstruction.lean} & infinite minimal contexts force an infinite depth-1 HML observable surface \\
\texttt{PNP/PresentMomentShattering.lean} & exact rho present-moment probes already realize arbitrary finite membership slices \\
\bottomrule
\end{tabular}
\end{center}

The point of the table is simple: the grassroots route is not starting from zero.
We already have a serious computability substrate, an observer-relative semantics
layer, and a new optimistic interface for small hypothesis classes.

\section{Encoded Hypothesis Families}

The key new grassroots abstraction is that a hypothesis class should be represented
by a finite code space together with a decoder.

\begin{quote}
code budget first, realized class second.
\end{quote}

Formally, \texttt{HypothesisClass.lean} introduces an \emph{encoded family}
\[
  \mathrm{decode} : \mathrm{Code} \to (\mathrm{Input} \to \mathrm{Output}),
\]
where \(\mathrm{Code}\) is finite.  The realized hypothesis class is the range of
\(\mathrm{decode}\).

\begin{theorem}[Realized class is bounded by code space]
For any encoded family \(H\),
\[
  |\mathrm{realized}(H)| \le |\mathrm{Code}(H)|.
\]
\end{theorem}

\noindent
This is the theorem \texttt{EncodedFamily.card\_realized\_le} in Lean.

The specialization relevant to the present route is the bit-coded case.

\begin{theorem}[Bit-budget bound]
If a family of Boolean classifiers is decoded from \(s\) bits, then it realizes at
most \(2^s\) distinct classifiers.
\end{theorem}

\noindent
This is formalized as
\texttt{BitEncodedClassifierFamily.card\_realized\_le}.

This theorem is elementary, but it captures the exact shape of the optimistic
repair that would be needed later.  A switching theorem cannot stop at ``the new
predictor is local.''  It must continue to a statement of the form:

\begin{quote}
the switched predictors lie in a family decoded from \(\mathrm{polylog}(m)\) bits.
\end{quote}

Only then does finite-class learning theory become genuinely available.

\section{Finite-Class ERM Kernel}

The next honest step after a finite encoded family is not yet a PAC theorem.
It is the finite combinatorial kernel of empirical risk minimization.

The new module \texttt{FiniteERM.lean} defines labeled samples as finite lists of
pairs \((x,y)\), empirical error as the number of misclassified examples in the
sample, and exact sample consistency as zero empirical error.

\begin{theorem}[Finite ERM existence]
Let \(H\) be a nonempty encoded family.  For every finite labeled sample \(S\),
there exists a realized predictor \(h^\star \in \mathrm{realized}(H)\) such that
\[
  \mathrm{err}_S(h^\star) \le \mathrm{err}_S(g)
  \qquad
  \text{for all } g \in \mathrm{realized}(H).
\]
\end{theorem}

\noindent
This is formalized as
\texttt{EncodedFamily.exists\_realized\_empiricalRiskMinimizer}.

\begin{theorem}[Zero-error transfer through ERM]
If some code in the family fits the sample exactly, then the chosen ERM
predictor also fits the sample exactly.
\end{theorem}

\noindent
This is formalized as
\texttt{EncodedFamily.empiricalRiskPredictor\_fitsSample\_of\_exists\_code\_fits}.

This is still modest, but it matters.  Combined with the code-budget theorem
from \texttt{HypothesisClass.lean}, it gives the first clean bridge from
``small encoded family'' to an actual learning-theoretic construction.

\section{Finite Uniform Consistency Bound}

The next step now formalized is a fully combinatorial finite-domain bound.  We
do not yet assume an arbitrary distribution or invoke measure theory.  Instead,
we count how many length-\(m\) input samples can make a wrong predictor look
perfectly consistent with the target labels.

\begin{theorem}[Wrong predictors fit only exponentially few samples]
Let \(X\) be a finite input domain.  If \(h \neq f\), then the number of
length-\(m\) samples on which \(h\) agrees with the target \(f\) at every sampled
point is at most
\[
  (|X|-1)^m.
\]
\end{theorem}

\noindent
This is formalized as
\texttt{card\_consistentSamples\_le\_of\_exists\_disagreement} in
\texttt{FiniteConsistencyBound.lean}.

\begin{theorem}[Finite encoded families have few deceptive samples]
Let \(H\) be an encoded family over a finite input domain \(X\).  The number of
length-\(m\) samples on which some wrong code in \(H\) still matches all target
labels is at most
\[
  |\mathrm{Code}(H)| \cdot (|X|-1)^m.
\]
\end{theorem}

\noindent
This is formalized as
\texttt{EncodedFamily.card\_deceptiveSamples\_le}.

\begin{theorem}[ERM recovers the target off the deceptive set]
If the target predictor is realized by the encoded family, then on every
non-deceptive sample the chosen ERM predictor equals the target predictor
exactly.
\end{theorem}

\noindent
This is formalized as
\texttt{EncodedFamily.empiricalRiskPredictor\_eq\_target\_of\_not\_deceptive}.

This is not yet the full finite-class generalization theorem one would state in
terms of \(\log |H|\) and an arbitrary data distribution.  But it is already the
right union-bound skeleton in a completely concrete finite setting.

\section{Finite Uniform Recovery Threshold}

The next corollary is the cleanest possible extraction from the deceptive-sample
bound.  If the deceptive set is strictly smaller than the whole length-\(m\)
sample space, then at least one non-deceptive sample exists.

\begin{theorem}[Threshold for existence of a good sample]
Let \(H\) be an encoded family over a finite input domain \(X\).  If
\[
  |\mathrm{Code}(H)| \cdot (|X|-1)^m < |X|^m,
\]
then there exists a length-\(m\) sample outside the deceptive set.
\end{theorem}

\noindent
This is formalized as
\texttt{EncodedFamily.exists\_nondeceptiveSample\_of\_bound\_lt} in
\texttt{FiniteUniformRecovery.lean}.

\begin{theorem}[Threshold corollary for exact ERM recovery]
If the target predictor is realized by the encoded family and
\[
  |\mathrm{Code}(H)| \cdot (|X|-1)^m < |X|^m,
\]
then there exists a length-\(m\) sample on which the chosen ERM predictor equals
the target predictor exactly.
\end{theorem}

\noindent
This is formalized as
\texttt{EncodedFamily.exists\_sample\_empiricalRiskPredictor\_eq\_target\_of\_bound\_lt}.

This is still finite and non-asymptotic, but it clarifies the next job.  To move
from existence to a real learning theorem, we now need a probability interface
that turns the combinatorial ratio behind the threshold into an explicit success
bound under sampling assumptions.

\section{Finite Uniform Rate Layer}

The next natural refinement is still finite, but no longer purely existential.
Instead of asking only whether some good sample exists, we measure how large the
good and bad regions are inside the finite sample space.

The module \texttt{FiniteUniformRate.lean} defines rational uniform densities on
the point-sample space:
\begin{enumerate}[leftmargin=1.5em]
  \item the deceptive rate,
  \item the non-deceptive rate,
  \item the exact-recovery rate for ERM.
\end{enumerate}

\begin{theorem}[Uniform deceptive-rate bound]
Let \(H\) be an encoded family over a nonempty finite input domain \(X\).  Then
the uniform fraction of length-\(m\) samples that are deceptive is at most
\[
  \frac{|\mathrm{Code}(H)| \cdot (|X|-1)^m}{|X|^m}.
\]
\end{theorem}

\noindent
This is formalized as
\texttt{EncodedFamily.deceptiveRate\_le\_bound} in
\texttt{FiniteUniformRate.lean}.

\begin{theorem}[Exact recovery dominates the non-deceptive region]
If the target predictor is realized by the encoded family, then the uniform exact
recovery rate of the chosen ERM predictor is at least
\[
  1 - \mathrm{rate}_{\mathrm{deceptive}}.
\]
\end{theorem}

\noindent
This is formalized as
\texttt{EncodedFamily.exactRecoveryRate\_ge\_one\_sub\_deceptiveRate}.

\begin{theorem}[Positive exact recovery rate under the threshold]
If the target predictor is realized and
\[
  |\mathrm{Code}(H)| \cdot (|X|-1)^m < |X|^m,
\]
then the exact recovery rate is strictly positive.
\end{theorem}

\noindent
This is formalized as
\texttt{EncodedFamily.exactRecoveryRate\_pos\_of\_bound\_lt}.

This is still not a full PAC-style theorem under arbitrary distributions.  But it
does close the uniform finite-sampling gap between ``some good sample exists'' and
``good samples occupy an explicitly bounded fraction of the sample space.''

\section{Finite Uniform Success Bounds}

The next cleanup step is to express the finite uniform rate bound using the
single-step miss ratio
\[
  \rho_X := \frac{|X|-1}{|X|}.
\]
For any nonempty finite input domain, \(\rho_X < 1\), so the finite deceptive
rate decays exponentially in the sample length \(m\).

The module \texttt{FiniteUniformSuccess.lean} packages the previous rate theorem
into this cleaner interface.

\begin{theorem}[Miss-ratio deceptive bound]
Let \(H\) be an encoded family over a nonempty finite input domain \(X\).  Then
\[
  \mathrm{rate}_{\mathrm{deceptive}}
  \le
  |\mathrm{Code}(H)| \cdot \rho_X^m.
\]
\end{theorem}

\noindent
This is formalized as
\texttt{EncodedFamily.deceptiveRate\_le\_codeMul\_uniformMissRatio\_pow}.

\begin{theorem}[Uniform exact recovery lower bound]
If the target predictor is realized by the encoded family, then
\[
  \mathrm{rate}_{\mathrm{exact\ recovery}}
  \ge
  1 - |\mathrm{Code}(H)| \cdot \rho_X^m.
\]
\end{theorem}

\noindent
This is formalized as
\texttt{EncodedFamily.exactRecoveryRate\_ge\_one\_sub\_codeMul\_uniformMissRatio\_pow}.

\begin{theorem}[\(\varepsilon\)-style corollary]
If
\[
  |\mathrm{Code}(H)| \cdot \rho_X^m \le \varepsilon,
\]
then the exact recovery rate is at least \(1 - \varepsilon\).
\end{theorem}

\noindent
This is formalized as
\texttt{EncodedFamily.exactRecoveryRate\_ge\_one\_sub\_of\_codeMul\_uniformMissRatio\_pow\_le}.

This is still uniform finite-sample learning rather than full distributional
generalization.  But it is the first version of the grassroots story with an
explicit success lower bound of the form ``one minus an exponentially decaying
failure term.''

\section{Finite Weighted Agreement and Family Bounds}

The next honest move is to relax the uniform-sampling assumption while still
staying fully finite.  Instead of sampling uniformly from a finite input domain,
we let one finite probability mass function \(\mu\) weight the inputs.

The module \texttt{FiniteIIDAgreement.lean} handles the one-predictor case.  For
one target \(f\) and one predictor \(h\), it defines:
\begin{enumerate}[leftmargin=1.5em]
  \item the one-step agreement mass of \(h\) with \(f\) under \(\mu\),
  \item the product mass of a length-\(m\) sample under independent draws from \(\mu\),
  \item the total mass of samples on which \(h\) agrees with \(f\) everywhere.
\end{enumerate}

\begin{theorem}[Exact weighted agreement law]
For any finite input type and finite PMF \(\mu\),
\[
  \mathrm{mass}_{\mu}(\text{all-$m$-point samples consistent with } h)
  =
  \bigl(\mathrm{agreementMass}_{\mu}(f,h)\bigr)^m.
\]
\end{theorem}

\noindent
This is formalized as
\texttt{consistentSampleMass\_eq\_agreementMass\_pow}.

\begin{theorem}[Weighted decay corollary]
If the one-step agreement mass is at most \(q\), then the total mass of
length-\(m\) consistent samples is at most \(q^m\).
\end{theorem}

\noindent
This is formalized as
\texttt{consistentSampleMass\_le\_pow\_of\_agreementMass\_le}.

The new module \texttt{FiniteIIDFamilyBound.lean} lifts this to the encoded-family
level.

\begin{theorem}[Weighted deceptive-family union bound]
Let \(H\) be a finite encoded family and \(\mu\) a finite PMF on the input type.
Then the total \(\mu^m\)-mass of deceptive samples is at most
\[
  \sum_{c \in \mathrm{BadCodes}(H,f)}
    \bigl(\mathrm{agreementMass}_{\mu}(f,\mathrm{decode}(c))\bigr)^m.
\]
\end{theorem}

\noindent
This is formalized as
\texttt{EncodedFamily.deceptiveSampleMass\_le\_badCodeAgreementMassSum}.

\begin{theorem}[Uniform weighted family corollary]
If every bad code in the family has one-step agreement mass at most \(q\), then
the total deceptive mass is at most
\[
  |\mathrm{Code}(H)| \cdot q^m.
\]
\end{theorem}

\noindent
This is formalized as
\texttt{EncodedFamily.deceptiveSampleMass\_le\_codeCard\_mul\_pow\_of\_agreementMass\_le}.

This is the first genuinely family-level finite distributional theorem in the
grassroots development.  The next real probabilistic step is no longer a union
bound itself, but a recovery or generalization theorem built on top of this
weighted family bound.

\section{Finite Weighted Recovery Bounds}

The module \texttt{FiniteIIDRecovery.lean} closes the next natural loop.  It
turns the weighted deceptive-mass bound into a weighted success statement for the
chosen ERM predictor.

\begin{theorem}[Weighted exact recovery dominates the nondeceptive mass]
If the target predictor is realized by the encoded family, then the total
\(\mu^m\)-mass of exact ERM recovery is at least
\[
  1 - \mathrm{deceptiveMass}_{\mu}.
\]
\end{theorem}

\noindent
This is formalized as
\texttt{EncodedFamily.exactRecoverySampleMass\_ge\_one\_sub\_deceptiveSampleMass}.

\begin{theorem}[Weighted exact recovery from bad-code agreement masses]
If the target predictor is realized by the encoded family, then the exact
recovery mass is at least
\[
  1 - \sum_{c \in \mathrm{BadCodes}(H,f)}
    \bigl(\mathrm{agreementMass}_{\mu}(f,\mathrm{decode}(c))\bigr)^m.
\]
\end{theorem}

\noindent
This is formalized as
\texttt{EncodedFamily.exactRecoverySampleMass\_ge\_one\_sub\_badCodeAgreementMassSum}.

\begin{theorem}[Uniform weighted exact recovery corollary]
If every bad code has one-step agreement mass at most \(q\), then the exact
recovery mass is at least
\[
  1 - |\mathrm{Code}(H)| \cdot q^m.
\]
\end{theorem}

\noindent
This is formalized as
\texttt{EncodedFamily.exactRecoverySampleMass\_ge\_one\_sub\_codeCard\_mul\_pow\_of\_agreementMass\_le}.

At this point the grassroots development has reached a fully finite,
distribution-weighted recovery theorem for ERM over encoded families.  The next
step is to lift this from exact recovery on a finite sample space to a genuine
finite-class generalization statement.

\section{What Still Counts as the Same Route}

At this stage the grassroots theory can also say something more conceptual, but
in a machine-checked way.  If a proposed rescue still wants to be \emph{the same
route} as the current switching/ERM program, then it cannot merely gesture at
weakness, GSLTs, or observer-relative semantics.  It must still instantiate the
existing finite learning interface.

The new module \texttt{SameRouteInterface.lean} packages exactly that point.  A
\emph{same-route family} is just:
\begin{enumerate}[leftmargin=1.5em]
  \item one available input surface \(\mathrm{AvailableInput}(P,B,G)\),
  \item one \(s\)-bit encoded boolean predictor family on that surface,
  \item one target predictor realized by some code in that family.
\end{enumerate}

\begin{theorem}[Same-route weighted recovery certificate]
Let \(F\) be an \(s\)-bit encoded classifier family and suppose the target
predictor is realized by some code in \(F\).  If every bad code has one-step
agreement mass at most \(q\), then the exact ERM recovery mass is at least
\[
  1 - 2^s q^m.
\]
\end{theorem}

\noindent
This is formalized as
\texttt{BitEncodedClassifierFamily.exactRecoverySampleMass\_ge\_one\_sub\_bitBudget\_mul\_pow\_of\_agreementMass\_le}.

The point is not that this theorem proves the current switching step.  The
point is that it now marks the boundary of the current route.  A weakness/GSLT
idea that still intends to be a \emph{repair} of the present program must hand
over this sort of certificate: an explicit visible input surface, an explicit
small code space, and explicit agreement-mass control for the bad codes.  If it
does not, then it is no longer filling a gap in the current route.  It is
starting a different route.

\section{Exact Post-Switch Summary Compression}

The next optimistic cleanup is to isolate the exact object that still needs to
be compressed.  The module \texttt{ExactSwitchedFamily.lean} now makes that
object explicit.

The manuscript-visible post-switch datum is the exact local surface
\[
  u = (z,a,b).
\]
The Lean file separates:
\begin{enumerate}[leftmargin=1.5em]
  \item the full exact visible surface \(u=(z,a,b)\),
  \item the invariant projection \((z,a)\),
  \item the fork projection \(((z,a),b)\).
\end{enumerate}

The main positive result is not yet a proof that the actual switched family is
small.  It is the next honest theorem underneath that goal.

\begin{theorem}[Finite summary implies explicit bit budget]
Let \(G\) be an indexed Boolean predictor family on some input type \(U\).  If
\(G\) factors through a map
\[
  \mathrm{view} : U \to V
\]
to a finite summary surface \(V\), then \(G\) admits a concrete truth-table bit
budget of size \(|V|\).
\end{theorem}

\noindent
This is formalized as
\texttt{IndexedPredictorFamily.hasBitBudget\_card\_of\_factorsThrough\_fintype}.

So the optimistic burden is now cleaner.  It is no longer
\begin{quote}
compress the switched family somehow.
\end{quote}
It is:
\begin{quote}
prove that the switched family depends only on a genuinely small finite summary
surface.
\end{quote}

The exact post-switch specializations are now also explicit.

\begin{theorem}[Invariant-view compression kernel]
If the exact switched family on \(u=(z,a,b)\) factors through the invariant
projection \((z,a)\), then it has an exact visible compression target with bit
budget
\[
  |Z| \cdot 2^k.
\]
\end{theorem}

\noindent
This is formalized by
\texttt{card\_invariantVisiblePostSwitchSurface} together with
\texttt{exactVisibleCompressionTarget\_of\_factorsThrough\_invariant\_fintype}.

\begin{theorem}[Fork-view compression kernel]
If the exact switched family factors through the fork projection
\(((z,a),b)\), then it has an exact visible compression target with bit budget
\[
  |Z| \cdot 2^k \cdot 2^k.
\]
\end{theorem}

\noindent
This is formalized by
\texttt{card\_forkVisiblePostSwitchSurface} together with
\texttt{exactVisibleCompressionTarget\_of\_factorsThrough\_fork\_fintype}.

This does \emph{not} defeat the crux.  Those summary surfaces may still be far
too large.  But it does give the ground-up reduction we wanted: a successful
repair must now prove an actually small finite summary theorem, not merely a
slogan about switched locality.

The new module \texttt{GlobalWeaknessObstruction.lean} makes the first exact
Meredith-side pressure point equally explicit.  In the current Lean bridge,
\texttt{distinctionWeakness ev} and \texttt{nonDistinctionWeakness ev} are
global scalars attached to the whole finite evidence assignment \(ev\).  They do
not vary with the underlying state \(u \in U\).

\begin{theorem}[Weakness-only factoring is constant]
If a boolean predictor \(h : U \to \{\mathrm{false},\mathrm{true}\}\) factors
through either of the current global summaries
\[
  \mathrm{distinctionWeakness}(ev)
  \qquad \text{or} \qquad
  \mathrm{nonDistinctionWeakness}(ev),
\]
then \(h\) is constant on \(U\):
\[
  \forall u,v \in U,\; h(u) = h(v).
\]
\end{theorem}

\noindent
This is formalized as
\texttt{factorsThroughDistinctionWeakness\_constant} and
\texttt{factorsThroughNonDistinctionWeakness\_constant}.  So the current
weakness bridge is real semantic infrastructure, but by itself it does not yet
instantiate the same-route learning interface.  Any GSLT/weakness rescue that
still wants to count as the current switching/ERM route must add genuinely
per-instance observables or side information beyond these present global
summaries.

\section{Hyperseed as a Local-View Surface}

The existing Hyperseed formalization already provides a clean observer-relative
semantics.

Given a perspective \(P\), a budget \(B\), and a guard region \(G\), it defines
an \emph{available region}
\[
  \mathrm{availableRegion}(P,B,G)
  =
  \mathrm{observableUniverse}(P)
  \cap
  \mathrm{nearEurycosm}(P,B)
  \cap
  G.
\]

For the grassroots PNP program, this matters because it gives a principled way to
talk about what a decoder can even access.  Instead of treating locality as an
informal phrase, we can work with a typed input surface:
\[
  \mathrm{AvailableInput}(P,B,G)
  := \{x : \mathrm{World} \mid x \in \mathrm{availableRegion}(P,B,G)\}.
\]

The new Lean module proves the corresponding encoded-family bound on such an
available input surface:

\begin{theorem}[Available-region code-budget bound]
An \(s\)-bit classifier family over one Hyperseed available region still realizes
at most \(2^s\) classifiers.
\end{theorem}

\noindent
This is formalized as
\texttt{card\_realized\_availableFamily\_le\_of\_bitBudget}.

The theorem is intentionally modest.  It does not claim that the current local
view \emph{is} a Hyperseed perspective.  It says that Hyperseed already gives us a
good general language for bounded observational access, and that this language can
interoperate with the encoded-hypothesis-class infrastructure.

\section{Operational Distinction and Weakness}

My own mathematical and computationalist intuition is that if there is an elegant
route through the real blockers, it will likely come from a semantics of
\emph{distinguishability} rather than from ad hoc local-rule counting alone.
That is the reason the Meredith lane is worth treating as part of the grassroots
background theory, at least optimistically.

At the abstract level, the bridge has three layers:
\begin{enumerate}[leftmargin=1.5em]
  \item a GSLT supplies an operational state space together with its bisimulation
  quotient;
  \item context-decorated HML supplies a language of experiments or observations;
  \item quantale weakness measures the mass of distinction events on that quotient.
\end{enumerate}

In slogan form:
\begin{quote}
operational behavior gives the ontology, HML gives the observables, and weakness
turns distinguishability into evidence.
\end{quote}

This is now more than a philosophical analogy in the local Lean state.

\begin{theorem}[Depth-bounded observational equivalence]
For each modal depth \(n\), the Lean development defines a relation
\[
  \mathrm{hmlEquivUpTo}_n(t,u),
\]
meaning that no context-decorated HML formula of modal depth at most \(n\)
distinguishes \(t\) from \(u\).
\end{theorem}

\noindent
This is formalized in \texttt{GSLT/Logic/LogicalMetric.lean}.  The same file also
defines a binary-valued depth-\(n\) logical-distance approximation
\[
  d_n(t,u) \in \{0,1\},
\]
and proves the corresponding pseudoultrametric inequality.

\begin{theorem}[Finite-depth witness forces quotient-level distinction under adequacy]
Assume the sound direction of Meredith's adequacy theorem for a GSLT \(S\).  If
some depth-\(n\) context-decorated HML formula distinguishes \(t\) from \(u\),
then \(t\) and \(u\) are genuinely distinct in the bisimulation quotient of
\(S\).  Equivalently, if
\[
  d_n(t,u) = 1,
\]
then the bisimulation classes \([t]\) and \([u]\) are different.
\end{theorem}

\noindent
This is now formalized in \texttt{GSLT/Meredith/WeaknessBridge.lean} as
\texttt{distinguishingWitness\_implies\_distinguished\_of\_adequacySound} and
\texttt{logicalDistanceApprox\_eq\_one\_implies\_classes\_ne\_of\_adequacySound}.

\begin{theorem}[Weakness on the bisimulation quotient]
For a finite bisimulation quotient \(U\) equipped with a quantale-valued weight
assignment \(\mu\), the Lean bridge defines the weakness of the distinction event
\[
  \mathrm{weakness}_{\mu}(\mathrm{distinctionEvent}(U)),
\]
interpreting operational distinguishability as an evidence quantity.
\end{theorem}

\noindent
This is the content of \texttt{GSLT/Meredith/WeaknessBridge.lean}, building on
\texttt{GSLT/Meredith/Bisimulation.lean}.  In the finite probabilistic case, this
has the expected logical-entropy shape.

The new concrete step is that the abstract minimal-context interface now has an
actual process-calculus instance.

\begin{theorem}[Concrete rho-calculus minimal-context bridge]
The rho-calculus evaluation contexts from
\texttt{Languages/ProcessCalculi/RhoCalculus/Context.lean} now induce a concrete
\texttt{HasMinimalContexts} instance for the rho GSLT, and for each rho context
\(K\),
\[
  \mathrm{contextStep}(p,K,q)
  \iff
  p \mathrel{\rightsquigarrow[K]} q.
\]
\end{theorem}

\noindent
This is formalized in \texttt{GSLT/Meredith/RhoMinimalContext.lean}.  That file
also proves concrete HML diamonds for matching COMM interactions.

For the PNP grassroots route, the optimistic use of this material is not that it
proves the switching theorem by itself.  Rather, it offers a clean semantics for
coarse observational views.  If a decoder is only sensitive to finite-depth local
experiments, then one can hope to represent that by an HML fragment and then ask
how much evidence survives under the induced quotient.  That is exactly the sort
of elegant abstraction that may help disentangle the real compression/locality
story from the current proof's more brittle prose.  The bridge is now one step
tighter than before: finite-depth observational separation no longer just lives
as a metric-like approximation, but already certifies distinct quotient classes
once adequacy-soundness is available.

The honest limit is equally important: this is still background theory.  It does
\emph{not} yet produce the switched small class, the generalization theorem, or
the \(K_{\mathrm{poly}}\) contradiction step.  But it is now a real machine-checked
background layer, not merely an idea.
The new obstruction above sharpens this limit: the currently formalized
weakness summaries are quotient-level evidence scalars, not a per-instance
predictor surface.  So any semantic rescue that remains close to the
current switching/ERM route still needs an extra layer that turns operational
semantics into state-indexed encoded observables.

The next obstruction is sharper still.  The new module
\texttt{InfinitaryHMLObstruction.lean} proves:
\begin{enumerate}[leftmargin=1.5em]
  \item if \(\mathrm{MinimalContext}(S)\) is infinite, then \(\mathrm{HMLFormula}(S)\)
  is infinite;
  \item under the same hypothesis, even the depth-1 fragment
  \(\mathrm{DepthFragment}_1(S)\) is infinite;
  \item in the concrete rho instance, there is an explicit infinite ladder of
  certified minimal contexts, so the rho depth-1 HML fragment is already
  infinite;
  \item therefore no finite code space can surject onto the current exact
  depth-1 probe family, abstractly or in the concrete rho instance.
\end{enumerate}

So the current Meredith semantic layer is now pinned down even more tightly:
before one can talk about \(\mathrm{polylog}(m)\)-bit encoded predictor families,
one first needs a separate theorem that compresses or finitizes the relevant
observational probe family.  The present HML/context layer does not supply that
for free.

The Meredith-side trilemma is now simple.
\begin{enumerate}[leftmargin=1.5em]
  \item Use the current weakness summaries: then the predictor is constant.
  \item Use the current exact HML/context probes: then the observable family is
  already too large, even at depth \(1\).
  \item Use the current exact present-moment surface directly: then one already
  encounters powerset-like finite probe slices.
\end{enumerate}

So any close-form repair now has to do something genuinely new.  It is no
longer enough to say ``use current weakness'' or ``use current HML probes a bit
more cleverly.''  One needs a new compressed observational layer together with
an explicit theorem showing that the switched predictor factors through it.

The new module \texttt{PresentMomentShattering.lean} tightens this further on a
very concrete slice of the current rho semantics.  Let state \(i\) be a single
output on channel \(i\), and let environment \(S\) be a flat bag of input
guards on the channels in a finite list \(S\).  Then the file proves:
\begin{enumerate}[leftmargin=1.5em]
  \item channel \(i\) lies in
  \(\mathrm{surfaceChannels}(\mathrm{state}_i,\mathrm{env}_S)\) iff \(i \in S\);
  \item the concrete external-present-moment pair
  \((\mathrm{par}(\mathrm{env}_S,\Box), i)\) is present iff \(i \in S\).
\end{enumerate}

So even before passing to the exact HML layer, the currently formalized
state-indexed observational surface already realizes arbitrary finite
membership patterns on a natural probe family.  This is not yet a full
``depth-1 HML shattering'' theorem, but it means the large-observable-surface
pressure is already visible one layer earlier in the exact present-moment
semantics.  In fact, the file now packages this as an injective finite-signature
map \(S \mapsto (i \mapsto \mathrm{observable}_S(i))\) on
\(\mathrm{Finset}(\mathrm{Fin}\,n)\): distinct finite probe sets yield distinct
exact present-moment signatures on the \(n\) probe states.  Any close-form
Meredith rescue now has to compress not only an infinitary exact HML layer, but
already a powerset-like concrete probe family in the rho present-moment surface.

\section{Kolmogorov Foundations Already Present}

The local repository also already contains a substantial Kolmogorov-complexity seed
under \texttt{Mettapedia/Computability/KolmogorovComplexity/}.

In particular, \texttt{Basic.lean} defines:
\begin{enumerate}[leftmargin=1.5em]
  \item binary strings as \texttt{List Bool},
  \item partial algorithms,
  \item plain Kolmogorov complexity \(C[U](x)\),
  \item a universal algorithm built from mathlib's partial-recursive code,
  \item the usual invariance theorem for universal algorithms.
\end{enumerate}

This is not yet the time-bounded conditional complexity \(K_{\mathrm{poly}}\)
that the present route uses.  But it matters for the grassroots route because it
means the eventual \(K_{\mathrm{poly}}\) development should be an extension of an
existing computability layer rather than a fresh bespoke definition dropped into a
late-stage contradiction proof.

In practical terms, the next computability milestone is:

\begin{quote}
define conditional and polynomial-time-bounded complexity on top of the existing
universal-machine layer, instead of axiomatizing them only at the paper level.
\end{quote}

\section{What the Grassroots Program Needs Next}

The current background theory suggests a fairly crisp optimistic roadmap.

\subsection{Finite-Class Learning Layer}

The ERM existence layer, finite uniform consistency bound, threshold recovery
corollary, rational uniform-rate layer, miss-ratio success bound, and weighted
one-predictor and finite-family weighted bounds are now in place.  The next
finite-class learning
steps are:
\begin{enumerate}[leftmargin=1.5em]
  \item a generalization bound stated in terms of \(\log |H|\),
  \item a probability interface for i.i.d. or distributional assumptions at the family level,
  \item concentration-style control beyond the exact finite-space formulas,
  \item a clean interface separating statistical assumptions from coding facts.
\end{enumerate}

In other words, the current Lean state already handles:
\begin{enumerate}[leftmargin=1.5em]
  \item empirical error on finite samples,
  \item existence of ERM predictors in nonempty finite encoded families,
  \item exact-fit transfer when some code in the family is sample-consistent,
  \item a finite uniform bound on the deceptive sample set,
  \item exact ERM recovery of the target off that deceptive set,
  \item a strict-cardinality threshold guaranteeing the existence of a good sample,
  \item a rational upper bound on the deceptive sample rate,
  \item a rational lower bound on the exact recovery rate,
  \item strict positivity of that exact recovery rate under the threshold,
  \item an explicit miss-ratio success bound of the form \(1 - |H|\rho_X^m\),
  \item an exact weighted one-predictor law of the form \(\mathrm{mass} = \alpha^m\),
  \item a weighted family union bound of the form
  \(\mathrm{deceptiveMass} \le \sum_{h \in H_{\mathrm{bad}}}\alpha_h^m\),
  \item a weighted exact recovery bound of the form
  \(\mathrm{recoveryMass} \ge 1 - \sum_{h \in H_{\mathrm{bad}}}\alpha_h^m\).
\end{enumerate}

\noindent
What remains is to lift that finite uniform rate layer into a real finite-class
generalization theorem.

\subsection{Concrete Decoder Subclasses}

The encoded-family theorem is only an interface.  It still needs a nontrivial
source of small classes.  The likely optimistic candidates are concrete local
decoders on tree-like neighborhoods:
\begin{enumerate}[leftmargin=1.5em]
  \item belief-propagation style update rules,
  \item GF(2)-linear constraint summaries from the VV layer,
  \item bounded-depth compositions of those primitives.
\end{enumerate}

The first concrete instance of that second lane now exists.  The new module
\texttt{AffineColumnFamily.lean} formalizes affine GF(2)-style predictors on the
retained VV column bits and proves:
\begin{quote}
if the exact switched family collapses to affine parity tests on the column
view \(a\), then the exact visible compression target is only \(k+1\) bits.
\end{quote}

\noindent
The key formal statement is
\texttt{exactVisibleCompressionTarget\_of\_realizedByExactAffineColumnFamily}.
This does not prove that the real switched family is affine.  It does give the
optimistic route its first genuinely nontrivial concrete landing zone below the
full truth-table class.

The next concrete step is now also formalized.  The new module
\texttt{AffineFeatureFamily.lean} allows a bounded number \(r\) of affine
GF(2)-style VV-column features, followed by an arbitrary truth table on the
resulting \(r\) feature bits.  Its exact raw bit budget is
\[
  r(k+1) + 2^r.
\]
The key formal statement is
\texttt{exactVisibleCompressionTarget\_of\_realizedByExactAffineFeatureFamily}.
So if the switched family collapses not to a single affine probe but to a
bounded affine-feature algebra, the code budget is still explicit and still far
below the full truth-table class on \(u=(z,a,b)\).

The next concrete step is sharper and more useful.  The new module
\texttt{SparseThresholdAffineFamily.lean} keeps the same \(r\) affine
GF(2)-style VV-column features but replaces the full truth table by a much
smaller combiner: an \(r\)-bit mask selecting active features and an \(r\)-bit
threshold code.  Its exact raw bit budget is now linear:
\[
  r(k+3).
\]
The key formal statement is
\texttt{exactVisibleCompressionTarget\_of\_realizedByExactSparseThresholdAffineFamily}.
So if the switched family collapses to a sparse-threshold affine algebra on the
retained VV column view, the optimistic route now has a concrete candidate
class whose coding cost no longer contains the exponential \(2^r\) blow-up.

If one of these affine / sparse-threshold families can be shown to encode the
actual switched predictors uniformly in \(\mathrm{polylog}(m)\) bits, then the
switching theorem has a plausible landing zone.

The next refinement is now also formalized.  The new module
\texttt{AffineDecisionListFamily.lean} keeps the same \(r\) affine GF(2)-style
VV-column features, but replaces the combiner by a fixed-order decision list:
each feature carries one response bit, and there is one default output bit if
no feature fires.  Its exact raw bit budget is
\[
  r(k+2) + 1.
\]
The key formal statement is
\texttt{exactVisibleCompressionTarget\_of\_realizedByExactAffineDecisionListFamily}.
So if the switched family collapses to an ordered affine decision-list algebra
on the retained VV column view, the optimistic route now has an even smaller
explicit candidate class than the sparse-threshold family.

The next layer is now also formalized.  The new module
\texttt{ExactAffineRecovery.lean} turns those four exact-surface candidate
classes into direct weighted finite-sampling recovery theorems.  Once the
target lies in one of the concrete exact-surface affine families, the existing
recovery backbone from \texttt{SameRouteInterface.lean} applies immediately
with one of the four explicit budgets
\[
  k+1, \qquad r(k+1)+2^r, \qquad r(k+3), \qquad r(k+2)+1.
\]
So the current small-class ladder is no longer only a compression interface.
It is now a direct recovery interface too.

The next big refinement is now also formalized.  The new module
\texttt{SharedAffineFeatureFamilies.lean} isolates the regime where one fixed
affine feature basis is shared across the switched family and only the
downstream combiner varies.  In that case the feature extractor no longer has
to be re-encoded per predictor, so the exact visible compression target drops
to the combiner alone:
\[
  2^r, \qquad 2r, \qquad r+1
\]
for the shared truth-table, shared sparse-threshold, and shared decision-list
regimes respectively.  The same module also proves the corresponding weighted
exact-recovery lower bounds on the exact post-switch surface.

This is the first genuinely sharp improvement in the optimistic lane.  If the
real switched family uses a common low-complexity feature extractor and varies
only in a light combiner, then the coding burden is much smaller than in the
previous per-predictor affine-family models.

The next concrete step is now also formalized.  The new module
\texttt{ExactABDecisionListFamily.lean} fixes the shared extractor to the raw
exact visible bits themselves: concatenate the retained VV column bits \(a\)
with the side-channel bits \(b\), then run a fixed-order decision list on those
\(2k\) raw bits.  This yields a completely concrete exact-surface candidate
class with budget
\[
  2k + 1.
\]
The key point is that the feature extractor is no longer existential or
parameterized.  It is the literal raw exact visible surface restricted to the
bit coordinates \((a,b)\).  The same module also proves the corresponding
weighted exact-recovery lower bound for that family.

The route is now factored one step more cleanly.  The new modules
\texttt{ABVisibleSurface.lean} and \texttt{ABDecisionListRoute.lean} isolate the
reduced raw visible surface \((a,b)\) as an explicit quotient candidate and
prove the pullback statement one actually wants: if the switched family factors
through that reduced surface and the reduced family is realized by fixed-order
decision lists there, then the exact-surface compression and weighted recovery
theorems follow immediately.  This is also the natural insertion point for any
future observational-equivalence argument: a semantic quotient can help justify
the map to \((a,b)\), but it does not need to alter the learning-theoretic
backbone once that map is fixed.  The same route file now also proves the
slightly stronger lift: if an exact-surface family is invariant under the
\((a,b)\) quotient, then a canonical reduced family can be reconstructed and
the same \(2k+1\)-bit decision-list target follows from that invariance alone.

The raw \((a,b)\) route is now richer than a single fixed-order decision list.
The new module \texttt{ExactABAffineFamilies.lean} reuses the existing affine
machinery on the same raw extractor and proves exact-surface compression and
weighted recovery bounds for three further candidate classes on the raw
\((a,b)\) bits:
\[
  r((2k)+1)+2^r, \qquad r((2k)+3), \qquad r((2k)+2)+1
\]
for bounded affine features, sparse-threshold affine predictors, and affine
decision lists respectively.  So the reduced raw surface is now a whole
candidate-family ladder, not just one isolated family.

The next refinement is now in place as well.  The new module
\texttt{SharedExactABFeatureFamilies.lean} adds the shared-basis collapse on
that same raw \((a,b)\) surface.  If one fixed affine basis on the \(2k\) raw
bits is reused across the whole switched family and only the downstream
combiner varies, the exact visible compression target drops all the way to
combiner-only budgets:
\[
  2^r, \qquad 2r, \qquad r+1
\]
for arbitrary truth tables, sparse-threshold combiners, and fixed-order
decision lists respectively.  So the raw \((a,b)\) quotient now supports both a
per-predictor affine ladder and a sharper shared-basis ladder.

That bridge is now factored explicitly too.  The new module
\texttt{SharedABFeatureRoutes.lean} proves that if the switched family factors
through the reduced raw surface \((a,b)\) and the reduced family is realized by
one of those shared-basis raw affine classes, then the exact-surface family
inherits the same combiner-only budget immediately.  This is the sharpest
current handoff point between any future observational quotient theorem and the
small-class recovery machinery.  The same file now also proves the invariance
version: if the exact-surface family is already invariant under the raw
\((a,b)\) quotient, then the canonical reduced family inherits those same
shared-basis combiner-only budgets without any separately supplied
factorization witness.

The target statement itself is now packaged in that style too.  The new module
\texttt{SharedABTargetInterface.lean} bundles the current route bottleneck into
explicit data packages: one shared raw \((a,b)\) feature basis, one quotient
invariance hypothesis, and one realization hypothesis for the lifted reduced
family.  From that package it derives the exact visible compression target with
budget \(2^r\), \(2r\), or \(r+1\) depending on the combiner class.  This is
the cleanest current formal statement of what the optimistic route actually
needs.

There is now also a clean backward step from that generic target statement back
toward the most concrete current route.  The module
\texttt{CanonicalABTargetRoute.lean} specializes the shared raw \((a,b)\)
feature basis to the canonical coordinate probes on the raw visible bits.  In
that specialization, the generic shared-basis decision-list target collapses
exactly to the older raw \((a,b)\) decision-list route.  So the current formal
burden can now be read in two equivalent ways:
\begin{enumerate}[leftmargin=1.5em]
  \item as a generic shared-basis target interface on the reduced raw surface,
  \item or as quotient invariance plus realization by fixed-order decision
  lists on the literal raw visible bits.
\end{enumerate}
This is the clearest backward-chaining reach from the cleaned-up target
statement back to the existing concrete work.

That concrete work is now packaged directly as well.  The new module
\texttt{CanonicalABCandidateInterface.lean} takes the canonical raw-bit route
and bundles it into one explicit candidate-data object: quotient invariance
under the raw \((a,b)\) projection together with realization of the lifted
reduced family by fixed-order decision lists on the raw visible bits.  From
that single package it derives both the exact visible compression target with
budget \(2k+1\) and, for each indexed predictor in the family, the matching
weighted exact-recovery lower bound through the already-existing raw decision
list bit family.  So the optimistic burden is now even sharper: produce one
such candidate-data witness, then prove the needed agreement-mass estimate.

That burden has now been reduced one step further.  The new module
\texttt{CanonicalABCodeWitness.lean} proves that if the switched family is
exactly the family induced by an explicit assignment of one raw \((a,b)\)
decision-list code to each index, then the canonical candidate-data package,
the \(2k+1\) compression target, and the per-index recovery theorem all follow
automatically.  So the route bottleneck is now close to its simplest concrete
form: show that the real switched family equals one such code-induced family,
then prove the corresponding agreement-mass estimate.

The last generic layer is now also in place.  The new module
\texttt{CanonicalABRecoveryInterface.lean} packages exactly the remaining
generic information needed by the route:
\begin{enumerate}[leftmargin=1.5em]
  \item one explicit raw \((a,b)\) decision-list code assignment per index,
  \item equality between the switched family and the induced code family,
  \item one agreement-mass upper bound.
\end{enumerate}
From that single recovery-data package, the \(2k+1\) compression target and the
weighted exact-recovery lower bound both follow automatically.  At that point,
the remaining work is no longer background theory; it is the actual
candidate-specific switched-family theorem.

The next manuscript-facing extension is now in place as well.  The paper's real
local input is not just \((a,b)\), but \(u=(z,a,b)\), where \(z\) comes from one
fixed shared extractor on the masked block.  The new module
\texttt{ExactZABDecisionListFamily.lean} formalizes the corresponding candidate
class directly: for any shared extractor \(\mathsf{zfeat}: Z \to \{0,1\}^r\), one
concatenates \(\mathsf{zfeat}(z)\), \(a\), and \(b\), then applies the same
fixed-order decision-list family as before.  The exact-surface budget is then
\(r+2k+1\), and the matching weighted exact-recovery bound is proved on that
larger visible surface.

That new route is then packaged in \texttt{ExactZABTargetInterface.lean}.  The
burden is exactly what one would want it to be: prove that the switched family
is realized by fixed-order decision lists on \((\mathsf{zfeat}(z),a,b)\), then
prove one bad-code agreement-mass estimate.  No extra generic machinery is
needed beyond that package.

Finally, the manuscript's constructive wrapper now plugs into the route
directly.  The new module \texttt{BitFamilyERM.lean} turns any fixed bit-encoded
local class into an indexed ERM-selected family together with its selected code
at each sample.  Then \texttt{ExactZABERMRoute.lean} specializes that statement
to the shared \((\mathsf{zfeat}(z),a,b)\) decision-list class: if the wrapper is
honestly specified as ERM over that class, then the selected code itself is the
route witness, and the compression target follows immediately.  At that point,
the remaining burden is even sharper: identify the right shared extractor
\(\mathsf{zfeat}\), show the wrapper is using that class, and control the
agreement mass of the bad codes.

There is now one more collapse on top of that manuscript-facing route.  The new
module \texttt{SharedExactZABFeatureFamilies.lean} assumes that the local bits
\((\mathsf{zfeat}(z),a,b)\) first pass through one fixed shared affine-feature
basis of width \(p\), and only the final combiner varies across predictors.  On
that shared-basis route, the exact-surface budgets fall to \(2^p\) for arbitrary
truth tables, \(2p\) for sparse thresholds, and \(p+1\) for fixed-order
decision lists.  Then \texttt{SharedExactZABTargetInterface.lean} packages the
corresponding target and recovery interfaces, and
\texttt{SharedExactZABERMRoute.lean} shows that if the wrapper is honestly ERM
over that shared-basis decision-list class, then the selected code itself again
is the route witness, now with the sharper \(p+1\) budget.

So the remaining burden is now split cleanly into two candidate routes on the
real local surface \(u=(z,a,b)\):
\begin{enumerate}[leftmargin=1.5em]
  \item the direct decision-list route on \((\mathsf{zfeat}(z),a,b)\), with budget \(r+2k+1\),
  \item the shared-basis route, where one fixed affine basis on \((\mathsf{zfeat}(z),a,b)\) reduces the decision-list budget to \(p+1\).
\end{enumerate}
The second route is now the sharper optimistic landing point.

The direct route has now been backward-chained as far as it can go generically.
The new module \texttt{CanonicalZABTargetRoute.lean} specializes the
shared-basis statement to the canonical coordinate basis on the raw exact bits,
so the shared-basis route collapses back to the direct exact decision-list
route.  Then \texttt{CanonicalZABCandidateInterface.lean} packages that direct
route as one concrete candidate object, \texttt{CanonicalZABCodeWitness.lean}
reduces it to one explicit exact decision-list code assignment per index, and
\texttt{CanonicalZABRecoveryInterface.lean} packages the last direct-route
ingredient: one agreement-mass upper bound.  So even on the less sharp
\(r+2k+1\) route, the remaining burden is now completely explicit.  Moreover,
\texttt{CanonicalZABERMRoute.lean} removes the last administrative gap between
the honest ERM wrapper and that final direct route: once the wrapper is really
ERM over the raw exact \((\mathsf{zfeat}(z),a,b)\) decision-list class, the
canonical code witness is exactly the ERM-selected code.  The new module
\texttt{CanonicalZABERMInterface.lean} packages the same conclusion in the most
manuscript-facing form: one sample assignment, one proof that the switched
family is exactly that ERM wrapper, and one bad-code agreement bound.

There is now also one fully concrete specialization of that exact route.
\texttt{BitVecZABVisibleSurface.lean} fixes the case where the local datum is
already a bitvector \(z : \mathrm{BitVec}\, r\), so the full visible input
\((z,a,b)\) becomes one raw bitvector of length \(r+2k\).  Then
\texttt{BitVecZABERMInterface.lean} specializes the final exact ERM interface
to the identity extractor on that raw full visible bit surface.  So in that
case the remaining burden is no longer “choose \(\mathsf{zfeat}\)”; it is only
“choose the samples, prove equality to the ERM wrapper, and control the bad
codes.”  The new module \texttt{ProjectedZABERMInterface.lean} narrows the same
burden in a different concrete way: when \(z\) is bit-valued, it is enough to
choose finitely many coordinates of \(z\) as the local summary extractor.

\subsection{Bridging to the Full PNP Route}

Only after those layers are built does it make sense to reconnect to the full
proof program:
\begin{enumerate}[leftmargin=1.5em]
  \item define the distributional ensemble cleanly,
  \item define local views as typed accessible surfaces,
  \item prove the switched family lies in a small encoded subclass,
  \item connect success bounds to a properly formalized \(K_{\mathrm{poly}}\),
  \item finally compare with the upper bound under \(P=NP\).
\end{enumerate}

\section{Current Status}

At present, the grassroots side has several real machine-checked gains and one
major pre-existing base layer:

\begin{center}
\begin{tabular}{lll}
\toprule
\textbf{Layer} & \textbf{Status} & \textbf{Artifact} \\
\midrule
encoded hypothesis families & proved & \texttt{HypothesisClass.lean} \\
finite-class ERM kernel & proved & \texttt{FiniteERM.lean} \\
finite uniform consistency bound & proved & \texttt{FiniteConsistencyBound.lean} \\
finite uniform recovery threshold & proved & \texttt{FiniteUniformRecovery.lean} \\
finite uniform rate layer & proved & \texttt{FiniteUniformRate.lean} \\
finite uniform success bounds & proved & \texttt{FiniteUniformSuccess.lean} \\
finite weighted one-predictor agreement law & proved & \texttt{FiniteIIDAgreement.lean} \\
finite weighted family union bound & proved & \texttt{FiniteIIDFamilyBound.lean} \\
finite weighted recovery bounds & proved & \texttt{FiniteIIDRecovery.lean} \\
same-route switching/ERM certificate & proved & \texttt{SameRouteInterface.lean} \\
exact post-switch finite-summary kernel & proved & \texttt{ExactSwitchedFamily.lean} \\
reduced raw visible surface $(a,b)$ & proved & \texttt{ABVisibleSurface.lean} \\
affine GF(2)-style column candidate class & proved & \texttt{AffineColumnFamily.lean} \\
bounded affine-feature candidate class & proved & \texttt{AffineFeatureFamily.lean} \\
sparse-threshold affine candidate class & proved & \texttt{SparseThresholdAffineFamily.lean} \\
affine decision-list candidate class & proved & \texttt{AffineDecisionListFamily.lean} \\
exact-surface affine recovery ladder & proved & \texttt{ExactAffineRecovery.lean} \\
shared-basis affine family ladder & proved & \texttt{SharedAffineFeatureFamilies.lean} \\
raw exact-surface decision-list family on $(a,b)$ & proved & \texttt{ExactABDecisionListFamily.lean} \\
pullback route from reduced $(a,b)$ families & proved & \texttt{ABDecisionListRoute.lean} \\
affine candidate ladder on raw $(a,b)$ bits & proved & \texttt{ExactABAffineFamilies.lean} \\
shared-basis affine ladder on raw $(a,b)$ bits & proved & \texttt{SharedExactABFeatureFamilies.lean} \\
quotient-route transfer to shared raw $(a,b)$ ladders & proved & \texttt{SharedABFeatureRoutes.lean} \\
unified shared raw $(a,b)$ target interface & proved & \texttt{SharedABTargetInterface.lean} \\
canonical raw $(a,b)$ specialization of the target interface & proved & \texttt{CanonicalABTargetRoute.lean} \\
canonical raw $(a,b)$ candidate-data interface with recovery consequence & proved & \texttt{CanonicalABCandidateInterface.lean} \\
explicit canonical raw $(a,b)$ code-witness reduction & proved & \texttt{CanonicalABCodeWitness.lean} \\
final canonical raw $(a,b)$ recovery interface & proved & \texttt{CanonicalABRecoveryInterface.lean} \\
ERM transport layer for fixed bit-encoded local classes & proved & \texttt{BitFamilyERM.lean} \\
exact decision-list family on $(zfeat(z),a,b)$ & proved & \texttt{ExactZABDecisionListFamily.lean} \\
shared $(zfeat(z),a,b)$ target and recovery interfaces & proved & \texttt{ExactZABTargetInterface.lean} \\
ERM specialization of the shared $(zfeat(z),a,b)$ route & proved & \texttt{ExactZABERMRoute.lean} \\
shared-basis affine ladder on $(zfeat(z),a,b)$ & proved & \texttt{SharedExactZABFeatureFamilies.lean} \\
shared-basis $(zfeat(z),a,b)$ target and recovery interfaces & proved & \texttt{SharedExactZABTargetInterface.lean} \\
ERM specialization of the shared-basis $(zfeat(z),a,b)$ route & proved & \texttt{SharedExactZABERMRoute.lean} \\
canonical specialization of the shared-basis exact route & proved & \texttt{CanonicalZABTargetRoute.lean} \\
canonical direct exact-route candidate interface & proved & \texttt{CanonicalZABCandidateInterface.lean} \\
explicit direct exact-route code-witness reduction & proved & \texttt{CanonicalZABCodeWitness.lean} \\
final direct exact-route recovery interface & proved & \texttt{CanonicalZABRecoveryInterface.lean} \\
ERM-to-canonical bridge for the exact $(zfeat(z),a,b)$ route & proved & \texttt{CanonicalZABERMRoute.lean} \\
final ERM interface for the canonical exact $(zfeat(z),a,b)$ route & proved & \texttt{CanonicalZABERMInterface.lean} \\
raw full visible bit surface when $z : \mathrm{BitVec}\, r$ & proved & \texttt{BitVecZABVisibleSurface.lean} \\
identity-extractor ERM interface on the raw full visible bit surface & proved & \texttt{BitVecZABERMInterface.lean} \\
coordinate-projection ERM interface on bit-valued local data & proved & \texttt{ProjectedZABERMInterface.lean} \\
weakness-globality obstruction & proved & \texttt{GlobalWeaknessObstruction.lean} \\
infinitary HML obstruction & proved & \texttt{InfinitaryHMLObstruction.lean} \\
present-moment shattering slice & proved & \texttt{PresentMomentShattering.lean} \\
depth-bounded HML distinction kernel & proved & \texttt{LogicalMetric.lean} \\
GSLT bisimulation-to-weakness bridge & proved & \texttt{WeaknessBridge.lean} \\
concrete rho minimal-context instance & proved & \texttt{RhoMinimalContext.lean} \\
Hyperseed available-region reuse & proved & \texttt{HypothesisClass.lean} + \texttt{Hyperseed} \\
plain Kolmogorov infrastructure & already present & \texttt{KolmogorovComplexity/Basic.lean} \\
finite-class generalization bounds & not yet formalized & next target \\
\(K_{\mathrm{poly}}\) layer & not yet formalized & later target \\
\bottomrule
\end{tabular}
\end{center}

This is not the final proof.  It is the right sort of groundwork.

\section{Conclusion}

The grassroots program should stay separate from the crux program.

The crux note asks where the present paper overreaches.  This note asks what
general machinery we would still want even if the overreach is patched later.
On that constructive side, the current Lean result is that small encoded
hypothesis families, finite empirical-risk minimizers, finite deceptive-sample
bounds, finite uniform recovery thresholds, rational exact-recovery rates,
explicit miss-ratio success bounds, weighted one-predictor agreement laws,
weighted family deceptive-mass bounds, weighted exact recovery bounds, exact
post-switch finite-summary compression kernels, one concrete affine GF(2)-style
candidate family, one bounded affine-feature candidate family, and
observer-relative input surfaces can already be handled cleanly inside one
coherent Mettapedia/mathlib formal background layer.  Optimistically, the same is
now becoming true for the observational-semantics side: finite-depth HML
distinction, weakness on operational quotients, and a concrete rho-calculus
minimal-context instance are landing as reusable infrastructure rather than as
one-off prose.  But the current Meredith bridge is now also honest about its own
shape: it produces global evidence summaries, not yet a nontrivial per-state
learning surface.  And even once one moves from those global summaries to the
HML/context layer, the current observable family is infinitary by default and
already concretely infinitary in the rho instance.

That is enough to sharpen the next optimistic task.  We should not try to prove
\(P \neq NP\) in one jump.  We should first prove that the intended switched
predictors live in a genuinely small encoded class.  The current best concrete
landing point is now sharper than the earlier raw \((a,b)\) route and sharper
than the first direct \((\mathsf{zfeat}(z),a,b)\) route: prove that the real
wrapper operates inside one shared-basis decision-list class on
\((\mathsf{zfeat}(z),a,b)\), and then prove the corresponding bad-code
agreement bound.  The direct \(r+2k+1\) route remains available, but the
shared-basis \(p+1\) route is now the best optimistic target.  Still, the
direct route is no longer vague: it has been reduced all the way to an explicit
code assignment plus an agreement bound.  Only after one of those routes is
instantiated should the rest of the argument be lifted on top of the
foundation.

\end{document}
